### 2024-10-25 index.qmd
Getting started

---
title: Regional-scale Base-flow Index of Gauged and Ungauged Basins in Drylands
author:
  - name: Caelum Mroczek
    orcid: 0009-0005-6695-0134
    corresponding: true
    email: csm428@nau.edu
    roles:
      - Conceptualization
      - Investigation
      - Methodology
      - Formal analysis
      - Visualization
      - Software
      - Data curation
      - Writing - Original Draft
    affiliations:
      - School of Earth and Sustainability, Northern Arizona University, Flagstaff, AZ, USA

  - name: Abraham Springer
    orcid: 0000-0003-4826-9124
    corresponding: false
    roles:
      - Conceptualization
      - Supervision
      - Project administration
      - Funding acquisition
    affiliations:
      - School of Earth and Sustainability, Northern Arizona University, Flagstaff, AZ, USA

  - name: Benjamin Lucas
    orcid: 0000-0002-2021-3076
    corresponding: false
    roles:
      - Software
      - Supervision
    affiliations:
      - Department of Mathematics and Statistics, Northern Arizona University, Flagstaff, AZ, USA

keywords:
  - Base Flow
  - Base-flow Index
  - Regionalization

abstract: |
  In September 2021, ...
plain-language-summary: |
  Earthquake data for the island of La Palma from the September 2021 eruption is found ...

key-points:
  - Base flow sustains dryland river ecology, but assessing its groundwater contribution faces data limitations
  - Recursive filtering and machine learning predict BFI and streamflow-associated recharge

date: last-modified
bibliography: references.bib
citation:
  container-title: Earth and Space Science
number-sections: true
---

## Introduction {#sec-intro}

Dryland regions, encompassing arid, semi-arid, hyper-arid, and dry sub-humid systems, account for 40% of the Earth’s land surface. They are home to approximately 2 billion people worldwide and represent the largest terrestrial biome (**iucn-drylands-2019)**. Despite supporting diverse ecosystems and human populations, dryland regions face mounting challenges exacerbated by increasing urbanization, expanding agricultural activities, and climate-induced amplification of precipitation patterns [@taylor2013]. This water scarcity is intensifying due to the compounding effects of climate variability and increased groundwater extraction [@taylor2013]. In drylands, groundwater serves as a vital resource for sustaining ecosystems and meeting human needs [@scanlon2006; @yao2018].

Base flow is the sustained portion of streamflow in the absence of runoff that is derived from groundwater discharge **(usgs-glossary-2018).** Base flow is critical to maintaining seasonal low-flow regimes, supporting aquatic ecosystems, and facilitating the transport of nutrients and chemicals. Base-flow contribution to streamflow can be highly variable spatially [@singh2018; @bosch2017; @beck2013], and temporally [@ficklin2016; @tan2020]. Increasing groundwater extraction, changes in land cover/land use, and changes in precipitation patterns due to climate change affect the timing and volumes of base flow [@tan2020; @taylor2013]. Effective management of water quantity and quality depends on understanding seasonal and interannual base flow patterns and long-term changes in base flow behavior.

## Data & Methods {#sec-data-methods}

## Conclusion {#sec-conclusion}

## References {.unnumbered}

::: {#refs}
:::

### 2024-10-26 index.qmd
Translating from Latex

@@
   - Regionalization
+  - Machine Learning
 
 abstract: |
-  In September 2021, ...
+  LOREM IPSUM ######
 plain-language-summary: |
-  Earthquake data for the island of La Palma from the September 2021 eruption is found ...
+  LOREM IPSUM ######
 
@@
   - Recursive filtering and machine learning predict BFI and streamflow-associated recharge
+  -
 
@@
 citation:
-  container-title: Earth and Space Science
+  container-title: Water Resources Research
 number-sections: true
@@
 Dryland regions, encompassing arid, semi-arid, hyper-arid, and dry sub-humid systems, account for 40% of the Earth’s land surface.
-They are home to approximately 2 billion people worldwide and represent the largest terrestrial biome (**iucn-drylands-2019)**.
+These regions are home to approximately 2 billion people globally and constitute the largest terrestrial biome (**iucn-drylands-2019)**.
 Despite supporting diverse ecosystems and human populations, dryland regions face mounting challenges exacerbated by increasing urbanization, expanding agricultural activities, and climate-induced amplification of precipitation patterns [@taylor2013].
@@
 
+The Base-Flow Index (BFI) is the ratio of the long-term mean base-flow volume to the long-term total streamflow volume expressed as a percentage.
+BFI serves as a normalized measure of groundwater contribution interannually or between basins.
+BFI is determined by hydrograph separation and is influenced by the climate and physiographic characteristics of a catchment [@neff2005; @beck2013; @singh2018].
+Between catchments, base flow fluctuates according to changes in the moisture content of the vadose zone, influenced by varying levels of evapotranspiration and aquifer storage dynamics [@bosch2017].
+Since BFI calculations rely on instrumented stream records, it remains unknown for ungauged catchments, which encompass most of the earth's land surface [@fekete2007].
+Addressing this information gap is integral to approaching a comprehensive understanding of groundwater dynamics globally.
+
+To address the challenge of quantifying base flow in ungauged catchments, numerous studies have applied regression and machine learning methods.
+[@ahiablame2013] found that using a regression model to estimate annual base flow of ungauged catchments was reasonably easy and accurate.
+@beck2013 overcame the nonlinearity of basin characteristics and improved results of multivariate analyses by using artificial neural networks to estimate BFI globally.
+@singh2018 implemented a random forest algorithm to predict long-term BFI for ungauged catchments across New Zealand.
+The application of machine learning in past studies highlights its versatility and effectiveness in addressing ecohydrologic dynamics.
+
+This study develops a technique for estimating BFI in ungauged basins in Arizona and determines the state’s long-term mean BFI from 1991 to 2020.
+Due to interannual variability in BFI, a 30-year period of record is used to evaluate the characteristic base-flow dynamics of Arizona.
+Using a machine learning model trained with the hydrogeologic characteristics of surface water basins, we first determine annual BFIs for each year in the 1991-2020 study period for ungauged river reaches and physiographic regions of Arizona.
+Long-term mean BFI data for the study period are then estimated.
+Regional patterns and trends in base flow of instrumented sites are subsequently investigated.
+These results provide novel analyses of low flow processes in Arizona...
+
 ## Data & Methods {#sec-data-methods}
+
+### Study Area {#sec-study-area}
+
+The state of Arizona, located in the southwest United States, covers a total area of 295,253 km$^2$.
+Arizona is divided into two primary physiographic provinces: the Colorado Plateau in the northeast, and the Basin-and-Range region in the northwest and south.
+The Central Highlands is a transition zone consisting of scattered basins separated by the mountainous foothills of the Mogollon Rim.
+The Colorado Plateau is dominated by high-elevation desert with an average elevation of 1,936 masl (6,352 ft).
+Average temperatures range from -6$^\circ$C (20$^\circ$F) to 26$^\circ$C (80$^\circ$F) and it averages 580 mm (23 in) of precipitation.
+The Basin-and-Range region has a semi-arid to arid climate with an average elevation of 490 masl (1600 ft).
+The average precipitation is 200 mm (8 in) annually with average temperatures ranging from 15$^\circ$C (60$^\circ$F) to 43$^\circ$C (110$^\circ$F).
+
+Arizona's hydrology varies seasonally and spatially between its physiographic regions.
+In the summer, localized and intense convective precipitation events are driven by the North American Monsoon, while in the winter, orographic precipitation comes from Pacific frontal systems **(eastoe2019mtnblock).** While monsoonal precipitation can account for up to 50% of annual precipitation, evaporation and dry preceding soil properties leads to most precipitation becoming runoff [@sheppard2002].
+As such, 94% of streams in Arizona are ephemeral or intermittent **(epa_AZephemeral).** Much of the hydrology of Arizona is snow-melt derived, driven by spring melt from the high-elevation Colorado Plateau winter snowpack.
+While winter precipitation provides only 30% of annual averages, it provides the bulk of water for natural reservoirs [@sheppard2002].
+
+### Data {#sec-data}
+
+Daily observed streamflow data obtained from the United States Geological Survey (USGS) National Water Information System (NWIS) were used in this study.
+Streamgages were selected depending on criteria to ensure the applicability of each site.
+@odonnell2016 found that 8-10 years of calibration data were necessary to account for climate variability in paired watershed studies in the region.
+Following this criteria, a minimum of 10 years of record was chosen.
+Any years missing more than 30 days of streamflow were removed from the possible record period.
+This analysis focuses on natural, streamflow-influencing recharge, excluding streamgages affected by regulation or diversions.
+These gages were identified through annual reports on water data published by the USGS **(usgs2010)**.
+Finally, streamgages on the Colorado River were excluded from the dataset on the basis that they represent managed flows shared by Colorado River Compact states.
+Following these parameters, 206 USGS streamgages with acceptable periods of record were used in this study (Figure \###### study-area).
+
+Watersheds across the United States are delineated by the USGS using a hydrologically-defined network.
+This system delineates the country using hierarchical hydrologic unit codes (HUCs), where each subsequent basin includes the digits of the enclosing basin.
+Here, 8-digit HUCs (HUC 8s) are used to divide Arizona into 84 sub-basins that are fully or partially in the state (Figure \###### study-area).
+These HUC 8 sub-basins are analogous to medium-sized river basins and are defined by surface water characteristics.
+\######
+
+The annual precipitation and temperature data came from the PRISM climate group at Oregon State University at a resolution of 4 km (<https://prism.oregonstate.edu;> [@daly2008]).
+The PRISM dataset provides valuable insights into regional climate in ungauged regions and has been shown to perform well across the southwestern US **(buban_PRISM).** Instead of the water year, PRISM data uses a calendar-year format, which was adopted for consistency in the water balance.
+Although this may introduce challenges in the annual estimates due to inter-annual snow storage, the use of long-term annual averages is likely to reduce any potential errors [@reitz2017].
+
+Annual evapotranspiration (ET) data came from TerraClimate, a 4-km grid climatalogical data set [@abatzoglou2018].
+TerraClimate uses a Penman-Monteith approach to generate a reference potential evapotranspiration (PET).
+The PET values were calculated assuming a reference grass surface across the landscape with unlimited water.
+In the drylands of the southwestern US, PET typically exceeds precipitation annually [@zomer2022].
+
+A 30-meter resolution Digital Elevation Model (DEM) of Arizona was used to derive key basin characteristics: basin area, average slope, and the proportion of each basin oriented toward north or south aspects.
+Various geospatial variables, such as aspect, were disaggregated then averaged to assess the areal percentage of each sub-variable within individual HUC 8 basins.
+By calculating these percentages, we aimed to get a more comprehensive understanding of landscape composition across space.
+Land cover from USGS-NLCD (National Land Cover Database), hydrologic soil group from SSURGO (Soil Survey Geographic Database), and underlying geology/karst from USGS were all similarly averaged across the basins.
+Aggregating variables to align with the HUC 8 boundaries allowed for more precise predictions of BFI by integrating spatial variations within each basin.
+
+### Base-flow separation {#sec-bf_sep}
+
+Directly measuring base flow and BFI presents unique challenges [@eckhardt2008].
+The technique chosen to separate base flow has been shown to affect results, and the choice of base-flow separation method is subjective since ‘true’ BFI values are not known [@beck2013].
+However, many methods have been developed to estimate these values.
+These methods include the use of tracers [@gonzales2009], graphical interpolation [@instituteofhydrology1980; @sloto1996], and digital filters [@arnold1995; @eckhardt2005; @lyne1979; @nathan1990].
+These techniques have varying levels of applicability depending on the spatial scale, time span, and the scope of the study.
+Comparisons of various base-flow separation techniques have been made in [@eckhardt2005; @eckhardt2008; @nathan1990] , but this study does not explore the superiority of different methods.
+
+Base flow was calculated using a single-parameter, recursive digital filter technique from [@nathan1990].
+This base-flow separation technique is based on a recursive digital filter used in signal analysis that separates high-frequency signals (quickflow) from low-frequency signals (base flow) [@lyne1979].
+This technique has been used in [@ARNOLD200021; @santhi2008; @bloomfield2009; @singh2018].
+It takes the form of
+
+$$
+b = \alpha b_{k-1} + \frac{1-\alpha}{2}(Q_k + Q_{k-1})
+$$ {#eq-bf-separate}
+
+where $b$ is base flow, $\alpha$ is the filter parameter, $Q$ is the total streamflow, and $k$ is the time step.
+A filter parameter $\alpha$ of 0.925 was used as in @nathan1990 and **\@fuka2014ecohydrology**.
+The filter was run three times (forward, backward, forward) to attenuate the base-flow signal.
+@eckhardt2023 noted that recursive digital filters lack a physical basis, but as the method is easy to automate, objective, and repeatable, it is appropriate for a regional-scale study.
+
+### Machine Learning
+
+The implementation of machine learning models to predict hydrologic indices has been successful in past studies [@singh2018] **(schmidt_ML_2020, Rozos2021Machine)**.
+In this work, we used the eXtreme Gradient Boosting (XGBoost) algorithm **(chen2016xgboost)** to predict BFI at ungauged locations using catchment characteristics as predictors.
+The XGBoost algorithm is a decision tree-based ensemble algorithm, which can be adapted for either regression or classification problems.
+It has grown in popularity in recent years due to its rapid, efficient, and scalable nature, and has proven effective in similar environmental modeling tasks such as streamflow forecasting **(szczepanek2022daily, ni2020streamflow)** and classifying land use/land cover data **(Georganos2018Very)**.
+
+XGBoost uses gradient boosting on decision tree algorithms to leverage the combination of multiple very low variance models into one overall prediction.
+Gradient boosting works sequentially, with the first tree being learned on the known target values, and each subsequent tree being learned on the errors of the tree prior.
+The algorithm then applies a weighting to each tree based on this boosting to determine how much the tree will contribute towards the final output.
+The final prediction is obtained by the input data traversing all $n$ weighted trees in the learned ensemble.
+The trained XGBoost model is used to predict BFI in ungauged catchments based on geospatial and hydroclimate predictor variables (Table \ref{tab-######}.
+Certain predictors were subdivided according to their areal coverage of each basin (e.g.
+land cover was divided into 16 subdivisions).
+
+|              | Variable                   | Source | Geoprocessing          |
+|--------------|----------------------------|--------|------------------------|
+| Hydroclimate | Precipitation              | PRISM  | Basin average          |
+|              | Mean Temperature           | PRISM  | Basin average          |
+|              | Average Evapotranspiration | MODIS  | Basin average          |
+| Geospatial   | Elevation                  | DEM    | Basin average          |
+|              | Area                       | DEM    | Basin average          |
+|              | Slope                      | DEM    | Basin average          |
+|              | Aspect                     | DEM    | Percent areal coverage |
+|              | Land Cover                 | NLCD   | Percent areal coverage |
+|              | Hydrologic Soil Group      | SSURGO | Percent areal coverage |
+|              | Geology                    | USGS   | Percent areal coverage |
+|              | Karst                      | USGS   | Percent areal coverage |
+
+: **Table 1.** Basin-characteristic variables used as predictors in XGBoost model.
+
+## Results
+
+### BFI of Gauged Catchments
+
+### BFI of Ungauged Catchments
+
+#### Predictor Importance
+
+#### Model Validation
+
+#### Predicted Long-term Mean BFI
 

### 2024-10-26 index.qmd
Update index.qmd

@@
 The final prediction is obtained by the input data traversing all $n$ weighted trees in the learned ensemble.
-The trained XGBoost model is used to predict BFI in ungauged catchments based on geospatial and hydroclimate predictor variables (Table \ref{tab-######}.
+The trained XGBoost model is used to predict BFI in ungauged catchments based on geospatial and hydroclimate predictor variables (Table \ref{tab-######}##}.
 Certain predictors were subdivided according to their areal coverage of each basin (e.g.

### 2024-10-31 index.qmd
Save

--- 2024-10-26 index.qmd
@@
 Dryland regions, encompassing arid, semi-arid, hyper-arid, and dry sub-humid systems, account for 40% of the Earth’s land surface.
-These regions are home to approximately 2 billion people globally and constitute the largest terrestrial biome (**iucn-drylands-2019)**.
+These regions are home to approximately 2 billion people globally and constitute the largest terrestrial biome [@iucn-drylands-2019].
 Despite supporting diverse ecosystems and human populations, dryland regions face mounting challenges exacerbated by increasing urbanization, expanding agricultural activities, and climate-induced amplification of precipitation patterns [@taylor2013].
@@
 
-Base flow is the sustained portion of streamflow in the absence of runoff that is derived from groundwater discharge **(usgs-glossary-2018).** Base flow is critical to maintaining seasonal low-flow regimes, supporting aquatic ecosystems, and facilitating the transport of nutrients and chemicals.
+Base flow is the sustained portion of streamflow in the absence of runoff that is derived from groundwater discharge [@usgs-glossary-2018].
+Base flow is critical to maintaining seasonal low-flow regimes, supporting aquatic ecosystems, and facilitating the transport of nutrients and chemicals.
 Base-flow contribution to streamflow can be highly variable spatially [@singh2018; @bosch2017; @beck2013], and temporally [@ficklin2016; @tan2020].
@@
 Regional patterns and trends in base flow of instrumented sites are subsequently investigated.
-These results provide novel analyses of low flow processes in Arizona...
+These results provide novel analyses of low flow processes in Arizona...######
 
@@
 Arizona's hydrology varies seasonally and spatially between its physiographic regions.
-In the summer, localized and intense convective precipitation events are driven by the North American Monsoon, while in the winter, orographic precipitation comes from Pacific frontal systems **(eastoe2019mtnblock).** While monsoonal precipitation can account for up to 50% of annual precipitation, evaporation and dry preceding soil properties leads to most precipitation becoming runoff [@sheppard2002].
-As such, 94% of streams in Arizona are ephemeral or intermittent **(epa_AZephemeral).** Much of the hydrology of Arizona is snow-melt derived, driven by spring melt from the high-elevation Colorado Plateau winter snowpack.
+In the summer, localized and intense convective precipitation events are driven by the North American Monsoon, while in the winter, orographic precipitation comes from Pacific frontal systems [@eastoe2019mtnblock].
+While monsoonal precipitation can account for up to 50% of annual precipitation, evaporation and dry preceding soil properties leads to most precipitation becoming runoff [@sheppard2002].
+As such, 94% of streams in Arizona are ephemeral or intermittent [@epa_AZephemeral].
+Much of the hydrology of Arizona is snow-melt derived, driven by spring melt from the high-elevation Colorado Plateau winter snowpack.
 While winter precipitation provides only 30% of annual averages, it provides the bulk of water for natural reservoirs [@sheppard2002].
@@
 This analysis focuses on natural, streamflow-influencing recharge, excluding streamgages affected by regulation or diversions.
-These gages were identified through annual reports on water data published by the USGS **(usgs2010)**.
+These gages were identified through annual reports on water data published by the USGS [@usgs2010].
 Finally, streamgages on the Colorado River were excluded from the dataset on the basis that they represent managed flows shared by Colorado River Compact states.
@@
 The annual precipitation and temperature data came from the PRISM climate group at Oregon State University at a resolution of 4 km (<https://prism.oregonstate.edu;> [@daly2008]).
-The PRISM dataset provides valuable insights into regional climate in ungauged regions and has been shown to perform well across the southwestern US **(buban_PRISM).** Instead of the water year, PRISM data uses a calendar-year format, which was adopted for consistency in the water balance.
+The PRISM dataset provides valuable insights into regional climate in ungauged regions and has been shown to perform well across the southwestern US [@buban_PRISM]**.** Instead of the water year, PRISM data uses a calendar-year format, which was adopted for consistency in the water balance.
 Although this may introduce challenges in the annual estimates due to inter-annual snow storage, the use of long-term annual averages is likely to reduce any potential errors [@reitz2017].
@@
 where $b$ is base flow, $\alpha$ is the filter parameter, $Q$ is the total streamflow, and $k$ is the time step.
-A filter parameter $\alpha$ of 0.925 was used as in @nathan1990 and **\@fuka2014ecohydrology**.
+A filter parameter $\alpha$ of 0.925 was used as in @nathan1990 and @fuka2014ecohydrology.
 The filter was run three times (forward, backward, forward) to attenuate the base-flow signal.
@@
 
-The implementation of machine learning models to predict hydrologic indices has been successful in past studies [@singh2018] **(schmidt_ML_2020, Rozos2021Machine)**.
+The implementation of machine learning models to predict hydrologic indices has been successful in past studies [@singh2018; @schmidt_ML_2020; @Rozos2021Machine].
 In this work, we used the eXtreme Gradient Boosting (XGBoost) algorithm **(chen2016xgboost)** to predict BFI at ungauged locations using catchment characteristics as predictors.
 The XGBoost algorithm is a decision tree-based ensemble algorithm, which can be adapted for either regression or classification problems.
-It has grown in popularity in recent years due to its rapid, efficient, and scalable nature, and has proven effective in similar environmental modeling tasks such as streamflow forecasting **(szczepanek2022daily, ni2020streamflow)** and classifying land use/land cover data **(Georganos2018Very)**.
+It has grown in popularity in recent years due to its rapid, efficient, and scalable nature, and has proven effective in similar environmental modeling tasks such as streamflow forecasting [@szczepanek2022daily**; @ni2020streamflow**] and classifying land use/land cover data [@Georganos2018Very].
 
@@
 The final prediction is obtained by the input data traversing all $n$ weighted trees in the learned ensemble.
-The trained XGBoost model is used to predict BFI in ungauged catchments based on geospatial and hydroclimate predictor variables (Table \ref{tab-######}.
+The trained XGBoost model is used to predict BFI in ungauged catchments based on geospatial and hydroclimate predictor variables (Table \ref{tab-######}##}##}##}.
 Certain predictors were subdivided according to their areal coverage of each basin (e.g.

### 2024-11-12 index.qmd
Update index.qmd

@@
 The final prediction is obtained by the input data traversing all $n$ weighted trees in the learned ensemble.
-The trained XGBoost model is used to predict BFI in ungauged catchments based on geospatial and hydroclimate predictor variables (Table \ref{tab-######}##}##}##}.
+The trained XGBoost model is used to predict BFI in ungauged catchments based on geospatial and hydroclimate predictor variables (Table \ref{tab-######}##}##}##}##}##}.
 Certain predictors were subdivided according to their areal coverage of each basin (e.g.
@@
 
+|                              | n    | R2    | MSE    | RMSE  | MAE   | Nash-Sutcliffe | pbias |
+|---------|---------|---------|---------|---------|---------|---------|---------|
+| Climate - Monsoon Dominated  | 3345 | 0.707 | 0.0166 | 0.129 | 0.077 | 0.694          | -12.6 |
+| Climate - Snowmelt Dominated | 6152 | 0.723 | 0.017  | 0.131 | 0.094 | 0.715          | -3.7  |
+| PhysRegion - Basin&Range     | 7754 | 0.749 | 0.018  | 0.133 | 0.089 | 0.741          | -5.9  |
+| PhysRegion - CO Plateau      | 1743 | 0.806 | 0.013  | 0.115 | 0.081 | 0.801          | -4.4  |
+| Climate - Warm-Wet           | 1613 | 0.690 | 0.015  | 0.124 | 0.081 | 0.679          | -9.9  |
+| Climate - Warm-Dry           | 3134 | 0.758 | 0.023  | 0.151 | 0.097 | 0.745          | -8.1  |
+| Climate - Cool-Wet           | 3134 | 0.710 | 0.013  | 0.115 | 0.085 | 0.708          | -2.5  |
+| Climate - Cool-Dry           | 1616 | 0.809 | 0.014  | 0.118 | 0.082 | 0.803          | -5    |
+| Slope - High                 | 4398 | 0.753 | 0.014  | 0.120 | 0.084 | 0.746          | -4.4  |
+| Slope - Low                  | 5099 | 0.759 | 0.019  | 0.138 | 0.090 | 0.750          | -6.9  |
+
 ### BFI of Ungauged Catchments

### 2024-11-13 index.qmd
writing

--- 2024-10-31 index.qmd
@@
   - Base flow sustains dryland river ecology, but assessing its groundwater contribution faces data limitations
-  - Recursive filtering and machine learning predict BFI and streamflow-associated recharge
+  - Recursive filtering and machine learning predict BFI 
   -
@@
 The implementation of machine learning models to predict hydrologic indices has been successful in past studies [@singh2018; @schmidt_ML_2020; @Rozos2021Machine].
-In this work, we used the eXtreme Gradient Boosting (XGBoost) algorithm **(chen2016xgboost)** to predict BFI at ungauged locations using catchment characteristics as predictors.
+In this work, we used the eXtreme Gradient Boosting (XGBoost) algorithm [@chen2016xgboost] to predict BFI at ungauged locations using catchment characteristics as predictors.
 The XGBoost algorithm is a decision tree-based ensemble algorithm, which can be adapted for either regression or classification problems.
@@
 The final prediction is obtained by the input data traversing all $n$ weighted trees in the learned ensemble.
-The trained XGBoost model is used to predict BFI in ungauged catchments based on geospatial and hydroclimate predictor variables (Table \ref{tab-######}##}##}##}.
+The trained XGBoost model is used to predict BFI in ungauged catchments based on geospatial and hydroclimate predictor variables (@tbl-predictors).
 Certain predictors were subdivided according to their areal coverage of each basin (e.g.
@@
 
-: **Table 1.** Basin-characteristic variables used as predictors in XGBoost model.
+: **Table 1.** Basin-characteristic variables used as predictors in XGBoost model. {#tbl-predictors}
+
+### Statistical Analyses
+
+Statisical analyses were performed on mean annual BFI values of instrumented streamgages.
+We examined the presence of temporal trends in mean annual BFI using the Mann-Kendall nonparametric trend test [@kendall1970; @mann1945nonparametric].
+The Mann-Kendall was used to determine trends in BFI and trend significance of gauged reaches during their period of record.
+
+Trends with a $\rho \le 0.05$ are considered significant.
 
@@
 
+The long-term average BFI for the 245 gauged reaches across Arizona is illustrated in **FIG BFI-INSTRUMENTED.** The highest BFI values (\>0.9) are found along the Grand Canyon in northwestern Arizona.
+The highly karstic geology of this region facilitates the rapid movement of subsurface flow to surface water and spring outlets [@chambless_deep-karst_2023].
+Relatively high BFI values (\>0.8) are found at the spring-fed headwaters of the Verde River (Del Rio Spring) and the spring-fed headwaters of Fossil Creek.
+These results are consistent with interpolated BFI values reported by @wolock2003.
+
+The stream reaches of the Little Colorado River Basin (northeastern Arizona) indicate consisently low BFI values (\< 0.2).
+This is likely due to low-yielding perched aquifers underlying the Defiance Uplift in northeastern Arizona, which are hydrologically connected to surface streams, while the high-yield, confined regional aquifer is much deeper [@blanchard02].
+This region also typifies a pattern of seen along most major rivers in the study area: upstream reaches indicate higher BFI values while the downstream reaches display lower BFI values.
+This is assumed to be due to greater groundwater-surface water interaction at the headwaters of streams (likely due to spring outlets) and the dilution of base flow while moving down stream (**CITATION**).
+This pattern is seen along the Gila River, Verde River, and Little Colorado Rivers in Arizona.
+
+#### Trends in BFI {#sec-bfi-trends}
+
 ### BFI of Ungauged Catchments
 
-#### Predictor Importance
+#### Model Validation
 
-#### Model Validation
+|                              | n    | R^2^  | MSE    | RMSE  | MAE   | NSE   | pbias |
+|------------------------------|------|-------|--------|-------|-------|-------|-------|
+| Climate - Monsoon Dominated  | 3345 | 0.707 | 0.0166 | 0.129 | 0.077 | 0.694 | -12.6 |
+| Climate - Snowmelt Dominated | 6152 | 0.723 | 0.017  | 0.131 | 0.094 | 0.715 | -3.7  |
+| PhysRegion - Basin&Range     | 7754 | 0.749 | 0.018  | 0.133 | 0.089 | 0.741 | -5.9  |
+| PhysRegion - CO Plateau      | 1743 | 0.806 | 0.013  | 0.115 | 0.081 | 0.801 | -4.4  |
+| Climate - Warm-Wet           | 1613 | 0.690 | 0.015  | 0.124 | 0.081 | 0.679 | -9.9  |
+| Climate - Warm-Dry           | 3134 | 0.758 | 0.023  | 0.151 | 0.097 | 0.745 | -8.1  |
+| Climate - Cool-Wet           | 3134 | 0.710 | 0.013  | 0.115 | 0.085 | 0.708 | -2.5  |
+| Climate - Cool-Dry           | 1616 | 0.809 | 0.014  | 0.118 | 0.082 | 0.803 | -5    |
+| Slope - High                 | 4398 | 0.753 | 0.014  | 0.120 | 0.084 | 0.746 | -4.4  |
+| Slope - Low                  | 5099 | 0.759 | 0.019  | 0.138 | 0.090 | 0.750 | -6.9  |
+
+: Table 2.
+Performance of model predictions for BFI for all sites split by various classifications.
+n is number of sites, R^2^ is the coefficient of determination of a linear regression, MSE is mean-squared-error, RMSE is root-mean-squared-error, MAE is mean-absolute-error, NSE is Nash-Sucliffe efficiency, and pbias is percent bias.
+{#tbl-performance}
+
+#### Predictor Importance {#sec-predictor-importance}
 

### 2024-11-19 index.qmd
adding figures

@@
 
+![Map of Arizona indicating US Geological Survey (USGS) streamgages used in this study. 8-digit HUC subbasin boundaries and physiographic regions shown.](images/StudyArea.png){#fig-study-area}
+
 Arizona's hydrology varies seasonally and spatially between its physiographic regions.
@@
 Finally, streamgages on the Colorado River were excluded from the dataset on the basis that they represent managed flows shared by Colorado River Compact states.
-Following these parameters, 206 USGS streamgages with acceptable periods of record were used in this study (Figure \###### study-area).
+Following these parameters, 206 USGS streamgages with acceptable periods of record were used in this study (@fig-study-area).
 
@@
 This system delineates the country using hierarchical hydrologic unit codes (HUCs), where each subsequent basin includes the digits of the enclosing basin.
-Here, 8-digit HUCs (HUC 8s) are used to divide Arizona into 84 sub-basins that are fully or partially in the state (Figure \###### study-area).
+Here, 8-digit HUCs (HUC 8s) are used to divide Arizona into 84 sub-basins that are fully or partially in the state (@fig-study-area).
 These HUC 8 sub-basins are analogous to medium-sized river basins and are defined by surface water characteristics.
@@
 
-The long-term average BFI for the 245 gauged reaches across Arizona is illustrated in **FIG BFI-INSTRUMENTED.** The highest BFI values (\>0.9) are found along the Grand Canyon in northwestern Arizona.
+![Mean annual BFI for the period of study (1991-2020) from instrumented stream flow data.](images/BFI_Instrumented.png){#fig-instrumented-bfi}
+
+The long-term average BFI for the 245 gauged reaches across Arizona is illustrated in @fig-instrumented-bfi.The highest BFI values (\>0.9) are found along the Grand Canyon in northwestern Arizona.
 The highly karstic geology of this region facilitates the rapid movement of subsurface flow to surface water and spring outlets [@chambless_deep-karst_2023].
@@
 #### Trends in BFI {#sec-bfi-trends}
+
+![Trends in BFI over full period of record for instrumented sites used in this study.
+Red upward (blue downward) arrows indicate an increasing (decreasing) trend at a significance level of 5%.
+Nonsignificant trends are shown by colors of circles.](images/BFITrends.png){#fig-instrumented-trend}
 
@@
 
+![Predicted mean annual BFI values for 8-digit HUC (1991-2020)](images/BFI_HUC8.png){#fig-bfi-huc}
+
+![Predicted mean annual BFI values for all reaches of Strahler stream order 3 and greater](images/BFI_AllStream.pdf){#fig-bfi-streams}
+
 ## Conclusion {#sec-conclusion}

### 2024-11-19 index.tex
adding figures

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
]{agujournal2019}

\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else  
    % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\setcounter{secnumdepth}{5}
% Make \paragraph and \subparagraph free-standing
\makeatletter
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}{
    \@ifstar
      \xxxParagraphStar
      \xxxParagraphNoStar
  }
  \newcommand{\xxxParagraphStar}[1]{\oldparagraph*{#1}\mbox{}}
  \newcommand{\xxxParagraphNoStar}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}{
    \@ifstar
      \xxxSubParagraphStar
      \xxxSubParagraphNoStar
  }
  \newcommand{\xxxSubParagraphStar}[1]{\oldsubparagraph*{#1}\mbox{}}
  \newcommand{\xxxSubParagraphNoStar}[1]{\oldsubparagraph{#1}\mbox{}}
\fi
\makeatother

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
% definitions for citeproc citations
\NewDocumentCommand\citeproctext{}{}
\NewDocumentCommand\citeproc{mm}{%
  \begingroup\def\citeproctext{#2}\cite{#1}\endgroup}
\makeatletter
 % allow citations to break across lines
 \let\@cite@ofmt\@firstofone
 % avoid brackets around text for \cite:
 \def\@biblabel#1{}
 \def\@cite#1#2{{#1\if@tempswa , #2\fi}}
\makeatother
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newenvironment{CSLReferences}[2] % #1 hanging-indent, #2 entry-spacing
 {\begin{list}{}{%
  \setlength{\itemindent}{0pt}
  \setlength{\leftmargin}{0pt}
  \setlength{\parsep}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
   \setlength{\leftmargin}{\cslhangindent}
   \setlength{\itemindent}{-1\cslhangindent}
  \fi
  % set entry spacing
  \setlength{\itemsep}{#2\baselineskip}}}
 {\end{list}}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{\hfill\break\parbox[t]{\linewidth}{\strut\ignorespaces#1\strut}}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

\usepackage{url} %this package should fix any errors with URLs in refs.
\usepackage{lineno}
\usepackage[inline]{trackchanges} %for better track changes. finalnew option will compile document with changes incorporated.
\usepackage{soul}
\linenumbers
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother

\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage{bookmark}

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Regional-scale Base-flow Index of Gauged and Ungauged Basins in Drylands},
  pdfauthor={Caelum Mroczek; Abraham Springer; Benjamin Lucas},
  pdfkeywords={Base Flow, Base-flow Index, Regionalization, Machine
Learning},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\journalname{Water Resources Research}

\draftfalse

\begin{document}
\title{Regional-scale Base-flow Index of Gauged and Ungauged Basins in
Drylands}

\authors{Caelum Mroczek\affil{1}, Abraham Springer\affil{1}, Benjamin
Lucas\affil{2}}
\affiliation{1}{School of Earth and Sustainability, Northern Arizona
University, Flagstaff, AZ, USA, }\affiliation{2}{Department of
Mathematics and Statistics, Northern Arizona University, Flagstaff, AZ,
USA, }
\correspondingauthor{Caelum Mroczek}{csm428@nau.edu}

\begin{abstract}
LOREM IPSUM \#\#\#\#\#\#
\end{abstract}

\section*{Plain Language Summary}
LOREM IPSUM \#\#\#\#\#\#

\section{Introduction}\label{sec-intro}

Dryland regions, encompassing arid, semi-arid, hyper-arid, and dry
sub-humid systems, account for 40\% of the Earth's land surface. These
regions are home to approximately 2 billion people globally and
constitute the largest terrestrial biome (IUCN, 2019). Despite
supporting diverse ecosystems and human populations, dryland regions
face mounting challenges exacerbated by increasing urbanization,
expanding agricultural activities, and climate-induced amplification of
precipitation patterns (Taylor et al., 2013). This water scarcity is
intensifying due to the compounding effects of climate variability and
increased groundwater extraction (Taylor et al., 2013). In drylands,
groundwater serves as a vital resource for sustaining ecosystems and
meeting human needs (Scanlon et al., 2006; Yao et al., 2018).

Base flow is the sustained portion of streamflow in the absence of
runoff that is derived from groundwater discharge (USGS, 2018). Base
flow is critical to maintaining seasonal low-flow regimes, supporting
aquatic ecosystems, and facilitating the transport of nutrients and
chemicals. Base-flow contribution to streamflow can be highly variable
spatially (Beck et al., 2013; Bosch et al., 2017; Singh et al., 2018),
and temporally (Ficklin et al., 2016; Tan et al., 2020). Increasing
groundwater extraction, changes in land cover/land use, and changes in
precipitation patterns due to climate change affect the timing and
volumes of base flow (Tan et al., 2020; Taylor et al., 2013). Effective
management of water quantity and quality depends on understanding
seasonal and interannual base flow patterns and long-term changes in
base flow behavior.

The Base-Flow Index (BFI) is the ratio of the long-term mean base-flow
volume to the long-term total streamflow volume expressed as a
percentage. BFI serves as a normalized measure of groundwater
contribution interannually or between basins. BFI is determined by
hydrograph separation and is influenced by the climate and physiographic
characteristics of a catchment (Beck et al., 2013; Neff et al., 2005;
Singh et al., 2018). Between catchments, base flow fluctuates according
to changes in the moisture content of the vadose zone, influenced by
varying levels of evapotranspiration and aquifer storage dynamics (Bosch
et al., 2017). Since BFI calculations rely on instrumented stream
records, it remains unknown for ungauged catchments, which encompass
most of the earth's land surface (Fekete et al., 2007). Addressing this
information gap is integral to approaching a comprehensive understanding
of groundwater dynamics globally.

To address the challenge of quantifying base flow in ungauged
catchments, numerous studies have applied regression and machine
learning methods. (Ahiablame et al., 2013) found that using a regression
model to estimate annual base flow of ungauged catchments was reasonably
easy and accurate. Beck et al. (2013) overcame the nonlinearity of basin
characteristics and improved results of multivariate analyses by using
artificial neural networks to estimate BFI globally. Singh et al. (2018)
implemented a random forest algorithm to predict long-term BFI for
ungauged catchments across New Zealand. The application of machine
learning in past studies highlights its versatility and effectiveness in
addressing ecohydrologic dynamics.

This study develops a technique for estimating BFI in ungauged basins in
Arizona and determines the state's long-term mean BFI from 1991 to 2020.
Due to interannual variability in BFI, a 30-year period of record is
used to evaluate the characteristic base-flow dynamics of Arizona. Using
a machine learning model trained with the hydrogeologic characteristics
of surface water basins, we first determine annual BFIs for each year in
the 1991-2020 study period for ungauged river reaches and physiographic
regions of Arizona. Long-term mean BFI data for the study period are
then estimated. Regional patterns and trends in base flow of
instrumented sites are subsequently investigated. These results provide
novel analyses of low flow processes in Arizona\ldots\#\#\#\#\#\#

\section{Data \& Methods}\label{sec-data-methods}

\subsection{Study Area}\label{sec-study-area}

The state of Arizona, located in the southwest United States, covers a
total area of 295,253 km\(^2\). Arizona is divided into two primary
physiographic provinces: the Colorado Plateau in the northeast, and the
Basin-and-Range region in the northwest and south. The Central Highlands
is a transition zone consisting of scattered basins separated by the
mountainous foothills of the Mogollon Rim. The Colorado Plateau is
dominated by high-elevation desert with an average elevation of 1,936
masl (6,352 ft). Average temperatures range from -6\(^\circ\)C
(20\(^\circ\)F) to 26\(^\circ\)C (80\(^\circ\)F) and it averages 580 mm
(23 in) of precipitation. The Basin-and-Range region has a semi-arid to
arid climate with an average elevation of 490 masl (1600 ft). The
average precipitation is 200 mm (8 in) annually with average
temperatures ranging from 15\(^\circ\)C (60\(^\circ\)F) to 43\(^\circ\)C
(110\(^\circ\)F).

\begin{figure}

\centering{

\includegraphics{images/StudyArea.png}

}

\caption{\label{fig-study-area}Map of Arizona indicating US Geological
Survey (USGS) streamgages used in this study. 8-digit HUC subbasin
boundaries and physiographic regions shown.}

\end{figure}%

Arizona's hydrology varies seasonally and spatially between its
physiographic regions. In the summer, localized and intense convective
precipitation events are driven by the North American Monsoon, while in
the winter, orographic precipitation comes from Pacific frontal systems
(Eastoe et al., 2019). While monsoonal precipitation can account for up
to 50\% of annual precipitation, evaporation and dry preceding soil
properties leads to most precipitation becoming runoff (Sheppard et al.,
2002). As such, 94\% of streams in Arizona are ephemeral or intermittent
(Levick et al., 2008). Much of the hydrology of Arizona is snow-melt
derived, driven by spring melt from the high-elevation Colorado Plateau
winter snowpack. While winter precipitation provides only 30\% of annual
averages, it provides the bulk of water for natural reservoirs (Sheppard
et al., 2002).

\subsection{Data}\label{sec-data}

Daily observed streamflow data obtained from the United States
Geological Survey (USGS) National Water Information System (NWIS) were
used in this study. Streamgages were selected depending on criteria to
ensure the applicability of each site. O'Donnell et al. (2016) found
that 8-10 years of calibration data were necessary to account for
climate variability in paired watershed studies in the region. Following
this criteria, a minimum of 10 years of record was chosen. Any years
missing more than 30 days of streamflow were removed from the possible
record period. This analysis focuses on natural, streamflow-influencing
recharge, excluding streamgages affected by regulation or diversions.
These gages were identified through annual reports on water data
published by the USGS (USGS, 2010). Finally, streamgages on the Colorado
River were excluded from the dataset on the basis that they represent
managed flows shared by Colorado River Compact states. Following these
parameters, 206 USGS streamgages with acceptable periods of record were
used in this study (Figure~\ref{fig-study-area}).

Watersheds across the United States are delineated by the USGS using a
hydrologically-defined network. This system delineates the country using
hierarchical hydrologic unit codes (HUCs), where each subsequent basin
includes the digits of the enclosing basin. Here, 8-digit HUCs (HUC 8s)
are used to divide Arizona into 84 sub-basins that are fully or
partially in the state (Figure~\ref{fig-study-area}). These HUC 8
sub-basins are analogous to medium-sized river basins and are defined by
surface water characteristics. \#\#\#\#\#\#

The annual precipitation and temperature data came from the PRISM
climate group at Oregon State University at a resolution of 4 km
(\url{https://prism.oregonstate.edu;} (Daly et al., 2008)). The PRISM
dataset provides valuable insights into regional climate in ungauged
regions and has been shown to perform well across the southwestern US
(Buban et al., 2020)\textbf{.} Instead of the water year, PRISM data
uses a calendar-year format, which was adopted for consistency in the
water balance. Although this may introduce challenges in the annual
estimates due to inter-annual snow storage, the use of long-term annual
averages is likely to reduce any potential errors (Reitz et al., 2017).

Annual evapotranspiration (ET) data came from TerraClimate, a 4-km grid
climatalogical data set (Abatzoglou et al., 2018). TerraClimate uses a
Penman-Monteith approach to generate a reference potential
evapotranspiration (PET). The PET values were calculated assuming a
reference grass surface across the landscape with unlimited water. In
the drylands of the southwestern US, PET typically exceeds precipitation
annually (Zomer et al., 2022).

A 30-meter resolution Digital Elevation Model (DEM) of Arizona was used
to derive key basin characteristics: basin area, average slope, and the
proportion of each basin oriented toward north or south aspects. Various
geospatial variables, such as aspect, were disaggregated then averaged
to assess the areal percentage of each sub-variable within individual
HUC 8 basins. By calculating these percentages, we aimed to get a more
comprehensive understanding of landscape composition across space. Land
cover from USGS-NLCD (National Land Cover Database), hydrologic soil
group from SSURGO (Soil Survey Geographic Database), and underlying
geology/karst from USGS were all similarly averaged across the basins.
Aggregating variables to align with the HUC 8 boundaries allowed for
more precise predictions of BFI by integrating spatial variations within
each basin.

\subsection{Base-flow separation}\label{sec-bf_sep}

Directly measuring base flow and BFI presents unique challenges
(Eckhardt, 2008). The technique chosen to separate base flow has been
shown to affect results, and the choice of base-flow separation method
is subjective since `true' BFI values are not known (Beck et al., 2013).
However, many methods have been developed to estimate these values.
These methods include the use of tracers (Gonzales et al., 2009),
graphical interpolation (Institute of Hydrology, 1980; Sloto et al.,
1996), and digital filters (Arnold et al., 1995; Eckhardt, 2005; Lyne et
al., 1979; Nathan et al., 1990). These techniques have varying levels of
applicability depending on the spatial scale, time span, and the scope
of the study. Comparisons of various base-flow separation techniques
have been made in (Eckhardt, 2005, 2008; Nathan et al., 1990) , but this
study does not explore the superiority of different methods.

Base flow was calculated using a single-parameter, recursive digital
filter technique from (Nathan et al., 1990). This base-flow separation
technique is based on a recursive digital filter used in signal analysis
that separates high-frequency signals (quickflow) from low-frequency
signals (base flow) (Lyne et al., 1979). This technique has been used in
(Arnold et al., 2000; Bloomfield et al., 2009; Santhi et al., 2008;
Singh et al., 2018). It takes the form of

\begin{equation}\phantomsection\label{eq-bf-separate}{
b = \alpha b_{k-1} + \frac{1-\alpha}{2}(Q_k + Q_{k-1})
}\end{equation}

where \(b\) is base flow, \(\alpha\) is the filter parameter, \(Q\) is
the total streamflow, and \(k\) is the time step. A filter parameter
\(\alpha\) of 0.925 was used as in Nathan et al. (1990) and Fuka et al.
(2014). The filter was run three times (forward, backward, forward) to
attenuate the base-flow signal. Eckhardt (2023) noted that recursive
digital filters lack a physical basis, but as the method is easy to
automate, objective, and repeatable, it is appropriate for a
regional-scale study.

\subsection{Machine Learning}\label{machine-learning}

The implementation of machine learning models to predict hydrologic
indices has been successful in past studies (Rozos et al., 2021; Schmidt
et al., 2020; Singh et al., 2018). In this work, we used the eXtreme
Gradient Boosting (XGBoost) algorithm (Chen et al., 2016) to predict BFI
at ungauged locations using catchment characteristics as predictors. The
XGBoost algorithm is a decision tree-based ensemble algorithm, which can
be adapted for either regression or classification problems. It has
grown in popularity in recent years due to its rapid, efficient, and
scalable nature, and has proven effective in similar environmental
modeling tasks such as streamflow forecasting Ni et al. (2020) and
classifying land use/land cover data (Georganos et al., 2018).

XGBoost uses gradient boosting on decision tree algorithms to leverage
the combination of multiple very low variance models into one overall
prediction. Gradient boosting works sequentially, with the first tree
being learned on the known target values, and each subsequent tree being
learned on the errors of the tree prior. The algorithm then applies a
weighting to each tree based on this boosting to determine how much the
tree will contribute towards the final output. The final prediction is
obtained by the input data traversing all \(n\) weighted trees in the
learned ensemble. The trained XGBoost model is used to predict BFI in
ungauged catchments based on geospatial and hydroclimate predictor
variables (Table~\ref{tbl-predictors}). Certain predictors were
subdivided according to their areal coverage of each basin (e.g.~land
cover was divided into 16 subdivisions).

\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.1892}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.3784}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.1081}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.3243}}@{}}
\caption{\textbf{Table 1.} Basin-characteristic variables used as
predictors in XGBoost model.}\label{tbl-predictors}\tabularnewline
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Variable
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Source
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Geoprocessing
\end{minipage} \\
\midrule\noalign{}
\endfirsthead
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Variable
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Source
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Geoprocessing
\end{minipage} \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
Hydroclimate & Precipitation & PRISM & Basin average \\
& Mean Temperature & PRISM & Basin average \\
& Average Evapotranspiration & MODIS & Basin average \\
Geospatial & Elevation & DEM & Basin average \\
& Area & DEM & Basin average \\
& Slope & DEM & Basin average \\
& Aspect & DEM & Percent areal coverage \\
& Land Cover & NLCD & Percent areal coverage \\
& Hydrologic Soil Group & SSURGO & Percent areal coverage \\
& Geology & USGS & Percent areal coverage \\
& Karst & USGS & Percent areal coverage \\
\end{longtable}

\subsection{Statistical Analyses}\label{statistical-analyses}

Statisical analyses were performed on mean annual BFI values of
instrumented streamgages. We examined the presence of temporal trends in
mean annual BFI using the Mann-Kendall nonparametric trend test
(Kendall, 1970; Mann, 1945). The Mann-Kendall was used to determine
trends in BFI and trend significance of gauged reaches during their
period of record.

Trends with a \(\rho \le 0.05\) are considered significant.

\section{Results}\label{results}

\subsection{BFI of Gauged Catchments}\label{bfi-of-gauged-catchments}

\begin{figure}

\centering{

\includegraphics{images/BFI_Instrumented.png}

}

\caption{\label{fig-instrumented-bfi}Mean annual BFI for the period of
study (1991-2020) from instrumented stream flow data.}

\end{figure}%

The long-term average BFI for the 245 gauged reaches across Arizona is
illustrated in \textbf{?@fig-instrumented-bfi.The} highest BFI values
(\textgreater0.9) are found along the Grand Canyon in northwestern
Arizona. The highly karstic geology of this region facilitates the rapid
movement of subsurface flow to surface water and spring outlets
(Chambless et al., 2023). Relatively high BFI values (\textgreater0.8)
are found at the spring-fed headwaters of the Verde River (Del Rio
Spring) and the spring-fed headwaters of Fossil Creek. These results are
consistent with interpolated BFI values reported by Wolock (2003).

The stream reaches of the Little Colorado River Basin (northeastern
Arizona) indicate consisently low BFI values (\textless{} 0.2). This is
likely due to low-yielding perched aquifers underlying the Defiance
Uplift in northeastern Arizona, which are hydrologically connected to
surface streams, while the high-yield, confined regional aquifer is much
deeper (Blanchard, 2002). This region also typifies a pattern of seen
along most major rivers in the study area: upstream reaches indicate
higher BFI values while the downstream reaches display lower BFI values.
This is assumed to be due to greater groundwater-surface water
interaction at the headwaters of streams (likely due to spring outlets)
and the dilution of base flow while moving down stream
(\textbf{CITATION}). This pattern is seen along the Gila River, Verde
River, and Little Colorado Rivers in Arizona.

\subsubsection{Trends in BFI}\label{sec-bfi-trends}

\begin{figure}

\centering{

\includegraphics{images/BFITrends.png}

}

\caption{\label{fig-instrumented-trend}Trends in BFI over full period of
record for instrumented sites used in this study. Red upward (blue
downward) arrows indicate an increasing (decreasing) trend at a
significance level of 5\%. Nonsignificant trends are shown by colors of
circles.}

\end{figure}%

\subsection{BFI of Ungauged
Catchments}\label{bfi-of-ungauged-catchments}

\subsubsection{Model Validation}\label{model-validation}

\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.3797}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.0759}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.0886}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.1013}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.0886}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.0886}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.0886}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.0886}}@{}}
\caption{Table 2. Performance of model predictions for BFI for all sites
split by various classifications. n is number of sites,
R\textsuperscript{2} is the coefficient of determination of a linear
regression, MSE is mean-squared-error, RMSE is root-mean-squared-error,
MAE is mean-absolute-error, NSE is Nash-Sucliffe efficiency, and pbias
is percent bias.}\label{tbl-performance}\tabularnewline
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
n
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
R\textsuperscript{2}
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
MSE
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
RMSE
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
MAE
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
NSE
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
pbias
\end{minipage} \\
\midrule\noalign{}
\endfirsthead
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
n
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
R\textsuperscript{2}
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
MSE
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
RMSE
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
MAE
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
NSE
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
pbias
\end{minipage} \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
Climate - Monsoon Dominated & 3345 & 0.707 & 0.0166 & 0.129 & 0.077 &
0.694 & -12.6 \\
Climate - Snowmelt Dominated & 6152 & 0.723 & 0.017 & 0.131 & 0.094 &
0.715 & -3.7 \\
PhysRegion - Basin\&Range & 7754 & 0.749 & 0.018 & 0.133 & 0.089 & 0.741
& -5.9 \\
PhysRegion - CO Plateau & 1743 & 0.806 & 0.013 & 0.115 & 0.081 & 0.801 &
-4.4 \\
Climate - Warm-Wet & 1613 & 0.690 & 0.015 & 0.124 & 0.081 & 0.679 &
-9.9 \\
Climate - Warm-Dry & 3134 & 0.758 & 0.023 & 0.151 & 0.097 & 0.745 &
-8.1 \\
Climate - Cool-Wet & 3134 & 0.710 & 0.013 & 0.115 & 0.085 & 0.708 &
-2.5 \\
Climate - Cool-Dry & 1616 & 0.809 & 0.014 & 0.118 & 0.082 & 0.803 &
-5 \\
Slope - High & 4398 & 0.753 & 0.014 & 0.120 & 0.084 & 0.746 & -4.4 \\
Slope - Low & 5099 & 0.759 & 0.019 & 0.138 & 0.090 & 0.750 & -6.9 \\
\end{longtable}

\subsubsection{Predictor Importance}\label{sec-predictor-importance}

\subsubsection{Predicted Long-term Mean
BFI}\label{predicted-long-term-mean-bfi}

\begin{figure}

\centering{

\includegraphics{images/BFI_HUC8.png}

}

\caption{\label{fig-bfi-huc}Predicted mean annual BFI values for 8-digit
HUC (1991-2020)}

\end{figure}%

\begin{figure}

\centering{

\includegraphics{images/BFI_AllStream.pdf}

}

\caption{\label{fig-bfi-streams}Predicted mean annual BFI values for all
reaches of Strahler stream order 3 and greater}

\end{figure}%

\section{Conclusion}\label{sec-conclusion}

\section*{References}\label{references}
\addcontentsline{toc}{section}{References}

\phantomsection\label{refs}
\begin{CSLReferences}{1}{0}
\vspace{1em}

\bibitem[\citeproctext]{ref-abatzoglou2018}
Abatzoglou, J. T. et al. (2018). TerraClimate, a high-resolution global
dataset of monthly climate and climatic water balance from
1958{\textendash}2015. \emph{Scientific Data}, \emph{5}(1), 170191.
\url{https://doi.org/10.1038/sdata.2017.191}

\bibitem[\citeproctext]{ref-ahiablame2013}
Ahiablame, L. et al. (2013). Estimation of annual baseflow at ungauged
sites in Indiana USA. \emph{Journal of Hydrology}, \emph{476}, 13--27.
\url{https://doi.org/10.1016/j.jhydrol.2012.10.002}

\bibitem[\citeproctext]{ref-arnold1995}
Arnold, J. G. et al. (1995). Automated Base Flow Separation and
Recession Analysis Techniques. \emph{Ground Water}, \emph{33}(6),
1010--1018. \url{https://doi.org/10.1111/j.1745-6584.1995.tb00046.x}

\bibitem[\citeproctext]{ref-ARNOLD200021}
Arnold, J. G. et al. (2000). Regional estimation of base flow and
groundwater recharge in the upper mississippi river basin. \emph{Journal
of Hydrology}, \emph{227}(1), 21--40.
https://doi.org/\url{https://doi.org/10.1016/S0022-1694(99)00139-0}

\bibitem[\citeproctext]{ref-beck2013}
Beck, H. E. et al. (2013). Global patterns in base flow index and
recession based on streamflow observations from 3394 catchments: Global
Patterns in Base Flow Characteristics. \emph{Water Resources Research},
\emph{49}(12), 7843--7863. \url{https://doi.org/10.1002/2013WR013918}

\bibitem[\citeproctext]{ref-blanchard02}
Blanchard, P. J. (2002). \emph{Assessments of aquifer sensitivity on
{Navajo Nation} and adjacent lands and ground-water vulnerability to
pesticide contamination on the {Navajo Indian Irrigation Project},
{Arizona}, {New Mexico}, and {Utah}} (No. Water-Resources Investigations
Report 02-4051). {USGS}. \url{https://doi.org/10.3133/wri024051}

\bibitem[\citeproctext]{ref-bloomfield2009}
Bloomfield, J. P. et al. (2009). Examining geological controls on
baseflow index (BFI) using regression analysis: An illustration from the
Thames Basin, UK. \emph{Journal of Hydrology}, \emph{373}(1-2),
164--176. \url{https://doi.org/10.1016/j.jhydrol.2009.04.025}

\bibitem[\citeproctext]{ref-bosch2017}
Bosch, D. D. et al. (2017). Temporal variations in baseflow for the
Little River experimental watershed in South Georgia, USA. \emph{Journal
of Hydrology: Regional Studies}, \emph{10}, 110--121.
\url{https://doi.org/10.1016/j.ejrh.2017.02.002}

\bibitem[\citeproctext]{ref-buban_PRISM}
Buban, M. S. et al. (2020). A comparison of the u.s. Climate reference
network precipitation data to the parameter-elevation regressions on
independent slopes model (PRISM). \emph{Journal of Hydrometeorology},
\emph{21}(10), 2391--2400. \url{https://doi.org/10.1175/JHM-D-19-0232.1}

\bibitem[\citeproctext]{ref-chambless_deep-karst_2023}
Chambless, H. E. et al. (2023). Deep-karst aquifer spring-flow trends in
a water-limited system, grand canyon national park, {USA}.
\emph{Hydrogeology Journal}, \emph{31}(7), 1755--1771.
\url{https://doi.org/10.1007/s10040-023-02702-w}

\bibitem[\citeproctext]{ref-chen2016xgboost}
Chen, T. et al. (2016). Xgboost: A scalable tree boosting system. In
\emph{Proceedings of the 22nd acm sigkdd international conference on
knowledge discovery and data mining} (pp. 785--794).

\bibitem[\citeproctext]{ref-daly2008}
Daly, C. et al. (2008). Physiographically sensitive mapping of
climatological temperature and precipitation across the conterminous
United States. \emph{International Journal of Climatology},
\emph{28}(15), 2031--2064. \url{https://doi.org/10.1002/joc.1688}

\bibitem[\citeproctext]{ref-eastoe2019mtnblock}
Eastoe, C. J. et al. (2019). Hydrology of mountain blocks in arizona and
new mexico as revealed by isotopes in groundwater and precipitation.
\emph{Geosciences}, \emph{9}(11).
\url{https://doi.org/10.3390/geosciences9110461}

\bibitem[\citeproctext]{ref-eckhardt2005}
Eckhardt, K. (2005). How to construct recursive digital filters for
baseflow separation. \emph{Hydrological Processes}, \emph{19}(2),
507--515. \url{https://doi.org/10.1002/hyp.5675}

\bibitem[\citeproctext]{ref-eckhardt2008}
Eckhardt, K. (2008). A comparison of baseflow indices, which were
calculated with seven different baseflow separation methods.
\emph{Journal of Hydrology}, \emph{352}(1-2), 168--173.
\url{https://doi.org/10.1016/j.jhydrol.2008.01.005}

\bibitem[\citeproctext]{ref-eckhardt2023}
Eckhardt, K. (2023). Technical note: How physically based is hydrograph
separation by recursive digital filtering? \emph{Hydrology and Earth
System Sciences}, \emph{27}(2), 495--499.
\url{https://doi.org/10.5194/hess-27-495-2023}

\bibitem[\citeproctext]{ref-fekete2007}
Fekete, B. M. et al. (2007). The current status of global river
discharge monitoring and potential new technologies complementing
traditional discharge measurements.

\bibitem[\citeproctext]{ref-ficklin2016}
Ficklin, D. L. et al. (2016). Impacts of recent climate change on trends
in baseflow and stormflow in United States watersheds. \emph{Geophysical
Research Letters}, \emph{43}(10), 5079--5088.
\url{https://doi.org/10.1002/2016GL069121}

\bibitem[\citeproctext]{ref-fuka2014ecohydrology}
Fuka, D. et al. (2014). EcoHydRology: A community modeling foundation
for eco-hydrology. \emph{R Package Version 0.4}, \emph{12}.

\bibitem[\citeproctext]{ref-Georganos2018Very}
Georganos, S. et al. (2018). Very high resolution object-based land
use--land cover urban classification using extreme gradient boosting.
\emph{IEEE Geoscience and Remote Sensing Letters}, \emph{15}, 607--611.
\url{https://doi.org/10.1109/LGRS.2018.2803259}

\bibitem[\citeproctext]{ref-gonzales2009}
Gonzales, A. L. et al. (2009). Comparison of different base flow
separation methods in a lowland catchment. \emph{Hydrology and Earth
System Sciences}, \emph{13}(11), 2055--2068.
\url{https://doi.org/10.5194/hess-13-2055-2009}

\bibitem[\citeproctext]{ref-instituteofhydrology1980}
Institute of Hydrology. (1980). \emph{Low flow studies report 3
catchment characteristic estimation manual}.

\bibitem[\citeproctext]{ref-iucn-drylands-2019}
IUCN. (2019, September). Drylands and climate change. Retrieved from
\url{https://www.iucn.org/resources/issues-brief/drylands-and-climate-change}

\bibitem[\citeproctext]{ref-kendall1970}
Kendall, M. G. (1970). \emph{Rank correlation methods} (4th ed). London:
Griffin.

\bibitem[\citeproctext]{ref-epa_AZephemeral}
Levick, L. R. et al. (2008). {The ecological and hydrological
significance of ephemeral and intermittent streams in the arid and
semi-arid American Southwest}. \emph{U.S. Environmental Protection
Agency and USDA/ARS Southwest Watershed Research Center},
(EPA/600/R-08/134).

\bibitem[\citeproctext]{ref-lyne1979}
Lyne, V. et al. (1979). Institute of engineers australia national
conference. In (Vol. 79, pp. 89--93). Barton, Australia: Institute of
Engineers Australia.

\bibitem[\citeproctext]{ref-mann1945nonparametric}
Mann, H. B. (1945). Nonparametric tests against trend.
\emph{Econometrica: Journal of the Econometric Society}, 245--259.

\bibitem[\citeproctext]{ref-nathan1990}
Nathan, R. J. et al. (1990). Evaluation of automated techniques for base
flow and recession analyses. \emph{Water Resources Research},
\emph{26}(7), 1465--1473. \url{https://doi.org/10.1029/WR026i007p01465}

\bibitem[\citeproctext]{ref-neff2005}
Neff, B. P. et al. (2005). \emph{Base flow in the great lakes basin}
(Report No. 2005-5217). Reston, VA.
\url{https://doi.org/10.3133/sir20055217}

\bibitem[\citeproctext]{ref-ni2020streamflow}
Ni, L. et al. (2020). Streamflow forecasting using extreme gradient
boosting model coupled with gaussian mixture model. \emph{Journal of
Hydrology}, \emph{586}, 124901.

\bibitem[\citeproctext]{ref-odonnell2016}
O'Donnell, F. C. et al. (2016). 12th Biennial Conference of Research on
the Colorado Plateau. In.

\bibitem[\citeproctext]{ref-reitz2017}
Reitz, M. et al. (2017). Annual Estimates of Recharge, Quick{-}Flow
Runoff, and Evapotranspiration for the Contiguous US Using Empirical
Regression Equations. \emph{JAWRA Journal of the American Water
Resources Association}, \emph{53}(4), 961--983.
\url{https://doi.org/10.1111/1752-1688.12546}

\bibitem[\citeproctext]{ref-Rozos2021Machine}
Rozos, E. et al. (2021). Machine learning in assessing the performance
of hydrological models. \emph{Hydrology}.
\url{https://doi.org/10.3390/hydrology9010005}

\bibitem[\citeproctext]{ref-santhi2008}
Santhi, C. et al. (2008). Regional estimation of base flow for the
conterminous United States by hydrologic landscape regions.
\emph{Journal of Hydrology}, \emph{351}(1-2), 139--153.
\url{https://doi.org/10.1016/j.jhydrol.2007.12.018}

\bibitem[\citeproctext]{ref-scanlon2006}
Scanlon, B. R. et al. (2006). Global synthesis of groundwater recharge
in semiarid and arid regions. \emph{Hydrological Processes},
\emph{20}(15), 3335--3370. \url{https://doi.org/10.1002/hyp.6335}

\bibitem[\citeproctext]{ref-schmidt_ML_2020}
Schmidt, L. et al. (2020). Challenges in applying machine learning
models for hydrological inference: A case study for flooding events
across germany. \emph{Water Resources Research}, \emph{56}(5),
e2019WR025924.
https://doi.org/\url{https://doi.org/10.1029/2019WR025924}

\bibitem[\citeproctext]{ref-sheppard2002}
Sheppard, P. et al. (2002). The climate of the US Southwest.
\emph{Climate Research}, \emph{21}, 219--238.
\url{https://doi.org/10.3354/cr021219}

\bibitem[\citeproctext]{ref-singh2018}
Singh, S. K. et al. (2018). Towards baseflow index characterisation at
national scale in New Zealand. \emph{Journal of Hydrology}, \emph{568},
646--657. \url{https://doi.org/10.1016/j.jhydrol.2018.11.025}

\bibitem[\citeproctext]{ref-sloto1996}
Sloto, R. et al. (1996). \emph{HYSEP: A Computer Program for Streamflow
Hydrograph Separation and Analysis} (No. 96-4040). {USGS}.
\url{https://doi.org/10.3133/wri964040}

\bibitem[\citeproctext]{ref-szczepanek2022daily}
Szczepanek, R. (2022). Daily streamflow forecasting in mountainous
catchment using XGBoost, LightGBM and CatBoost. \emph{Hydrology},
\emph{9}(12), 226.

\bibitem[\citeproctext]{ref-tan2020}
Tan, X. et al. (2020). Global Changes in Baseflow Under the Impacts of
Changing Climate and Vegetation. \emph{Water Resources Research},
\emph{56}(9), e2020WR027349. \url{https://doi.org/10.1029/2020WR027349}

\bibitem[\citeproctext]{ref-taylor2013}
Taylor, R. G. et al. (2013). Ground water and climate change.
\emph{Nature Climate Change}, \emph{3}(4), 322--329.
\url{https://doi.org/10.1038/nclimate1744}

\bibitem[\citeproctext]{ref-usgs2010}
USGS. (2010). Water resources data for the united states water year
2007. \url{http://wdr.water.usgs.gov/wy2007/search.jsp}.

\bibitem[\citeproctext]{ref-usgs-glossary-2018}
USGS. (2018). Water science glossary. Retrieved from
\url{https://www.usgs.gov/special-topics/water-science-school/science/water-science-glossary}

\bibitem[\citeproctext]{ref-wolock2003}
Wolock, D. M. (2003). Estimated mean annual natural ground-water
recharge in the conterminous united states.
\url{https://doi.org/10.5066/P9FSSVF3}.

\bibitem[\citeproctext]{ref-yao2018}
Yao, Y. et al. (2018). Role of Groundwater in the Dryland
Ecohydrological System: A Case Study of the Heihe River Basin.
\emph{Journal of Geophysical Research: Atmospheres}, \emph{123}(13),
6760--6776. \url{https://doi.org/10.1029/2018JD028432}

\bibitem[\citeproctext]{ref-zomer2022}
Zomer, R. J. et al. (2022). Version 3 of the Global Aridity Index and
Potential Evapotranspiration Database. \emph{Scientific Data},
\emph{9}(1), 409. \url{https://doi.org/10.1038/s41597-022-01493-1}

\end{CSLReferences}

\end{document}

### 2024-11-26 index.qmd
sdf
Update index.qmd

@@
 We examined the presence of temporal trends in mean annual BFI using the Mann-Kendall nonparametric trend test [@kendall1970; @mann1945nonparametric].
-The Mann-Kendall was used to determine trends in BFI and trend significance of gauged reaches during their period of record.
+The Mann-Kendall test was used to determine trends in BFI and trend significance of gauged reaches.
 
@@
 
+| Classification Group         |   n | n_pos | n_neg | perc_pos | perc_neg |
+|:-----------------------------|----:|------:|------:|---------:|---------:|
+| Climate - Monsoon Dominated  |  94 |    10 |    23 |    0.106 |    0.245 |
+| Climate - Snowmelt Dominated | 150 |    19 |    17 |    0.127 |    0.113 |
+| PhysRegion - Basin&Range     | 190 |    26 |    33 |    0.137 |    0.174 |
+| PhysRegion - CO Plateau      |  54 |     3 |     7 |    0.056 |    0.130 |
+| Climate - Warm-Wet           |  31 |     3 |     6 |    0.097 |    0.194 |
+| Climate - Warm-Dry           |  81 |    15 |    16 |    0.185 |    0.198 |
+| Climate - Cool-Wet           |  83 |     6 |    11 |    0.072 |    0.133 |
+| Climate - Cool-Dry           |  49 |     5 |     7 |    0.102 |    0.143 |
+| Slope - High                 | 111 |    12 |    15 |    0.108 |    0.135 |
+| Slope - Low                  | 133 |    17 |    25 |    0.128 |    0.188 |
+
+: Table 2.
+Comparison of trends for BFI for all sites split by various classifications.
+Only sites with a significant ($\rho \le 0.05$) trend are included here as established by a Mann-Kendall test for monotonic trends across the full period of record.
+n is the number of sites, n_pos (n_neg) is the number of sites with positive (negative) trends, perc_pos (perc_neg) is the percentage of n with a positive (negative) trend.
+{#tbl-trends}
+
 ### BFI of Ungauged Catchments
@@
 
-: Table 2.
+: Table 3.
 Performance of model predictions for BFI for all sites split by various classifications.
-n is number of sites, R^2^ is the coefficient of determination of a linear regression, MSE is mean-squared-error, RMSE is root-mean-squared-error, MAE is mean-absolute-error, NSE is Nash-Sucliffe efficiency, and pbias is percent bias.
+n is number of observatios, R^2^ is the coefficient of determination of a linear regression, MSE is mean-squared-error, RMSE is root-mean-squared-error, MAE is mean-absolute-error, NSE is Nash-Sucliffe efficiency, and pbias is percent bias.
 {#tbl-performance}

### 2024-11-26 index.qmd
New

@@
 
-: **Table 1.** Basin-characteristic variables used as predictors in XGBoost model. {#tbl-predictors}
+: Basin-characteristic variables used as predictors in XGBoost model. {#tbl-predictors}
 
@@
 
-The long-term average BFI for the 245 gauged reaches across Arizona is illustrated in @fig-instrumented-bfi.The highest BFI values (\>0.9) are found along the Grand Canyon in northwestern Arizona.
+The long-term average BFI for the 244 gauged reaches across Arizona is illustrated in @fig-instrumented-bfi .
+The highest BFI values (\>0.9) are found along the Grand Canyon in northwestern Arizona.
 The highly karstic geology of this region facilitates the rapid movement of subsurface flow to surface water and spring outlets [@chambless_deep-karst_2023].
@@
 
+![](images/BFITrends_significant.png)
+
 | Classification Group         |   n | n_pos | n_neg | perc_pos | perc_neg |
@@
 
-: Table 2.
-Comparison of trends for BFI for all sites split by various classifications.
+: Comparison of trends for BFI for all sites split by various classifications.
 Only sites with a significant ($\rho \le 0.05$) trend are included here as established by a Mann-Kendall test for monotonic trends across the full period of record.
@@
 
-: Table 3.
-Performance of model predictions for BFI for all sites split by various classifications.
+: Performance of model predictions for BFI for all sites split by various classifications.
 n is number of observatios, R^2^ is the coefficient of determination of a linear regression, MSE is mean-squared-error, RMSE is root-mean-squared-error, MAE is mean-absolute-error, NSE is Nash-Sucliffe efficiency, and pbias is percent bias.
 {#tbl-performance}
+
+Consistently negative pbias values across all classifications suggest an under-prediction of BFI.
 

### 2024-11-26 index.qmd
Updating figures

@@
 
-![Map of Arizona indicating US Geological Survey (USGS) streamgages used in this study. 8-digit HUC subbasin boundaries and physiographic regions shown.](images/StudyArea.png){#fig-study-area}
+![Map of Arizona indicating US Geological Survey (USGS) streamgages used in this study.
+8-digit HUC subbasin boundaries and physiographic regions shown.](images/StudyArea_20241126.png){#fig-study-area}
 
@@
 
-![Mean annual BFI for the period of study (1991-2020) from instrumented stream flow data.](images/BFI_Instrumented.png){#fig-instrumented-bfi}
+![Mean annual BFI for the period of study (1991-2020) from instrumented stream flow data.](images/BFI_Instrumented_20241126.png){#fig-instrumented-bfi}
 
@@
 Red upward (blue downward) arrows indicate an increasing (decreasing) trend at a significance level of 5%.
-Nonsignificant trends are shown by colors of circles.](images/BFITrends.png){#fig-instrumented-trend}
-
-![](images/BFITrends_significant.png)
+White circles represent sites with no statistically significant trends.](images/BFITrends_20241126.png){#fig-instrumented-trend}
 
@@
 
-![Predicted mean annual BFI values for 8-digit HUC (1991-2020)](images/BFI_HUC8.png){#fig-bfi-huc}
+![Predicted mean annual BFI values for 8-digit HUC (1991-2020)](images/BFI_HUC8_20241126.png){#fig-bfi-huc}
 

### 2024-11-26 index.tex
Updating figures

@@
 
-\includegraphics{images/StudyArea.png}
+\includegraphics{images/StudyArea_20241126.png}
 
@@
   >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.3243}}@{}}
-\caption{\textbf{Table 1.} Basin-characteristic variables used as
-predictors in XGBoost model.}\label{tbl-predictors}\tabularnewline
+\caption{Basin-characteristic variables used as predictors in XGBoost
+model.}\label{tbl-predictors}\tabularnewline
 \toprule\noalign{}
@@
 mean annual BFI using the Mann-Kendall nonparametric trend test
-(Kendall, 1970; Mann, 1945). The Mann-Kendall was used to determine
-trends in BFI and trend significance of gauged reaches during their
-period of record.
+(Kendall, 1970; Mann, 1945). The Mann-Kendall test was used to determine
+trends in BFI and trend significance of gauged reaches.
 
@@
 
-\includegraphics{images/BFI_Instrumented.png}
+\includegraphics{images/BFI_Instrumented_20241126.png}
 
@@
 
-The long-term average BFI for the 245 gauged reaches across Arizona is
-illustrated in \textbf{?@fig-instrumented-bfi.The} highest BFI values
-(\textgreater0.9) are found along the Grand Canyon in northwestern
-Arizona. The highly karstic geology of this region facilitates the rapid
-movement of subsurface flow to surface water and spring outlets
-(Chambless et al., 2023). Relatively high BFI values (\textgreater0.8)
-are found at the spring-fed headwaters of the Verde River (Del Rio
-Spring) and the spring-fed headwaters of Fossil Creek. These results are
-consistent with interpolated BFI values reported by Wolock (2003).
+The long-term average BFI for the 244 gauged reaches across Arizona is
+illustrated in Figure~\ref{fig-instrumented-bfi} . The highest BFI
+values (\textgreater0.9) are found along the Grand Canyon in
+northwestern Arizona. The highly karstic geology of this region
+facilitates the rapid movement of subsurface flow to surface water and
+spring outlets (Chambless et al., 2023). Relatively high BFI values
+(\textgreater0.8) are found at the spring-fed headwaters of the Verde
+River (Del Rio Spring) and the spring-fed headwaters of Fossil Creek.
+These results are consistent with interpolated BFI values reported by
+Wolock (2003).
 
@@
 
-\includegraphics{images/BFITrends.png}
+\includegraphics{images/BFITrends_20241126.png}
 
@@
 downward) arrows indicate an increasing (decreasing) trend at a
-significance level of 5\%. Nonsignificant trends are shown by colors of
-circles.}
+significance level of 5\%. White circles represent sites with no
+statistically significant trends.}
 
 \end{figure}%
+
+\begin{longtable}[]{@{}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.4348}}
+  >{\raggedleft\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.0725}}
+  >{\raggedleft\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1014}}
+  >{\raggedleft\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1014}}
+  >{\raggedleft\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1449}}
+  >{\raggedleft\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1449}}@{}}
+\caption{Comparison of trends for BFI for all sites split by various
+classifications. Only sites with a significant (\(\rho \le 0.05\)) trend
+are included here as established by a Mann-Kendall test for monotonic
+trends across the full period of record. n is the number of sites,
+n\_pos (n\_neg) is the number of sites with positive (negative) trends,
+perc\_pos (perc\_neg) is the percentage of n with a positive (negative)
+trend.}\label{tbl-trends}\tabularnewline
+\toprule\noalign{}
+\begin{minipage}[b]{\linewidth}\raggedright
+Classification Group
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
+n
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
+n\_pos
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
+n\_neg
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
+perc\_pos
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
+perc\_neg
+\end{minipage} \\
+\midrule\noalign{}
+\endfirsthead
+\toprule\noalign{}
+\begin{minipage}[b]{\linewidth}\raggedright
+Classification Group
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
+n
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
+n\_pos
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
+n\_neg
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
+perc\_pos
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
+perc\_neg
+\end{minipage} \\
+\midrule\noalign{}
+\endhead
+\bottomrule\noalign{}
+\endlastfoot
+Climate - Monsoon Dominated & 94 & 10 & 23 & 0.106 & 0.245 \\
+Climate - Snowmelt Dominated & 150 & 19 & 17 & 0.127 & 0.113 \\
+PhysRegion - Basin\&Range & 190 & 26 & 33 & 0.137 & 0.174 \\
+PhysRegion - CO Plateau & 54 & 3 & 7 & 0.056 & 0.130 \\
+Climate - Warm-Wet & 31 & 3 & 6 & 0.097 & 0.194 \\
+Climate - Warm-Dry & 81 & 15 & 16 & 0.185 & 0.198 \\
+Climate - Cool-Wet & 83 & 6 & 11 & 0.072 & 0.133 \\
+Climate - Cool-Dry & 49 & 5 & 7 & 0.102 & 0.143 \\
+Slope - High & 111 & 12 & 15 & 0.108 & 0.135 \\
+Slope - Low & 133 & 17 & 25 & 0.128 & 0.188 \\
+\end{longtable}
 
@@
   >{\raggedright\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.0886}}@{}}
-\caption{Table 2. Performance of model predictions for BFI for all sites
-split by various classifications. n is number of sites,
+\caption{Performance of model predictions for BFI for all sites split by
+various classifications. n is number of observatios,
 R\textsuperscript{2} is the coefficient of determination of a linear
@@
 
+Consistently negative pbias values across all classifications suggest an
+under-prediction of BFI.
+
 \subsubsection{Predictor Importance}\label{sec-predictor-importance}
@@
 
-\includegraphics{images/BFI_HUC8.png}
+\includegraphics{images/BFI_HUC8_20241126.png}
 

### 2024-11-29 index.qmd
new words
updating trend stuff

@@
 
+These trends were produced using a Mann-Kendall test for monotonic trends.
+This test expects that the data is non-parametric and that the observations are not autocorrelated over time.
+Observations of BFI at 4 streamgages show significant autocorrelation, though only one streamgage (09486500 - Santa Cruz River at Cortaro, AZ) indicated a significant trend in BFI.
+Due to this autocorrelation, the streamgage was removed from the trend analysis as the autocorrelation may inflate the variance of the Mann-Kendall statistic leading to an over/underestimate of the trend (CITATION).
+
 Trends with a $\rho \le 0.05$ are considered significant.
@@
 
+Trends in BFI over the period of record for each streamgage in this analysis are illustrated in @fig-instrumented-trend and @tbl-trends .
+
 ![Trends in BFI over full period of record for instrumented sites used in this study.
@@
 |:-----------------------------|----:|------:|------:|---------:|---------:|
-| Climate - Monsoon Dominated  |  94 |    10 |    23 |    0.106 |    0.245 |
+| Climate - Monsoon Dominated  |  94 |     9 |    23 |    0.096 |    0.245 |
 | Climate - Snowmelt Dominated | 150 |    19 |    17 |    0.127 |    0.113 |
-| PhysRegion - Basin&Range     | 190 |    26 |    33 |    0.137 |    0.174 |
+| PhysRegion - Basin&Range     | 190 |    25 |    33 |    0.132 |    0.174 |
 | PhysRegion - CO Plateau      |  54 |     3 |     7 |    0.056 |    0.130 |
-| Climate - Warm-Wet           |  31 |     3 |     6 |    0.097 |    0.194 |
+| Climate - Warm-Wet           |  31 |     2 |     6 |    0.065 |    0.194 |
 | Climate - Warm-Dry           |  81 |    15 |    16 |    0.185 |    0.198 |
@@
 | Slope - High                 | 111 |    12 |    15 |    0.108 |    0.135 |
-| Slope - Low                  | 133 |    17 |    25 |    0.128 |    0.188 |
+| Slope - Low                  | 133 |    16 |    25 |    0.120 |    0.188 |
 

### 2024-12-02 index.qmd
Update index.qmd

@@
 ![Map of Arizona indicating US Geological Survey (USGS) streamgages used in this study.
-8-digit HUC subbasin boundaries and physiographic regions shown.](images/StudyArea_20241126.png){#fig-study-area}
+8-digit HUC subbasin boundaries and physiographic regions shown.](images/StudyArea_20241127.png){#fig-study-area}
 
@@
 This base-flow separation technique is based on a recursive digital filter used in signal analysis that separates high-frequency signals (quickflow) from low-frequency signals (base flow) [@lyne1979].
+@eckhardt2023 noted that recursive digital filters lack a physical basis, but as the method is easy to automate, objective, and repeatable, it is appropriate for a regional-scale study.
 This technique has been used in [@ARNOLD200021; @santhi2008; @bloomfield2009; @singh2018].
@@
 The filter was run three times (forward, backward, forward) to attenuate the base-flow signal.
-@eckhardt2023 noted that recursive digital filters lack a physical basis, but as the method is easy to automate, objective, and repeatable, it is appropriate for a regional-scale study.
 
@@
 
-![Mean annual BFI for the period of study (1991-2020) from instrumented stream flow data.](images/BFI_Instrumented_20241126.png){#fig-instrumented-bfi}
+![Mean annual BFI for the period of record from instrumented stream flow data.](images/BFI_Instrumented_20241126.png){#fig-instrumented-bfi}
 
@@
 
-| Classification Group         |   n | n_pos | n_neg | perc_pos | perc_neg |
-|:-----------------------------|----:|------:|------:|---------:|---------:|
-| Climate - Monsoon Dominated  |  94 |     9 |    23 |    0.096 |    0.245 |
-| Climate - Snowmelt Dominated | 150 |    19 |    17 |    0.127 |    0.113 |
-| PhysRegion - Basin&Range     | 190 |    25 |    33 |    0.132 |    0.174 |
-| PhysRegion - CO Plateau      |  54 |     3 |     7 |    0.056 |    0.130 |
+| Group                              | n   | n_pos | n_neg | perc_pos | perc_neg |
+|------------------------------------|-----|-------|-------|----------|----------|
+| Precipitation - Monsoon Dominated  | 87  | 8     | 21    | 0.092    | 0.241    |
+| Precipitation - Snowmelt Dominated | 118 | 9     | 12    | 0.076    | 0.102    |
+| PhysRegion - Basin&Range           | 156 | 14    | 26    | 0.090    | 0.167    |
+| PhysRegion - CO Plateau            | 49  | 3     | 7     | 0.061    | 0.143    |
 | Climate - Warm-Wet                 | 31  | 2     | 6     | 0.065    | 0.194    |
-| Climate - Warm-Dry           |  81 |    15 |    16 |    0.185 |    0.198 |
-| Climate - Cool-Wet           |  83 |     6 |    11 |    0.072 |    0.133 |
-| Climate - Cool-Dry           |  49 |     5 |     7 |    0.102 |    0.143 |
-| Slope - High                 | 111 |    12 |    15 |    0.108 |    0.135 |
-| Slope - Low                  | 133 |    16 |    25 |    0.120 |    0.188 |
+| Climate - Warm-Dry                 | 55  | 6     | 11    | 0.109    | 0.200    |
+| Climate - Cool-Wet                 | 74  | 4     | 9     | 0.054    | 0.122    |
+| Climate - Cool-Dry                 | 45  | 5     | 7     | 0.111    | 0.156    |
+| Slope - High                       | 102 | 10    | 12    | 0.098    | 0.118    |
+| Slope - Low                        | 103 | 7     | 21    | 0.068    | 0.204    |
 
@@
 
-|                              | n    | R^2^  | MSE    | RMSE  | MAE   | NSE   | pbias |
-|------------------------------|------|-------|--------|-------|-------|-------|-------|
-| Climate - Monsoon Dominated  | 3345 | 0.707 | 0.0166 | 0.129 | 0.077 | 0.694 | -12.6 |
-| Climate - Snowmelt Dominated | 6152 | 0.723 | 0.017  | 0.131 | 0.094 | 0.715 | -3.7  |
+| Classification Group               | n    | R^2^  | MSE    | RMSE  | MAE   | NSE   | pbias |
+|----------------|--------|--------|--------|--------|--------|--------|--------|
+| Precipitation - Monsoon Dominated  | 3345 | 0.707 | 0.0166 | 0.129 | 0.077 | 0.694 | -12.6 |
+| Precipitation - Snowmelt Dominated | 6152 | 0.723 | 0.017  | 0.131 | 0.094 | 0.715 | -3.7  |
 | PhysRegion - Basin&Range           | 7754 | 0.749 | 0.018  | 0.133 | 0.089 | 0.741 | -5.9  |

### 2024-12-02 index.qmd
Update index.qmd

@@
 
-| Group                              | n   | n_pos | n_neg | perc_pos | perc_neg |
+| Classification Group               | n   | n_pos | n_neg | perc_pos | perc_neg |
 |------------------------------------|-----|-------|-------|----------|----------|
@@
 
-| Classification Group               | n    | R^2^  | MSE    | RMSE  | MAE   | NSE   | pbias |
-|----------------|--------|--------|--------|--------|--------|--------|--------|
-| Precipitation - Monsoon Dominated  | 3345 | 0.707 | 0.0166 | 0.129 | 0.077 | 0.694 | -12.6 |
-| Precipitation - Snowmelt Dominated | 6152 | 0.723 | 0.017  | 0.131 | 0.094 | 0.715 | -3.7  |
-| PhysRegion - Basin&Range           | 7754 | 0.749 | 0.018  | 0.133 | 0.089 | 0.741 | -5.9  |
-| PhysRegion - CO Plateau            | 1743 | 0.806 | 0.013  | 0.115 | 0.081 | 0.801 | -4.4  |
-| Climate - Warm-Wet                 | 1613 | 0.690 | 0.015  | 0.124 | 0.081 | 0.679 | -9.9  |
-| Climate - Warm-Dry                 | 3134 | 0.758 | 0.023  | 0.151 | 0.097 | 0.745 | -8.1  |
-| Climate - Cool-Wet                 | 3134 | 0.710 | 0.013  | 0.115 | 0.085 | 0.708 | -2.5  |
-| Climate - Cool-Dry                 | 1616 | 0.809 | 0.014  | 0.118 | 0.082 | 0.803 | -5    |
-| Slope - High                       | 4398 | 0.753 | 0.014  | 0.120 | 0.084 | 0.746 | -4.4  |
-| Slope - Low                        | 5099 | 0.759 | 0.019  | 0.138 | 0.090 | 0.750 | -6.9  |
+| Classification Group           |  n    |   R2  |  MSE  |  RMSE |  MAE  | NSE            | pbias |
+|---------------------------------|------|-------|-------|-------|-------|----------------|-------|
+| Climate - Monsoon Dominated     | 3039 | 0.633 | 0.016 | 0.126 | 0.074 | 0.619          | -13.7 |
+| Climate - Snowmelt Dominated    | 4685 | 0.733 | 0.015 | 0.121 | 0.087 | 0.725          |  -3.5 |
+| PhysRegion - Basin&Range        | 6147 | 0.733 | 0.016 | 0.127 | 0.084 | 0.724          |  -6.3 |
+| PhysRegion - CO Plateau         | 1577 | 0.846 | 0.011 | 0.104 | 0.073 | 0.843          |  -3.8 |
+| Climate - Warm-Wet              | 1506 | 0.693 | 0.014 | 0.117 | 0.077 | 0.685          |  -8.2 |
+| Climate - Warm-Dry              | 2351 | 0.693 | 0.022 | 0.147 | 0.092 | 0.675          | -11.9 |
+| Climate - Cool-Wet              | 2350 | 0.738 | 0.011 | 0.106 | 0.078 | 0.736          |  -1.7 |
+| Climate - Cool-Dry              | 1517 | 0.831 | 0.012 | 0.111 | 0.078 | 0.827          |  -4.3 |
+| Slope - High                    | 3795 | 0.776 | 0.012 | 0.111 | 0.079 | 0.771          |  -3.3 |
+| Slope - Low                     | 3929 | 0.724 | 0.018 | 0.133 | 0.085 | 0.713          |  -9.1 |
 

### 2024-12-02 index.qmd
Update index.qmd

@@
 Finally, streamgages on the Colorado River were excluded from the dataset on the basis that they represent managed flows shared by Colorado River Compact states.
-Following these parameters, 206 USGS streamgages with acceptable periods of record were used in this study (@fig-study-area).
+Following these parameters, 205 USGS streamgages with acceptable periods of record were used in this study (@fig-study-area).
 
@@
 
-The long-term average BFI for the 244 gauged reaches across Arizona is illustrated in @fig-instrumented-bfi .
+The long-term average BFI for the 205 gauged reaches across Arizona is illustrated in @fig-instrumented-bfi .
 The highest BFI values (\>0.9) are found along the Grand Canyon in northwestern Arizona.

### 2024-12-03 index.qmd
updating

@@
 ![Map of Arizona indicating US Geological Survey (USGS) streamgages used in this study.
-8-digit HUC subbasin boundaries and physiographic regions shown.](images/StudyArea_20241127.png){#fig-study-area}
+8-digit HUC subbasin boundaries and physiographic regions shown.](images/StudyArea_20241203.png){#fig-study-area}
 
@@
 
-![Mean annual BFI for the period of record from instrumented stream flow data.](images/BFI_Instrumented_20241126.png){#fig-instrumented-bfi}
+![Mean annual BFI for the period of record from instrumented stream flow data.](images/BFI_Instrumented_20241203.png){#fig-instrumented-bfi}
 
@@
 Red upward (blue downward) arrows indicate an increasing (decreasing) trend at a significance level of 5%.
-White circles represent sites with no statistically significant trends.](images/BFITrends_20241126.png){#fig-instrumented-trend}
+White circles represent sites with no statistically significant trends.](images/BFITrends_20241203.png){#fig-instrumented-trend}
 
@@
 
+![Linear relationship between observed BFI and predicted BFI. The solid line is the 1:1 line, the dashed line is regressed to the data.](images/actual-predicted.png){#fig-actual_predicted}
+
 #### Model Validation
@@
 | Classification Group         | n    | R2    | MSE   | RMSE  | MAE   | NSE   | pbias |
-|---------------------------------|------|-------|-------|-------|-------|----------------|-------|
+|------------------------------|------|-------|-------|-------|-------|-------|-------|
 | Climate - Monsoon Dominated  | 3039 | 0.633 | 0.016 | 0.126 | 0.074 | 0.619 | -13.7 |
@@
 
-![Predicted mean annual BFI values for 8-digit HUC (1991-2020)](images/BFI_HUC8_20241126.png){#fig-bfi-huc}
+![Predicted mean annual BFI values for 8-digit HUC (1991-2020)](images/BFI_HUC8_20241203.png){#fig-bfi-huc}
 
-![Predicted mean annual BFI values for all reaches of Strahler stream order 3 and greater](images/BFI_AllStream.pdf){#fig-bfi-streams}
+![Predicted mean annual BFI values for all reaches of Strahler stream order 3 and greater](images/BFI_AllStream.png){#fig-bfi-streams}
 

### 2024-12-07 index.qmd
k

@@
 
-This study develops a technique for estimating BFI in ungauged basins in Arizona and determines the state’s long-term mean BFI from 1991 to 2020.
-Due to interannual variability in BFI, a 30-year period of record is used to evaluate the characteristic base-flow dynamics of Arizona.
-Using a machine learning model trained with the hydrogeologic characteristics of surface water basins, we first determine annual BFIs for each year in the 1991-2020 study period for ungauged river reaches and physiographic regions of Arizona.
-Long-term mean BFI data for the study period are then estimated.
-Regional patterns and trends in base flow of instrumented sites are subsequently investigated.
-These results provide novel analyses of low flow processes in Arizona...######
+This study develops a technique for estimating BFI in ungauged basins in Arizona and determines the state’s long-term mean BFI.
+Using a machine learning model trained with the hydrogeologic characteristics of surface water basins, we estimate mean BFI in ungauged basins over a period of record from 1991-2020.
+Regional patterns and trends in base flow and BFI of instrumented sites are subsequently investigated.
+These results provide novel analyses of low flow processes in Arizona, furthering our understanding of climate controls on consistent flows in dryland basins.
 
@@
 The state of Arizona, located in the southwest United States, covers a total area of 295,253 km$^2$.
-Arizona is divided into two primary physiographic provinces: the Colorado Plateau in the northeast, and the Basin-and-Range region in the northwest and south.
+Arizona is divided into two primary physiographic provinces: the Colorado Plateau in the northeast, and the Basin-and-Range region in the west and south.
 The Central Highlands is a transition zone consisting of scattered basins separated by the mountainous foothills of the Mogollon Rim.
 The Colorado Plateau is dominated by high-elevation desert with an average elevation of 1,936 masl (6,352 ft).
-Average temperatures range from -6$^\circ$C (20$^\circ$F) to 26$^\circ$C (80$^\circ$F) and it averages 580 mm (23 in) of precipitation.
+Mean temperatures range from -6$^\circ$C (20$^\circ$F) to 26$^\circ$C (80$^\circ$F) and it averages 580 mm (23 in) of precipitation.
 The Basin-and-Range region has a semi-arid to arid climate with an average elevation of 490 masl (1600 ft).
-The average precipitation is 200 mm (8 in) annually with average temperatures ranging from 15$^\circ$C (60$^\circ$F) to 43$^\circ$C (110$^\circ$F).
+Average temperature ranges from 15$^\circ$C (60$^\circ$F) to 43$^\circ$C (110$^\circ$F) and the region averages 200 mm (8 in) of precipitation annually.
+###CITATION###
 
@@
 Much of the hydrology of Arizona is snow-melt derived, driven by spring melt from the high-elevation Colorado Plateau winter snowpack.
-While winter precipitation provides only 30% of annual averages, it provides the bulk of water for natural reservoirs [@sheppard2002].
+While winter precipitation provides only 30% of annual averages, it provides the majority of water for natural reservoirs [@sheppard2002].
 
@@
 Following these parameters, 205 USGS streamgages with acceptable periods of record were used in this study (@fig-study-area).
+Periods of record ranged from 10 to 112 years, with a median of 28 years.
 
@@
 These HUC 8 sub-basins are analogous to medium-sized river basins and are defined by surface water characteristics.
-\######
 
-The annual precipitation and temperature data came from the PRISM climate group at Oregon State University at a resolution of 4 km (<https://prism.oregonstate.edu;> [@daly2008]).
+Annual precipitation and temperature data came from the PRISM climate group at Oregon State University at a resolution of 4 km (<https://prism.oregonstate.edu;> [@daly2008]).
 The PRISM dataset provides valuable insights into regional climate in ungauged regions and has been shown to perform well across the southwestern US [@buban_PRISM]**.** Instead of the water year, PRISM data uses a calendar-year format, which was adopted for consistency in the water balance.
@@
 By calculating these percentages, we aimed to get a more comprehensive understanding of landscape composition across space.
-Land cover from USGS-NLCD (National Land Cover Database), hydrologic soil group from SSURGO (Soil Survey Geographic Database), and underlying geology/karst from USGS were all similarly averaged across the basins.
+Land cover from USGS-NLCD (National Land Cover Database), hydrologic soil group from SSURGO (Soil Survey Geographic Database), and underlying geology and karst from USGS were all similarly averaged across the basins.
 Aggregating variables to align with the HUC 8 boundaries allowed for more precise predictions of BFI by integrating spatial variations within each basin.
@@
 @eckhardt2023 noted that recursive digital filters lack a physical basis, but as the method is easy to automate, objective, and repeatable, it is appropriate for a regional-scale study.
-This technique has been used in [@ARNOLD200021; @santhi2008; @bloomfield2009; @singh2018].
+The Lyne-Hollick filter has been used in [@ARNOLD200021; @santhi2008; @bloomfield2009; @singh2018].
 It takes the form of
@@
 
-Statisical analyses were performed on mean annual BFI values of instrumented streamgages.
-We examined the presence of temporal trends in mean annual BFI using the Mann-Kendall nonparametric trend test [@kendall1970; @mann1945nonparametric].
+Statisical analyses were performed on annual BFI and base flow values of instrumented streamgages.
+We examined temporal trends using the Mann-Kendall nonparametric trend test to determine the presence of monotonic trends [@kendall1970; @mann1945nonparametric].
 The Mann-Kendall test was used to determine trends in BFI and trend significance of gauged reaches.
+This test is widely used in studies of this kind [@ficklin2016; @ayers2019; @woodhouse_udall22].
 
-These trends were produced using a Mann-Kendall test for monotonic trends.
+Trends in BFI and base flow were produced using a Mann-Kendall test for monotonic trends.
 This test expects that the data is non-parametric and that the observations are not autocorrelated over time.
-Observations of BFI at 4 streamgages show significant autocorrelation, though only one streamgage (09486500 - Santa Cruz River at Cortaro, AZ) indicated a significant trend in BFI.
+Observations at 4 streamgages show significant autocorrelation, though only one streamgage (09486500 - Santa Cruz River at Cortaro, AZ) indicated a significant trend in BFI.
 Due to this autocorrelation, the streamgage was removed from the trend analysis as the autocorrelation may inflate the variance of the Mann-Kendall statistic leading to an over/underestimate of the trend (CITATION).

### 2024-12-08 MANUSCRIPT.qmd
Writing
Methods - ML + Feature Selection

Results - Ungauged model validation + importance

--- 2024-12-07 index.qmd
@@
 The trained XGBoost model is used to predict BFI in ungauged catchments based on geospatial and hydroclimate predictor variables (@tbl-predictors).
-Certain predictors were subdivided according to their areal coverage of each basin (e.g.
+Certain features were subdivided according to their areal coverage of each basin (e.g.
 land cover was divided into 16 subdivisions).
@@
 |              | Variable                  | Source       | Geoprocessing          |
-|--------------|----------------------------|--------|------------------------|
-| Hydroclimate | Precipitation              | PRISM  | Basin average          |
-|              | Mean Temperature           | PRISM  | Basin average          |
-|              | Average Evapotranspiration | MODIS  | Basin average          |
-| Geospatial   | Elevation                  | DEM    | Basin average          |
+|------------------|-------------------|------------------|------------------|
+| Hydroclimate | Precipitation\*           | PRISM        | Basin average          |
+|              | Mean Temperature\*        | PRISM        | Basin average          |
+|              | Mean Evapotranspiration\* | TerraClimate | Basin average          |
+| Geospatial   | Elevation\*               | DEM          | Basin average          |
 |              | Area                      | DEM          | Basin average          |
@@
 |              | Aspect                    | DEM          | Percent areal coverage |
-|              | Land Cover                 | NLCD   | Percent areal coverage |
-|              | Hydrologic Soil Group      | SSURGO | Percent areal coverage |
+|              | Land Cover\*              | NLCD         | Percent areal coverage |
+|              | Hydrologic Soil Group\*   | SSURGO       | Percent areal coverage |
 |              | Geology                   | USGS         | Percent areal coverage |
@@
 
-: Basin-characteristic variables used as predictors in XGBoost model. {#tbl-predictors}
+: Basin-characteristic variables used as initial features in XGBoost model. Starred features are maintained in the final, dimensionality-reduced model. {#tbl-predictors}
+
+Our training dataset consisted of 7724 observations of 45 variables.
+We first determined the optimal values for our hyperparameters using an exhaustive grid search and 5-fold cross-validation.
+The hyperparameters explored were: learning rate (referred to as $\eta$), minimum split loss ($\gamma$), maximum tree depth, minimum child weight, and number of trees.
+While these do not represent all possible hyperparameters of XGBoost, we are confident that the range of values explored represent sufficient variation to ensure that we have found a high-performing model.
+The optimal hyperparameters were found to be 700 trees, an $\eta$ of 0.05, a $\gamma$ of 0.075, maximum tree depth of 7, and a minimum child weight of 5.
+Using these values, the XGBoost model was then trained on the dataset using 10-fold cross-validation.
+$K$-fold cross-validation provides an unbiased estimate of a model's accuracy on unseen data, while also insuring against overfitting or underfitting.
+It does this by randomly splitting the data into $k$ folds of equal size, training the model on all-but-one fold, and testing on the one held-out fold (referred to as the validation set).
+This process is then repeated $k-1$ more times.
+We note that the training process starts fresh for each repeat of the process, meaning we are effectively training $k$ independent models, with each having the same hyperparameters but different training data.
+By averaging the performance of the model over $k$ folds we ensure robustness in our prediction accuracy.
+All of our models were optimized and assessed using root mean squared error as the performance metric.
+
+#### Feature Selection
+
+Machine learning models can face issues with overfitting when provided with a large set of predicting features.
+This can cause reductions in performance on unseen data and requires greater computational power and memory storage [@feat_selec2017].
+Dimensionality reduction is a powerful tool in addressing these issues which can be divided into two broad categories: feature extraction and feature selection.
+Feature extraction involves projecting the higher dimensional dataset to a feature space with less dimensions, though this entails creating a new set of features that lose the physical meaning of the original feature space.
+On the other hand, feature selection maintains a subset of the original features, all of which retain their physical meaning providing better readability and interpretability.
+Here, a supervised features selection was used to reduce the number of predictors and, as a result, increased learning performance, decreased computational costs, and reduced overfitting.
+
+An initial model was trained on the full feature set of 45 predictors (@tbl-predictors).
+We then used a feature importance method of feature selection to remove noise and less-important features.
+Feature importance scores are an approach where importance scores are assigned to individual features indicating how much the feature contributed (positively or negatively) to the model [@ML_interpretability2019].
+Here, we use SHAP (SHapley Additive exPlanations) values as the method of assigning importance values to each feature [@SHAP_values].
+The top-10 most-important features (@suppfig-shap_values) were subset and used in training a subsequent model which saw performance increases and reduced computing time.
+This subsequent model was used as the final model in this analysis.
 
@@
 This test expects that the data is non-parametric and that the observations are not autocorrelated over time.
-Observations at 4 streamgages show significant autocorrelation, though only one streamgage (09486500 - Santa Cruz River at Cortaro, AZ) indicated a significant trend in BFI.
-Due to this autocorrelation, the streamgage was removed from the trend analysis as the autocorrelation may inflate the variance of the Mann-Kendall statistic leading to an over/underestimate of the trend (CITATION).
+Using a Durbin-Watson test, observations at 4 streamgages show significant autocorrelation, though only one streamgage (09486500 - Santa Cruz River at Cortaro, AZ) indicated a significant trend in BFI.
+Due to this autocorrelation, the streamgage was removed from the trend analysis as the autocorrelation may inflate the variance of the Mann-Kendall statistic leading to an over/underestimate of the trend [@MK_Hamed1998].
 
@@
 
+#### Model Validation
+
+Predicted values of BFI are plotted against observed values for the entire period of record of the instrumented dataset in @fig-actual_predicted .
+The agreement between 'out of bag' (blind cross-validation of the model as if each site were ungauged) predicted and observed values is acceptable (R^2^ = 0.764) and indicates that the model performs well across the full dataset.
+The overall Root Mean Square Error (RMSE) is 0.129 and the overall percent bias (pbias) is -5.6.
+Model performance metrics across various classifications are summarized in @tbl-performance .
+The various performance metrics indicate that this regional model performs well across different spatial and climate classifications.
+Negative pbias values across all classifications, along with the overall pbias, suggest that the model consistently under-predicts BFI.
+Categories with relatively poor performances of R^2^ and Nash-Sutcliffe Efficiency (NSE) also show relatively large biases.
+
 ![Linear relationship between observed BFI and predicted BFI. The solid line is the 1:1 line, the dashed line is regressed to the data.](images/actual-predicted.png){#fig-actual_predicted}
-
-#### Model Validation
 
@@
 
-Consistently negative pbias values across all classifications suggest an under-prediction of BFI.
+#### Predictor Importance {#sec-predictor-importance}
 
-#### Predictor Importance {#sec-predictor-importance}
+The predictors used to determine ungauged BFI were assessed to determine their importance.
+@suppfig-shap_values indicates the relative importance of predictors in the final XGBoost model along with the values of the .
+The most important feature in predicting long-term BFI is the elevation of the basin.
+While elevation has no physical impact on base flow characteristics, elevation has been shown to be one of the most important predictors in previous BFI studies [@singh2018; @beck2013].
+The importance of catchment elevation is likely cumulative due to lower temperatures, orographic precipitation, and the persistence of snowpack at higher elevations.
+Seasonal snowpack duration has been shown to have a significant relationship to springflow, and thus groundwater recharge, in the region [@donovan_karst_2022].
+This relationship is evident in further hydroclimate features.
+Higher temperatures tend to have a negative influence on BFI, while precipitation shows a mixed influence with some higher values negatively influencing BFI.
+This is likely due to a greater proportion of this precipitation leading to runoff, thus resulting in lower BFI values.
+
+The features describing land cover and land use indicate that a higher percentages of evergreen forest have a positive influence on predicted BFI.
+On the other hand, higher percentages shrub-land and developed land have negative influences on BFI.
+Hydrologic soil types also show strong trends in their influence on BFI.
+Soil-type C indicates soils with moderately high runoff potential when wetted, typically having 20-40% clay.
+Soil-type A has low runoff potential, facilitating rapid transmission of water through the soil profile [@nrcs2009].
+With this being the case, higher percentages of soil-type C tend to have a negative influence on BFI while soil-type A percentages show a mixed influence.
+This mixed influence is likely due to SHAP values reflecting interactions between features in the model as opposed to direct relationships between feature and prediction.
+Regions where soil-type A dominate may be found along steep slopes or low vegetation which dominate the model prediction.
+Furthermore, the relationship between soil-type A and BFI may be non-linear and more complex in regional base-flow dynamics.
 
@@
 
-![Predicted mean annual BFI values for 8-digit HUC (1991-2020)](images/BFI_HUC8_20241203.png){#fig-bfi-huc}
+The long-term BFI (1991-2020) for all stream reaches in Arizona of Strahler stream order 3 and greater are shown in @fig-bfi-streams and the regionalized (HUC-8) long-term BFI is shown in @fig-bfi-huc .
 
-![Predicted mean annual BFI values for all reaches of Strahler stream order 3 and greater](images/BFI_AllStream.png){#fig-bfi-streams}
+![Predicted long-term BFI values for 8-digit HUC (1991-2020)](images/BFI_HUC8_20241203.png){#fig-bfi-huc}
+
+![Predicted long-term BFI values for all stream reaches of Strahler stream order 3 and greater (1991-2020)](images/BFI_AllStream.png){#fig-bfi-streams}
 
@@
 
+-   This regional model, trained on a relatively sparse dataset over a region spanning 295,253 km^2^, is shown to predict BFI well spatially and across climate classifications.
+
+-
+
 ## References {.unnumbered}
 
-::: {#refs}
-:::
+## Supporting Information
+
+![SHAP value plot of features used in final model. LC indicates land cover, and soil types are defined by hydrologic soil group.](images/supplemental/shap_summary.png){#suppfig-shap_values}

### 2024-12-08 MANUSCRIPT.tex
Writing
Methods - ML + Feature Selection

Results - Ungauged model validation + importance

--- 2024-11-26 index.tex
@@
 \makeatletter
+\@ifpackageloaded{float}{}{\usepackage{float}}
+\floatstyle{plain}
+\@ifundefined{c@chapter}{\newfloat{suppfig}{h}{losuppfig}}{\newfloat{suppfig}{h}{losuppfig}[chapter]}
+\floatname{suppfig}{Figure S}
+\newcommand*\quartosuppfigref[1]{Figure \hyperref[#1]{S\ref{#1}}}
+\@ifpackageloaded{caption}{}{\usepackage{caption}}
+\DeclareCaptionLabelFormat{quartosuppfigreflabelformat}{#1#2}
+\captionsetup[suppfig]{labelformat=quartosuppfigreflabelformat}
+\newcommand*\listofsuppfigs{\listof{suppfig}{List of Supplementary Figures}}
+\makeatother
+\makeatletter
 \@ifpackageloaded{caption}{}{\usepackage{caption}}
@@
 This study develops a technique for estimating BFI in ungauged basins in
-Arizona and determines the state's long-term mean BFI from 1991 to 2020.
-Due to interannual variability in BFI, a 30-year period of record is
-used to evaluate the characteristic base-flow dynamics of Arizona. Using
-a machine learning model trained with the hydrogeologic characteristics
-of surface water basins, we first determine annual BFIs for each year in
-the 1991-2020 study period for ungauged river reaches and physiographic
-regions of Arizona. Long-term mean BFI data for the study period are
-then estimated. Regional patterns and trends in base flow of
-instrumented sites are subsequently investigated. These results provide
-novel analyses of low flow processes in Arizona\ldots\#\#\#\#\#\#
+Arizona and determines the state's long-term mean BFI. Using a machine
+learning model trained with the hydrogeologic characteristics of surface
+water basins, we estimate mean BFI in ungauged basins over a period of
+record from 1991-2020. Regional patterns and trends in base flow and BFI
+of instrumented sites are subsequently investigated. These results
+provide novel analyses of low flow processes in Arizona, furthering our
+understanding of climate controls on consistent flows in dryland basins.
 
@@
 physiographic provinces: the Colorado Plateau in the northeast, and the
-Basin-and-Range region in the northwest and south. The Central Highlands
-is a transition zone consisting of scattered basins separated by the
+Basin-and-Range region in the west and south. The Central Highlands is a
+transition zone consisting of scattered basins separated by the
 mountainous foothills of the Mogollon Rim. The Colorado Plateau is
 dominated by high-elevation desert with an average elevation of 1,936
-masl (6,352 ft). Average temperatures range from -6\(^\circ\)C
+masl (6,352 ft). Mean temperatures range from -6\(^\circ\)C
 (20\(^\circ\)F) to 26\(^\circ\)C (80\(^\circ\)F) and it averages 580 mm
 (23 in) of precipitation. The Basin-and-Range region has a semi-arid to
-arid climate with an average elevation of 490 masl (1600 ft). The
-average precipitation is 200 mm (8 in) annually with average
-temperatures ranging from 15\(^\circ\)C (60\(^\circ\)F) to 43\(^\circ\)C
-(110\(^\circ\)F).
+arid climate with an average elevation of 490 masl (1600 ft). Average
+temperature ranges from 15\(^\circ\)C (60\(^\circ\)F) to 43\(^\circ\)C
+(110\(^\circ\)F) and the region averages 200 mm (8 in) of precipitation
+annually. \#\#\#CITATION\#\#\#
 
@@
 
-\includegraphics{images/StudyArea_20241126.png}
+\includegraphics{images/StudyArea_20241203.png}
 
@@
 winter snowpack. While winter precipitation provides only 30\% of annual
-averages, it provides the bulk of water for natural reservoirs (Sheppard
-et al., 2002).
+averages, it provides the majority of water for natural reservoirs
+(Sheppard et al., 2002).
 
@@
 managed flows shared by Colorado River Compact states. Following these
-parameters, 206 USGS streamgages with acceptable periods of record were
-used in this study (Figure~\ref{fig-study-area}).
+parameters, 205 USGS streamgages with acceptable periods of record were
+used in this study (Figure~\ref{fig-study-area}). Periods of record
+ranged from 10 to 112 years, with a median of 28 years.
 
@@
 sub-basins are analogous to medium-sized river basins and are defined by
-surface water characteristics. \#\#\#\#\#\#
-
-The annual precipitation and temperature data came from the PRISM
-climate group at Oregon State University at a resolution of 4 km
+surface water characteristics.
+
+Annual precipitation and temperature data came from the PRISM climate
+group at Oregon State University at a resolution of 4 km
 (\url{https://prism.oregonstate.edu;} (Daly et al., 2008)). The PRISM
@@
 group from SSURGO (Soil Survey Geographic Database), and underlying
-geology/karst from USGS were all similarly averaged across the basins.
-Aggregating variables to align with the HUC 8 boundaries allowed for
-more precise predictions of BFI by integrating spatial variations within
-each basin.
+geology and karst from USGS were all similarly averaged across the
+basins. Aggregating variables to align with the HUC 8 boundaries allowed
+for more precise predictions of BFI by integrating spatial variations
+within each basin.
 
@@
 that separates high-frequency signals (quickflow) from low-frequency
-signals (base flow) (Lyne et al., 1979). This technique has been used in
-(Arnold et al., 2000; Bloomfield et al., 2009; Santhi et al., 2008;
-Singh et al., 2018). It takes the form of
+signals (base flow) (Lyne et al., 1979). Eckhardt (2023) noted that
+recursive digital filters lack a physical basis, but as the method is
+easy to automate, objective, and repeatable, it is appropriate for a
+regional-scale study. The Lyne-Hollick filter has been used in (Arnold
+et al., 2000; Bloomfield et al., 2009; Santhi et al., 2008; Singh et
+al., 2018). It takes the form of
 
@@
 (2014). The filter was run three times (forward, backward, forward) to
-attenuate the base-flow signal. Eckhardt (2023) noted that recursive
-digital filters lack a physical basis, but as the method is easy to
-automate, objective, and repeatable, it is appropriate for a
-regional-scale study.
+attenuate the base-flow signal.
 
@@
 ungauged catchments based on geospatial and hydroclimate predictor
-variables (Table~\ref{tbl-predictors}). Certain predictors were
-subdivided according to their areal coverage of each basin (e.g.~land
-cover was divided into 16 subdivisions).
+variables (Table~\ref{tbl-predictors}). Certain features were subdivided
+according to their areal coverage of each basin (e.g.~land cover was
+divided into 16 subdivisions).
 
 \begin{longtable}[]{@{}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.1892}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.3784}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.1081}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.3243}}@{}}
-\caption{Basin-characteristic variables used as predictors in XGBoost
-model.}\label{tbl-predictors}\tabularnewline
+  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.2329}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.2877}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.2329}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.2466}}@{}}
+\caption{Basin-characteristic variables used as initial features in
+XGBoost model. Starred features are maintained in the final,
+dimensionality-reduced model.}\label{tbl-predictors}\tabularnewline
 \toprule\noalign{}
@@
 \endlastfoot
-Hydroclimate & Precipitation & PRISM & Basin average \\
-& Mean Temperature & PRISM & Basin average \\
-& Average Evapotranspiration & MODIS & Basin average \\
-Geospatial & Elevation & DEM & Basin average \\
+Hydroclimate & Precipitation* & PRISM & Basin average \\
+& Mean Temperature* & PRISM & Basin average \\
+& Mean Evapotranspiration* & TerraClimate & Basin average \\
+Geospatial & Elevation* & DEM & Basin average \\
 & Area & DEM & Basin average \\
@@
 & Aspect & DEM & Percent areal coverage \\
-& Land Cover & NLCD & Percent areal coverage \\
-& Hydrologic Soil Group & SSURGO & Percent areal coverage \\
+& Land Cover* & NLCD & Percent areal coverage \\
+& Hydrologic Soil Group* & SSURGO & Percent areal coverage \\
 & Geology & USGS & Percent areal coverage \\
@@
 
+\subsubsection{Feature Selection}\label{feature-selection}
+
+Machine learning models can face issues with overfitting when provided
+with a large set of predicting features. This can cause reductions in
+performance on unseen data and requires greater computational power and
+memory storage (Li et al., 2017). Dimensionality reduction is a powerful
+tool in addressing these issues which can be divided into two broad
+categories: feature extraction and feature selection. Feature extraction
+involves projecting the higher dimensional dataset to a feature space
+with less dimensions, though this entails creating a new set of features
+that lose the physical meaning of the original feature space. On the
+other hand, feature selection maintains a subset of the original
+features, all of which retain their physical meaning providing better
+readability and interpretability. Here, a supervised features selection
+was used to reduce the number of predictors and, as a result, increased
+learning performance, decreased computational costs, and reduced
+overfitting.
+
+An initial model was trained on the full feature set of 45 predictors
+(Table~\ref{tbl-predictors}). We then used a feature importance method
+of feature selection to remove noise and less-important features.
+Feature importance scores are an approach where importance scores are
+assigned to individual features indicating how much the feature
+contributed (positively or negatively) to the model (Murdoch et al.,
+2019). Here, we use SHAP (SHapley Additive exPlanations) values as the
+method of assigning importance values to each feature (Lundberg \& Lee,
+2017). The top-10 most-important features
+(\textbf{?@suppfig-shap\_values}) were subset and used in training a
+subsequent model which saw performance increases and reduced computing
+time. This subsequent model was used as the final model in this
+analysis.
+
 \subsection{Statistical Analyses}\label{statistical-analyses}
 
-Statisical analyses were performed on mean annual BFI values of
-instrumented streamgages. We examined the presence of temporal trends in
-mean annual BFI using the Mann-Kendall nonparametric trend test
-(Kendall, 1970; Mann, 1945). The Mann-Kendall test was used to determine
-trends in BFI and trend significance of gauged reaches.
+Statisical analyses were performed on annual BFI and base flow values of
+instrumented streamgages. We examined temporal trends using the
+Mann-Kendall nonparametric trend test to determine the presence of
+monotonic trends (Kendall, 1970; Mann, 1945). The Mann-Kendall test was
+used to determine trends in BFI and trend significance of gauged
+reaches. This test is widely used in studies of this kind (Ayers et al.,
+2019; Ficklin et al., 2016; Woodhouse et al., 2022).
+
+Trends in BFI and base flow were produced using a Mann-Kendall test for
+monotonic trends. This test expects that the data is non-parametric and
+that the observations are not autocorrelated over time. Observations at
+4 streamgages show significant autocorrelation, though only one
+streamgage (09486500 - Santa Cruz River at Cortaro, AZ) indicated a
+significant trend in BFI. Due to this autocorrelation, the streamgage
+was removed from the trend analysis as the autocorrelation may inflate
+the variance of the Mann-Kendall statistic leading to an
+over/underestimate of the trend (Hamed \& Rao, 1998).
 
@@
 
-\includegraphics{images/BFI_Instrumented_20241126.png}
+\includegraphics{images/BFI_Instrumented_20241203.png}
 
@@
 \caption{\label{fig-instrumented-bfi}Mean annual BFI for the period of
-study (1991-2020) from instrumented stream flow data.}
+record from instrumented stream flow data.}
 
@@
 
-The long-term average BFI for the 244 gauged reaches across Arizona is
+The long-term average BFI for the 205 gauged reaches across Arizona is
 illustrated in Figure~\ref{fig-instrumented-bfi} . The highest BFI
@@
 
+Trends in BFI over the period of record for each streamgage in this
+analysis are illustrated in Figure~\ref{fig-instrumented-trend} and
+Table~\ref{tbl-trends} .
+
 \begin{figure}
@@
 
-\includegraphics{images/BFITrends_20241126.png}
+\includegraphics{images/BFITrends_20241203.png}
 
@@
 \begin{longtable}[]{@{}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.4348}}
-  >{\raggedleft\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.0725}}
-  >{\raggedleft\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1014}}
-  >{\raggedleft\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1014}}
-  >{\raggedleft\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1449}}
-  >{\raggedleft\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1449}}@{}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.4800}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.0667}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.0933}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.0933}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1333}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1333}}@{}}
 \caption{Comparison of trends for BFI for all sites split by various
@@
 Classification Group
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
 n
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
 n\_pos
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
 n\_neg
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
 perc\_pos
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
 perc\_neg
@@
 \endlastfoot
-Climate - Monsoon Dominated & 94 & 10 & 23 & 0.106 & 0.245 \\
-Climate - Snowmelt Dominated & 150 & 19 & 17 & 0.127 & 0.113 \\
-PhysRegion - Basin\&Range & 190 & 26 & 33 & 0.137 & 0.174 \\
-PhysRegion - CO Plateau & 54 & 3 & 7 & 0.056 & 0.130 \\
-Climate - Warm-Wet & 31 & 3 & 6 & 0.097 & 0.194 \\
-Climate - Warm-Dry & 81 & 15 & 16 & 0.185 & 0.198 \\
-Climate - Cool-Wet & 83 & 6 & 11 & 0.072 & 0.133 \\
-Climate - Cool-Dry & 49 & 5 & 7 & 0.102 & 0.143 \\
-Slope - High & 111 & 12 & 15 & 0.108 & 0.135 \\
-Slope - Low & 133 & 17 & 25 & 0.128 & 0.188 \\
+Precipitation - Monsoon Dominated & 87 & 8 & 21 & 0.092 & 0.241 \\
+Precipitation - Snowmelt Dominated & 118 & 9 & 12 & 0.076 & 0.102 \\
+PhysRegion - Basin\&Range & 156 & 14 & 26 & 0.090 & 0.167 \\
+PhysRegion - CO Plateau & 49 & 3 & 7 & 0.061 & 0.143 \\
+Climate - Warm-Wet & 31 & 2 & 6 & 0.065 & 0.194 \\
+Climate - Warm-Dry & 55 & 6 & 11 & 0.109 & 0.200 \\
+Climate - Cool-Wet & 74 & 4 & 9 & 0.054 & 0.122 \\
+Climate - Cool-Dry & 45 & 5 & 7 & 0.111 & 0.156 \\
+Slope - High & 102 & 10 & 12 & 0.098 & 0.118 \\
+Slope - Low & 103 & 7 & 21 & 0.068 & 0.204 \\
 \end{longtable}
@@
 
+\subsubsection{Predictor Importance}\label{predictor-importance}
+
+\begin{figure}
+
+\centering{
+
+\includegraphics{images/actual-predicted.png}
+
+}
+
+\caption{\label{fig-actual_predicted}Linear relationship between
+observed BFI and predicted BFI. The solid line is the 1:1 line, the
+dashed line is regressed to the data.}
+
+\end{figure}%
+
 \subsubsection{Model Validation}\label{model-validation}
@@
 \begin{longtable}[]{@{}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.3797}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.0759}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.0886}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.1013}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.0886}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.0886}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.0886}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.0886}}@{}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.3846}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.0769}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.0897}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.0897}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.0897}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.0897}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.0897}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.0897}}@{}}
 \caption{Performance of model predictions for BFI for all sites split by
@@
 \begin{minipage}[b]{\linewidth}\raggedright
+Classification Group
 \end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
@@
 \end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
-R\textsuperscript{2}
+R2
 \end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
@@
 \endlastfoot
-Climate - Monsoon Dominated & 3345 & 0.707 & 0.0166 & 0.129 & 0.077 &
-0.694 & -12.6 \\
-Climate - Snowmelt Dominated & 6152 & 0.723 & 0.017 & 0.131 & 0.094 &
-0.715 & -3.7 \\
-PhysRegion - Basin\&Range & 7754 & 0.749 & 0.018 & 0.133 & 0.089 & 0.741
-& -5.9 \\
-PhysRegion - CO Plateau & 1743 & 0.806 & 0.013 & 0.115 & 0.081 & 0.801 &
--4.4 \\
-Climate - Warm-Wet & 1613 & 0.690 & 0.015 & 0.124 & 0.081 & 0.679 &
--9.9 \\
-Climate - Warm-Dry & 3134 & 0.758 & 0.023 & 0.151 & 0.097 & 0.745 &
--8.1 \\
-Climate - Cool-Wet & 3134 & 0.710 & 0.013 & 0.115 & 0.085 & 0.708 &
--2.5 \\
-Climate - Cool-Dry & 1616 & 0.809 & 0.014 & 0.118 & 0.082 & 0.803 &
--5 \\
-Slope - High & 4398 & 0.753 & 0.014 & 0.120 & 0.084 & 0.746 & -4.4 \\
-Slope - Low & 5099 & 0.759 & 0.019 & 0.138 & 0.090 & 0.750 & -6.9 \\
+Climate - Monsoon Dominated & 3039 & 0.633 & 0.016 & 0.126 & 0.074 &
+0.619 & -13.7 \\
+Climate - Snowmelt Dominated & 4685 & 0.733 & 0.015 & 0.121 & 0.087 &
+0.725 & -3.5 \\
+PhysRegion - Basin\&Range & 6147 & 0.733 & 0.016 & 0.127 & 0.084 & 0.724
+& -6.3 \\
+PhysRegion - CO Plateau & 1577 & 0.846 & 0.011 & 0.104 & 0.073 & 0.843 &
+-3.8 \\
+Climate - Warm-Wet & 1506 & 0.693 & 0.014 & 0.117 & 0.077 & 0.685 &
+-8.2 \\
+Climate - Warm-Dry & 2351 & 0.693 & 0.022 & 0.147 & 0.092 & 0.675 &
+-11.9 \\
+Climate - Cool-Wet & 2350 & 0.738 & 0.011 & 0.106 & 0.078 & 0.736 &
+-1.7 \\
+Climate - Cool-Dry & 1517 & 0.831 & 0.012 & 0.111 & 0.078 & 0.827 &
+-4.3 \\
+Slope - High & 3795 & 0.776 & 0.012 & 0.111 & 0.079 & 0.771 & -3.3 \\
+Slope - Low & 3929 & 0.724 & 0.018 & 0.133 & 0.085 & 0.713 & -9.1 \\
 \end{longtable}
@@
 
-\includegraphics{images/BFI_HUC8_20241126.png}
+\includegraphics{images/BFI_HUC8_20241203.png}
 
@@
 
-\includegraphics{images/BFI_AllStream.pdf}
+\includegraphics{images/BFI_AllStream.png}
 
@@
 
+\section{Supporting Information}\label{supporting-information}
+
+\includegraphics{images/supplemental/shap_summary.png}
+\{\#suppfig-shap\_values\}
+
 \phantomsection\label{refs}
 \begin{CSLReferences}{1}{0}
-\vspace{1em}
-
 \bibitem[\citeproctext]{ref-abatzoglou2018}
@@
 https://doi.org/\url{https://doi.org/10.1016/S0022-1694(99)00139-0}
+
+\bibitem[\citeproctext]{ref-ayers2019}
+Ayers, J. R., Villarini, G., Jones, C., \& Schilling, K. (2019). Changes
+in monthly baseflow across the U.S. Midwest. \emph{Hydrological
+Processes}, \emph{33}(5), 748--758.
+\url{https://doi.org/10.1002/hyp.13359}
 
@@
 
+\bibitem[\citeproctext]{ref-MK_Hamed1998}
+Hamed, K. H., \& Rao, A. (1998). A modified mann-kendall trend test for
+autocorrelated data. \emph{Journal of Hydrology}, \emph{204}(1),
+182--196.
+https://doi.org/\url{https://doi.org/10.1016/S0022-1694(97)00125-X}
+
 \bibitem[\citeproctext]{ref-instituteofhydrology1980}
@@
 
+\bibitem[\citeproctext]{ref-feat_selec2017}
+Li, J., Cheng, K., Wang, S., Morstatter, F., Trevino, R. P., Tang, J.,
+\& Liu, H. (2017). Feature selection: A data perspective. \emph{ACM
+Comput. Surv.}, \emph{50}(6). \url{https://doi.org/10.1145/3136625}
+
+\bibitem[\citeproctext]{ref-SHAP_values}
+Lundberg, S. M., \& Lee, S.-I. (2017). A unified approach to
+interpreting model predictions, \emph{30}. Retrieved from
+\url{https://proceedings.neurips.cc/paper_files/paper/2017/file/8a20a8621978632d76c43dfd28b67767-Paper.pdf}
+
 \bibitem[\citeproctext]{ref-lyne1979}
@@
 \emph{Econometrica: Journal of the Econometric Society}, 245--259.
+
+\bibitem[\citeproctext]{ref-ML_interpretability2019}
+Murdoch, W. J., Singh, C., Kumbier, K., Abbasi-Asl, R., \& Yu, B.
+(2019). Definitions, methods, and applications in interpretable machine
+learning. \emph{Proceedings of the National Academy of Sciences},
+\emph{116}(44), 22071--22080.
+\url{https://doi.org/10.1073/pnas.1900654116}
 
@@
 
+\bibitem[\citeproctext]{ref-woodhouse_udall22}
+Woodhouse, C. A. et al. (2022). Upper {Gila}, {Salt}, and {Verde
+Rivers}: {Arid Land Rivers} in a {Changing Climate}. \emph{Earth
+Interactions}, \emph{26}(1), 1--14.
+\url{https://doi.org/10.1175/EI-D-21-0014.1}
+
 \bibitem[\citeproctext]{ref-yao2018}

### 2024-12-11 MANUSCRIPT.qmd
Update MANUSCRIPT.qmd

@@
 |              | Variable                  | Source       | Geoprocessing          |
-|------------------|-------------------|------------------|------------------|
+|--------------|---------------------------|--------------|------------------------|
 | Hydroclimate | Precipitation\*           | PRISM        | Basin average          |
@@
 Machine learning models can face issues with overfitting when provided with a large set of predicting features.
-This can cause reductions in performance on unseen data and requires greater computational power and memory storage [@feat_selec2017].
+(CITATION) This can cause reductions in performance on unseen data and requires greater computational power and memory storage [@feat_selec2017].
 Dimensionality reduction is a powerful tool in addressing these issues which can be divided into two broad categories: feature extraction and feature selection.

### 2024-12-11 MANUSCRIPT.qmd
Update MANUSCRIPT.qmd

--- 2024-12-07 index.qmd
@@
 Streamgages were selected depending on criteria to ensure the applicability of each site.
-@odonnell2016 found that 8-10 years of calibration data were necessary to account for climate variability in paired watershed studies in the region.
-Following this criteria, a minimum of 10 years of record was chosen.
-Any years missing more than 30 days of streamflow were removed from the possible record period.
-This analysis focuses on natural, streamflow-influencing recharge, excluding streamgages affected by regulation or diversions.
+Following the findings of @odonnell2016, which determined that 8–10 years of calibration data are necessary to account for climate variability in paired watershed studies in the region, a minimum record length of 10 years was required.
+Additionally, years with more than 30 missing days of streamflow data were excluded from the analysis.
+This study focuses on natural, streamflow-influencing dynamics, so streamgages affected by regulation or diversions were excluded.
 These gages were identified through annual reports on water data published by the USGS [@usgs2010].
-Finally, streamgages on the Colorado River were excluded from the dataset on the basis that they represent managed flows shared by Colorado River Compact states.
-Following these parameters, 205 USGS streamgages with acceptable periods of record were used in this study (@fig-study-area).
+Furthermore, streamgages along the Colorado River were omitted because they represent managed flows governed by the Colorado River Compact.
+After applying these selection criteria, 205 USGS streamgages with acceptable periods of record were included in the study (@fig-study-area).
 Periods of record ranged from 10 to 112 years, with a median of 28 years.
@@
 The XGBoost algorithm is a decision tree-based ensemble algorithm, which can be adapted for either regression or classification problems.
-It has grown in popularity in recent years due to its rapid, efficient, and scalable nature, and has proven effective in similar environmental modeling tasks such as streamflow forecasting [@szczepanek2022daily**; @ni2020streamflow**] and classifying land use/land cover data [@Georganos2018Very].
+Its efficiency, scalability, and robustness have made it increasingly popular in recent years, with successful applications in environmental modeling tasks such as streamflow forecasting [@szczepanek2022daily; @ni2020streamflow] and land use/land cover classification [@Georganos2018Very].
 
-XGBoost uses gradient boosting on decision tree algorithms to leverage the combination of multiple very low variance models into one overall prediction.
-Gradient boosting works sequentially, with the first tree being learned on the known target values, and each subsequent tree being learned on the errors of the tree prior.
-The algorithm then applies a weighting to each tree based on this boosting to determine how much the tree will contribute towards the final output.
-The final prediction is obtained by the input data traversing all $n$ weighted trees in the learned ensemble.
-The trained XGBoost model is used to predict BFI in ungauged catchments based on geospatial and hydroclimate predictor variables (@tbl-predictors).
-Certain predictors were subdivided according to their areal coverage of each basin (e.g.
+XGBoost operates by leveraging gradient boosting on decision tree algorithms, combining multiple low-variance models to produce a robust overall prediction.
+Gradient boosting works iteratively: the initial tree is trained on the target values, while subsequent trees are trained on the residual errors of the preceding tree.
+Each tree is assigned a weight based on its contribution to reducing error, and these weights are used to determine the influence of each tree in the final model.
+The ultimate prediction is made by aggregating the outputs of all $n$ weighted trees in the ensemble.
+In this study, the trained XGBoost model is used to predict BFI in ungauged catchments based on geospatial and hydroclimate predictor variables (@tbl-predictors).
+Certain features were further subdivided according to their areal coverage within each basin (e.g.
 land cover was divided into 16 subdivisions).
+This approach allowed the model to capture finer-scale spatial variability and improve predictive accuracy.
 
 |              | Variable                  | Source       | Geoprocessing          |
-|--------------|----------------------------|--------|------------------------|
-| Hydroclimate | Precipitation              | PRISM  | Basin average          |
-|              | Mean Temperature           | PRISM  | Basin average          |
-|              | Average Evapotranspiration | MODIS  | Basin average          |
-| Geospatial   | Elevation                  | DEM    | Basin average          |
+|------------------|--------------------|------------------|------------------|
+| Hydroclimate | Precipitation\*           | PRISM        | Basin average          |
+|              | Mean Temperature\*        | PRISM        | Basin average          |
+|              | Mean Evapotranspiration\* | TerraClimate | Basin average          |
+| Geospatial   | Elevation\*               | DEM          | Basin average          |
 |              | Area                      | DEM          | Basin average          |
@@
 
-: Basin-characteristic variables used as predictors in XGBoost model. {#tbl-predictors}
+: Basin-characteristic variables used as initial features in XGBoost model. Starred features are maintained in the final, dimensionality-reduced model. {#tbl-predictors}
+
+Our initial training dataset comprised 7,724 observations across 45 variables.
+To optimize the model's performance, we first conducted an exhaustive grid search combined with 5-fold cross-validation to identify the optimal hyperparameter values.
+The hyperparameters evaluated included the learning rate ($\eta$), minimum split loss ($\gamma$), maximum tree depth, minimum child weight, and the number of trees.
+While not an exhaustive list of all possible XGBoost hyperparameters, the range of values explored provided sufficient variation to ensure the selection of a high-performing model.The optimal hyperparameters were determined to be: 700 trees, a learning rate ($\eta$) of 0.05, a minimum split loss ($\gamma$) of 0.075, a maximum tree depth of 7, and a minimum child weight of 5.
+Using these values, the XGBoost model was trained on the dataset with 10-fold cross-validation.
+
+$K$-fold cross-validation provides an unbiased estimate of a model's accuracy on unseen data, while also insuring against overfitting or underfitting.
+In this approach, the data is randomly divided into $k$ folds of equal size.
+The model is trained on $k-1$ folds and tested on the remaining fold, referred to as the validation set.
+This process is repeated $k$ times, with each fold serving as the validation set exactly once.
+Each iteration trains an independent model with the same hyperparameters but using a different subset of training data.
+By averaging the model performance across all $k$ folds, we achieve a robust and reliable estimate of predictive accuracy.
+Root mean squared error (RMSE) was used as the performance metric for both model optimization and evaluation.
+RMSE provided a consistent and interpretable measure of the model’s accuracy throughout the training and validation process.
+
+#### Feature Selection
+
+Machine learning models are prone to overfitting, especially when provided with a large set of predictive features (CITATION).
+Overfitting can degrade model performance on unseen data and increase the demand for computational resources and memory storage [@feat_selec2017].
+Dimensionality reduction offers a robust solution to these challenges and generally falls into two broad categories: feature extraction and feature selection.
+
+Feature extraction involves transforming the original dataset into a lower-dimensional feature space.
+However, this process generates new features that often lack the physical interpretability of the original variables.
+In contrast, feature selection identifies a subset of the original features, preserving their physical meaning while improving model readability and interpretability.
+In this study, supervised feature selection was employed to reduce the number of predictors, which enhanced learning performance, reduced computational costs, and mitigated overfitting.
+
+To begin, an initial model was trained using the full feature set of 45 predictors (@tbl-predictors).
+A feature selection method based on feature importance was then applied to identify and remove less relevant and noisy features.
+Feature importance scores quantify the contribution of individual features—either positively or negatively—to the model's predictions [@ML_interpretability2019].
+In this analysis, SHAP (SHapley Additive exPlanations) values were used to compute feature importance scores [@SHAP_values].
+
+The ten most important features (\@suppfig-shap_values) were selected based on their SHAP values and used to train a subsequent model.
+This refined model demonstrated improved performance and reduced computational time compared to the initial model.
+The final model, trained on this optimized feature subset, was ultimately used for the analysis presented here.
 
@@
 
-Statisical analyses were performed on annual BFI and base flow values of instrumented streamgages.
-We examined temporal trends using the Mann-Kendall nonparametric trend test to determine the presence of monotonic trends [@kendall1970; @mann1945nonparametric].
-The Mann-Kendall test was used to determine trends in BFI and trend significance of gauged reaches.
+Statistical analyses were conducted on annual BFI and base flow values from instrumented streamgages to identify temporal trends using the Mann-Kendall nonparametric trend test [@kendall1970; @mann1945nonparametric].
+This test detects monotonic trends in datasets that are non-parametric and assumes the absence of autocorrelation among observations.
 This test is widely used in studies of this kind [@ficklin2016; @ayers2019; @woodhouse_udall22].
 
-Trends in BFI and base flow were produced using a Mann-Kendall test for monotonic trends.
-This test expects that the data is non-parametric and that the observations are not autocorrelated over time.
-Observations at 4 streamgages show significant autocorrelation, though only one streamgage (09486500 - Santa Cruz River at Cortaro, AZ) indicated a significant trend in BFI.
-Due to this autocorrelation, the streamgage was removed from the trend analysis as the autocorrelation may inflate the variance of the Mann-Kendall statistic leading to an over/underestimate of the trend (CITATION).
-
+To check for autocorrelation, we applied the Durbin-Watson test, which revealed significant autocorrelation at four streamgages.
+Of these, only one streamgage (09486500 - Santa Cruz River at Cortaro, AZ) showed a significant trend in BFI.
+This streamgage was excluded from the trend analysis, as autocorrelation could inflate the variance of the Mann-Kendall statistic, potentially leading to biased trend estimates [@MK_Hamed1998].
 Trends with a $\rho \le 0.05$ are considered significant.
@@
 This is likely due to low-yielding perched aquifers underlying the Defiance Uplift in northeastern Arizona, which are hydrologically connected to surface streams, while the high-yield, confined regional aquifer is much deeper [@blanchard02].
-This region also typifies a pattern of seen along most major rivers in the study area: upstream reaches indicate higher BFI values while the downstream reaches display lower BFI values.
-This is assumed to be due to greater groundwater-surface water interaction at the headwaters of streams (likely due to spring outlets) and the dilution of base flow while moving down stream (**CITATION**).
-This pattern is seen along the Gila River, Verde River, and Little Colorado Rivers in Arizona.
+A notable trend emerges along most major rivers in the study area: upstream reaches tend to exhibit higher BFI values, while downstream reaches display lower values.
+This pattern is presumed to result from greater groundwater-surface water interactions at stream headwaters, often influenced by spring outlets, and the dilution of base flow as water moves downstream (**CITATION**).
+This trend is particularly evident along the Gila River, Verde River, and Little Colorado River in Arizona.
 
@@
 
+#### Model Validation
+
+Predicted values of BFI are plotted against observed values for the entire period of record of the instrumented dataset in @fig-actual_predicted .
+The agreement between "out-of-bag" predictions (blind cross-validation, treating each site as ungauged) and observed values is acceptable (R^2^ = 0.764) indicating that the model performs well across the full dataset.
+The overall Root Mean Square Error (RMSE) is 0.129 and the overall percent bias (pbias) is -5.6.
+Model performance metrics across various classifications are summarized in @tbl-performance.
+These performance metrics demonstrate that the regional model performs consistently well across different spatial and climatic classifications.
+However, the negative pbias values across all classifications, along with the overall pbias, indicate a systematic underprediction of BFI by the model.
+Categories with relatively lower R^2^ and Nash-Sutcliffe Efficiency (NSE) values also exhibit higher biases, reflecting weaker model performance in those specific contexts.
+
 ![Linear relationship between observed BFI and predicted BFI. The solid line is the 1:1 line, the dashed line is regressed to the data.](images/actual-predicted.png){#fig-actual_predicted}
-
-#### Model Validation
 
@@
 
-Consistently negative pbias values across all classifications suggest an under-prediction of BFI.
+#### Predictor Importance {#sec-predictor-importance}
 
-#### Predictor Importance {#sec-predictor-importance}
+The predictors used to estimate BFI at ungauged sites were evaluated for their importance in the final XGBoost model, as illustrated in @suppfig-shap_values.
+The most influential feature for predicting long-term BFI is basin elevation.
+While elevation itself does not directly affect base flow characteristics, it has consistently been identified as a key predictor in previous BFI studies [@singh2018; @beck2013].
+The significance of elevation likely stems from its cumulative effects, such as lower temperatures, increased orographic precipitation, and prolonged snowpack persistence at higher altitudes.
+Seasonal snowpack duration, in particular, has been shown to strongly correlate with springflow and groundwater recharge in this region [@donovan_karst_2022].
+This relationship is further supported by hydroclimate features.
+Higher temperatures tend to negatively influence BFI, while precipitation exhibits a mixed influence.
+In some cases, higher precipitation values correlate with lower BFI, likely due to a larger proportion of precipitation contributing to runoff rather than infiltration and groundwater recharge.
+
+Land cover and land use predictors also play a significant role in BFI estimation.
+Analysis of SHAP values indicates that a higher percentage of evergreen forest positively influences BFI predictions, while higher proportions of shrubland and developed land exert a negative influence.
+Similarly, hydrologic soil types show distinct trends in their impact on BFI.
+Soil Type C, characterized by moderately high runoff potential (20-40% clay), tends to negatively influence BFI.
+In contrast, Soil Type A, which has low runoff potential and facilitates rapid water infiltration, exhibits a mixed influence [@nrcs2009].
+The mixed effects of Soil Type A are likely due to SHAP values capturing interactions between features rather than direct relationships.
+For example, regions dominated by Soil Type A may also have steep slopes or sparse vegetation.
+Additionally, the relationship between Soil Type A and BFI is likely non-linear and influenced by the complex dynamics of regional base-flow processes.
 

### 2024-12-13 MANUSCRIPT.qmd
Update MANUSCRIPT.qmd
updates

@@
 |              | Variable                  | Source       | Geoprocessing          |
-|------------------|--------------------|------------------|------------------|
+|------------------|-------------------|------------------|------------------|
 | Hydroclimate | Precipitation\*           | PRISM        | Basin average          |
@@
 
-To check for autocorrelation, we applied the Durbin-Watson test, which revealed significant autocorrelation at four streamgages.
+To check for autocorrelation, we applied the Durbin-Watson test, which revealed significant autocorrelation at four streamgages on an annual basis.
 Of these, only one streamgage (09486500 - Santa Cruz River at Cortaro, AZ) showed a significant trend in BFI.
@@
 
-![Mean annual BFI for the period of record from instrumented stream flow data.](images/BFI_Instrumented_20241203.png){#fig-instrumented-bfi}
+![Long-term BFI for the period of record from instrumented stream flow data.](images/BFI_Instrumented_20241203.png){#fig-instrumented-bfi}
 
@@
 
-#### Trends in BFI {#sec-bfi-trends}
+#### Trends in Base Flow & BFI {#sec-bfi-trends}
 
-Trends in BFI over the period of record for each streamgage in this analysis are illustrated in @fig-instrumented-trend and @tbl-trends .
+Trends in BFI over the period of record for each streamgage in this analysis are illustrated in @fig-instrumented-trend and @tbl-trends.
+Base flow and BFI trends were analyzed across all instrumented sites over their respective periods of record using the Mann–Kendall test.
+Statistically significant trends were observed in both metrics, with a 72.20% coincidence rate between significant base flow and BFI trends, indicating a strong dependence of BFI on base-flow dynamics.
+
+#### Spatial Trends
+
+TALK ABOUT THE SPATIAL TRENDS IN THE MAP
 
@@
 
+#### Classification Trends
+
+Classifications presented in @tbl-trends were determined based on precipitation regime, physiographic region, climate, and slope.
+The dominant precipitation regime (monsoon vs.
+snowmelt) was identified by analyzing streamflow hydrographs for each station, focusing on peak flow periods during the monsoon season (July–September) and the snowmelt season (March–June).
+Physiographic region was assigned based on which region the streamgage is in.
+Climate classifications were defined as warm (above the median temperature of Arizona), cool (below the median temperature), wet (above the median precipitation), and dry (below the median precipitation).
+Slope was categorized as high (above the median slope) and low (below the median slope).
+
+Statistically significant decreasing trends in BFI were more common than increasing trends across all site classifications (@tbl-trends).
+While decreasing trends dominate, both increasing and decreasing trends are observed within each classification.
+Monsoon-dominated regions exhibit a higher proportion of significant negative trends (24.1%) compared to snowmelt-dominated regions (10.2%), suggesting that monsoon-dominated systems may be more vulnerable to declining baseflows.
+Among climate classifications, warm-dry climates have the highest proportion of negative trends (20.0%), followed by warm-wet climates (19.4%), indicating that regions with higher temperatures are more prone to baseflow declines.
+Low-slope regions show a greater prevalence of negative trends (20.4%) compared to high-slope regions (11.8%).
+This suggests that flatter areas may be more susceptible to baseflow reductions, potentially due to differences in hydrologic connectivity and recharge dynamics.
+
 | Classification Group                    | n   | n_pos | n_neg | perc_pos | perc_neg |
-|------------------------------------|-----|-------|-------|----------|----------|
+|-------------------------|----------|----------|----------|----------|----------|
 | Precipitation - Monsoon Dominated       | 87  | 8     | 21    | 0.092    | 0.241    |
 | Precipitation - Snowmelt Dominated      | 118 | 9     | 12    | 0.076    | 0.102    |
-| PhysRegion - Basin&Range           | 156 | 14    | 26    | 0.090    | 0.167    |
-| PhysRegion - CO Plateau            | 49  | 3     | 7     | 0.061    | 0.143    |
+| Physiographic Region - Basin and Range  | 156 | 14    | 26    | 0.090    | 0.167    |
+| Physiographic Region - Colorado Plateau | 49  | 3     | 7     | 0.061    | 0.143    |
 | Climate - Warm-Wet                      | 31  | 2     | 6     | 0.065    | 0.194    |
@@
 {#tbl-trends}
+
+#### Coincident Climate Trends
+
+Given the variations in the period of record across the instrumented network (see SUPPLEMENTAL - Period of Record figure), we analyzed trends in baseflow and BFI in relation to coincident trends in climate variables.
+Trends are classified as coincident when the direction of the climate variable trend (positive or negative) aligns with the trend observed in baseflow or BFI (@tbl-coincident-trends).
+This analysis includes both significant and non-significant trends, which is appropriate in environmental studies where the influence of complex, interconnected processes may not always manifest as statistically significant patterns over limited observational periods.
+
+The analysis shows that baseflow and BFI trends most frequently align with precipitation trends (64.88% and 53.17%, respectively), emphasizing precipitation as the primary driver of local groundwater recharge and discharge.
+Coincidence with AET (actual evapotranspiration) trends (55.12% for baseflow and 44.88% for BFI) suggests that evapotranspiration also plays a significant role, particularly in arid regions where it can reduce recharge or baseflow during dry periods.
+In contrast, temperature trends show lower percentages of coincidence, often opposing baseflow and BFI trends.
+Specifically, positive (negative) temperature trends are frequently associated with negative (positive) baseflow trends (60.98%) and BFI trends (52.68%).
+These results highlight the complex interplay between climatic variables and hydrological processes.
+While precipitation exerts the strongest influence on baseflow and BFI, evapotranspiration and temperature add further variability in specific environmental contexts.
+
+|           | **Climate Variable** | **Coincidence Percentage (%)** |
+|-----------|:--------------------:|:------------------------------:|
+| BFI       |         AET          |             44.88              |
+|           |    Precipitation     |             53.17              |
+|           |     Temperature      |             47.32              |
+| Base Flow |         AET          |             55.12              |
+|           |    Precipitation     |             64.88              |
+|           |     Temperature      |             39.02              |
+
+: Coincident trends of climate variables (AET, precipitation, temperature) with base flow and BFI trends. {#tbl-coincident-trends}
 

### 2024-12-16 MANUSCRIPT.qmd
finishing up

@@
 ---
-title: Regional-scale Base-flow Index of Gauged and Ungauged Basins in Drylands
+title: Regional-scale Base-flow Index of Gauged and Ungauged Basins in Drylands | Regional Base-Flow Index in Arid Landscapes Using Machine Learning and Instrumented Records | Estimating Base-Flow Index in Gauged and Ungauged Dryland Basins
 author:
@@
     affiliations:
-      - School of Earth and Sustainability, Northern Arizona University, Flagstaff, AZ, USA
+      - 'School of Earth and Sustainability, Northern Arizona University, Flagstaff, AZ, USA'
 
@@
     corresponding: false
-    roles:
-      - Conceptualization
-      - Supervision
-      - Project administration
-      - Funding acquisition
+    roles: []
     affiliations:
-      - School of Earth and Sustainability, Northern Arizona University, Flagstaff, AZ, USA
+      - 'School of Earth and Sustainability, Northern Arizona University, Flagstaff, AZ, USA'
+
+  - name: Neha Gupta
+    orcid: 0000-0002-2247-3509
+    corresponding: false
+    roles: []
+    affiliations:
+      - 'Department of Hydrology and Atmospheric Sciences, University of Arizona, Tucson, AZ, USA'
 
@@
     corresponding: false
-    roles:
-      - Software
-      - Supervision
+    roles: []
     affiliations:
-      - Department of Mathematics and Statistics, Northern Arizona University, Flagstaff, AZ, USA
+      - 'Department of Mathematics and Statistics, Northern Arizona University, Flagstaff, AZ, USA'
 
@@
 abstract: |
-  LOREM IPSUM ######
+  Base flow, sustained by groundwater discharge, is a vital component of river ecosystems, particularly in drylands, where water resources are limited.
+This study analyzes the instrumented streamflow record in Arizona to assess long-term base-flow index (BFI) trends across gauged catchments.
+Results indicate that approximately 32% of Arizona’s streamflow originates from groundwater discharge, with significant spatial variability driven by landscape and climatic factors.
+Base flow  relationships are analyzed with coincident trends in climate variables such as precipitation, evapotranspiration, and temperature.
+Spatial and climatic trends reveal  variability in base flow contributions, providing insight into groundwater-surface water interactions in arid and semi-arid landscapes.
+Building on this analysis, we applied machine learning methods to predict BFI in ungauged basins, addressing the challenges of Arizona’s sparse streamgage network.
+Using the eXtreme Gradient Boosting (XGBoost) algorithm trained on hydroclimate and physiographic predictors, we estimated long-term BFI from 1991 to 2020.
+This combined approach integrates observational data with predictive modeling to enhance our understanding of base flow processes and provide a framework for water resource management in data-limited regions.
+
 plain-language-summary: |
-  LOREM IPSUM ######
+  Rivers in drylands, such as Arizona, depend on groundwater contributions to maintain flow during dry periods.
+This portion of streamflow, called base flow, plays a critical role in supporting ecosystems and water availability in these regions.
+This study examines long-term patterns of base flow in Arizona using data from streamflow monitoring stations.
+We found that about 32% of Arizona's river flow comes from groundwater, though this varies significantly across the state depending on local climate and landscape features.
+We also analyzed how base flow is influenced by changes in precipitation, evapotranspiration, and temperature, providing insights into how groundwater and surface water interact in these complex environments.
+To estimate base flow in basins without monitoring stations, we used machine learning techniques.
+By training a model on data from monitored sites and corresponding hydroclimate data, we predicted long-term base flow for ungauged areas across Arizona.
+This innovative approach addresses the challenge of Arizona's sparse monitoring network and provides a valuable tool for understanding and managing water resources in regions with limited data.
 
 key-points:
-  - Base flow sustains dryland river ecology, but assessing its groundwater contribution faces data limitations
-  - Recursive filtering and machine learning predict BFI 
-  -
+  - Approximately 32% of Arizona's streamflow originates from groundwater discharge, with significant spatial and climatic variability.
+  - A machine learning model (XGBoost) successfully predicted long-term BFI in ungauged basins, addressing gaps in Arizonas sparse streamgage network. 
+  - Base-flow trends align most closely with precipitation changes, emphasizing the critical role of climate in shaping groundwater-surface water dynamics in dryland regions.
 
@@
 These regions are home to approximately 2 billion people globally and constitute the largest terrestrial biome [@iucn-drylands-2019].
-Despite supporting diverse ecosystems and human populations, dryland regions face mounting challenges exacerbated by increasing urbanization, expanding agricultural activities, and climate-induced amplification of precipitation patterns [@taylor2013].
+Despite supporting diverse ecosystems and human populations, dryland regions face mounting hydrologic challenges exacerbated by increasing urbanization, expanding agricultural activities, and climate-induced amplification of precipitation patterns [@taylor2013].
 This water scarcity is intensifying due to the compounding effects of climate variability and increased groundwater extraction [@taylor2013].
@@
 
-To address the challenge of quantifying base flow in ungauged catchments, numerous studies have applied regression and machine learning methods.
+Advancements in machine learning provide tools to predict hydrologic indices in ungauged basins, addressing the limitations of sparse streamgage networks.
+To tackle the challenge of quantifying base flow in ungauged catchments, numerous studies have applied both regression and machine learning methods.
 [@ahiablame2013] found that using a regression model to estimate annual base flow of ungauged catchments was reasonably easy and accurate.
@@
 @singh2018 implemented a random forest algorithm to predict long-term BFI for ungauged catchments across New Zealand.
-The application of machine learning in past studies highlights its versatility and effectiveness in addressing ecohydrologic dynamics.
+These applications demonstrate the versatility and effectiveness of machine learning in capturing complex ecohydrologic dynamics and improving our understanding of groundwater contributions to streamflow.
 
-This study develops a technique for estimating BFI in ungauged basins in Arizona and determines the state’s long-term mean BFI.
-Using a machine learning model trained with the hydrogeologic characteristics of surface water basins, we estimate mean BFI in ungauged basins over a period of record from 1991-2020.
-Regional patterns and trends in base flow and BFI of instrumented sites are subsequently investigated.
-These results provide novel analyses of low flow processes in Arizona, furthering our understanding of climate controls on consistent flows in dryland basins.
+This study develops a technique for estimating BFI in ungauged basins across Arizona and evaluates the state’s long-term BFI.
+Regional trends in baseflow and BFI at instrumented sites are also analyzed, with these trends being linked to coincident trends in precipitation, evapotranspiration (ET~O~​), and temperature over the same periods.
+Using a machine learning model trained on the hydrogeologic characteristics of surface water basins, we estimate the mean BFI for ungauged basins from 1991 to 2020.
+This approach helps address the spatial gaps in the streamgage network, which is relatively sparse across the state.
+The results offer novel insights into low-flow processes in both gauged and ungauged basins, enhancing our understanding of climate controls on consistent flows in Arizona.
+This study contributes to a more comprehensive view of hydrological dynamics in the context of arid and semi-arid landscapes.
 
@@
 The Basin-and-Range region has a semi-arid to arid climate with an average elevation of 490 masl (1600 ft).
-Average temperature ranges from 15$^\circ$C (60$^\circ$F) to 43$^\circ$C (110$^\circ$F) and the region averages 200 mm (8 in) of precipitation annually.
-###CITATION###
+Average temperature ranges from 15$^\circ$C (60$^\circ$F) to 43$^\circ$C (110$^\circ$F) and the region averages 200 mm (8 in) of precipitation annually [@az_climate2024].
 
@@
 
+While our data selection ensures a robust record for this analysis, it also highlights gaps in spatial coverage that the machine learning model aims to address.
+Arizona has 184 active USGS streamflow stations (as of 2024) across an area of 295,253 km², whereas Indiana has 189 active stations within an area of 94,326 km².
+This translates to a streamgage density of approximately 2.004 gages per 1,000 km² in Indiana, over three times higher than Arizona's density of 0.623 gages per 1,000 km².
+This stark difference underscores the relative sparsity of Arizona's streamgage network, particularly in the context of its larger geographic area and the unique hydrological challenges posed by its arid and semi-arid landscapes.
+
 Watersheds across the United States are delineated by the USGS using a hydrologically-defined network.
@@
 Annual precipitation and temperature data came from the PRISM climate group at Oregon State University at a resolution of 4 km (<https://prism.oregonstate.edu;> [@daly2008]).
-The PRISM dataset provides valuable insights into regional climate in ungauged regions and has been shown to perform well across the southwestern US [@buban_PRISM]**.** Instead of the water year, PRISM data uses a calendar-year format, which was adopted for consistency in the water balance.
+The PRISM dataset provides valuable insights into regional climate in ungauged regions and has been shown to perform well across the southwestern US [@buban_PRISM].
+Instead of the water year, PRISM data uses a calendar-year format, which was adopted for consistency in the water balance.
 Although this may introduce challenges in the annual estimates due to inter-annual snow storage, the use of long-term annual averages is likely to reduce any potential errors [@reitz2017].
 
-Annual evapotranspiration (ET) data came from TerraClimate, a 4-km grid climatalogical data set [@abatzoglou2018].
-TerraClimate uses a Penman-Monteith approach to generate a reference potential evapotranspiration (PET).
-The PET values were calculated assuming a reference grass surface across the landscape with unlimited water.
-In the drylands of the southwestern US, PET typically exceeds precipitation annually [@zomer2022].
+Annual reference evapotranspiration (ET~O~) data came from TerraClimate, a 4-km grid climatalogical data set [@abatzoglou2018].
+TerraClimate uses a Penman-Monteith approach to generate a reference evapotranspiration.
+The ET~O~ values were calculated assuming a reference grass surface across the landscape with unlimited water.
+In the drylands of the southwestern US, ET~O~ typically exceeds precipitation annually [@zomer2022].
 
@@
 The XGBoost algorithm is a decision tree-based ensemble algorithm, which can be adapted for either regression or classification problems.
+This algorithm iteratively builds an ensemble of decision trees, where each tree corrects errors from previous trees to improve predictions [@chen2016xgboost].
 Its efficiency, scalability, and robustness have made it increasingly popular in recent years, with successful applications in environmental modeling tasks such as streamflow forecasting [@szczepanek2022daily; @ni2020streamflow] and land use/land cover classification [@Georganos2018Very].
@@
 
-Machine learning models are prone to overfitting, especially when provided with a large set of predictive features (CITATION).
+Machine learning models are prone to overfitting, especially when provided with a large set of predictive features [@Ying_2019].
 Overfitting can degrade model performance on unseen data and increase the demand for computational resources and memory storage [@feat_selec2017].
@@
 
-The ten most important features (\@suppfig-shap_values) were selected based on their SHAP values and used to train a subsequent model.
+SHAP values is a method to explain the prediction of an individual instance by calculating the contribution of each feature to that prediction.
+The method is based on coalition game theory and is discussed further in @SHAP_values.
+Here, SHAP values are used for global interpretation of feature importance and feature effects on the model.
+Global feature importance is produced by the absolute Shapley values of each feature across the dataset, providing a list of features in order of most to least important.
+Feature effects provide an indication of the relationship between the value of a predicting feature and its impact on the prediction.
+
+The ten most important features [@fig-shap_values] were selected based on their SHAP values and used to train a subsequent model [@tbl-predictors].
 This refined model demonstrated improved performance and reduced computational time compared to the initial model.
@@
 
-The long-term average BFI for the 205 gauged reaches across Arizona is illustrated in @fig-instrumented-bfi .
+The long-term BFI for the 205 gauged reaches across Arizona is illustrated in @fig-instrumented-bfi .
+The long-term mean BFI is 0.32, indicating that \~32% of long-term streamflow in Arizona likely originates from groundwater discharge and other delayed sources.
 The highest BFI values (\>0.9) are found along the Grand Canyon in northwestern Arizona.
@@
 
-The stream reaches of the Little Colorado River Basin (northeastern Arizona) indicate consisently low BFI values (\< 0.2).
+The stream reaches of the Little Colorado River Basin (northeastern Arizona) indicate consistently low BFI values (\< 0.2).
 This is likely due to low-yielding perched aquifers underlying the Defiance Uplift in northeastern Arizona, which are hydrologically connected to surface streams, while the high-yield, confined regional aquifer is much deeper [@blanchard02].
-A notable trend emerges along most major rivers in the study area: upstream reaches tend to exhibit higher BFI values, while downstream reaches display lower values.
-This pattern is presumed to result from greater groundwater-surface water interactions at stream headwaters, often influenced by spring outlets, and the dilution of base flow as water moves downstream (**CITATION**).
+A notable tendency emerges along most major rivers in the study area: upstream reaches tend to exhibit higher BFI values, while downstream reaches display lower values.
+This pattern is presumed to result from greater groundwater-surface water interactions at stream headwaters, influenced by spring outlets, and the dilution of base flow as water moves downstream.
 This trend is particularly evident along the Gila River, Verde River, and Little Colorado River in Arizona.
@@
 
-#### Spatial Trends
-
-TALK ABOUT THE SPATIAL TRENDS IN THE MAP
+@fig-instrumented-trend illustrates the spatial variation in BFI trends across the study area.
+Statistically significant decreasing trends are observed at 16.1% of sites, while increasing trends are found at 8.8% of sites, with no clear regional patterns for either.
+In the Basin and Range physiographic region, 9% of sites show increasing trends, while 16.7% exhibit decreasing trends.
+In the Colorado Plateau region, increasing trends occur at 6.1% of sites, and decreasing trends are observed at 14.3% of sites.
 
@@
 
-| Classification Group                    | n   | n_pos | n_neg | perc_pos | perc_neg |
-|-------------------------|----------|----------|----------|----------|----------|
+| **Classification Group** | **n** | **n_pos** | **n_neg** | **perc_pos** | **perc_neg** |
+|------------|------------|------------|------------|------------|------------|
 | Precipitation - Monsoon Dominated | 87 | 8 | 21 | 0.092 | 0.241 |
@@
 Trends are classified as coincident when the direction of the climate variable trend (positive or negative) aligns with the trend observed in baseflow or BFI (@tbl-coincident-trends).
-This analysis includes both significant and non-significant trends, which is appropriate in environmental studies where the influence of complex, interconnected processes may not always manifest as statistically significant patterns over limited observational periods.
+This analysis includes both significant and non-significant trends, which is appropriate where the influence of complex, interconnected processes may not always manifest as statistically significant patterns over limited observational periods [@ficklin2016].
 
@@
 |-----------|:--------------------:|:------------------------------:|
-| BFI       |         AET          |             44.88              |
+| BFI       |        ET~O~         |             44.88              |
 |           |    Precipitation     |             53.17              |
 |           |     Temperature      |             47.32              |
-| Base Flow |         AET          |             55.12              |
+| Base Flow |        ET~O~         |             55.12              |
 |           |    Precipitation     |             64.88              |
@@
 
-| Classification Group         | n    | R2    | MSE   | RMSE  | MAE   | NSE   | pbias |
-|------------------------------|------|-------|-------|-------|-------|-------|-------|
+| **Classification Group** | **n** | **R2** | **MSE** | **RMSE** | **MAE** | **NSE** | **pbias** |
+|---------|---------|---------|---------|---------|---------|---------|---------|
 | Climate - Monsoon Dominated | 3039 | 0.633 | 0.016 | 0.126 | 0.074 | 0.619 | -13.7 |
@@
 
-The predictors used to estimate BFI at ungauged sites were evaluated for their importance in the final XGBoost model, as illustrated in @suppfig-shap_values.
+The predictors used to estimate BFI at ungauged sites were evaluated for their importance in the final XGBoost model, as illustrated in @fig-shap_values.
 The most influential feature for predicting long-term BFI is basin elevation.
 While elevation itself does not directly affect base flow characteristics, it has consistently been identified as a key predictor in previous BFI studies [@singh2018; @beck2013].
-The significance of elevation likely stems from its cumulative effects, such as lower temperatures, increased orographic precipitation, and prolonged snowpack persistence at higher altitudes.
+The importance of elevation aligns with findings from [@beck2013], highlighting its role as a proxy for climate variables such as temperature, precipitation, and snowpack duration.
 Seasonal snowpack duration, in particular, has been shown to strongly correlate with springflow and groundwater recharge in this region [@donovan_karst_2022].
@@
 In some cases, higher precipitation values correlate with lower BFI, likely due to a larger proportion of precipitation contributing to runoff rather than infiltration and groundwater recharge.
+
+![SHAP value plot of features used in final model. LC indicates land cover, and soil types are defined by hydrologic soil group.](images/shap_summary.png){#fig-shap_values}
 
@@
 
-#### Predicted Long-term Mean BFI
+#### Predicted Long-term BFI
 
-The long-term BFI (1991-2020) for all stream reaches in Arizona of Strahler stream order 3 and greater are shown in @fig-bfi-streams and the regionalized (HUC-8) long-term BFI is shown in @fig-bfi-huc .
+The regionalized (HUC-8) long-term BFI (1991–2020) is shown in \@fig-bfi-huc, while the long-term BFI for stream reaches with a Strahler stream order of 3 or greater is presented in @fig-bfi-streams.
+Basins with high BFIs, such as those along the Grand Canyon in the northwestern part of the study area, indicate greater surface water and groundwater interaction.
+Elevated BFI values are also observed along portions of the Mogollon Rim, a heavily forested region with high precipitation that marks the transition between physiographic regions.
+Additionally, headwater regions of perennial rivers tend to exhibit higher BFI values.
+In contrast, low BFI values are found in areas like the Defiance Uplift in northeastern Arizona and the arid southern regions of the state.
 
@@
 
-## Conclusion {#sec-conclusion}
+@fig-bfi-streams indicates the predicted long-term BFI for stream reaches in Arizona of a Strahler stream order of 3 or greater.
+Stream-level BFI values often correspond to the long-term BFI observed in HUC-8 basins, as geospatial predictor data were aggregated at the basin level.
+@fig-bfi-streams highlights stream density across the region, revealing areas with higher concentrations of perennial streams, particularly in regions with favorable hydrogeologic conditions, such as spring-fed systems in the Verde River and the Salt River headwaters.
+Conversely, lower BFI values are observed in arid regions and those with lower stream densities such as parts of the Little Colorado River Basin, where groundwater recharge is limited [@flint_flint07].
+While BFI values and their spatial patterns may align with certain terrain features or climatic patterns, it is important to recognize that BFI is driven by complex interactions among multiple factors [@fig-shap_values].
+
+## Summary & Conclusions {#sec-conclusion}
 
@@
 ## References {.unnumbered}
-
-## Supporting Information
-
-![SHAP value plot of features used in final model. LC indicates land cover, and soil types are defined by hydrologic soil group.](images/supplemental/shap_summary.png){#suppfig-shap_values}

### 2024-12-16 MANUSCRIPT.qmd
draft

@@
 Base flow  relationships are analyzed with coincident trends in climate variables such as precipitation, evapotranspiration, and temperature.
-Spatial and climatic trends reveal  variability in base flow contributions, providing insight into groundwater-surface water interactions in arid and semi-arid landscapes.
+Spatial and climatic trends reveal  variability in base-flow contributions, providing insight into groundwater-surface water interactions in arid and semi-arid landscapes.
 Building on this analysis, we applied machine learning methods to predict BFI in ungauged basins, addressing the challenges of Arizona’s sparse streamgage network.
 Using the eXtreme Gradient Boosting (XGBoost) algorithm trained on hydroclimate and physiographic predictors, we estimated long-term BFI from 1991 to 2020.
-This combined approach integrates observational data with predictive modeling to enhance our understanding of base flow processes and provide a framework for water resource management in data-limited regions.
+This combined approach integrates observational data with predictive modeling to enhance our understanding of base-flow processes and provide a framework for water resource management in data-limited regions.
 
@@
 Increasing groundwater extraction, changes in land cover/land use, and changes in precipitation patterns due to climate change affect the timing and volumes of base flow [@tan2020; @taylor2013].
-Effective management of water quantity and quality depends on understanding seasonal and interannual base flow patterns and long-term changes in base flow behavior.
+Effective management of water quantity and quality depends on understanding seasonal and interannual base-flow patterns and long-term changes in base-flow behavior.
 
@@
 This study develops a technique for estimating BFI in ungauged basins across Arizona and evaluates the state’s long-term BFI.
-Regional trends in baseflow and BFI at instrumented sites are also analyzed, with these trends being linked to coincident trends in precipitation, evapotranspiration (ET~O~​), and temperature over the same periods.
+Regional trends in base flow and BFI at instrumented sites are also analyzed, with these trends being linked to coincident trends in precipitation, evapotranspiration (ET~O~​), and temperature over the same periods.
 Using a machine learning model trained on the hydrogeologic characteristics of surface water basins, we estimate the mean BFI for ungauged basins from 1991 to 2020.
@@
 This study focuses on natural, streamflow-influencing dynamics, so streamgages affected by regulation or diversions were excluded.
-These gages were identified through annual reports on water data published by the USGS [@usgs2010].
+These streamgages were identified through annual reports on water data published by the USGS [@usgs2010].
 Furthermore, streamgages along the Colorado River were omitted because they represent managed flows governed by the Colorado River Compact.
@@
 
-Statistical analyses were conducted on annual BFI and base flow values from instrumented streamgages to identify temporal trends using the Mann-Kendall nonparametric trend test [@kendall1970; @mann1945nonparametric].
+Statistical analyses were conducted on annual BFI and base-flow values from instrumented streamgages to identify temporal trends using the Mann-Kendall nonparametric trend test [@kendall1970; @mann1945nonparametric].
 This test detects monotonic trends in datasets that are non-parametric and assumes the absence of autocorrelation among observations.
@@
 While decreasing trends dominate, both increasing and decreasing trends are observed within each classification.
-Monsoon-dominated regions exhibit a higher proportion of significant negative trends (24.1%) compared to snowmelt-dominated regions (10.2%), suggesting that monsoon-dominated systems may be more vulnerable to declining baseflows.
-Among climate classifications, warm-dry climates have the highest proportion of negative trends (20.0%), followed by warm-wet climates (19.4%), indicating that regions with higher temperatures are more prone to baseflow declines.
+Monsoon-dominated regions exhibit a higher proportion of significant negative trends (24.1%) compared to snowmelt-dominated regions (10.2%), suggesting that monsoon-dominated systems may be more vulnerable to declining base flow.
+Among climate classifications, warm-dry climates have the highest proportion of negative trends (20.0%), followed by warm-wet climates (19.4%), indicating that regions with higher temperatures are more prone to base flow declines.
 Low-slope regions show a greater prevalence of negative trends (20.4%) compared to high-slope regions (11.8%).
-This suggests that flatter areas may be more susceptible to baseflow reductions, potentially due to differences in hydrologic connectivity and recharge dynamics.
+This suggests that flatter areas may be more susceptible to base-flow reductions, potentially due to differences in hydrologic connectivity and recharge dynamics.
 
 | **Classification Group** | **n** | **n_pos** | **n_neg** | **perc_pos** | **perc_neg** |
-|------------|------------|------------|------------|------------|------------|
+|----|----|----|----|----|----|
 | Precipitation - Monsoon Dominated | 87 | 8 | 21 | 0.092 | 0.241 |
@@
 
-Given the variations in the period of record across the instrumented network (see SUPPLEMENTAL - Period of Record figure), we analyzed trends in baseflow and BFI in relation to coincident trends in climate variables.
-Trends are classified as coincident when the direction of the climate variable trend (positive or negative) aligns with the trend observed in baseflow or BFI (@tbl-coincident-trends).
+Given the variations in the period of record across the instrumented network (see SUPPLEMENTAL - Period of Record figure), we analyzed trends in base flow and BFI in relation to coincident trends in climate variables.
+Trends are classified as coincident when the direction of the climate variable trend (positive or negative) aligns with the trend observed in base flow or BFI (@tbl-coincident-trends).
 This analysis includes both significant and non-significant trends, which is appropriate where the influence of complex, interconnected processes may not always manifest as statistically significant patterns over limited observational periods [@ficklin2016].
 
-The analysis shows that baseflow and BFI trends most frequently align with precipitation trends (64.88% and 53.17%, respectively), emphasizing precipitation as the primary driver of local groundwater recharge and discharge.
-Coincidence with AET (actual evapotranspiration) trends (55.12% for baseflow and 44.88% for BFI) suggests that evapotranspiration also plays a significant role, particularly in arid regions where it can reduce recharge or baseflow during dry periods.
-In contrast, temperature trends show lower percentages of coincidence, often opposing baseflow and BFI trends.
-Specifically, positive (negative) temperature trends are frequently associated with negative (positive) baseflow trends (60.98%) and BFI trends (52.68%).
+The analysis shows that base flow and BFI trends most frequently align with precipitation trends (64.88% and 53.17%, respectively), emphasizing precipitation as the primary driver of local groundwater recharge and discharge.
+Coincidence with AET (actual evapotranspiration) trends (55.12% for base flow and 44.88% for BFI) suggests that evapotranspiration also plays a significant role, particularly in arid regions where it can reduce recharge or base flow during dry periods.
+In contrast, temperature trends show lower percentages of coincidence, often opposing base flow and BFI trends.
+Specifically, positive (negative) temperature trends are frequently associated with negative (positive) base-flow trends (60.98%) and BFI trends (52.68%).
 These results highlight the complex interplay between climatic variables and hydrological processes.
-While precipitation exerts the strongest influence on baseflow and BFI, evapotranspiration and temperature add further variability in specific environmental contexts.
+While precipitation exerts the strongest influence on base flow and BFI, evapotranspiration and temperature add further variability in specific environmental contexts.
 
@@
 | **Classification Group** | **n** | **R2** | **MSE** | **RMSE** | **MAE** | **NSE** | **pbias** |
-|---------|---------|---------|---------|---------|---------|---------|---------|
+|----|----|----|----|----|----|----|----|
 | Climate - Monsoon Dominated | 3039 | 0.633 | 0.016 | 0.126 | 0.074 | 0.619 | -13.7 |
@@
 The most influential feature for predicting long-term BFI is basin elevation.
-While elevation itself does not directly affect base flow characteristics, it has consistently been identified as a key predictor in previous BFI studies [@singh2018; @beck2013].
+While elevation itself does not directly affect base-flow characteristics, it has consistently been identified as a key predictor in previous BFI studies [@singh2018; @beck2013].
 The importance of elevation aligns with findings from [@beck2013], highlighting its role as a proxy for climate variables such as temperature, precipitation, and snowpack duration.
@@
 
-The regionalized (HUC-8) long-term BFI (1991–2020) is shown in \@fig-bfi-huc, while the long-term BFI for stream reaches with a Strahler stream order of 3 or greater is presented in @fig-bfi-streams.
+The regionalized (HUC-8) long-term BFI (1991–2020) is shown in @fig-bfi-huc, while the long-term BFI for stream reaches with a Strahler stream order of 3 or greater is presented in @fig-bfi-streams.
 Basins with high BFIs, such as those along the Grand Canyon in the northwestern part of the study area, indicate greater surface water and groundwater interaction.
@@
 
--   This regional model, trained on a relatively sparse dataset over a region spanning 295,253 km^2^, is shown to predict BFI well spatially and across climate classifications.
+This study provides new insights into base-flow dynamics and groundwater contributions in Arizona’s dryland rivers by combining an analysis of instrumented streamflow records with machine learning predictions for ungauged basins.
+The results highlight significant spatial variability in BFI, with approximately 32% of Arizona’s long-term streamflow originating from groundwater discharge.
+Regions such as the Grand Canyon and Mogollon Rim demonstrate high BFI values due to strong groundwater-surface water interactions, while areas like the Little Colorado River Basin exhibit low BFI values, reflecting limited groundwater recharge.
 
--
+Our analysis of BFI trends in gauged catchments revealed that precipitation is the primary driver of base-flow variability, with evapotranspiration and temperature contributing additional complexity.
+These findings emphasize the critical role of climate-hydrology interactions in shaping groundwater contributions to streamflow.
+Inverse trends between temperature and BFI suggest that further warming could reduce groundwater contributions to streamflow.
+Coincident trends in precipitation and BFI further underscore the importance understanding recharge processes in arid and semi-arid landscapes.
+
+Using an XGBoost machine learning algorithm, we successfully predicted long-term BFI in ungauged basins, achieving strong model performance (R² = 0.764, RMSE = 0.129).
+The model performed well across all classifications, demonstrating its robustness in capturing base-flow dynamics across a region with substantial variability in climate, elevation, and physiographic characteristics.
+Key predictors included elevation, land cover, and soil type, highlighting the importance of integrating hydroclimate and physiographic characteristics into regional hydrological models.
+These predictions address the limitations posed by Arizona’s sparse streamgage network, offering a scalable approach to estimate BFI in data-limited regions.
+
+DATA LIMITATIONS - ABE
+
+This study demonstrates the benefits of combining observational records with machine learning to improve our understanding of streamflow processes in drylands.
+Future work should explore the projected effects of climate change on base-flow processes and developing models for other data-poor regions.
+The framework presented here has broad applicability to other arid and semi-arid regions worldwide and can inform water resource management strategies aimed at addressing water scarcity and adapting to climate variability.
 

### 2024-12-16 MANUSCRIPT.qmd
Update MANUSCRIPT.qmd

@@
 |              | Variable                  | Source       | Geoprocessing          |
-|--------------|---------------------------|--------------|------------------------|
+|----------------|----------------------|----------------|-------------------|
 | Hydroclimate | Precipitation\*           | PRISM        | Basin average          |
@@
 | **Classification Group** | **n** | **n_pos** | **n_neg** | **perc_pos** | **perc_neg** |
-|----|----|----|----|----|----|
+|------------|------------|------------|------------|------------|------------|
 | Precipitation - Monsoon Dominated | 87 | 8 | 21 | 0.092 | 0.241 |
@@
 | **Classification Group** | **n** | **R2** | **MSE** | **RMSE** | **MAE** | **NSE** | **pbias** |
-|----|----|----|----|----|----|----|----|
+|---------|---------|---------|---------|---------|---------|---------|---------|
 | Climate - Monsoon Dominated | 3039 | 0.633 | 0.016 | 0.126 | 0.074 | 0.619 | -13.7 |

### 2025-01-07 MANUSCRIPT.qmd
Abe Edits Draft

@@
 ---
-title: Regional-scale Base-flow Index of Gauged and Ungauged Basins in Drylands | Regional Base-Flow Index in Arid Landscapes Using Machine Learning and Instrumented Records | Estimating Base-Flow Index in Gauged and Ungauged Dryland Basins
+title: Regional Base-Flow Index in Arid Landscapes Using Machine Learning and Instrumented Records
+subtitle: Regional-scale Base-flow Index of Gauged and Ungauged Basins in Drylands |  | Estimating Base-Flow Index in Gauged and Ungauged Dryland Basins
 author:
@@
 To tackle the challenge of quantifying base flow in ungauged catchments, numerous studies have applied both regression and machine learning methods.
-[@ahiablame2013] found that using a regression model to estimate annual base flow of ungauged catchments was reasonably easy and accurate.
+@ahiablame2013 found that using a regression model to estimate annual base flow of ungauged catchments was reasonably easy and accurate.
 @beck2013 overcame the nonlinearity of basin characteristics and improved results of multivariate analyses by using artificial neural networks to estimate BFI globally.
@@
 This study develops a technique for estimating BFI in ungauged basins across Arizona and evaluates the state’s long-term BFI.
-Regional trends in base flow and BFI at instrumented sites are also analyzed, with these trends being linked to coincident trends in precipitation, evapotranspiration (ET~O~​), and temperature over the same periods.
+Regional trends in base flow and BFI at instrumented sites are also analyzed, with these trends being linked to coincident trends in precipitation, evapotranspiration (ET~O~), and temperature over the same periods.
 Using a machine learning model trained on the hydrogeologic characteristics of surface water basins, we estimate the mean BFI for ungauged basins from 1991 to 2020.
@@
 This study focuses on natural, streamflow-influencing dynamics, so streamgages affected by regulation or diversions were excluded.
-These streamgages were identified through annual reports on water data published by the USGS [@usgs2010].
+The regulated river streamgages were identified through annual reports on water data published by the USGS [@usgs2010].
 Furthermore, streamgages along the Colorado River were omitted because they represent managed flows governed by the Colorado River Compact.
@@
 
-While our data selection ensures a robust record for this analysis, it also highlights gaps in spatial coverage that the machine learning model aims to address.
-Arizona has 184 active USGS streamflow stations (as of 2024) across an area of 295,253 km², whereas Indiana has 189 active stations within an area of 94,326 km².
+While our data selection ensures a robust record for this analysis, it also highlights gaps in spatial coverage that the machine learning model can address.
+Arizona has 184 active USGS streamflow stations (as of 2024) across an area of 295,253 km², whereas Indiana (a humid state in the U.S.) has 189 active stations within an area of 94,326 km².
 This translates to a streamgage density of approximately 2.004 gages per 1,000 km² in Indiana, over three times higher than Arizona's density of 0.623 gages per 1,000 km².
@@
 By calculating these percentages, we aimed to get a more comprehensive understanding of landscape composition across space.
-Land cover from USGS-NLCD (National Land Cover Database), hydrologic soil group from SSURGO (Soil Survey Geographic Database), and underlying geology and karst from USGS were all similarly averaged across the basins.
+Land cover was acquired from USGS-NLCD (National Land Cover Database), hydrologic soil group from SSURGO (Soil Survey Geographic Database), and underlying geology and karst from USGS were all similarly averaged across the basins.
 Aggregating variables to align with the HUC 8 boundaries allowed for more precise predictions of BFI by integrating spatial variations within each basin.
@@
 
-DATA LIMITATIONS - ABE
+While there has been an increase of stream gages in Arizona to comply with other imperatives, such as the Clean Water Act, these gages are not of sufficient quality for base-flow analyses.
+New, non-USGS stream gages are typically installed by flood control districts for flood flow monitoring, and do not accurately track base flow.
+New USGS stream gages have been added over the past 10 years to address instream flow rights and to improve the density of monitoring in the future.
 

### 2025-01-13 MANUSCRIPT.qmd
Update MANUSCRIPT.qmd

@@
 |              | Variable                  | Source       | Geoprocessing          |
-|----------------|----------------------|----------------|-------------------|
+|-----------------|---------------------|-----------------|------------------|
 | Hydroclimate | Precipitation\*           | PRISM        | Basin average          |

### 2025-01-15 MANUSCRIPT.qmd
new figs and Neha edits

@@
 These techniques have varying levels of applicability depending on the spatial scale, time span, and the scope of the study.
-Comparisons of various base-flow separation techniques have been made in [@eckhardt2005; @eckhardt2008; @nathan1990] , but this study does not explore the superiority of different methods.
+Comparisons of various base-flow separation techniques have been made in previous studies [ e.g.
+@eckhardt2005; @eckhardt2008; @nathan1990]; this study does not explore the superiority of different methods.
 
-Base flow was calculated using a single-parameter, recursive digital filter technique from [@nathan1990].
+Base flow was calculated using a single-parameter, recursive digital filter technique from @nathan1990.
 This base-flow separation technique is based on a recursive digital filter used in signal analysis that separates high-frequency signals (quickflow) from low-frequency signals (base flow) [@lyne1979].
 @eckhardt2023 noted that recursive digital filters lack a physical basis, but as the method is easy to automate, objective, and repeatable, it is appropriate for a regional-scale study.
-The Lyne-Hollick filter has been used in [@ARNOLD200021; @santhi2008; @bloomfield2009; @singh2018].
-It takes the form of
+The Lyne-Hollick filter has been used in multiple studies [@ARNOLD200021; @santhi2008; @bloomfield2009; @singh2018], and it takes the form of
 
@@
 The implementation of machine learning models to predict hydrologic indices has been successful in past studies [@singh2018; @schmidt_ML_2020; @Rozos2021Machine].
-In this work, we used the eXtreme Gradient Boosting (XGBoost) algorithm [@chen2016xgboost] to predict BFI at ungauged locations using catchment characteristics as predictors.
+In this work, we used the eXtreme Gradient Boosting (XGBoost) algorithm [@chen2016xgboost] to predict BFI at ungauged locations using catchment characteristics as predictors [@tbl-predictors].
 The XGBoost algorithm is a decision tree-based ensemble algorithm, which can be adapted for either regression or classification problems.
@@
 |              | Variable                  | Source       | Geoprocessing          |
-|-----------------|---------------------|-----------------|------------------|
+|------------------|--------------------|------------------|------------------|
 | Hydroclimate | Precipitation\*           | PRISM        | Basin average          |
@@
 However, this process generates new features that often lack the physical interpretability of the original variables.
-In contrast, feature selection identifies a subset of the original features, preserving their physical meaning while improving model readability and interpretability.
+In contrast, feature selection identifies a subset of the original features, preserving their physical meaning while improving model readability and interpretability [@feat_selec2017].
 In this study, supervised feature selection was employed to reduce the number of predictors, which enhanced learning performance, reduced computational costs, and mitigated overfitting.
@@
 This test detects monotonic trends in datasets that are non-parametric and assumes the absence of autocorrelation among observations.
-This test is widely used in studies of this kind [@ficklin2016; @ayers2019; @woodhouse_udall22].
+This test is widely used in studies of this nature [@ficklin2016; @ayers2019; @woodhouse_udall22].
 
@@
 snowmelt) was identified by analyzing streamflow hydrographs for each station, focusing on peak flow periods during the monsoon season (July–September) and the snowmelt season (March–June).
-Physiographic region was assigned based on which region the streamgage is in.
-Climate classifications were defined as warm (above the median temperature of Arizona), cool (below the median temperature), wet (above the median precipitation), and dry (below the median precipitation).
+Physiographic region was assigned based on which region the streamgage is located.
+Climate classifications were defined as warm (above the long-term median temperature of Arizona), cool (below the long-term median temperature), wet (above the long-term median precipitation), and dry (below the long-term median precipitation).
 Slope was categorized as high (above the median slope) and low (below the median slope).
@@
 While decreasing trends dominate, both increasing and decreasing trends are observed within each classification.
-Monsoon-dominated regions exhibit a higher proportion of significant negative trends (24.1%) compared to snowmelt-dominated regions (10.2%), suggesting that monsoon-dominated systems may be more vulnerable to declining base flow.
+Monsoon-dominated regions exhibit a higher proportion of significant negative trends (24.1%) compared to snowmelt-dominated regions (10.2%), suggesting that monsoon-dominated systems are more consistently correlated with declining base flow.
 Among climate classifications, warm-dry climates have the highest proportion of negative trends (20.0%), followed by warm-wet climates (19.4%), indicating that regions with higher temperatures are more prone to base flow declines.
@@
 The analysis shows that base flow and BFI trends most frequently align with precipitation trends (64.88% and 53.17%, respectively), emphasizing precipitation as the primary driver of local groundwater recharge and discharge.
-Coincidence with AET (actual evapotranspiration) trends (55.12% for base flow and 44.88% for BFI) suggests that evapotranspiration also plays a significant role, particularly in arid regions where it can reduce recharge or base flow during dry periods.
+Coincidence with ET~O~ (reference evapotranspiration) trends (55.12% for base flow and 44.88% for BFI) suggests that evapotranspiration also plays a significant role, particularly in arid regions where it can reduce recharge or base flow during dry periods.
 In contrast, temperature trends show lower percentages of coincidence, often opposing base flow and BFI trends.
@@
 
-: Coincident trends of climate variables (AET, precipitation, temperature) with base flow and BFI trends. {#tbl-coincident-trends}
+: Coincident trends of climate variables (ET~O~, precipitation, temperature) with base flow and BFI trends. {#tbl-coincident-trends}
 
@@
 The agreement between "out-of-bag" predictions (blind cross-validation, treating each site as ungauged) and observed values is acceptable (R^2^ = 0.764) indicating that the model performs well across the full dataset.
-The overall Root Mean Square Error (RMSE) is 0.129 and the overall percent bias (pbias) is -5.6.
+The overall RMSE is 0.129 and the overall percent bias (pbias) is -5.6%.
 Model performance metrics across various classifications are summarized in @tbl-performance.
@@
 While elevation itself does not directly affect base-flow characteristics, it has consistently been identified as a key predictor in previous BFI studies [@singh2018; @beck2013].
-The importance of elevation aligns with findings from [@beck2013], highlighting its role as a proxy for climate variables such as temperature, precipitation, and snowpack duration.
+The importance of elevation aligns with findings from @beck2013, highlighting its role as a proxy for climate variables such as temperature, precipitation, and snowpack duration.
 Seasonal snowpack duration, in particular, has been shown to strongly correlate with springflow and groundwater recharge in this region [@donovan_karst_2022].
-This relationship is further supported by hydroclimate features.
-Higher temperatures tend to negatively influence BFI, while precipitation exhibits a mixed influence.
+This relationship is further supported by hydroclimate features where higher temperatures tend to negatively influence BFI, while precipitation exhibits a mixed influence.
 In some cases, higher precipitation values correlate with lower BFI, likely due to a larger proportion of precipitation contributing to runoff rather than infiltration and groundwater recharge.
@@
 
-While there has been an increase of stream gages in Arizona to comply with other imperatives, such as the Clean Water Act, these gages are not of sufficient quality for base-flow analyses.
+While there has been an increase of stream gages in Arizona to comply with regulatory imperatives, such as the Clean Water Act, these gages are not of sufficient quality for base-flow analyses.
 New, non-USGS stream gages are typically installed by flood control districts for flood flow monitoring, and do not accurately track base flow.

### 2025-01-15 MANUSCRIPT.qmd
Update MANUSCRIPT.qmd

@@
 These techniques have varying levels of applicability depending on the spatial scale, time span, and the scope of the study.
-Comparisons of various base-flow separation techniques have been made in previous studies [ e.g.
+Comparisons of various base-flow separation techniques have been made in previous studies [e.g.
 @eckhardt2005; @eckhardt2008; @nathan1990]; this study does not explore the superiority of different methods.
@@
 RMSE provided a consistent and interpretable measure of the model’s accuracy throughout the training and validation process.
+
+![Schematic of k-fold cross validation.
+The dataset is randomly divided into k stratified folds.
+Each fold serves as the validation set once, while the remaining folds are combined to create a training set for model development.
+Performance metrics for the test set are calculated and recorded, and this process is repeated for all k folds.](images/k-fold_fig.png){#fig-k-fold width="6000"}
 
@@
 ## Results
+
+### BFI of Ungauged Catchments
+
+#### Model Validation
+
+Predicted values of BFI are plotted against observed values for the entire period of record of the instrumented dataset in @fig-actual_predicted .
+The agreement between "out-of-bag" predictions (blind cross-validation, treating each site as ungauged) and observed values is acceptable (R^2^ = 0.764) indicating that the model performs well across the full dataset.
+The overall RMSE is 0.129 and the overall percent bias (pbias) is -5.6%.
+Model performance metrics across various classifications are summarized in @tbl-performance.
+These performance metrics demonstrate that the regional model performs consistently well across different spatial and climatic classifications.
+However, the negative pbias values across all classifications, along with the overall pbias, indicate a systematic underprediction of BFI by the model.
+Categories with relatively lower R^2^ and Nash-Sutcliffe Efficiency (NSE) values also exhibit higher biases, reflecting weaker model performance in those specific contexts.
+
+![Linear relationship between observed BFI and predicted BFI. The solid line is the 1:1 line, the dashed line is regressed to the data.](images/actual-predicted.png){#fig-actual_predicted}
+
+| **Classification Group** | **n** | **R2** | **MSE** | **RMSE** | **MAE** | **NSE** | **pbias** |
+|---------|---------|---------|---------|---------|---------|---------|---------|
+| Climate - Monsoon Dominated | 3039 | 0.633 | 0.016 | 0.126 | 0.074 | 0.619 | -13.7 |
+| Climate - Snowmelt Dominated | 4685 | 0.733 | 0.015 | 0.121 | 0.087 | 0.725 | -3.5 |
+| PhysRegion - Basin&Range | 6147 | 0.733 | 0.016 | 0.127 | 0.084 | 0.724 | -6.3 |
+| PhysRegion - CO Plateau | 1577 | 0.846 | 0.011 | 0.104 | 0.073 | 0.843 | -3.8 |
+| Climate - Warm-Wet | 1506 | 0.693 | 0.014 | 0.117 | 0.077 | 0.685 | -8.2 |
+| Climate - Warm-Dry | 2351 | 0.693 | 0.022 | 0.147 | 0.092 | 0.675 | -11.9 |
+| Climate - Cool-Wet | 2350 | 0.738 | 0.011 | 0.106 | 0.078 | 0.736 | -1.7 |
+| Climate - Cool-Dry | 1517 | 0.831 | 0.012 | 0.111 | 0.078 | 0.827 | -4.3 |
+| Slope - High | 3795 | 0.776 | 0.012 | 0.111 | 0.079 | 0.771 | -3.3 |
+| Slope - Low | 3929 | 0.724 | 0.018 | 0.133 | 0.085 | 0.713 | -9.1 |
+
+: Performance of model predictions for BFI for all sites split by various classifications.
+n is number of observatios, R^2^ is the coefficient of determination of a linear regression, MSE is mean-squared-error, RMSE is root-mean-squared-error, MAE is mean-absolute-error, NSE is Nash-Sucliffe efficiency, and pbias is percent bias.
+{#tbl-performance}
+
+#### Predictor Importance {#sec-predictor-importance}
+
+The predictors used to estimate BFI at ungauged sites were evaluated for their importance in the final XGBoost model, as illustrated in @fig-shap_values.
+The most influential feature for predicting long-term BFI is basin elevation.
+While elevation itself does not directly affect base-flow characteristics, it has consistently been identified as a key predictor in previous BFI studies [@singh2018; @beck2013].
+The importance of elevation aligns with findings from @beck2013, highlighting its role as a proxy for climate variables such as temperature, precipitation, and snowpack duration.
+Seasonal snowpack duration, in particular, has been shown to strongly correlate with springflow and groundwater recharge in this region [@donovan_karst_2022].
+This relationship is further supported by hydroclimate features where higher temperatures tend to negatively influence BFI, while precipitation exhibits a mixed influence.
+In some cases, higher precipitation values correlate with lower BFI, likely due to a larger proportion of precipitation contributing to runoff rather than infiltration and groundwater recharge.
+
+![SHAP value plot of features used in final model. LC indicates land cover, and soil types are defined by hydrologic soil group.](images/shap_summary.png){#fig-shap_values}
+
+Land cover and land use predictors also play a significant role in BFI estimation.
+Analysis of SHAP values indicates that a higher percentage of evergreen forest positively influences BFI predictions, while higher proportions of shrubland and developed land exert a negative influence.
+Similarly, hydrologic soil types show distinct trends in their impact on BFI.
+Soil Type C, characterized by moderately high runoff potential (20-40% clay), tends to negatively influence BFI.
+In contrast, Soil Type A, which has low runoff potential and facilitates rapid water infiltration, exhibits a mixed influence [@nrcs2009].
+The mixed effects of Soil Type A are likely due to SHAP values capturing interactions between features rather than direct relationships.
+For example, regions dominated by Soil Type A may also have steep slopes or sparse vegetation.
+Additionally, the relationship between Soil Type A and BFI is likely non-linear and influenced by the complex dynamics of regional base-flow processes.
+
+#### Predicted Long-term BFI
+
+The regionalized (HUC-8) long-term BFI (1991–2020) is shown in @fig-bfi-huc, while the long-term BFI for stream reaches with a Strahler stream order of 3 or greater is presented in @fig-bfi-streams.
+Basins with high BFIs, such as those along the Grand Canyon in the northwestern part of the study area, indicate greater surface water and groundwater interaction.
+Elevated BFI values are also observed along portions of the Mogollon Rim, a heavily forested region with high precipitation that marks the transition between physiographic regions.
+Additionally, headwater regions of perennial rivers tend to exhibit higher BFI values.
+In contrast, low BFI values are found in areas like the Defiance Uplift in northeastern Arizona and the arid southern regions of the state.
+
+![Predicted long-term BFI values for 8-digit HUC (1991-2020)](images/BFI_HUC8_20241203.png){#fig-bfi-huc}
+
+![Predicted long-term BFI values for all stream reaches of Strahler stream order 3 and greater (1991-2020)](images/BFI_AllStream.png){#fig-bfi-streams}
+
+@fig-bfi-streams indicates the predicted long-term BFI for stream reaches in Arizona of a Strahler stream order of 3 or greater.
+Stream-level BFI values often correspond to the long-term BFI observed in HUC-8 basins, as geospatial predictor data were aggregated at the basin level.
+@fig-bfi-streams highlights stream density across the region, revealing areas with higher concentrations of perennial streams, particularly in regions with favorable hydrogeologic conditions, such as spring-fed systems in the Verde River and the Salt River headwaters.
+Conversely, lower BFI values are observed in arid regions and those with lower stream densities such as parts of the Little Colorado River Basin, where groundwater recharge is limited [@flint_flint07].
+While BFI values and their spatial patterns may align with certain terrain features or climatic patterns, it is important to recognize that BFI is driven by complex interactions among multiple factors [@fig-shap_values].
 
@@
 
-### BFI of Ungauged Catchments
-
-#### Model Validation
-
-Predicted values of BFI are plotted against observed values for the entire period of record of the instrumented dataset in @fig-actual_predicted .
-The agreement between "out-of-bag" predictions (blind cross-validation, treating each site as ungauged) and observed values is acceptable (R^2^ = 0.764) indicating that the model performs well across the full dataset.
-The overall RMSE is 0.129 and the overall percent bias (pbias) is -5.6%.
-Model performance metrics across various classifications are summarized in @tbl-performance.
-These performance metrics demonstrate that the regional model performs consistently well across different spatial and climatic classifications.
-However, the negative pbias values across all classifications, along with the overall pbias, indicate a systematic underprediction of BFI by the model.
-Categories with relatively lower R^2^ and Nash-Sutcliffe Efficiency (NSE) values also exhibit higher biases, reflecting weaker model performance in those specific contexts.
-
-![Linear relationship between observed BFI and predicted BFI. The solid line is the 1:1 line, the dashed line is regressed to the data.](images/actual-predicted.png){#fig-actual_predicted}
-
-| **Classification Group** | **n** | **R2** | **MSE** | **RMSE** | **MAE** | **NSE** | **pbias** |
-|---------|---------|---------|---------|---------|---------|---------|---------|
-| Climate - Monsoon Dominated | 3039 | 0.633 | 0.016 | 0.126 | 0.074 | 0.619 | -13.7 |
-| Climate - Snowmelt Dominated | 4685 | 0.733 | 0.015 | 0.121 | 0.087 | 0.725 | -3.5 |
-| PhysRegion - Basin&Range | 6147 | 0.733 | 0.016 | 0.127 | 0.084 | 0.724 | -6.3 |
-| PhysRegion - CO Plateau | 1577 | 0.846 | 0.011 | 0.104 | 0.073 | 0.843 | -3.8 |
-| Climate - Warm-Wet | 1506 | 0.693 | 0.014 | 0.117 | 0.077 | 0.685 | -8.2 |
-| Climate - Warm-Dry | 2351 | 0.693 | 0.022 | 0.147 | 0.092 | 0.675 | -11.9 |
-| Climate - Cool-Wet | 2350 | 0.738 | 0.011 | 0.106 | 0.078 | 0.736 | -1.7 |
-| Climate - Cool-Dry | 1517 | 0.831 | 0.012 | 0.111 | 0.078 | 0.827 | -4.3 |
-| Slope - High | 3795 | 0.776 | 0.012 | 0.111 | 0.079 | 0.771 | -3.3 |
-| Slope - Low | 3929 | 0.724 | 0.018 | 0.133 | 0.085 | 0.713 | -9.1 |
-
-: Performance of model predictions for BFI for all sites split by various classifications.
-n is number of observatios, R^2^ is the coefficient of determination of a linear regression, MSE is mean-squared-error, RMSE is root-mean-squared-error, MAE is mean-absolute-error, NSE is Nash-Sucliffe efficiency, and pbias is percent bias.
-{#tbl-performance}
-
-#### Predictor Importance {#sec-predictor-importance}
-
-The predictors used to estimate BFI at ungauged sites were evaluated for their importance in the final XGBoost model, as illustrated in @fig-shap_values.
-The most influential feature for predicting long-term BFI is basin elevation.
-While elevation itself does not directly affect base-flow characteristics, it has consistently been identified as a key predictor in previous BFI studies [@singh2018; @beck2013].
-The importance of elevation aligns with findings from @beck2013, highlighting its role as a proxy for climate variables such as temperature, precipitation, and snowpack duration.
-Seasonal snowpack duration, in particular, has been shown to strongly correlate with springflow and groundwater recharge in this region [@donovan_karst_2022].
-This relationship is further supported by hydroclimate features where higher temperatures tend to negatively influence BFI, while precipitation exhibits a mixed influence.
-In some cases, higher precipitation values correlate with lower BFI, likely due to a larger proportion of precipitation contributing to runoff rather than infiltration and groundwater recharge.
-
-![SHAP value plot of features used in final model. LC indicates land cover, and soil types are defined by hydrologic soil group.](images/shap_summary.png){#fig-shap_values}
-
-Land cover and land use predictors also play a significant role in BFI estimation.
-Analysis of SHAP values indicates that a higher percentage of evergreen forest positively influences BFI predictions, while higher proportions of shrubland and developed land exert a negative influence.
-Similarly, hydrologic soil types show distinct trends in their impact on BFI.
-Soil Type C, characterized by moderately high runoff potential (20-40% clay), tends to negatively influence BFI.
-In contrast, Soil Type A, which has low runoff potential and facilitates rapid water infiltration, exhibits a mixed influence [@nrcs2009].
-The mixed effects of Soil Type A are likely due to SHAP values capturing interactions between features rather than direct relationships.
-For example, regions dominated by Soil Type A may also have steep slopes or sparse vegetation.
-Additionally, the relationship between Soil Type A and BFI is likely non-linear and influenced by the complex dynamics of regional base-flow processes.
-
-#### Predicted Long-term BFI
-
-The regionalized (HUC-8) long-term BFI (1991–2020) is shown in @fig-bfi-huc, while the long-term BFI for stream reaches with a Strahler stream order of 3 or greater is presented in @fig-bfi-streams.
-Basins with high BFIs, such as those along the Grand Canyon in the northwestern part of the study area, indicate greater surface water and groundwater interaction.
-Elevated BFI values are also observed along portions of the Mogollon Rim, a heavily forested region with high precipitation that marks the transition between physiographic regions.
-Additionally, headwater regions of perennial rivers tend to exhibit higher BFI values.
-In contrast, low BFI values are found in areas like the Defiance Uplift in northeastern Arizona and the arid southern regions of the state.
-
-![Predicted long-term BFI values for 8-digit HUC (1991-2020)](images/BFI_HUC8_20241203.png){#fig-bfi-huc}
-
-![Predicted long-term BFI values for all stream reaches of Strahler stream order 3 and greater (1991-2020)](images/BFI_AllStream.png){#fig-bfi-streams}
-
-@fig-bfi-streams indicates the predicted long-term BFI for stream reaches in Arizona of a Strahler stream order of 3 or greater.
-Stream-level BFI values often correspond to the long-term BFI observed in HUC-8 basins, as geospatial predictor data were aggregated at the basin level.
-@fig-bfi-streams highlights stream density across the region, revealing areas with higher concentrations of perennial streams, particularly in regions with favorable hydrogeologic conditions, such as spring-fed systems in the Verde River and the Salt River headwaters.
-Conversely, lower BFI values are observed in arid regions and those with lower stream densities such as parts of the Little Colorado River Basin, where groundwater recharge is limited [@flint_flint07].
-While BFI values and their spatial patterns may align with certain terrain features or climatic patterns, it is important to recognize that BFI is driven by complex interactions among multiple factors [@fig-shap_values].
-
 ## Summary & Conclusions {#sec-conclusion}
@@
 
+Using an XGBoost machine learning algorithm, we successfully predicted long-term BFI in ungauged basins, achieving strong model performance (R² = 0.764, RMSE = 0.129).
+The model performed well across all classifications, demonstrating its robustness in capturing base-flow dynamics across a region with substantial variability in climate, elevation, and physiographic characteristics.
+Key predictors included elevation, land cover, and soil type, highlighting the importance of integrating hydroclimate and physiographic characteristics into regional hydrological models.
+These predictions address the limitations posed by Arizona’s sparse streamgage network, offering a scalable approach to estimate BFI in data-limited regions.
+
 Our analysis of BFI trends in gauged catchments revealed that precipitation is the primary driver of base-flow variability, with evapotranspiration and temperature contributing additional complexity.
@@
 Coincident trends in precipitation and BFI further underscore the importance understanding recharge processes in arid and semi-arid landscapes.
-
-Using an XGBoost machine learning algorithm, we successfully predicted long-term BFI in ungauged basins, achieving strong model performance (R² = 0.764, RMSE = 0.129).
-The model performed well across all classifications, demonstrating its robustness in capturing base-flow dynamics across a region with substantial variability in climate, elevation, and physiographic characteristics.
-Key predictors included elevation, land cover, and soil type, highlighting the importance of integrating hydroclimate and physiographic characteristics into regional hydrological models.
-These predictions address the limitations posed by Arizona’s sparse streamgage network, offering a scalable approach to estimate BFI in data-limited regions.
 

### 2025-01-18 MANUSCRIPT.qmd
Update MANUSCRIPT.qmd

@@
 
-![Schematic of k-fold cross validation.
-The dataset is randomly divided into k stratified folds.
+![Schematic of $k$-fold cross validation.
+The dataset is randomly divided into $k$ stratified folds.
 Each fold serves as the validation set once, while the remaining folds are combined to create a training set for model development.
-Performance metrics for the test set are calculated and recorded, and this process is repeated for all k folds.](images/k-fold_fig.png){#fig-k-fold width="6000"}
+Performance metrics for the test set are calculated and recorded, and this process is repeated for all $k$ folds.](images/k-fold_fig.png){#fig-k-fold width="6000"}
 
@@
 
-The analysis shows that base flow and BFI trends most frequently align with precipitation trends (64.88% and 53.17%, respectively), emphasizing precipitation as the primary driver of local groundwater recharge and discharge.
-Coincidence with ET~O~ (reference evapotranspiration) trends (55.12% for base flow and 44.88% for BFI) suggests that evapotranspiration also plays a significant role, particularly in arid regions where it can reduce recharge or base flow during dry periods.
-In contrast, temperature trends show lower percentages of coincidence, often opposing base flow and BFI trends.
-Specifically, positive (negative) temperature trends are frequently associated with negative (positive) base-flow trends (60.98%) and BFI trends (52.68%).
-These results highlight the complex interplay between climatic variables and hydrological processes.
-While precipitation exerts the strongest influence on base flow and BFI, evapotranspiration and temperature add further variability in specific environmental contexts.
-
 |           | **Climate Variable** | **Coincidence Percentage (%)** |
@@
 : Coincident trends of climate variables (ET~O~, precipitation, temperature) with base flow and BFI trends. {#tbl-coincident-trends}
+
+The analysis shows that base flow and BFI trends most frequently align with precipitation trends (64.88% and 53.17%, respectively), emphasizing precipitation as the primary driver of local groundwater recharge and discharge.
+Coincidence with ET~O~ (reference evapotranspiration) trends (55.12% for base flow and 44.88% for BFI) suggests that evapotranspiration also plays a significant role, particularly in arid regions where it can reduce recharge or base flow during dry periods.
+In contrast, temperature trends show lower percentages of coincidence, often opposing base flow and BFI trends.
+Specifically, positive (negative) temperature trends are frequently associated with negative (positive) base-flow trends (60.98%) and BFI trends (52.68%).
+These results highlight the complex interplay between climatic variables and hydrological processes.
+While precipitation exerts the strongest influence on base flow and BFI, evapotranspiration and temperature add further variability in specific environmental contexts.
 
@@
 Inverse trends between temperature and BFI suggest that further warming could reduce groundwater contributions to streamflow.
-Coincident trends in precipitation and BFI further underscore the importance understanding recharge processes in arid and semi-arid landscapes.
+Coincident trends in precipitation and BFI further underscore the importance of understanding recharge processes, especially in arid and semi-arid landscapes, where precipitation events play key roles in replenishing groundwater.
 
-While there has been an increase of stream gages in Arizona to comply with regulatory imperatives, such as the Clean Water Act, these gages are not of sufficient quality for base-flow analyses.
-New, non-USGS stream gages are typically installed by flood control districts for flood flow monitoring, and do not accurately track base flow.
-New USGS stream gages have been added over the past 10 years to address instream flow rights and to improve the density of monitoring in the future.
+However, the utility of the streamgage networks for base-flow analyses in Arizona is limited by their design and focus.There has been an increase in the number of streamgages in the region to meet regulatory imperatives, such as the Clean Water Act.
+Even so, many of these gages are not suited for base-flow studies because they emphasize flood monitoring and lack the ability to accurately measure low-flow dynamics.
+New, non-USGS streamgages are typically installed by flood control districts (e.g.
+the ALERT system) and are designed for tracking flood flows rather than base flow.
+New USGS streamgages have been added over the past decade to address instream flow rights and to improve the density of monitoring in the future.
 

### 2025-01-18 MANUSCRIPT.qmd
Update MANUSCRIPT.qmd

@@
       - Writing - Original Draft
+      - Writing – Review & Editing 
     affiliations:
@@
     corresponding: false
-    roles: []
+    roles:
+      - Conceptualization
+      - Funding acquisition
+      - Project administration
+      - Resources
+      - Supervision
+      - Writing – Review & Editing 
     affiliations:
@@
     corresponding: false
-    roles: []
+    roles: 
+      - Conceptualization
+      - Funding acquisition
+      - Project administration
+      - Resources
+      - Supervision
+      - Writing – Review & Editing 
     affiliations:
@@
     corresponding: false
-    roles: []
+    roles: 
+      - Conceptualization
+      - Resources
+      - Supervision
+      - Software
+      - Writing – Review & Editing 
     affiliations:

### 2025-01-19 MANUSCRIPT.qmd
edits

@@
 title: Regional Base-Flow Index in Arid Landscapes Using Machine Learning and Instrumented Records
-subtitle: Regional-scale Base-flow Index of Gauged and Ungauged Basins in Drylands |  | Estimating Base-Flow Index in Gauged and Ungauged Dryland Basins
 author:
@@
 |  | Mean Temperature\* | PRISM | Basin average |
-|              | Mean Evapotranspiration\* | TerraClimate | Basin average          |
+|  | Reference Evapotranspiration\* | TerraClimate | Basin average |
 | Geospatial | Elevation\* | DEM | Basin average |
@@
 
-![SHAP value plot of features used in final model. LC indicates land cover, and soil types are defined by hydrologic soil group.](images/shap_summary.png){#fig-shap_values}
+![SHAP value plot of features used in final model.
+Land cover features are indicated by the percentage of cover by each land cover type and soil types are defined by hydrologic soil group.](images/shap_summary.png){#fig-shap_values}
 
@@
 
-However, the utility of the streamgage networks for base-flow analyses in Arizona is limited by their design and focus.There has been an increase in the number of streamgages in the region to meet regulatory imperatives, such as the Clean Water Act.
-Even so, many of these gages are not suited for base-flow studies because they emphasize flood monitoring and lack the ability to accurately measure low-flow dynamics.
+However, the utility of the streamgage networks for base-flow analyses in Arizona is limited by their design and focus.
+There has been an increase in the number of streamgages in the region to meet regulatory imperatives, such as the Clean Water Act.
+Even so, many of these gages are not suited for base-flow studies because they emphasize peak flow monitoring and lack the ability to accurately measure low-flow dynamics [@flood-control2020].
 New, non-USGS streamgages are typically installed by flood control districts (e.g.

### 2025-01-19 MANUSCRIPT.tex
edits

@@
 \hypersetup{
-  pdftitle={Regional-scale Base-flow Index of Gauged and Ungauged Basins in Drylands},
-  pdfauthor={Caelum Mroczek; Abraham Springer; Benjamin Lucas},
+  pdftitle={Regional Base-Flow Index in Arid Landscapes Using Machine Learning and Instrumented Records},
+  pdfauthor={Caelum Mroczek; Abraham Springer; Neha Gupta; Benjamin Lucas},
   pdfkeywords={Base Flow, Base-flow Index, Regionalization, Machine
@@
 \begin{document}
-\title{Regional-scale Base-flow Index of Gauged and Ungauged Basins in
-Drylands}
-
-\authors{Caelum Mroczek\affil{1}, Abraham Springer\affil{1}, Benjamin
-Lucas\affil{2}}
+\title{Regional Base-Flow Index in Arid Landscapes Using Machine
+Learning and Instrumented Records}
+
+\authors{Caelum Mroczek\affil{1}, Abraham Springer\affil{1}, Neha
+Gupta\affil{2}, Benjamin Lucas\affil{3}}
 \affiliation{1}{School of Earth and Sustainability, Northern Arizona
-University, Flagstaff, AZ, USA, }\affiliation{2}{Department of
-Mathematics and Statistics, Northern Arizona University, Flagstaff, AZ,
-USA, }
+University, Flagstaff, AZ, USA, }\affiliation{2}{Department of Hydrology
+and Atmospheric Sciences, University of Arizona, Tucson, AZ,
+USA, }\affiliation{3}{Department of Mathematics and Statistics, Northern
+Arizona University, Flagstaff, AZ, USA, }
 \correspondingauthor{Caelum Mroczek}{csm428@nau.edu}
@@
 \begin{abstract}
-LOREM IPSUM \#\#\#\#\#\#
+Base flow, sustained by groundwater discharge, is a vital component of
+river ecosystems, particularly in drylands, where water resources are
+limited. This study analyzes the instrumented streamflow record in
+Arizona to assess long-term base-flow index (BFI) trends across gauged
+catchments. Results indicate that approximately 32\% of Arizona's
+streamflow originates from groundwater discharge, with significant
+spatial variability driven by landscape and climatic factors. Base flow
+relationships are analyzed with coincident trends in climate variables
+such as precipitation, evapotranspiration, and temperature. Spatial and
+climatic trends reveal variability in base-flow contributions, providing
+insight into groundwater-surface water interactions in arid and
+semi-arid landscapes. Building on this analysis, we applied machine
+learning methods to predict BFI in ungauged basins, addressing the
+challenges of Arizona's sparse streamgage network. Using the eXtreme
+Gradient Boosting (XGBoost) algorithm trained on hydroclimate and
+physiographic predictors, we estimated long-term BFI from 1991 to 2020.
+This combined approach integrates observational data with predictive
+modeling to enhance our understanding of base-flow processes and provide
+a framework for water resource management in data-limited regions.
 \end{abstract}
@@
 \section*{Plain Language Summary}
-LOREM IPSUM \#\#\#\#\#\#
+Rivers in drylands, such as Arizona, depend on groundwater contributions
+to maintain flow during dry periods. This portion of streamflow, called
+base flow, plays a critical role in supporting ecosystems and water
+availability in these regions. This study examines long-term patterns of
+base flow in Arizona using data from streamflow monitoring stations. We
+found that about 32\% of Arizona's river flow comes from groundwater,
+though this varies significantly across the state depending on local
+climate and landscape features. We also analyzed how base flow is
+influenced by changes in precipitation, evapotranspiration, and
+temperature, providing insights into how groundwater and surface water
+interact in these complex environments. To estimate base flow in basins
+without monitoring stations, we used machine learning techniques. By
+training a model on data from monitored sites and corresponding
+hydroclimate data, we predicted long-term base flow for ungauged areas
+across Arizona. This innovative approach addresses the challenge of
+Arizona's sparse monitoring network and provides a valuable tool for
+understanding and managing water resources in regions with limited data.
 
@@
 supporting diverse ecosystems and human populations, dryland regions
-face mounting challenges exacerbated by increasing urbanization,
-expanding agricultural activities, and climate-induced amplification of
-precipitation patterns (Taylor et al., 2013). This water scarcity is
-intensifying due to the compounding effects of climate variability and
-increased groundwater extraction (Taylor et al., 2013). In drylands,
-groundwater serves as a vital resource for sustaining ecosystems and
-meeting human needs (Scanlon et al., 2006; Yao et al., 2018).
+face mounting hydrologic challenges exacerbated by increasing
+urbanization, expanding agricultural activities, and climate-induced
+amplification of precipitation patterns (Taylor et al., 2013). This
+water scarcity is intensifying due to the compounding effects of climate
+variability and increased groundwater extraction (Taylor et al., 2013).
+In drylands, groundwater serves as a vital resource for sustaining
+ecosystems and meeting human needs (Scanlon et al., 2006; Yao et al.,
+2018).
 
@@
 management of water quantity and quality depends on understanding
-seasonal and interannual base flow patterns and long-term changes in
-base flow behavior.
+seasonal and interannual base-flow patterns and long-term changes in
+base-flow behavior.
 
@@
 
-To address the challenge of quantifying base flow in ungauged
-catchments, numerous studies have applied regression and machine
-learning methods. (Ahiablame et al., 2013) found that using a regression
-model to estimate annual base flow of ungauged catchments was reasonably
-easy and accurate. Beck et al. (2013) overcame the nonlinearity of basin
-characteristics and improved results of multivariate analyses by using
-artificial neural networks to estimate BFI globally. Singh et al. (2018)
-implemented a random forest algorithm to predict long-term BFI for
-ungauged catchments across New Zealand. The application of machine
-learning in past studies highlights its versatility and effectiveness in
-addressing ecohydrologic dynamics.
-
-This study develops a technique for estimating BFI in ungauged basins in
-Arizona and determines the state's long-term mean BFI. Using a machine
-learning model trained with the hydrogeologic characteristics of surface
-water basins, we estimate mean BFI in ungauged basins over a period of
-record from 1991-2020. Regional patterns and trends in base flow and BFI
-of instrumented sites are subsequently investigated. These results
-provide novel analyses of low flow processes in Arizona, furthering our
-understanding of climate controls on consistent flows in dryland basins.
+Advancements in machine learning provide tools to predict hydrologic
+indices in ungauged basins, addressing the limitations of sparse
+streamgage networks. To tackle the challenge of quantifying base flow in
+ungauged catchments, numerous studies have applied both regression and
+machine learning methods. Ahiablame et al. (2013) found that using a
+regression model to estimate annual base flow of ungauged catchments was
+reasonably easy and accurate. Beck et al. (2013) overcame the
+nonlinearity of basin characteristics and improved results of
+multivariate analyses by using artificial neural networks to estimate
+BFI globally. Singh et al. (2018) implemented a random forest algorithm
+to predict long-term BFI for ungauged catchments across New Zealand.
+These applications demonstrate the versatility and effectiveness of
+machine learning in capturing complex ecohydrologic dynamics and
+improving our understanding of groundwater contributions to streamflow.
+
+This study develops a technique for estimating BFI in ungauged basins
+across Arizona and evaluates the state's long-term BFI. Regional trends
+in base flow and BFI at instrumented sites are also analyzed, with these
+trends being linked to coincident trends in precipitation,
+evapotranspiration (ET\textsubscript{O}), and temperature over the same
+periods. Using a machine learning model trained on the hydrogeologic
+characteristics of surface water basins, we estimate the mean BFI for
+ungauged basins from 1991 to 2020. This approach helps address the
+spatial gaps in the streamgage network, which is relatively sparse
+across the state. The results offer novel insights into low-flow
+processes in both gauged and ungauged basins, enhancing our
+understanding of climate controls on consistent flows in Arizona. This
+study contributes to a more comprehensive view of hydrological dynamics
+in the context of arid and semi-arid landscapes.
 
@@
 (110\(^\circ\)F) and the region averages 200 mm (8 in) of precipitation
-annually. \#\#\#CITATION\#\#\#
+annually (Arizona State Climate Office, 2024).
 
@@
 used in this study. Streamgages were selected depending on criteria to
-ensure the applicability of each site. O'Donnell et al. (2016) found
-that 8-10 years of calibration data were necessary to account for
-climate variability in paired watershed studies in the region. Following
-this criteria, a minimum of 10 years of record was chosen. Any years
-missing more than 30 days of streamflow were removed from the possible
-record period. This analysis focuses on natural, streamflow-influencing
-recharge, excluding streamgages affected by regulation or diversions.
-These gages were identified through annual reports on water data
-published by the USGS (USGS, 2010). Finally, streamgages on the Colorado
-River were excluded from the dataset on the basis that they represent
-managed flows shared by Colorado River Compact states. Following these
-parameters, 205 USGS streamgages with acceptable periods of record were
-used in this study (Figure~\ref{fig-study-area}). Periods of record
-ranged from 10 to 112 years, with a median of 28 years.
+ensure the applicability of each site. Following the findings of
+O'Donnell et al. (2016), which determined that 8--10 years of
+calibration data are necessary to account for climate variability in
+paired watershed studies in the region, a minimum record length of 10
+years was required. Additionally, years with more than 30 missing days
+of streamflow data were excluded from the analysis. This study focuses
+on natural, streamflow-influencing dynamics, so streamgages affected by
+regulation or diversions were excluded. The regulated river streamgages
+were identified through annual reports on water data published by the
+USGS (USGS, 2010). Furthermore, streamgages along the Colorado River
+were omitted because they represent managed flows governed by the
+Colorado River Compact. After applying these selection criteria, 205
+USGS streamgages with acceptable periods of record were included in the
+study (Figure~\ref{fig-study-area}). Periods of record ranged from 10 to
+112 years, with a median of 28 years.
+
+While our data selection ensures a robust record for this analysis, it
+also highlights gaps in spatial coverage that the machine learning model
+can address. Arizona has 184 active USGS streamflow stations (as of
+2024) across an area of 295,253 km², whereas Indiana (a humid state in
+the U.S.) has 189 active stations within an area of 94,326 km². This
+translates to a streamgage density of approximately 2.004 gages per
+1,000 km² in Indiana, over three times higher than Arizona's density of
+0.623 gages per 1,000 km². This stark difference underscores the
+relative sparsity of Arizona's streamgage network, particularly in the
+context of its larger geographic area and the unique hydrological
+challenges posed by its arid and semi-arid landscapes.
 
@@
 regions and has been shown to perform well across the southwestern US
-(Buban et al., 2020)\textbf{.} Instead of the water year, PRISM data
-uses a calendar-year format, which was adopted for consistency in the
-water balance. Although this may introduce challenges in the annual
-estimates due to inter-annual snow storage, the use of long-term annual
-averages is likely to reduce any potential errors (Reitz et al., 2017).
-
-Annual evapotranspiration (ET) data came from TerraClimate, a 4-km grid
-climatalogical data set (Abatzoglou et al., 2018). TerraClimate uses a
-Penman-Monteith approach to generate a reference potential
-evapotranspiration (PET). The PET values were calculated assuming a
-reference grass surface across the landscape with unlimited water. In
-the drylands of the southwestern US, PET typically exceeds precipitation
-annually (Zomer et al., 2022).
+(Buban et al., 2020). Instead of the water year, PRISM data uses a
+calendar-year format, which was adopted for consistency in the water
+balance. Although this may introduce challenges in the annual estimates
+due to inter-annual snow storage, the use of long-term annual averages
+is likely to reduce any potential errors (Reitz et al., 2017).
+
+Annual reference evapotranspiration (ET\textsubscript{O}) data came from
+TerraClimate, a 4-km grid climatalogical data set (Abatzoglou et al.,
+2018). TerraClimate uses a Penman-Monteith approach to generate a
+reference evapotranspiration. The ET\textsubscript{O} values were
+calculated assuming a reference grass surface across the landscape with
+unlimited water. In the drylands of the southwestern US,
+ET\textsubscript{O} typically exceeds precipitation annually (Zomer et
+al., 2022).
 
@@
 comprehensive understanding of landscape composition across space. Land
-cover from USGS-NLCD (National Land Cover Database), hydrologic soil
-group from SSURGO (Soil Survey Geographic Database), and underlying
-geology and karst from USGS were all similarly averaged across the
-basins. Aggregating variables to align with the HUC 8 boundaries allowed
-for more precise predictions of BFI by integrating spatial variations
-within each basin.
+cover was acquired from USGS-NLCD (National Land Cover Database),
+hydrologic soil group from SSURGO (Soil Survey Geographic Database), and
+underlying geology and karst from USGS were all similarly averaged
+across the basins. Aggregating variables to align with the HUC 8
+boundaries allowed for more precise predictions of BFI by integrating
+spatial variations within each basin.
 
@@
 of the study. Comparisons of various base-flow separation techniques
-have been made in (Eckhardt, 2005, 2008; Nathan et al., 1990) , but this
-study does not explore the superiority of different methods.
+have been made in previous studies (e.g. Eckhardt, 2005, 2008; Nathan et
+al., 1990); this study does not explore the superiority of different
+methods.
 
 Base flow was calculated using a single-parameter, recursive digital
-filter technique from (Nathan et al., 1990). This base-flow separation
+filter technique from Nathan et al. (1990). This base-flow separation
 technique is based on a recursive digital filter used in signal analysis
@@
 easy to automate, objective, and repeatable, it is appropriate for a
-regional-scale study. The Lyne-Hollick filter has been used in (Arnold
-et al., 2000; Bloomfield et al., 2009; Santhi et al., 2008; Singh et
-al., 2018). It takes the form of
+regional-scale study. The Lyne-Hollick filter has been used in multiple
+studies (Arnold et al., 2000; Bloomfield et al., 2009; Santhi et al.,
+2008; Singh et al., 2018), and it takes the form of
 
@@
 Gradient Boosting (XGBoost) algorithm (Chen et al., 2016) to predict BFI
-at ungauged locations using catchment characteristics as predictors. The
-XGBoost algorithm is a decision tree-based ensemble algorithm, which can
-be adapted for either regression or classification problems. It has
-grown in popularity in recent years due to its rapid, efficient, and
-scalable nature, and has proven effective in similar environmental
-modeling tasks such as streamflow forecasting Ni et al. (2020) and
-classifying land use/land cover data (Georganos et al., 2018).
-
-XGBoost uses gradient boosting on decision tree algorithms to leverage
-the combination of multiple very low variance models into one overall
-prediction. Gradient boosting works sequentially, with the first tree
-being learned on the known target values, and each subsequent tree being
-learned on the errors of the tree prior. The algorithm then applies a
-weighting to each tree based on this boosting to determine how much the
-tree will contribute towards the final output. The final prediction is
-obtained by the input data traversing all \(n\) weighted trees in the
-learned ensemble. The trained XGBoost model is used to predict BFI in
-ungauged catchments based on geospatial and hydroclimate predictor
-variables (Table~\ref{tbl-predictors}). Certain features were subdivided
-according to their areal coverage of each basin (e.g.~land cover was
-divided into 16 subdivisions).
+at ungauged locations using catchment characteristics as predictors
+Table~\ref{tbl-predictors}. The XGBoost algorithm is a decision
+tree-based ensemble algorithm, which can be adapted for either
+regression or classification problems. This algorithm iteratively builds
+an ensemble of decision trees, where each tree corrects errors from
+previous trees to improve predictions (Chen et al., 2016). Its
+efficiency, scalability, and robustness have made it increasingly
+popular in recent years, with successful applications in environmental
+modeling tasks such as streamflow forecasting (Ni et al., 2020;
+Szczepanek, 2022) and land use/land cover classification (Georganos et
+al., 2018).
+
+XGBoost operates by leveraging gradient boosting on decision tree
+algorithms, combining multiple low-variance models to produce a robust
+overall prediction. Gradient boosting works iteratively: the initial
+tree is trained on the target values, while subsequent trees are trained
+on the residual errors of the preceding tree. Each tree is assigned a
+weight based on its contribution to reducing error, and these weights
+are used to determine the influence of each tree in the final model. The
+ultimate prediction is made by aggregating the outputs of all \(n\)
+weighted trees in the ensemble. In this study, the trained XGBoost model
+is used to predict BFI in ungauged catchments based on geospatial and
+hydroclimate predictor variables (Table~\ref{tbl-predictors}). Certain
+features were further subdivided according to their areal coverage
+within each basin (e.g.~land cover was divided into 16 subdivisions).
+This approach allowed the model to capture finer-scale spatial
+variability and improve predictive accuracy.
 
 \begin{longtable}[]{@{}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.2329}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.2877}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.2329}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.2466}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.2603}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.2466}}
   >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.2466}}@{}}
@@
 & Mean Temperature* & PRISM & Basin average \\
-& Mean Evapotranspiration* & TerraClimate & Basin average \\
+& Reference Evapotranspiration* & TerraClimate & Basin average \\
 Geospatial & Elevation* & DEM & Basin average \\
@@
 
+Our initial training dataset comprised 7,724 observations across 45
+variables. To optimize the model's performance, we first conducted an
+exhaustive grid search combined with 5-fold cross-validation to identify
+the optimal hyperparameter values. The hyperparameters evaluated
+included the learning rate (\(\eta\)), minimum split loss (\(\gamma\)),
+maximum tree depth, minimum child weight, and the number of trees. While
+not an exhaustive list of all possible XGBoost hyperparameters, the
+range of values explored provided sufficient variation to ensure the
+selection of a high-performing model.The optimal hyperparameters were
+determined to be: 700 trees, a learning rate (\(\eta\)) of 0.05, a
+minimum split loss (\(\gamma\)) of 0.075, a maximum tree depth of 7, and
+a minimum child weight of 5. Using these values, the XGBoost model was
+trained on the dataset with 10-fold cross-validation.
+
+\(K\)-fold cross-validation provides an unbiased estimate of a model's
+accuracy on unseen data, while also insuring against overfitting or
+underfitting. In this approach, the data is randomly divided into \(k\)
+folds of equal size. The model is trained on \(k-1\) folds and tested on
+the remaining fold, referred to as the validation set. This process is
+repeated \(k\) times, with each fold serving as the validation set
+exactly once. Each iteration trains an independent model with the same
+hyperparameters but using a different subset of training data. By
+averaging the model performance across all \(k\) folds, we achieve a
+robust and reliable estimate of predictive accuracy. Root mean squared
+error (RMSE) was used as the performance metric for both model
+optimization and evaluation. RMSE provided a consistent and
+interpretable measure of the model's accuracy throughout the training
+and validation process.
+
+\begin{figure}
+
+\centering{
+
+\includegraphics[width=62.5in,height=\textheight]{images/k-fold_fig.png}
+
+}
+
+\caption{\label{fig-k-fold}Schematic of \(k\)-fold cross validation. The
+dataset is randomly divided into \(k\) stratified folds. Each fold
+serves as the validation set once, while the remaining folds are
+combined to create a training set for model development. Performance
+metrics for the test set are calculated and recorded, and this process
+is repeated for all \(k\) folds.}
+
+\end{figure}%
+
 \subsubsection{Feature Selection}\label{feature-selection}
 
-Machine learning models can face issues with overfitting when provided
-with a large set of predicting features. This can cause reductions in
-performance on unseen data and requires greater computational power and
-memory storage (Li et al., 2017). Dimensionality reduction is a powerful
-tool in addressing these issues which can be divided into two broad
-categories: feature extraction and feature selection. Feature extraction
-involves projecting the higher dimensional dataset to a feature space
-with less dimensions, though this entails creating a new set of features
-that lose the physical meaning of the original feature space. On the
-other hand, feature selection maintains a subset of the original
-features, all of which retain their physical meaning providing better
-readability and interpretability. Here, a supervised features selection
-was used to reduce the number of predictors and, as a result, increased
-learning performance, decreased computational costs, and reduced
-overfitting.
-
-An initial model was trained on the full feature set of 45 predictors
-(Table~\ref{tbl-predictors}). We then used a feature importance method
-of feature selection to remove noise and less-important features.
-Feature importance scores are an approach where importance scores are
-assigned to individual features indicating how much the feature
-contributed (positively or negatively) to the model (Murdoch et al.,
-2019). Here, we use SHAP (SHapley Additive exPlanations) values as the
-method of assigning importance values to each feature (Lundberg \& Lee,
-2017). The top-10 most-important features
-(\textbf{?@suppfig-shap\_values}) were subset and used in training a
-subsequent model which saw performance increases and reduced computing
-time. This subsequent model was used as the final model in this
-analysis.
+Machine learning models are prone to overfitting, especially when
+provided with a large set of predictive features (Ying, 2019).
+Overfitting can degrade model performance on unseen data and increase
+the demand for computational resources and memory storage (Li et al.,
+2017). Dimensionality reduction offers a robust solution to these
+challenges and generally falls into two broad categories: feature
+extraction and feature selection.
+
+Feature extraction involves transforming the original dataset into a
+lower-dimensional feature space. However, this process generates new
+features that often lack the physical interpretability of the original
+variables. In contrast, feature selection identifies a subset of the
+original features, preserving their physical meaning while improving
+model readability and interpretability (Li et al., 2017). In this study,
+supervised feature selection was employed to reduce the number of
+predictors, which enhanced learning performance, reduced computational
+costs, and mitigated overfitting.
+
+To begin, an initial model was trained using the full feature set of 45
+predictors (Table~\ref{tbl-predictors}). A feature selection method
+based on feature importance was then applied to identify and remove less
+relevant and noisy features. Feature importance scores quantify the
+contribution of individual features---either positively or
+negatively---to the model's predictions (Murdoch et al., 2019). In this
+analysis, SHAP (SHapley Additive exPlanations) values were used to
+compute feature importance scores (Lundberg \& Lee, 2017).
+
+SHAP values is a method to explain the prediction of an individual
+instance by calculating the contribution of each feature to that
+prediction. The method is based on coalition game theory and is
+discussed further in Lundberg \& Lee (2017). Here, SHAP values are used
+for global interpretation of feature importance and feature effects on
+the model. Global feature importance is produced by the absolute Shapley
+values of each feature across the dataset, providing a list of features
+in order of most to least important. Feature effects provide an
+indication of the relationship between the value of a predicting feature
+and its impact on the prediction.
+
+The ten most important features Figure~\ref{fig-shap_values} were
+selected based on their SHAP values and used to train a subsequent model
+Table~\ref{tbl-predictors}. This refined model demonstrated improved
+performance and reduced computational time compared to the initial
+model. The final model, trained on this optimized feature subset, was
+ultimately used for the analysis presented here.
 
@@
 
-Statisical analyses were performed on annual BFI and base flow values of
-instrumented streamgages. We examined temporal trends using the
-Mann-Kendall nonparametric trend test to determine the presence of
-monotonic trends (Kendall, 1970; Mann, 1945). The Mann-Kendall test was
-used to determine trends in BFI and trend significance of gauged
-reaches. This test is widely used in studies of this kind (Ayers et al.,
-2019; Ficklin et al., 2016; Woodhouse et al., 2022).
-
-Trends in BFI and base flow were produced using a Mann-Kendall test for
-monotonic trends. This test expects that the data is non-parametric and
-that the observations are not autocorrelated over time. Observations at
-4 streamgages show significant autocorrelation, though only one
-streamgage (09486500 - Santa Cruz River at Cortaro, AZ) indicated a
-significant trend in BFI. Due to this autocorrelation, the streamgage
-was removed from the trend analysis as the autocorrelation may inflate
-the variance of the Mann-Kendall statistic leading to an
-over/underestimate of the trend (Hamed \& Rao, 1998).
-
-Trends with a \(\rho \le 0.05\) are considered significant.
+Statistical analyses were conducted on annual BFI and base-flow values
+from instrumented streamgages to identify temporal trends using the
+Mann-Kendall nonparametric trend test (Kendall, 1970; Mann, 1945). This
+test detects monotonic trends in datasets that are non-parametric and
+assumes the absence of autocorrelation among observations. This test is
+widely used in studies of this nature (Ayers et al., 2019; Ficklin et
+al., 2016; Woodhouse et al., 2022).
+
+To check for autocorrelation, we applied the Durbin-Watson test, which
+revealed significant autocorrelation at four streamgages on an annual
+basis. Of these, only one streamgage (09486500 - Santa Cruz River at
+Cortaro, AZ) showed a significant trend in BFI. This streamgage was
+excluded from the trend analysis, as autocorrelation could inflate the
+variance of the Mann-Kendall statistic, potentially leading to biased
+trend estimates (Hamed \& Rao, 1998). Trends with a \(\rho \le 0.05\)
+are considered significant.
 
 \section{Results}\label{results}
-
-\subsection{BFI of Gauged Catchments}\label{bfi-of-gauged-catchments}
-
-\begin{figure}
-
-\centering{
-
-\includegraphics{images/BFI_Instrumented_20241203.png}
-
-}
-
-\caption{\label{fig-instrumented-bfi}Mean annual BFI for the period of
-record from instrumented stream flow data.}
-
-\end{figure}%
-
-The long-term average BFI for the 205 gauged reaches across Arizona is
-illustrated in Figure~\ref{fig-instrumented-bfi} . The highest BFI
-values (\textgreater0.9) are found along the Grand Canyon in
-northwestern Arizona. The highly karstic geology of this region
-facilitates the rapid movement of subsurface flow to surface water and
-spring outlets (Chambless et al., 2023). Relatively high BFI values
-(\textgreater0.8) are found at the spring-fed headwaters of the Verde
-River (Del Rio Spring) and the spring-fed headwaters of Fossil Creek.
-These results are consistent with interpolated BFI values reported by
-Wolock (2003).
-
-The stream reaches of the Little Colorado River Basin (northeastern
-Arizona) indicate consisently low BFI values (\textless{} 0.2). This is
-likely due to low-yielding perched aquifers underlying the Defiance
-Uplift in northeastern Arizona, which are hydrologically connected to
-surface streams, while the high-yield, confined regional aquifer is much
-deeper (Blanchard, 2002). This region also typifies a pattern of seen
-along most major rivers in the study area: upstream reaches indicate
-higher BFI values while the downstream reaches display lower BFI values.
-This is assumed to be due to greater groundwater-surface water
-interaction at the headwaters of streams (likely due to spring outlets)
-and the dilution of base flow while moving down stream
-(\textbf{CITATION}). This pattern is seen along the Gila River, Verde
-River, and Little Colorado Rivers in Arizona.
-
-\subsubsection{Trends in BFI}\label{sec-bfi-trends}
-
-Trends in BFI over the period of record for each streamgage in this
-analysis are illustrated in Figure~\ref{fig-instrumented-trend} and
-Table~\ref{tbl-trends} .
-
-\begin{figure}
-
-\centering{
-
-\includegraphics{images/BFITrends_20241203.png}
-
-}
-
-\caption{\label{fig-instrumented-trend}Trends in BFI over full period of
-record for instrumented sites used in this study. Red upward (blue
-downward) arrows indicate an increasing (decreasing) trend at a
-significance level of 5\%. White circles represent sites with no
-statistically significant trends.}
-
-\end{figure}%
-
-\begin{longtable}[]{@{}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.4800}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.0667}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.0933}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.0933}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1333}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1333}}@{}}
-\caption{Comparison of trends for BFI for all sites split by various
-classifications. Only sites with a significant (\(\rho \le 0.05\)) trend
-are included here as established by a Mann-Kendall test for monotonic
-trends across the full period of record. n is the number of sites,
-n\_pos (n\_neg) is the number of sites with positive (negative) trends,
-perc\_pos (perc\_neg) is the percentage of n with a positive (negative)
-trend.}\label{tbl-trends}\tabularnewline
-\toprule\noalign{}
-\begin{minipage}[b]{\linewidth}\raggedright
-Classification Group
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
-n
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
-n\_pos
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
-n\_neg
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
-perc\_pos
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
-perc\_neg
-\end{minipage} \\
-\midrule\noalign{}
-\endfirsthead
-\toprule\noalign{}
-\begin{minipage}[b]{\linewidth}\raggedright
-Classification Group
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
-n
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
-n\_pos
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
-n\_neg
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
-perc\_pos
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
-perc\_neg
-\end{minipage} \\
-\midrule\noalign{}
-\endhead
-\bottomrule\noalign{}
-\endlastfoot
-Precipitation - Monsoon Dominated & 87 & 8 & 21 & 0.092 & 0.241 \\
-Precipitation - Snowmelt Dominated & 118 & 9 & 12 & 0.076 & 0.102 \\
-PhysRegion - Basin\&Range & 156 & 14 & 26 & 0.090 & 0.167 \\
-PhysRegion - CO Plateau & 49 & 3 & 7 & 0.061 & 0.143 \\
-Climate - Warm-Wet & 31 & 2 & 6 & 0.065 & 0.194 \\
-Climate - Warm-Dry & 55 & 6 & 11 & 0.109 & 0.200 \\
-Climate - Cool-Wet & 74 & 4 & 9 & 0.054 & 0.122 \\
-Climate - Cool-Dry & 45 & 5 & 7 & 0.111 & 0.156 \\
-Slope - High & 102 & 10 & 12 & 0.098 & 0.118 \\
-Slope - Low & 103 & 7 & 21 & 0.068 & 0.204 \\
-\end{longtable}
 
@@
 
-\subsubsection{Predictor Importance}\label{predictor-importance}
+\subsubsection{Model Validation}\label{model-validation}
+
+Predicted values of BFI are plotted against observed values for the
+entire period of record of the instrumented dataset in
+Figure~\ref{fig-actual_predicted} . The agreement between ``out-of-bag''
+predictions (blind cross-validation, treating each site as ungauged) and
+observed values is acceptable (R\textsuperscript{2} = 0.764) indicating
+that the model performs well across the full dataset. The overall RMSE
+is 0.129 and the overall percent bias (pbias) is -5.6\%. Model
+performance metrics across various classifications are summarized in
+Table~\ref{tbl-performance}. These performance metrics demonstrate that
+the regional model performs consistently well across different spatial
+and climatic classifications. However, the negative pbias values across
+all classifications, along with the overall pbias, indicate a systematic
+underprediction of BFI by the model. Categories with relatively lower
+R\textsuperscript{2} and Nash-Sutcliffe Efficiency (NSE) values also
+exhibit higher biases, reflecting weaker model performance in those
+specific contexts.
 
@@
 
-\subsubsection{Model Validation}\label{model-validation}
-
 \begin{longtable}[]{@{}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.3846}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.0769}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.0897}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.0897}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.0897}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.0897}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.0897}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.0897}}@{}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.1250}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.1250}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.1250}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.1250}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.1250}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.1250}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.1250}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.1250}}@{}}
 \caption{Performance of model predictions for BFI for all sites split by
@@
 \begin{minipage}[b]{\linewidth}\raggedright
-Classification Group
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
-n
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
-R2
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
-MSE
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
-RMSE
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
-MAE
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
-NSE
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
-pbias
+\textbf{Classification Group}
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
+\textbf{n}
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
+\textbf{R2}
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
+\textbf{MSE}
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
+\textbf{RMSE}
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
+\textbf{MAE}
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
+\textbf{NSE}
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
+\textbf{pbias}
 \end{minipage} \\
@@
 
-Consistently negative pbias values across all classifications suggest an
-under-prediction of BFI.
-
 \subsubsection{Predictor Importance}\label{sec-predictor-importance}
 
-\subsubsection{Predicted Long-term Mean
-BFI}\label{predicted-long-term-mean-bfi}
+The predictors used to estimate BFI at ungauged sites were evaluated for
+their importance in the final XGBoost model, as illustrated in
+Figure~\ref{fig-shap_values}. The most influential feature for
+predicting long-term BFI is basin elevation. While elevation itself does
+not directly affect base-flow characteristics, it has consistently been
+identified as a key predictor in previous BFI studies (Beck et al.,
+2013; Singh et al., 2018). The importance of elevation aligns with
+findings from Beck et al. (2013), highlighting its role as a proxy for
+climate variables such as temperature, precipitation, and snowpack
+duration. Seasonal snowpack duration, in particular, has been shown to
+strongly correlate with springflow and groundwater recharge in this
+region (Donovan et al., 2022). This relationship is further supported by
+hydroclimate features where higher temperatures tend to negatively
+influence BFI, while precipitation exhibits a mixed influence. In some
+cases, higher precipitation values correlate with lower BFI, likely due
+to a larger proportion of precipitation contributing to runoff rather
+than infiltration and groundwater recharge.
 
@@
 
+\includegraphics{images/shap_summary.png}
+
+}
+
+\caption{\label{fig-shap_values}SHAP value plot of features used in
+final model. Land cover features are indicated by the percentage of
+cover by each land cover type and soil types are defined by hydrologic
+soil group.}
+
+\end{figure}%
+
+Land cover and land use predictors also play a significant role in BFI
+estimation. Analysis of SHAP values indicates that a higher percentage
+of evergreen forest positively influences BFI predictions, while higher
+proportions of shrubland and developed land exert a negative influence.
+Similarly, hydrologic soil types show distinct trends in their impact on
+BFI. Soil Type C, characterized by moderately high runoff potential
+(20-40\% clay), tends to negatively influence BFI. In contrast, Soil
+Type A, which has low runoff potential and facilitates rapid water
+infiltration, exhibits a mixed influence (USDA, 2009). The mixed effects
+of Soil Type A are likely due to SHAP values capturing interactions
+between features rather than direct relationships. For example, regions
+dominated by Soil Type A may also have steep slopes or sparse
+vegetation. Additionally, the relationship between Soil Type A and BFI
+is likely non-linear and influenced by the complex dynamics of regional
+base-flow processes.
+
+\subsubsection{Predicted Long-term BFI}\label{predicted-long-term-bfi}
+
+The regionalized (HUC-8) long-term BFI (1991--2020) is shown in
+Figure~\ref{fig-bfi-huc}, while the long-term BFI for stream reaches
+with a Strahler stream order of 3 or greater is presented in
+Figure~\ref{fig-bfi-streams}. Basins with high BFIs, such as those along
+the Grand Canyon in the northwestern part of the study area, indicate
+greater surface water and groundwater interaction. Elevated BFI values
+are also observed along portions of the Mogollon Rim, a heavily forested
+region with high precipitation that marks the transition between
+physiographic regions. Additionally, headwater regions of perennial
+rivers tend to exhibit higher BFI values. In contrast, low BFI values
+are found in areas like the Defiance Uplift in northeastern Arizona and
+the arid southern regions of the state.
+
+\begin{figure}
+
+\centering{
+
 \includegraphics{images/BFI_HUC8_20241203.png}
@@
 
-\caption{\label{fig-bfi-huc}Predicted mean annual BFI values for 8-digit
+\caption{\label{fig-bfi-huc}Predicted long-term BFI values for 8-digit
 HUC (1991-2020)}
@@
 
-\caption{\label{fig-bfi-streams}Predicted mean annual BFI values for all
-reaches of Strahler stream order 3 and greater}
+\caption{\label{fig-bfi-streams}Predicted long-term BFI values for all
+stream reaches of Strahler stream order 3 and greater (1991-2020)}
 
@@
 
-\section{Conclusion}\label{sec-conclusion}
+Figure~\ref{fig-bfi-streams} indicates the predicted long-term BFI for
+stream reaches in Arizona of a Strahler stream order of 3 or greater.
+Stream-level BFI values often correspond to the long-term BFI observed
+in HUC-8 basins, as geospatial predictor data were aggregated at the
+basin level. Figure~\ref{fig-bfi-streams} highlights stream density
+across the region, revealing areas with higher concentrations of
+perennial streams, particularly in regions with favorable hydrogeologic
+conditions, such as spring-fed systems in the Verde River and the Salt
+River headwaters. Conversely, lower BFI values are observed in arid
+regions and those with lower stream densities such as parts of the
+Little Colorado River Basin, where groundwater recharge is limited
+(Flint \& Flint, 2007). While BFI values and their spatial patterns may
+align with certain terrain features or climatic patterns, it is
+important to recognize that BFI is driven by complex interactions among
+multiple factors Figure~\ref{fig-shap_values}.
+
+\subsection{BFI of Gauged Catchments}\label{bfi-of-gauged-catchments}
+
+\begin{figure}
+
+\centering{
+
+\includegraphics{images/BFI_Instrumented_20241203.png}
+
+}
+
+\caption{\label{fig-instrumented-bfi}Long-term BFI for the period of
+record from instrumented stream flow data.}
+
+\end{figure}%
+
+The long-term BFI for the 205 gauged reaches across Arizona is
+illustrated in Figure~\ref{fig-instrumented-bfi} . The long-term mean
+BFI is 0.32, indicating that \textasciitilde32\% of long-term streamflow
+in Arizona likely originates from groundwater discharge and other
+delayed sources. The highest BFI values (\textgreater0.9) are found
+along the Grand Canyon in northwestern Arizona. The highly karstic
+geology of this region facilitates the rapid movement of subsurface flow
+to surface water and spring outlets (Chambless et al., 2023). Relatively
+high BFI values (\textgreater0.8) are found at the spring-fed headwaters
+of the Verde River (Del Rio Spring) and the spring-fed headwaters of
+Fossil Creek. These results are consistent with interpolated BFI values
+reported by Wolock (2003).
+
+The stream reaches of the Little Colorado River Basin (northeastern
+Arizona) indicate consistently low BFI values (\textless{} 0.2). This is
+likely due to low-yielding perched aquifers underlying the Defiance
+Uplift in northeastern Arizona, which are hydrologically connected to
+surface streams, while the high-yield, confined regional aquifer is much
+deeper (Blanchard, 2002). A notable tendency emerges along most major
+rivers in the study area: upstream reaches tend to exhibit higher BFI
+values, while downstream reaches display lower values. This pattern is
+presumed to result from greater groundwater-surface water interactions
+at stream headwaters, influenced by spring outlets, and the dilution of
+base flow as water moves downstream. This trend is particularly evident
+along the Gila River, Verde River, and Little Colorado River in Arizona.
+
+\subsubsection{Trends in Base Flow \& BFI}\label{sec-bfi-trends}
+
+Trends in BFI over the period of record for each streamgage in this
+analysis are illustrated in Figure~\ref{fig-instrumented-trend} and
+Table~\ref{tbl-trends}. Base flow and BFI trends were analyzed across
+all instrumented sites over their respective periods of record using the
+Mann--Kendall test. Statistically significant trends were observed in
+both metrics, with a 72.20\% coincidence rate between significant base
+flow and BFI trends, indicating a strong dependence of BFI on base-flow
+dynamics.
+
+Figure~\ref{fig-instrumented-trend} illustrates the spatial variation in
+BFI trends across the study area. Statistically significant decreasing
+trends are observed at 16.1\% of sites, while increasing trends are
+found at 8.8\% of sites, with no clear regional patterns for either. In
+the Basin and Range physiographic region, 9\% of sites show increasing
+trends, while 16.7\% exhibit decreasing trends. In the Colorado Plateau
+region, increasing trends occur at 6.1\% of sites, and decreasing trends
+are observed at 14.3\% of sites.
+
+\begin{figure}
+
+\centering{
+
+\includegraphics{images/BFITrends_20241203.png}
+
+}
+
+\caption{\label{fig-instrumented-trend}Trends in BFI over full period of
+record for instrumented sites used in this study. Red upward (blue
+downward) arrows indicate an increasing (decreasing) trend at a
+significance level of 5\%. White circles represent sites with no
+statistically significant trends.}
+
+\end{figure}%
+
+\subsubsection{Classification Trends}\label{classification-trends}
+
+Classifications presented in Table~\ref{tbl-trends} were determined
+based on precipitation regime, physiographic region, climate, and slope.
+The dominant precipitation regime (monsoon vs.~snowmelt) was identified
+by analyzing streamflow hydrographs for each station, focusing on peak
+flow periods during the monsoon season (July--September) and the
+snowmelt season (March--June). Physiographic region was assigned based
+on which region the streamgage is located. Climate classifications were
+defined as warm (above the long-term median temperature of Arizona),
+cool (below the long-term median temperature), wet (above the long-term
+median precipitation), and dry (below the long-term median
+precipitation). Slope was categorized as high (above the median slope)
+and low (below the median slope).
+
+Statistically significant decreasing trends in BFI were more common than
+increasing trends across all site classifications
+(Table~\ref{tbl-trends}). While decreasing trends dominate, both
+increasing and decreasing trends are observed within each
+classification. Monsoon-dominated regions exhibit a higher proportion of
+significant negative trends (24.1\%) compared to snowmelt-dominated
+regions (10.2\%), suggesting that monsoon-dominated systems are more
+consistently correlated with declining base flow. Among climate
+classifications, warm-dry climates have the highest proportion of
+negative trends (20.0\%), followed by warm-wet climates (19.4\%),
+indicating that regions with higher temperatures are more prone to base
+flow declines. Low-slope regions show a greater prevalence of negative
+trends (20.4\%) compared to high-slope regions (11.8\%). This suggests
+that flatter areas may be more susceptible to base-flow reductions,
+potentially due to differences in hydrologic connectivity and recharge
+dynamics.
+
+\begin{longtable}[]{@{}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1667}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1667}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1667}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1667}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1667}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1667}}@{}}
+\caption{Comparison of trends for BFI for all sites split by various
+classifications. Only sites with a significant (\(\rho \le 0.05\)) trend
+are included here as established by a Mann-Kendall test for monotonic
+trends across the full period of record. n is the number of sites,
+n\_pos (n\_neg) is the number of sites with positive (negative) trends,
+perc\_pos (perc\_neg) is the percentage of n with a positive (negative)
+trend.}\label{tbl-trends}\tabularnewline
+\toprule\noalign{}
+\begin{minipage}[b]{\linewidth}\raggedright
+\textbf{Classification Group}
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
+\textbf{n}
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
+\textbf{n\_pos}
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
+\textbf{n\_neg}
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
+\textbf{perc\_pos}
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
+\textbf{perc\_neg}
+\end{minipage} \\
+\midrule\noalign{}
+\endfirsthead
+\toprule\noalign{}
+\begin{minipage}[b]{\linewidth}\raggedright
+\textbf{Classification Group}
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
+\textbf{n}
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
+\textbf{n\_pos}
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
+\textbf{n\_neg}
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
+\textbf{perc\_pos}
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
+\textbf{perc\_neg}
+\end{minipage} \\
+\midrule\noalign{}
+\endhead
+\bottomrule\noalign{}
+\endlastfoot
+Precipitation - Monsoon Dominated & 87 & 8 & 21 & 0.092 & 0.241 \\
+Precipitation - Snowmelt Dominated & 118 & 9 & 12 & 0.076 & 0.102 \\
+Physiographic Region - Basin and Range & 156 & 14 & 26 & 0.090 &
+0.167 \\
+Physiographic Region - Colorado Plateau & 49 & 3 & 7 & 0.061 & 0.143 \\
+Climate - Warm-Wet & 31 & 2 & 6 & 0.065 & 0.194 \\
+Climate - Warm-Dry & 55 & 6 & 11 & 0.109 & 0.200 \\
+Climate - Cool-Wet & 74 & 4 & 9 & 0.054 & 0.122 \\
+Climate - Cool-Dry & 45 & 5 & 7 & 0.111 & 0.156 \\
+Slope - High & 102 & 10 & 12 & 0.098 & 0.118 \\
+Slope - Low & 103 & 7 & 21 & 0.068 & 0.204 \\
+\end{longtable}
+
+\subsubsection{Coincident Climate
+Trends}\label{coincident-climate-trends}
+
+Given the variations in the period of record across the instrumented
+network (see SUPPLEMENTAL - Period of Record figure), we analyzed trends
+in base flow and BFI in relation to coincident trends in climate
+variables. Trends are classified as coincident when the direction of the
+climate variable trend (positive or negative) aligns with the trend
+observed in base flow or BFI (Table~\ref{tbl-coincident-trends}). This
+analysis includes both significant and non-significant trends, which is
+appropriate where the influence of complex, interconnected processes may
+not always manifest as statistically significant patterns over limited
+observational periods (Ficklin et al., 2016).
+
+\begin{longtable}[]{@{}lcc@{}}
+\caption{Coincident trends of climate variables (ET\textsubscript{O},
+precipitation, temperature) with base flow and BFI
+trends.}\label{tbl-coincident-trends}\tabularnewline
+\toprule\noalign{}
+& \textbf{Climate Variable} & \textbf{Coincidence Percentage (\%)} \\
+\midrule\noalign{}
+\endfirsthead
+\toprule\noalign{}
+& \textbf{Climate Variable} & \textbf{Coincidence Percentage (\%)} \\
+\midrule\noalign{}
+\endhead
+\bottomrule\noalign{}
+\endlastfoot
+BFI & ET\textsubscript{O} & 44.88 \\
+& Precipitation & 53.17 \\
+& Temperature & 47.32 \\
+Base Flow & ET\textsubscript{O} & 55.12 \\
+& Precipitation & 64.88 \\
+& Temperature & 39.02 \\
+\end{longtable}
+
+The analysis shows that base flow and BFI trends most frequently align
+with precipitation trends (64.88\% and 53.17\%, respectively),
+emphasizing precipitation as the primary driver of local groundwater
+recharge and discharge. Coincidence with ET\textsubscript{O} (reference
+evapotranspiration) trends (55.12\% for base flow and 44.88\% for BFI)
+suggests that evapotranspiration also plays a significant role,
+particularly in arid regions where it can reduce recharge or base flow
+during dry periods. In contrast, temperature trends show lower
+percentages of coincidence, often opposing base flow and BFI trends.
+Specifically, positive (negative) temperature trends are frequently
+associated with negative (positive) base-flow trends (60.98\%) and BFI
+trends (52.68\%). These results highlight the complex interplay between
+climatic variables and hydrological processes. While precipitation
+exerts the strongest influence on base flow and BFI, evapotranspiration
+and temperature add further variability in specific environmental
+contexts.
+
+\section{Summary \& Conclusions}\label{sec-conclusion}
+
+This study provides new insights into base-flow dynamics and groundwater
+contributions in Arizona's dryland rivers by combining an analysis of
+instrumented streamflow records with machine learning predictions for
+ungauged basins. The results highlight significant spatial variability
+in BFI, with approximately 32\% of Arizona's long-term streamflow
+originating from groundwater discharge. Regions such as the Grand Canyon
+and Mogollon Rim demonstrate high BFI values due to strong
+groundwater-surface water interactions, while areas like the Little
+Colorado River Basin exhibit low BFI values, reflecting limited
+groundwater recharge.
+
+Using an XGBoost machine learning algorithm, we successfully predicted
+long-term BFI in ungauged basins, achieving strong model performance (R²
+= 0.764, RMSE = 0.129). The model performed well across all
+classifications, demonstrating its robustness in capturing base-flow
+dynamics across a region with substantial variability in climate,
+elevation, and physiographic characteristics. Key predictors included
+elevation, land cover, and soil type, highlighting the importance of
+integrating hydroclimate and physiographic characteristics into regional
+hydrological models. These predictions address the limitations posed by
+Arizona's sparse streamgage network, offering a scalable approach to
+estimate BFI in data-limited regions.
+
+Our analysis of BFI trends in gauged catchments revealed that
+precipitation is the primary driver of base-flow variability, with
+evapotranspiration and temperature contributing additional complexity.
+These findings emphasize the critical role of climate-hydrology
+interactions in shaping groundwater contributions to streamflow. Inverse
+trends between temperature and BFI suggest that further warming could
+reduce groundwater contributions to streamflow. Coincident trends in
+precipitation and BFI further underscore the importance of understanding
+recharge processes, especially in arid and semi-arid landscapes, where
+precipitation events play key roles in replenishing groundwater.
+
+However, the utility of the streamgage networks for base-flow analyses
+in Arizona is limited by their design and focus. There has been an
+increase in the number of streamgages in the region to meet regulatory
+imperatives, such as the Clean Water Act. Even so, many of these gages
+are not suited for base-flow studies because they emphasize peak flow
+monitoring and lack the ability to accurately measure low-flow dynamics
+(Maricopa County, 2020). New, non-USGS streamgages are typically
+installed by flood control districts (e.g.~the ALERT system) and are
+designed for tracking flood flows rather than base flow. New USGS
+streamgages have been added over the past decade to address instream
+flow rights and to improve the density of monitoring in the future.
+
+This study demonstrates the benefits of combining observational records
+with machine learning to improve our understanding of streamflow
+processes in drylands. Future work should explore the projected effects
+of climate change on base-flow processes and developing models for other
+data-poor regions. The framework presented here has broad applicability
+to other arid and semi-arid regions worldwide and can inform water
+resource management strategies aimed at addressing water scarcity and
+adapting to climate variability.
 
@@
 \addcontentsline{toc}{section}{References}
-
-\section{Supporting Information}\label{supporting-information}
-
-\includegraphics{images/supplemental/shap_summary.png}
-\{\#suppfig-shap\_values\}
 
@@
 
+\bibitem[\citeproctext]{ref-az_climate2024}
+Arizona State Climate Office. (2024). {Climate of Arizona}. Retrieved
+from \url{https://globalfutures.asu.edu/azclimate/climate/}
+
 \bibitem[\citeproctext]{ref-arnold1995}
@@
 
+\bibitem[\citeproctext]{ref-donovan_karst_2022}
+Donovan, K. M. et al. (2022). Karst spring processes and storage
+implications in high elevation, semiarid southwestern united states. In
+M. J. Currell \& B. G. Katz (Eds.), \emph{Geophysical monograph series}
+(1st ed., pp. 35--50). Wiley.
+\url{https://doi.org/10.1002/9781119818625.ch4}
+
 \bibitem[\citeproctext]{ref-eastoe2019mtnblock}
@@
 \url{https://doi.org/10.1002/2016GL069121}
+
+\bibitem[\citeproctext]{ref-flint_flint07}
+Flint, L. E., \& Flint, A. L. (2007). \emph{Regional Analysis of
+Ground-Water Recharge}.
 
@@
 
+\bibitem[\citeproctext]{ref-flood-control2020}
+Maricopa County, F. C. D. of. (2020). \emph{{Flood Warning Program -
+Program Review}}. Maricopa County Flood Control District. Retrieved from
+\url{https://alert.fcd.maricopa.gov/alert/FWP_report_FINAL.pdf}
+
 \bibitem[\citeproctext]{ref-ML_interpretability2019}
@@
 
+\bibitem[\citeproctext]{ref-nrcs2009}
+USDA. (2009). \emph{Part 630 hydrology national engineering handbook,
+chapter 7: Hydrologic soil groups}. USDA Natural Resources Conservation
+Service.
+
 \bibitem[\citeproctext]{ref-usgs2010}
@@
 
+\bibitem[\citeproctext]{ref-Ying_2019}
+Ying, X. (2019). An overview of overfitting and its solutions.
+\emph{Journal of Physics: Conference Series}, \emph{1168}(2), 022022.
+\url{https://doi.org/10.1088/1742-6596/1168/2/022022}
+
 \bibitem[\citeproctext]{ref-zomer2022}

### 2025-01-20 MANUSCRIPT.qmd
supp info

@@
 |   | Variable | Source | Geoprocessing |
-|------------------|-------------------|------------------|------------------|
+|----|----|----|----|
 | Hydroclimate | Precipitation\* | PRISM | Basin average |
@@
 | **Classification Group** | **n** | **n_pos** | **n_neg** | **perc_pos** | **perc_neg** |
-|------------|------------|------------|------------|------------|------------|
+|----|----|----|----|----|----|
 | Precipitation - Monsoon Dominated | 87 | 8 | 21 | 0.092 | 0.241 |

### 2025-01-20 Supp_Info.qmd
supp info

---
title: "Supplemental Info"
editor: visual
---

![Histogram of BFI trends grouped by long-term instrumented BFI. Significant trends (\>0.05) are indicated with bolded borders. More significant negative trends are apparent in low-BFI gauges. Significant positive trends are equally distributed among BFIs.](images/supplemental/bfi-trend_histogram.png)

![Quadrant plot of Kendall's Tau trends in base flow vs. BFI. Significant trends in base flow and BFI indicate that variations in base flow volumes are coincident with trends in BFI. This indicates that trends in BFI typically follow trends in groundwater discharge (base flow) and that trends in BFI aren't typically due to an overall decrease/increase in streamflow. One site indicates a decrease in base flow volumes and an increase in BFI, suggesting that decreases in surface water inputs is of a greater magnitude than the increase of base flow.](images/supplemental/baseflow-bfi_quadrants.png)

![Plot of residuals of predicted BFI vs observed BFI.](images/supplemental/act-pred_residuals.png)

### 2025-01-25 MANUSCRIPT.qmd
starting submission docs

--- 2025-01-19 MANUSCRIPT.qmd
@@
   - Approximately 32% of Arizona's streamflow originates from groundwater discharge, with significant spatial and climatic variability. 
-  - A machine learning model (XGBoost) successfully predicted long-term BFI in ungauged basins, addressing gaps in Arizonas sparse streamgage network. 
-  - Base-flow trends align most closely with precipitation changes, emphasizing the critical role of climate in shaping groundwater-surface water dynamics in dryland regions.
+  - A machine learning model (XGBoost) successfully predicted long-term BFI in ungauged basins, addressing gaps in a sparse streamgage network. 
+  - Base-flow trends align most closely with precipitation changes, emphasizing the role of climate in shaping groundwater-surface water dynamics in dryland regions.
 
@@
 This study develops a technique for estimating BFI in ungauged basins across Arizona and evaluates the state’s long-term BFI.
-Regional trends in base flow and BFI at instrumented sites are also analyzed, with these trends being linked to coincident trends in precipitation, evapotranspiration (ET~O~), and temperature over the same periods.
+Regional trends in base flow and BFI at instrumented sites are also analyzed, with these trends being linked to coincident trends in precipitation, reference evapotranspiration (ET~O~), and temperature over the same periods.
 Using a machine learning model trained on the hydrogeologic characteristics of surface water basins, we estimate the mean BFI for ungauged basins from 1991 to 2020.
@@
 
-Annual reference evapotranspiration (ET~O~) data came from TerraClimate, a 4-km grid climatalogical data set [@abatzoglou2018].
+Annual reference evapotranspiration (ET~O~) data came from TerraClimate, a 4-km grid climatological data set [@abatzoglou2018].
 TerraClimate uses a Penman-Monteith approach to generate a reference evapotranspiration.
@@
 
-![Linear relationship between observed BFI and predicted BFI. The solid line is the 1:1 line, the dashed line is regressed to the data.](images/actual-predicted.png){#fig-actual_predicted}
+![Linear relationship between observed BFI and predicted BFI. The solid line is the 1:1 line, the dashed, green line is regressed to the data.](images/actual-predicted.png){#fig-actual_predicted}
 
@@
 : Performance of model predictions for BFI for all sites split by various classifications.
-n is number of observatios, R^2^ is the coefficient of determination of a linear regression, MSE is mean-squared-error, RMSE is root-mean-squared-error, MAE is mean-absolute-error, NSE is Nash-Sucliffe efficiency, and pbias is percent bias.
+n is number of observations, R^2^ is the coefficient of determination of a linear regression, MSE is mean-squared-error, RMSE is root-mean-squared-error, MAE is mean-absolute-error, NSE is Nash-Sutcliffe efficiency, and pbias is percent bias.
 {#tbl-performance}
@@
 
-Given the variations in the period of record across the instrumented network (see SUPPLEMENTAL - Period of Record figure), we analyzed trends in base flow and BFI in relation to coincident trends in climate variables.
+Given the variations in the period of record across the instrumented network (see Figure S2), we analyzed trends in base flow and BFI in relation to coincident trends in climate variables.
 Trends are classified as coincident when the direction of the climate variable trend (positive or negative) aligns with the trend observed in base flow or BFI (@tbl-coincident-trends).
@@
 
-However, the utility of the streamgage networks for base-flow analyses in Arizona is limited by their design and focus.
+The utility of the streamgage networks for base-flow analyses in Arizona is limited by their design and focus.
 There has been an increase in the number of streamgages in the region to meet regulatory imperatives, such as the Clean Water Act.
@@
 New, non-USGS streamgages are typically installed by flood control districts (e.g.
-the ALERT system) and are designed for tracking flood flows rather than base flow.
+the ALERT system) and are designed to track flood flows rather than base flow.
 New USGS streamgages have been added over the past decade to address instream flow rights and to improve the density of monitoring in the future.

### 2025-01-29 MANUSCRIPT.qmd

@@
 $K$-fold cross-validation provides an unbiased estimate of a model's accuracy on unseen data, while also insuring against overfitting or underfitting.
-In this approach, the data is randomly divided into $k$ folds of equal size.
+In this approach, the data are randomly divided into $k$ folds of equal size.
 The model is trained on $k-1$ folds and tested on the remaining fold, referred to as the validation set.
@@
 Each fold serves as the validation set once, while the remaining folds are combined to create a training set for model development.
-Performance metrics for the test set are calculated and recorded, and this process is repeated for all $k$ folds.](images/k-fold_fig.png){#fig-k-fold width="6000"}
+Performance metrics for the test set are calculated and recorded, and this process is repeated for all $k$ folds.](images/k-fold_fig.png){#fig-k-fold fig-align="center" width="100%"}
 
@@
 Additionally, headwater regions of perennial rivers tend to exhibit higher BFI values.
-In contrast, low BFI values are found in areas like the Defiance Uplift in northeastern Arizona and the arid southern regions of the state.
+In contrast, low BFI values are found in areas like the Defiance Plateau in northeastern Arizona and the arid southern regions of the state.
 
-![Predicted long-term BFI values for 8-digit HUC (1991-2020)](images/BFI_HUC8_20241203.png){#fig-bfi-huc}
+![Predicted long-term BFI values for 8-digit HUC (1991-2020)](images/BFI_HUC8_20250129.png){#fig-bfi-huc}
 
@@
 The stream reaches of the Little Colorado River Basin (northeastern Arizona) indicate consistently low BFI values (\< 0.2).
-This is likely due to low-yielding perched aquifers underlying the Defiance Uplift in northeastern Arizona, which are hydrologically connected to surface streams, while the high-yield, confined regional aquifer is much deeper [@blanchard02].
+This is likely due to low-yielding perched aquifers underlying the Defiance Plateau in northeastern Arizona, which are hydrologically connected to surface streams, while the high-yield, confined regional aquifer is much deeper [@blanchard02].
 A notable tendency emerges along most major rivers in the study area: upstream reaches tend to exhibit higher BFI values, while downstream reaches display lower values.
@@
 The results highlight significant spatial variability in BFI, with approximately 32% of Arizona’s long-term streamflow originating from groundwater discharge.
-Regions such as the Grand Canyon and Mogollon Rim demonstrate high BFI values due to strong groundwater-surface water interactions, while areas like the Little Colorado River Basin exhibit low BFI values, reflecting limited groundwater recharge.
+Regions such as the Grand Canyon and Mogollon Rim demonstrate high BFI values due to strong groundwater-surface water interactions, while areas like the Little Colorado River Basin exhibit low BFI values, reflecting limited groundwater recharge (@fig-bfi-huc).
 
 Using an XGBoost machine learning algorithm, we successfully predicted long-term BFI in ungauged basins, achieving strong model performance (R² = 0.764, RMSE = 0.129).
-The model performed well across all classifications, demonstrating its robustness in capturing base-flow dynamics across a region with substantial variability in climate, elevation, and physiographic characteristics.
-Key predictors included elevation, land cover, and soil type, highlighting the importance of integrating hydroclimate and physiographic characteristics into regional hydrological models.
+The model performed well across all classifications, demonstrating its robustness in capturing base-flow dynamics across a region with substantial variability in climate, elevation, and physiographic characteristics (@tbl-performance).
+Key predictors included elevation, land cover, and soil type, highlighting the importance of integrating hydroclimate and physiographic characteristics into regional hydrological models (@fig-shap_values).
 These predictions address the limitations posed by Arizona’s sparse streamgage network, offering a scalable approach to estimate BFI in data-limited regions.
 
-Our analysis of BFI trends in gauged catchments revealed that precipitation is the primary driver of base-flow variability, with evapotranspiration and temperature contributing additional complexity.
+Our analysis of BFI trends in gauged catchments revealed that precipitation is the primary driver of base-flow variability, with evapotranspiration and temperature contributing additional complexity (@tbl-coincident-trends).
 These findings emphasize the critical role of climate-hydrology interactions in shaping groundwater contributions to streamflow.

### 2025-01-31 MANUSCRIPT.qmd
finalizing

@@
       - 'Department of Hydrology and Atmospheric Sciences, University of Arizona, Tucson, AZ, USA'
+
+  - name: Temuulen Sankey
+    orcid: 0000-0002-6919-8428
+    corresponding: false
+    roles: 
+      - Writing – Review & Editing 
+    affiliations:
+      - 'School of Informatics, Computing, and Cyber Systems, Northern Arizona University, Flagstaff, AZ, US'
 
@@
 
-![Map of Arizona indicating US Geological Survey (USGS) streamgages used in this study.
-8-digit HUC subbasin boundaries and physiographic regions shown.](images/StudyArea_20241203.png){#fig-study-area}
+![Map of Arizona and US Geological Survey (USGS) streamgages used in this study. 8-digit HUC subbasin boundaries and physiographic regions shown.](images/StudyArea_20241203.png){#fig-study-area}
 
@@
 While our data selection ensures a robust record for this analysis, it also highlights gaps in spatial coverage that the machine learning model can address.
-Arizona has 184 active USGS streamflow stations (as of 2024) across an area of 295,253 km², whereas Indiana (a humid state in the U.S.) has 189 active stations within an area of 94,326 km².
+Arizona has 184 active USGS streamflow stations (as of 2024) across an area of 295,253 km², whereas Indiana (a humid state in the U.S.), for example, has 189 active stations within an area of 94,326 km².
 This translates to a streamgage density of approximately 2.004 gages per 1,000 km² in Indiana, over three times higher than Arizona's density of 0.623 gages per 1,000 km².
@@
 The hyperparameters evaluated included the learning rate ($\eta$), minimum split loss ($\gamma$), maximum tree depth, minimum child weight, and the number of trees.
-While not an exhaustive list of all possible XGBoost hyperparameters, the range of values explored provided sufficient variation to ensure the selection of a high-performing model.The optimal hyperparameters were determined to be: 700 trees, a learning rate ($\eta$) of 0.05, a minimum split loss ($\gamma$) of 0.075, a maximum tree depth of 7, and a minimum child weight of 5.
+While not an exhaustive list of all possible XGBoost hyperparameters, the range of values explored provided sufficient variation to ensure the selection of a high-performing model.
+The optimal hyperparameters were determined to be: 700 trees, a learning rate ($\eta$) of 0.05, a minimum split loss ($\gamma$) of 0.075, a maximum tree depth of 7, and a minimum child weight of 5.
 Using these values, the XGBoost model was trained on the dataset with 10-fold cross-validation.
@@
 
-The regionalized (HUC-8) long-term BFI (1991–2020) is shown in @fig-bfi-huc, while the long-term BFI for stream reaches with a Strahler stream order of 3 or greater is presented in @fig-bfi-streams.
+![Predicted long-term BFI values for 8-digit HUC (1991-2020)](images/BFI_HUC8_20250129.png){#fig-bfi-huc}
+
+The regionalized (HUC-8) long-term BFI (1991–2020) is shown in @fig-bfi-huc.
 Basins with high BFIs, such as those along the Grand Canyon in the northwestern part of the study area, indicate greater surface water and groundwater interaction.
@@
 In contrast, low BFI values are found in areas like the Defiance Plateau in northeastern Arizona and the arid southern regions of the state.
-
-![Predicted long-term BFI values for 8-digit HUC (1991-2020)](images/BFI_HUC8_20250129.png){#fig-bfi-huc}
-
-![Predicted long-term BFI values for all stream reaches of Strahler stream order 3 and greater (1991-2020)](images/BFI_AllStream.png){#fig-bfi-streams}
-
-@fig-bfi-streams indicates the predicted long-term BFI for stream reaches in Arizona of a Strahler stream order of 3 or greater.
-Stream-level BFI values often correspond to the long-term BFI observed in HUC-8 basins, as geospatial predictor data were aggregated at the basin level.
-@fig-bfi-streams highlights stream density across the region, revealing areas with higher concentrations of perennial streams, particularly in regions with favorable hydrogeologic conditions, such as spring-fed systems in the Verde River and the Salt River headwaters.
-Conversely, lower BFI values are observed in arid regions and those with lower stream densities such as parts of the Little Colorado River Basin, where groundwater recharge is limited [@flint_flint07].
 While BFI values and their spatial patterns may align with certain terrain features or climatic patterns, it is important to recognize that BFI is driven by complex interactions among multiple factors [@fig-shap_values].

### 2025-01-31 MANUSCRIPT.qmd
submission version

@@
 
+## Open Research
+
+All code and data necessary to reproduce the analyses presented in this manuscript are available at: <https://github.com/CaelumMroczek/baseflow-AZ>.
+
+## Acknowledgements
+
+The authors would like to express gratitude to the following people for helpful discussion and suggestions: Cole Denver, Aidan Houlgate, Rayni Lewis, Ryan Lima, Quentin McCalla, Isaiah Mesa, Georgia Roberts.
+This research was supported by an Arizona Board of Regents’ Grant from the Technology and Research Initiative Fund governed by A.R.S.
+§ 15-1648.
+The authors acknowledge the financial support provided by this grant, which made this study possible.
+
 ## References {.unnumbered}

### 2025-03-20 elsevier_MANUSCRIPT.qmd
adding elsevier formatting

---
title: "elsevier_MANUSCRIPT"
format: html
---

### 2025-03-20 elsevier_MANUSCRIPT.qmd
Update elsevier_MANUSCRIPT.qmd

@@
 title: "elsevier_MANUSCRIPT"
-format: html
+format:
+  pdf: default
+  elsevier-pdf:
+    keep-tex: true
+    journal:
+      name: Journal of Hydrology
+      formatting: preprint
+      model: 1p
+      layout: onecolumn
+      cite-style: number
+      highlights:
+        - Approximately 32% of Arizona's streamflow originates from groundwater discharge, with significant spatial and climatic variability. 
+        - A machine learning model (XGBoost) successfully predicted long-term BFI in ungauged basins, addressing gaps in a sparse streamgage network. 
+        - Base-flow trends align most closely with precipitation changes, emphasizing the role of climate in shaping groundwater-surface water dynamics in dryland regions.
 ---
+
+## Introduction {#sec-intro}
+
+Dryland regions, encompassing arid, semi-arid, hyper-arid, and dry sub-humid systems, account for 40% of the Earth’s land surface.
+These regions are home to approximately 2 billion people globally and constitute the largest terrestrial biome [@iucn-drylands-2019].
+Despite supporting diverse ecosystems and human populations, dryland regions face mounting hydrologic challenges exacerbated by increasing urbanization, expanding agricultural activities, and climate-induced amplification of precipitation patterns [@taylor2013].
+This water scarcity is intensifying due to the compounding effects of climate variability and increased groundwater extraction [@taylor2013].
+In drylands, groundwater serves as a vital resource for sustaining ecosystems and meeting human needs [@scanlon2006; @yao2018].
+
+Base flow is the sustained portion of streamflow in the absence of runoff that is derived from groundwater discharge [@usgs-glossary-2018].
+Base flow is critical to maintaining seasonal low-flow regimes, supporting aquatic ecosystems, and facilitating the transport of nutrients and chemicals.
+Base-flow contribution to streamflow can be highly variable spatially [@singh2018; @bosch2017; @beck2013], and temporally [@ficklin2016; @tan2020].
+Increasing groundwater extraction, changes in land cover/land use, and changes in precipitation patterns due to climate change affect the timing and volumes of base flow [@tan2020; @taylor2013].
+Effective management of water quantity and quality depends on understanding seasonal and interannual base-flow patterns and long-term changes in base-flow behavior.
+
+The Base-Flow Index (BFI) is the ratio of the long-term mean base-flow volume to the long-term total streamflow volume expressed as a percentage.
+BFI serves as a normalized measure of groundwater contribution interannually or between basins.
+BFI is determined by hydrograph separation and is influenced by the climate and physiographic characteristics of a catchment [@neff2005; @beck2013; @singh2018].
+Between catchments, base flow fluctuates according to changes in the moisture content of the vadose zone, influenced by varying levels of evapotranspiration and aquifer storage dynamics [@bosch2017].
+Since BFI calculations rely on instrumented stream records, it remains unknown for ungauged catchments, which encompass most of the earth's land surface [@fekete2007].
+Addressing this information gap is integral to approaching a comprehensive understanding of groundwater dynamics globally.
+
+Advancements in machine learning provide tools to predict hydrologic indices in ungauged basins, addressing the limitations of sparse streamgage networks.
+To tackle the challenge of quantifying base flow in ungauged catchments, numerous studies have applied both regression and machine learning methods.
+@ahiablame2013 found that using a regression model to estimate annual base flow of ungauged catchments was reasonably easy and accurate.
+@beck2013 overcame the nonlinearity of basin characteristics and improved results of multivariate analyses by using artificial neural networks to estimate BFI globally.
+@singh2018 implemented a random forest algorithm to predict long-term BFI for ungauged catchments across New Zealand.
+These applications demonstrate the versatility and effectiveness of machine learning in capturing complex ecohydrologic dynamics and improving our understanding of groundwater contributions to streamflow.
+
+This study develops a technique for estimating BFI in ungauged basins across Arizona and evaluates the state’s long-term BFI.
+Regional trends in base flow and BFI at instrumented sites are also analyzed, with these trends being linked to coincident trends in precipitation, reference evapotranspiration (ET~O~), and temperature over the same periods.
+Using a machine learning model trained on the hydrogeologic characteristics of surface water basins, we estimate the mean BFI for ungauged basins from 1991 to 2020.
+This approach helps address the spatial gaps in the streamgage network, which is relatively sparse across the state.
+The results offer novel insights into low-flow processes in both gauged and ungauged basins, enhancing our understanding of climate controls on consistent flows in Arizona.
+This study contributes to a more comprehensive view of hydrological dynamics in the context of arid and semi-arid landscapes.
+
+## Methods {#sec-methods}
+
+### Study Area {#sec-study-area}
+
+The state of Arizona, located in the southwest United States, covers a total area of 295,253 km$^2$.
+Arizona is divided into two primary physiographic provinces: the Colorado Plateau in the northeast, and the Basin-and-Range region in the west and south.
+The Central Highlands is a transition zone consisting of scattered basins separated by the mountainous foothills of the Mogollon Rim.
+The Colorado Plateau is dominated by high-elevation desert with an average elevation of 1,936 masl (6,352 ft).
+Mean temperatures range from -6$^\circ$C (20$^\circ$F) to 26$^\circ$C (80$^\circ$F) and it averages 580 mm (23 in) of precipitation.
+The Basin-and-Range region has a semi-arid to arid climate with an average elevation of 490 masl (1600 ft).
+Average temperature ranges from 15$^\circ$C (60$^\circ$F) to 43$^\circ$C (110$^\circ$F) and the region averages 200 mm (8 in) of precipitation annually [@az_climate2024].
+
+![Map of Arizona and US Geological Survey (USGS) streamgages used in this study. 8-digit HUC subbasin boundaries and physiographic regions shown.](images/StudyArea_20241203.png){#fig-study-area}
+
+Arizona's hydrology varies seasonally and spatially between its physiographic regions.
+In the summer, localized and intense convective precipitation events are driven by the North American Monsoon, while in the winter, orographic precipitation comes from Pacific frontal systems [@eastoe2019mtnblock].
+While monsoonal precipitation can account for up to 50% of annual precipitation, evaporation and dry preceding soil properties leads to most precipitation becoming runoff [@sheppard2002].
+As such, 94% of streams in Arizona are ephemeral or intermittent [@epa_AZephemeral].
+Much of the hydrology of Arizona is snow-melt derived, driven by spring melt from the high-elevation Colorado Plateau winter snowpack.
+While winter precipitation provides only 30% of annual averages, it provides the majority of water for natural reservoirs [@sheppard2002].
+
+### Data {#sec-data}
+
+Daily observed streamflow data obtained from the United States Geological Survey (USGS) National Water Information System (NWIS) were used in this study.
+Streamgages were selected depending on criteria to ensure the applicability of each site.
+Following the findings of @odonnell2016, which determined that 8–10 years of calibration data are necessary to account for climate variability in paired watershed studies in the region, a minimum record length of 10 years was required.
+Additionally, years with more than 30 missing days of streamflow data were excluded from the analysis.
+This study focuses on natural, streamflow-influencing dynamics, so streamgages affected by regulation or diversions were excluded.
+The regulated river streamgages were identified through annual reports on water data published by the USGS [@usgs2010].
+Furthermore, streamgages along the Colorado River were omitted because they represent managed flows governed by the Colorado River Compact.
+After applying these selection criteria, 205 USGS streamgages with acceptable periods of record were included in the study (@fig-study-area).
+Periods of record ranged from 10 to 112 years, with a median of 28 years.
+
+While our data selection ensures a robust record for this analysis, it also highlights gaps in spatial coverage that the machine learning model can address.
+Arizona has 184 active USGS streamflow stations (as of 2024) across an area of 295,253 km², whereas Indiana (a humid state in the U.S.), for example, has 189 active stations within an area of 94,326 km².
+This translates to a streamgage density of approximately 2.004 gages per 1,000 km² in Indiana, over three times higher than Arizona's density of 0.623 gages per 1,000 km².
+This stark difference underscores the relative sparsity of Arizona's streamgage network, particularly in the context of its larger geographic area and the unique hydrological challenges posed by its arid and semi-arid landscapes.
+<!--# Add something here to discuss the application in broader semi-arid low streamgage regions globally -->
+
+Watersheds across the United States are delineated by the USGS using a hydrologically-defined network.
+This system delineates the country using hierarchical hydrologic unit codes (HUCs), where each subsequent basin includes the digits of the enclosing basin.
+Here, 8-digit HUCs (HUC 8s) are used to divide Arizona into 84 sub-basins that are fully or partially in the state (@fig-study-area).
+These HUC 8 sub-basins are analogous to medium-sized river basins and are defined by surface water characteristics.
+
+Annual precipitation and temperature data came from the PRISM climate group at Oregon State University at a resolution of 4 km (<https://prism.oregonstate.edu;> [@daly2008]).
+The PRISM dataset provides valuable insights into regional climate in ungauged regions and has been shown to perform well across the southwestern US [@buban_PRISM].
+Instead of the water year, PRISM data uses a calendar-year format, which was adopted for consistency in the water balance.
+Although this may introduce challenges in the annual estimates due to inter-annual snow storage, the use of long-term annual averages is likely to reduce any potential errors [@reitz2017].
+
+Annual reference evapotranspiration (ET~O~) data came from TerraClimate, a 4-km grid climatological data set [@abatzoglou2018].
+TerraClimate uses a Penman-Monteith approach to generate a reference evapotranspiration.
+The ET~O~ values were calculated assuming a reference grass surface across the landscape with unlimited water.
+In the drylands of the southwestern US, ET~O~ typically exceeds precipitation annually [@zomer2022].
+
+A 30-meter resolution Digital Elevation Model (DEM) of Arizona was used to derive key basin characteristics: basin area, average slope, and the proportion of each basin oriented toward north or south aspects.
+Various geospatial variables, such as aspect, were disaggregated then averaged to assess the areal percentage of each sub-variable within individual HUC 8 basins.
+By calculating these percentages, we aimed to get a more comprehensive understanding of landscape composition across space.
+Land cover was acquired from USGS-NLCD (National Land Cover Database), hydrologic soil group from SSURGO (Soil Survey Geographic Database), and underlying geology and karst from USGS were all similarly averaged across the basins.
+Aggregating variables to align with the HUC 8 boundaries allowed for more precise predictions of BFI by integrating spatial variations within each basin.
+
+### Base-flow separation {#sec-bf_sep}
+
+Directly measuring base flow and BFI presents unique challenges [@eckhardt2008].
+The technique chosen to separate base flow has been shown to affect results, and the choice of base-flow separation method is subjective since ‘true’ BFI values are not known [@beck2013].
+However, many methods have been developed to estimate these values.
+These methods include the use of tracers [@gonzales2009], graphical interpolation [@instituteofhydrology1980; @sloto1996], and digital filters [@arnold1995; @eckhardt2005; @lyne1979; @nathan1990].
+These techniques have varying levels of applicability depending on the spatial scale, time span, and the scope of the study.
+Comparisons of various base-flow separation techniques have been made in previous studies [e.g.
+@eckhardt2005; @eckhardt2008; @nathan1990]; this study does not explore the superiority of different methods.
+
+Base flow was calculated using a single-parameter, recursive digital filter technique from @nathan1990.
+This base-flow separation technique is based on a recursive digital filter used in signal analysis that separates high-frequency signals (quickflow) from low-frequency signals (base flow) [@lyne1979].
+@eckhardt2023 noted that recursive digital filters lack a physical basis, but as the method is easy to automate, objective, and repeatable, it is appropriate for a regional-scale study.
+The Lyne-Hollick filter has been used in multiple studies [@ARNOLD200021; @santhi2008; @bloomfield2009; @singh2018], and it takes the form of
+
+$$
+b = \alpha b_{k-1} + \frac{1-\alpha}{2}(Q_k + Q_{k-1})
+$$ {#eq-bf-separate}
+
+where $b$ is base flow, $\alpha$ is the filter parameter, $Q$ is the total streamflow, and $k$ is the time step.
+A filter parameter $\alpha$ of 0.925 was used as in @nathan1990 and @fuka2014ecohydrology.
+The filter was run three times (forward, backward, forward) to attenuate the base-flow signal.
+
+### Machine Learning
+
+The implementation of machine learning models to predict hydrologic indices has been successful in past studies [@singh2018; @schmidt_ML_2020; @Rozos2021Machine].
+In this work, we used the eXtreme Gradient Boosting (XGBoost) algorithm [@chen2016xgboost] to predict BFI at ungauged locations using catchment characteristics as predictors [@tbl-predictors].
+The XGBoost algorithm is a decision tree-based ensemble algorithm, which can be adapted for either regression or classification problems.
+This algorithm iteratively builds an ensemble of decision trees, where each tree corrects errors from previous trees to improve predictions [@chen2016xgboost].
+Its efficiency, scalability, and robustness have made it increasingly popular in recent years, with successful applications in environmental modeling tasks such as streamflow forecasting [@szczepanek2022daily; @ni2020streamflow] and land use/land cover classification [@Georganos2018Very].
+
+XGBoost operates by leveraging gradient boosting on decision tree algorithms, combining multiple low-variance models to produce a robust overall prediction.
+Gradient boosting works iteratively: the initial tree is trained on the target values, while subsequent trees are trained on the residual errors of the preceding tree.
+Each tree is assigned a weight based on its contribution to reducing error, and these weights are used to determine the influence of each tree in the final model.
+The ultimate prediction is made by aggregating the outputs of all $n$ weighted trees in the ensemble.
+In this study, the trained XGBoost model is used to predict BFI in ungauged catchments based on geospatial and hydroclimate predictor variables (@tbl-predictors).
+Certain features were further subdivided according to their areal coverage within each basin (e.g.
+land cover was divided into 16 subdivisions).
+This approach allowed the model to capture finer-scale spatial variability and improve predictive accuracy.
+
+|   | Variable | Source | Geoprocessing |
+|----|----|----|----|
+| Hydroclimate | Precipitation\* | PRISM | Basin average |
+|  | Mean Temperature\* | PRISM | Basin average |
+|  | Reference Evapotranspiration\* | TerraClimate | Basin average |
+| Geospatial | Elevation\* | DEM | Basin average |
+|  | Area | DEM | Basin average |
+|  | Slope | DEM | Basin average |
+|  | Aspect | DEM | Percent areal coverage |
+|  | Land Cover\* | NLCD | Percent areal coverage |
+|  | Hydrologic Soil Group\* | SSURGO | Percent areal coverage |
+|  | Geology | USGS | Percent areal coverage |
+|  | Karst | USGS | Percent areal coverage |
+
+: Basin-characteristic variables used as initial features in XGBoost model. Starred features are maintained in the final, dimensionality-reduced model. {#tbl-predictors}
+
+Our initial training dataset comprised 7,724 observations across 45 variables.
+To optimize the model's performance, we first conducted an exhaustive grid search combined with 5-fold cross-validation to identify the optimal hyperparameter values.
+The hyperparameters evaluated included the learning rate ($\eta$), minimum split loss ($\gamma$), maximum tree depth, minimum child weight, and the number of trees.
+While not an exhaustive list of all possible XGBoost hyperparameters, the range of values explored provided sufficient variation to ensure the selection of a high-performing model.
+The optimal hyperparameters were determined to be: 700 trees, a learning rate ($\eta$) of 0.05, a minimum split loss ($\gamma$) of 0.075, a maximum tree depth of 7, and a minimum child weight of 5.
+Using these values, the XGBoost model was trained on the dataset with 10-fold cross-validation.
+
+$K$-fold cross-validation provides an unbiased estimate of a model's accuracy on unseen data, while also insuring against overfitting or underfitting.
+In this approach, the data are randomly divided into $k$ folds of equal size.
+The model is trained on $k-1$ folds and tested on the remaining fold, referred to as the validation set.
+This process is repeated $k$ times, with each fold serving as the validation set exactly once.
+Each iteration trains an independent model with the same hyperparameters but using a different subset of training data.
+By averaging the model performance across all $k$ folds, we achieve a robust and reliable estimate of predictive accuracy.
+Root mean squared error (RMSE) was used as the performance metric for both model optimization and evaluation.
+RMSE provided a consistent and interpretable measure of the model’s accuracy throughout the training and validation process.
+
+![Schematic of $k$-fold cross validation.
+The dataset is randomly divided into $k$ stratified folds.
+Each fold serves as the validation set once, while the remaining folds are combined to create a training set for model development.
+Performance metrics for the test set are calculated and recorded, and this process is repeated for all $k$ folds.](images/k-fold_fig.png){#fig-k-fold fig-align="center" width="100%"}
+
+#### Feature Selection {#sec-feat-selection}
+
+Machine learning models are prone to overfitting, especially when provided with a large set of predictive features [@Ying_2019].
+Overfitting can degrade model performance on unseen data and increase the demand for computational resources and memory storage [@feat_selec2017].
+Dimensionality reduction offers a robust solution to these challenges and generally falls into two broad categories: feature extraction and feature selection.
+
+Feature extraction involves transforming the original dataset into a lower-dimensional feature space.
+However, this process generates new features that often lack the physical interpretability of the original variables.
+In contrast, feature selection identifies a subset of the original features, preserving their physical meaning while improving model readability and interpretability [@feat_selec2017].
+In this study, supervised feature selection was employed to reduce the number of predictors, which enhanced learning performance, reduced computational costs, and mitigated overfitting.
+
+To begin, an initial model was trained using the full feature set of 45 predictors (@tbl-predictors).
+A feature selection method based on feature importance was then applied to identify and remove less relevant and noisy features.
+Feature importance scores quantify the contribution of individual features—either positively or negatively—to the model's predictions [@ML_interpretability2019].
+In this analysis, SHAP (SHapley Additive exPlanations) values were used to compute feature importance scores [@SHAP_values].
+
+SHAP values is a method to explain the prediction of an individual instance by calculating the contribution of each feature to that prediction.
+The method is based on coalition game theory and is discussed further in @SHAP_values.
+Here, SHAP values are used for global interpretation of feature importance and feature effects on the model.
+Global feature importance is produced by the absolute Shapley values of each feature across the dataset, providing a list of features in order of most to least important.
+Feature effects provide an indication of the relationship between the value of a predicting feature and its impact on the prediction.
+
+The ten most important features [@fig-shap_values] were selected based on their SHAP values and used to train a subsequent model [@tbl-predictors].
+This refined model demonstrated improved performance and reduced computational time compared to the initial model.
+The final model, trained on this optimized feature subset, was ultimately used for the analysis presented here.
+
+### Statistical Analyses
+
+Statistical analyses were conducted on annual BFI and base-flow values from instrumented streamgages to identify temporal trends using the Mann-Kendall nonparametric trend test [@kendall1970; @mann1945nonparametric].
+This test detects monotonic trends in datasets that are non-parametric and assumes the absence of autocorrelation among observations.
+This test is widely used in studies of this nature [@ficklin2016; @ayers2019; @woodhouse_udall22].
+
+To check for autocorrelation, we applied the Durbin-Watson test, which revealed significant autocorrelation at four streamgages on an annual basis.
+Of these, only one streamgage (09486500 - Santa Cruz River at Cortaro, AZ) showed a significant trend in BFI.
+This streamgage was excluded from the trend analysis, as autocorrelation could inflate the variance of the Mann-Kendall statistic, potentially leading to biased trend estimates [@MK_Hamed1998].
+Trends with a $\rho \le 0.05$ are considered significant.
+
+## Results
+
+### BFI of Ungauged Catchments
+
+#### Model Validation
+
+Predicted values of BFI are plotted against observed values for the entire period of record of the instrumented dataset in @fig-actual_predicted .
+The agreement between "out-of-bag" predictions (blind cross-validation, treating each site as ungauged) and observed values is acceptable (R^2^ = 0.764) indicating that the model performs well across the full dataset.
+The overall RMSE is 0.129 and the overall percent bias (pbias) is -5.6%.
+Model performance metrics across various classifications are summarized in @tbl-performance.
+These performance metrics demonstrate that the regional model performs consistently well across different spatial and climatic classifications.
+However, the negative pbias values across all classifications, along with the overall pbias, indicate a systematic underprediction of BFI by the model.
+Categories with relatively lower R^2^ and Nash-Sutcliffe Efficiency (NSE) values also exhibit higher biases, reflecting weaker model performance in those specific contexts.
+
+![Linear relationship between observed BFI and predicted BFI. The solid line is the 1:1 line, the dashed, green line is regressed to the data.](images/actual-predicted.png){#fig-actual_predicted}
+
+| **Classification Group** | **n** | **R2** | **MSE** | **RMSE** | **MAE** | **NSE** | **pbias** |
+|---------|---------|---------|---------|---------|---------|---------|---------|
+| Climate - Monsoon Dominated | 3039 | 0.633 | 0.016 | 0.126 | 0.074 | 0.619 | -13.7 |
+| Climate - Snowmelt Dominated | 4685 | 0.733 | 0.015 | 0.121 | 0.087 | 0.725 | -3.5 |
+| PhysRegion - Basin&Range | 6147 | 0.733 | 0.016 | 0.127 | 0.084 | 0.724 | -6.3 |
+| PhysRegion - CO Plateau | 1577 | 0.846 | 0.011 | 0.104 | 0.073 | 0.843 | -3.8 |
+| Climate - Warm-Wet | 1506 | 0.693 | 0.014 | 0.117 | 0.077 | 0.685 | -8.2 |
+| Climate - Warm-Dry | 2351 | 0.693 | 0.022 | 0.147 | 0.092 | 0.675 | -11.9 |
+| Climate - Cool-Wet | 2350 | 0.738 | 0.011 | 0.106 | 0.078 | 0.736 | -1.7 |
+| Climate - Cool-Dry | 1517 | 0.831 | 0.012 | 0.111 | 0.078 | 0.827 | -4.3 |
+| Slope - High | 3795 | 0.776 | 0.012 | 0.111 | 0.079 | 0.771 | -3.3 |
+| Slope - Low | 3929 | 0.724 | 0.018 | 0.133 | 0.085 | 0.713 | -9.1 |
+
+: Performance of model predictions for BFI for all sites split by various classifications.
+n is number of observations, R^2^ is the coefficient of determination of a linear regression, MSE is mean-squared-error, RMSE is root-mean-squared-error, MAE is mean-absolute-error, NSE is Nash-Sutcliffe efficiency, and pbias is percent bias.
+{#tbl-performance}
+
+#### Predictor Importance {#sec-predictor-importance}
+
+The predictors used to estimate BFI at ungauged sites were evaluated for their importance in the final XGBoost model, as illustrated in @fig-shap_values.
+The most influential feature for predicting long-term BFI is basin elevation.
+While elevation itself does not directly affect base-flow characteristics, it has consistently been identified as a key predictor in previous BFI studies [@singh2018; @beck2013].
+The importance of elevation aligns with findings from @beck2013, highlighting its role as a proxy for climate variables such as temperature, precipitation, and snowpack duration.
+Seasonal snowpack duration, in particular, has been shown to strongly correlate with springflow and groundwater recharge in this region [@donovan_karst_2022].
+This relationship is further supported by hydroclimate features where higher temperatures tend to negatively influence BFI, while precipitation exhibits a mixed influence.
+In some cases, higher precipitation values correlate with lower BFI, likely due to a larger proportion of precipitation contributing to runoff rather than infiltration and groundwater recharge.
+
+![SHAP value plot of features used in final model.
+Land cover features are indicated by the percentage of cover by each land cover type and soil types are defined by hydrologic soil group.](images/shap_summary.png){#fig-shap_values}
+
+Land cover and land use predictors also play a significant role in BFI estimation.
+Analysis of SHAP values indicates that a higher percentage of evergreen forest positively influences BFI predictions, while higher proportions of shrubland and developed land exert a negative influence.
+Similarly, hydrologic soil types show distinct trends in their impact on BFI.
+Soil Type C, characterized by moderately high runoff potential (20-40% clay), tends to negatively influence BFI.
+In contrast, Soil Type A, which has low runoff potential and facilitates rapid water infiltration, exhibits a mixed influence [@nrcs2009].
+The mixed effects of Soil Type A are likely due to SHAP values capturing interactions between features rather than direct relationships.
+For example, regions dominated by Soil Type A may also have steep slopes or sparse vegetation.
+Additionally, the relationship between Soil Type A and BFI is likely non-linear and influenced by the complex dynamics of regional base-flow processes.
+
+#### Predicted Long-term BFI
+
+![Predicted long-term BFI values for 8-digit HUC (1991-2020)](images/BFI_HUC8_20250129.png){#fig-bfi-huc}
+
+The regionalized (HUC-8) long-term BFI (1991–2020) is shown in @fig-bfi-huc.
+Basins with high BFIs, such as those along the Grand Canyon in the northwestern part of the study area, indicate greater surface water and groundwater interaction.
+Elevated BFI values are also observed along portions of the Mogollon Rim, a heavily forested region with high precipitation that marks the transition between physiographic regions.
+Additionally, headwater regions of perennial rivers tend to exhibit higher BFI values.
+In contrast, low BFI values are found in areas like the Defiance Plateau in northeastern Arizona and the arid southern regions of the state.
+While BFI values and their spatial patterns may align with certain terrain features or climatic patterns, it is important to recognize that BFI is driven by complex interactions among multiple factors [@fig-shap_values].
+
+### BFI of Gauged Catchments
+
+![Long-term BFI for the period of record from instrumented stream flow data.](images/BFI_Instrumented_20241203.png){#fig-instrumented-bfi}
+
+The long-term BFI for the 205 gauged reaches across Arizona is illustrated in @fig-instrumented-bfi .
+The long-term mean BFI is 0.32, indicating that \~32% of long-term streamflow in Arizona likely originates from groundwater discharge and other delayed sources.
+The highest BFI values (\>0.9) are found along the Grand Canyon in northwestern Arizona.
+The highly karstic geology of this region facilitates the rapid movement of subsurface flow to surface water and spring outlets [@chambless_deep-karst_2023].
+Relatively high BFI values (\>0.8) are found at the spring-fed headwaters of the Verde River (Del Rio Spring) and the spring-fed headwaters of Fossil Creek.
+These results are consistent with interpolated BFI values reported by @wolock2003.
+
+The stream reaches of the Little Colorado River Basin (northeastern Arizona) indicate consistently low BFI values (\< 0.2).
+This is likely due to low-yielding perched aquifers underlying the Defiance Plateau in northeastern Arizona, which are hydrologically connected to surface streams, while the high-yield, confined regional aquifer is much deeper [@blanchard02].
+A notable tendency emerges along most major rivers in the study area: upstream reaches tend to exhibit higher BFI values, while downstream reaches display lower values.
+This pattern is presumed to result from greater groundwater-surface water interactions at stream headwaters, influenced by spring outlets, and the dilution of base flow as water moves downstream.
+This trend is particularly evident along the Gila River, Verde River, and Little Colorado River in Arizona.
+
+#### Trends in Base Flow & BFI {#sec-bfi-trends}
+
+Trends in BFI over the period of record for each streamgage in this analysis are illustrated in @fig-instrumented-trend and @tbl-trends.
+Base flow and BFI trends were analyzed across all instrumented sites over their respective periods of record using the Mann–Kendall test.
+Statistically significant trends were observed in both metrics, with a 72.20% coincidence rate between significant base flow and BFI trends, indicating a strong dependence of BFI on base-flow dynamics.
+
+@fig-instrumented-trend illustrates the spatial variation in BFI trends across the study area.
+Statistically significant decreasing trends are observed at 16.1% of sites, while increasing trends are found at 8.8% of sites, with no clear regional patterns for either.
+In the Basin and Range physiographic region, 9% of sites show increasing trends, while 16.7% exhibit decreasing trends.
+In the Colorado Plateau region, increasing trends occur at 6.1% of sites, and decreasing trends are observed at 14.3% of sites.
+
+![Trends in BFI over full period of record for instrumented sites used in this study.
+Red upward (blue downward) arrows indicate an increasing (decreasing) trend at a significance level of 5%.
+White circles represent sites with no statistically significant trends.](images/BFITrends_20241203.png){#fig-instrumented-trend}
+
+#### Classification Trends
+
+Classifications presented in @tbl-trends were determined based on precipitation regime, physiographic region, climate, and slope.
+The dominant precipitation regime (monsoon vs.
+snowmelt) was identified by analyzing streamflow hydrographs for each station, focusing on peak flow periods during the monsoon season (July–September) and the snowmelt season (March–June).
+Physiographic region was assigned based on which region the streamgage is located.
+Climate classifications were defined as warm (above the long-term median temperature of Arizona), cool (below the long-term median temperature), wet (above the long-term median precipitation), and dry (below the long-term median precipitation).
+Slope was categorized as high (above the median slope) and low (below the median slope).
+
+Statistically significant decreasing trends in BFI were more common than increasing trends across all site classifications (@tbl-trends).
+While decreasing trends dominate, both increasing and decreasing trends are observed within each classification.
+Monsoon-dominated regions exhibit a higher proportion of significant negative trends (24.1%) compared to snowmelt-dominated regions (10.2%), suggesting that monsoon-dominated systems are more consistently correlated with declining base flow.
+Among climate classifications, warm-dry climates have the highest proportion of negative trends (20.0%), followed by warm-wet climates (19.4%), indicating that regions with higher temperatures are more prone to base flow declines.
+Low-slope regions show a greater prevalence of negative trends (20.4%) compared to high-slope regions (11.8%).
+This suggests that flatter areas may be more susceptible to base-flow reductions, potentially due to differences in hydrologic connectivity and recharge dynamics.
+
+| **Classification Group** | **n** | **n_pos** | **n_neg** | **perc_pos** | **perc_neg** |
+|------------|------------|------------|------------|------------|------------|
+| Precipitation - Monsoon Dominated | 87 | 8 | 21 | 0.092 | 0.241 |
+| Precipitation - Snowmelt Dominated | 118 | 9 | 12 | 0.076 | 0.102 |
+| Physiographic Region - Basin and Range | 156 | 14 | 26 | 0.090 | 0.167 |
+| Physiographic Region - Colorado Plateau | 49 | 3 | 7 | 0.061 | 0.143 |
+| Climate - Warm-Wet | 31 | 2 | 6 | 0.065 | 0.194 |
+| Climate - Warm-Dry | 55 | 6 | 11 | 0.109 | 0.200 |
+| Climate - Cool-Wet | 74 | 4 | 9 | 0.054 | 0.122 |
+| Climate - Cool-Dry | 45 | 5 | 7 | 0.111 | 0.156 |
+| Slope - High | 102 | 10 | 12 | 0.098 | 0.118 |
+| Slope - Low | 103 | 7 | 21 | 0.068 | 0.204 |
+
+: Comparison of trends for BFI for all sites split by various classifications.
+Only sites with a significant ($\rho \le 0.05$) trend are included here as established by a Mann-Kendall test for monotonic trends across the full period of record.
+n is the number of sites, n_pos (n_neg) is the number of sites with positive (negative) trends, perc_pos (perc_neg) is the percentage of n with a positive (negative) trend.
+{#tbl-trends}
+
+#### Coincident Climate Trends
+
+Given the variations in the period of record across the instrumented network (see Figure S2), we analyzed trends in base flow and BFI in relation to coincident trends in climate variables.
+Trends are classified as coincident when the direction of the climate variable trend (positive or negative) aligns with the trend observed in base flow or BFI (@tbl-coincident-trends).
+This analysis includes both significant and non-significant trends, which is appropriate where the influence of complex, interconnected processes may not always manifest as statistically significant patterns over limited observational periods [@ficklin2016].
+
+|           | **Climate Variable** | **Coincidence Percentage (%)** |
+|-----------|:--------------------:|:------------------------------:|
+| BFI       |        ET~O~         |             44.88              |
+|           |    Precipitation     |             53.17              |
+|           |     Temperature      |             47.32              |
+| Base Flow |        ET~O~         |             55.12              |
+|           |    Precipitation     |             64.88              |
+|           |     Temperature      |             39.02              |
+
+: Coincident trends of climate variables (ET~O~, precipitation, temperature) with base flow and BFI trends. {#tbl-coincident-trends}
+
+The analysis shows that base flow and BFI trends most frequently align with precipitation trends (64.88% and 53.17%, respectively), emphasizing precipitation as the primary driver of local groundwater recharge and discharge.
+Coincidence with ET~O~ (reference evapotranspiration) trends (55.12% for base flow and 44.88% for BFI) suggests that evapotranspiration also plays a significant role, particularly in arid regions where it can reduce recharge or base flow during dry periods.
+In contrast, temperature trends show lower percentages of coincidence, often opposing base flow and BFI trends.
+Specifically, positive (negative) temperature trends are frequently associated with negative (positive) base-flow trends (60.98%) and BFI trends (52.68%).
+These results highlight the complex interplay between climatic variables and hydrological processes.
+While precipitation exerts the strongest influence on base flow and BFI, evapotranspiration and temperature add further variability in specific environmental contexts.
+
+## Discussion
+### Model Performance and Implications
+
+The regional model demonstrated strong predictive capability (overall R²=0.764) but showed systematic underprediction (negative pbias).
+The systematic bias suggests potential limitations in the model or underlying predictor data, particularly in monsoon-dominated and warm-dry climates.
+The performance differences across climatic and physiographic classifications highlight the need for targeted improvements, such as integrating additional predictors or refining the representation of regional hydrological processes.
+
+### Interpretation of Predictor Variables
+
+Elevation emerged as a critical predictor, reinforcing its effectiveness as a proxy for climate conditions influencing groundwater recharge.
+The mixed influences of Soil Type A may indicate complexities in how infiltration processes are represented in the model or variability in hydrologic conditions within similar soil classes.
+Future research could benefit from explicit inclusion of temporal variables such as snowpack duration, soil moisture storage, and finer-scale land cover attributes.
+
+### Spatial Patterns of BFI
+
+The spatial distribution of predicted and observed BFI reflects known hydrogeological and climatic conditions.
+High BFI values correlate strongly with spring-fed streams and physiographic transitions, while low values align with arid regions.
+This spatial consistency validates the model’s utility for regional water resource assessments and highlights areas vulnerable to declining baseflows.
+
+### Trends and Climate Implications
+
+The prevalence of negative BFI trends, especially in warm-dry and low-slope regions, suggests that climate warming and reduced precipitation could disproportionately impact baseflow.
+The relatively weaker alignment of temperature trends with BFI suggests that temperature impacts are mediated through complex processes, potentially offset by concurrent precipitation or land-use changes.
+This complexity emphasizes the importance of integrated climate and land-management strategies to sustain groundwater-dependent ecosystems.

### 2025-03-24 elsevier_MANUSCRIPT.qmd
editting

@@
         - Base-flow trends align most closely with precipitation changes, emphasizing the role of climate in shaping groundwater-surface water dynamics in dryland regions.
+bibliography: references.bib
 ---
@@
 Predicted values of BFI are plotted against observed values for the entire period of record of the instrumented dataset in @fig-actual_predicted .
-The agreement between "out-of-bag" predictions (blind cross-validation, treating each site as ungauged) and observed values is acceptable (R^2^ = 0.764) indicating that the model performs well across the full dataset.
+The agreement between "out-of-bag" predictions (blind cross-validation, treating each site as ungauged) and observed values indicate strong model performance across the full dataset (R^2^ = 0.764).
 The overall RMSE is 0.129 and the overall percent bias (pbias) is -5.6%.
+
 Model performance metrics across various classifications are summarized in @tbl-performance.
-These performance metrics demonstrate that the regional model performs consistently well across different spatial and climatic classifications.
-However, the negative pbias values across all classifications, along with the overall pbias, indicate a systematic underprediction of BFI by the model.
-Categories with relatively lower R^2^ and Nash-Sutcliffe Efficiency (NSE) values also exhibit higher biases, reflecting weaker model performance in those specific contexts.
+These metrics show consistent model performance across spatial and climatic classifications.
+However, the negative pbias values across all classifications indicate a systematic underprediction of BFI.
+Categories with relatively lower R^2^ and Nash-Sutcliffe Efficiency (NSE) values also exhibit higher biases, reflecting weaker model performance in those contexts.
 
@@
 Seasonal snowpack duration, in particular, has been shown to strongly correlate with springflow and groundwater recharge in this region [@donovan_karst_2022].
-This relationship is further supported by hydroclimate features where higher temperatures tend to negatively influence BFI, while precipitation exhibits a mixed influence.
-In some cases, higher precipitation values correlate with lower BFI, likely due to a larger proportion of precipitation contributing to runoff rather than infiltration and groundwater recharge.
 
@@
 In contrast, Soil Type A, which has low runoff potential and facilitates rapid water infiltration, exhibits a mixed influence [@nrcs2009].
-The mixed effects of Soil Type A are likely due to SHAP values capturing interactions between features rather than direct relationships.
-For example, regions dominated by Soil Type A may also have steep slopes or sparse vegetation.
-Additionally, the relationship between Soil Type A and BFI is likely non-linear and influenced by the complex dynamics of regional base-flow processes.
 
-#### Predicted Long-term BFI
-
-![Predicted long-term BFI values for 8-digit HUC (1991-2020)](images/BFI_HUC8_20250129.png){#fig-bfi-huc}
+#### Predicted BFI in Ungauged Basins
 
@@
 In contrast, low BFI values are found in areas like the Defiance Plateau in northeastern Arizona and the arid southern regions of the state.
-While BFI values and their spatial patterns may align with certain terrain features or climatic patterns, it is important to recognize that BFI is driven by complex interactions among multiple factors [@fig-shap_values].
 
-### BFI of Gauged Catchments
+![Predicted long-term BFI values for 8-digit HUC (1991-2020)](images/BFI_HUC8_20250129.png){#fig-bfi-huc}
+
+### Observed BFI in Gauged Catchments
+
+The long-term BFI for the 205 gauged reaches across Arizona is illustrated in @fig-instrumented-bfi .
+The long-term mean BFI is 0.32, indicating that \~32% of long-term streamflow in Arizona likely originates from groundwater discharge and other delayed sources.
+The highest BFI values (\>0.9) are found along the Grand Canyon in northwestern Arizona, where highly karstic geology facilitates rapid subsurface flow to surface water and spring outlets [@chambless_deep-karst_2023].
+High BFI values (\>0.8) are also found at the spring-fed headwaters of the Verde River and Fossil Creek.
+
+The stream reaches of the Little Colorado River Basin (northeastern Arizona) indicate consistently low BFI values (\< 0.2).
+This is likely due to low-yielding perched aquifers underlying the Defiance Plateau in northeastern Arizona, which are hydrologically connected to surface streams, while the high-yield, confined regional aquifer is much deeper [@blanchard02].
 
@@
 
-The long-term BFI for the 205 gauged reaches across Arizona is illustrated in @fig-instrumented-bfi .
-The long-term mean BFI is 0.32, indicating that \~32% of long-term streamflow in Arizona likely originates from groundwater discharge and other delayed sources.
-The highest BFI values (\>0.9) are found along the Grand Canyon in northwestern Arizona.
-The highly karstic geology of this region facilitates the rapid movement of subsurface flow to surface water and spring outlets [@chambless_deep-karst_2023].
-Relatively high BFI values (\>0.8) are found at the spring-fed headwaters of the Verde River (Del Rio Spring) and the spring-fed headwaters of Fossil Creek.
-These results are consistent with interpolated BFI values reported by @wolock2003.
+#### Trends in Base Flow and BFI {#sec-bfi-trends}
 
-The stream reaches of the Little Colorado River Basin (northeastern Arizona) indicate consistently low BFI values (\< 0.2).
-This is likely due to low-yielding perched aquifers underlying the Defiance Plateau in northeastern Arizona, which are hydrologically connected to surface streams, while the high-yield, confined regional aquifer is much deeper [@blanchard02].
-A notable tendency emerges along most major rivers in the study area: upstream reaches tend to exhibit higher BFI values, while downstream reaches display lower values.
-This pattern is presumed to result from greater groundwater-surface water interactions at stream headwaters, influenced by spring outlets, and the dilution of base flow as water moves downstream.
-This trend is particularly evident along the Gila River, Verde River, and Little Colorado River in Arizona.
-
-#### Trends in Base Flow & BFI {#sec-bfi-trends}
-
-Trends in BFI over the period of record for each streamgage in this analysis are illustrated in @fig-instrumented-trend and @tbl-trends.
+Trends in BFI over the period of record for each streamgage are illustrated in @fig-instrumented-trend and summarized in @tbl-trends.
 Base flow and BFI trends were analyzed across all instrumented sites over their respective periods of record using the Mann–Kendall test.
-Statistically significant trends were observed in both metrics, with a 72.20% coincidence rate between significant base flow and BFI trends, indicating a strong dependence of BFI on base-flow dynamics.
+Statistically significant trends were observed in both metrics, with a 72.2% coincidence rate between significant base flow and BFI trends, indicating a strong dependence of BFI on base-flow dynamics.
 
 @fig-instrumented-trend illustrates the spatial variation in BFI trends across the study area.
-Statistically significant decreasing trends are observed at 16.1% of sites, while increasing trends are found at 8.8% of sites, with no clear regional patterns for either.
-In the Basin and Range physiographic region, 9% of sites show increasing trends, while 16.7% exhibit decreasing trends.
-In the Colorado Plateau region, increasing trends occur at 6.1% of sites, and decreasing trends are observed at 14.3% of sites.
+Statistically significant decreasing trends are observed at 16.1% of sites, while increasing trends are found at 8.8% of sites.
+No consistent regional patterns are evident.
 
@@
 
+Coincident trends are defined as those in which base flow or BFI trends match in direction with climate variable trends over the same period.
+Results are summarized in @tbl-coincident-trends.
+This analysis includes both significant and non-significant trends, which is appropriate where the influence of complex, interconnected processes may not always manifest as statistically significant patterns over limited observational periods [@ficklin2016].
+Coincidence was highest with precipitation (64.88% for base flow; 53.17% for BFI), followed by ETO and then temperature.
+Temperature trends most often show an inverse relationship—i.e., warming associated with declining BFI and base flow.
+
 Given the variations in the period of record across the instrumented network (see Figure S2), we analyzed trends in base flow and BFI in relation to coincident trends in climate variables.
 Trends are classified as coincident when the direction of the climate variable trend (positive or negative) aligns with the trend observed in base flow or BFI (@tbl-coincident-trends).
-This analysis includes both significant and non-significant trends, which is appropriate where the influence of complex, interconnected processes may not always manifest as statistically significant patterns over limited observational periods [@ficklin2016].
 
@@
 
-The analysis shows that base flow and BFI trends most frequently align with precipitation trends (64.88% and 53.17%, respectively), emphasizing precipitation as the primary driver of local groundwater recharge and discharge.
-Coincidence with ET~O~ (reference evapotranspiration) trends (55.12% for base flow and 44.88% for BFI) suggests that evapotranspiration also plays a significant role, particularly in arid regions where it can reduce recharge or base flow during dry periods.
-In contrast, temperature trends show lower percentages of coincidence, often opposing base flow and BFI trends.
-Specifically, positive (negative) temperature trends are frequently associated with negative (positive) base-flow trends (60.98%) and BFI trends (52.68%).
-These results highlight the complex interplay between climatic variables and hydrological processes.
-While precipitation exerts the strongest influence on base flow and BFI, evapotranspiration and temperature add further variability in specific environmental contexts.
+## Discussion
 
-## Discussion
 ### Model Performance and Implications
@@
 The performance differences across climatic and physiographic classifications highlight the need for targeted improvements, such as integrating additional predictors or refining the representation of regional hydrological processes.
+<!--# Discuss that we tested 3 models: regional (by phys region), full predictor set, and trimmed set.
+Trimmed set was the best performing, indicating that the regional separation isn't as beneficial as a more robust predictor dataset --> In the development of the final XGBoost model, we tested making two regional models (one for each physiographic region) to ensure best predictions.
+We found that the state-wide model outperformed the regional model, indicating that regional separation isn't as beneficial as a larger, state-wide dataset.
 
@@
 
-Elevation emerged as a critical predictor, reinforcing its effectiveness as a proxy for climate conditions influencing groundwater recharge.
+Elevation was found to be a critical predictor, reinforcing its effectiveness as a proxy for climate conditions influencing groundwater recharge.
+This finding agrees with previous studies across varied landscapes and scales.
+@singh2018 indicated that mean basin elevation was the most imporant predictor in their random forest model predicting BFI nationally in New Zealand.
+@santhi2008 found that relief (basin max elevation minus minimum elevation) was a key parameter in the best fit of their stepwise regression across hydrologic landscape regions in the conterminous US.
+@beck2013 indicated that basin elevation was the 3rd most important predictor of their artificial neural network in predicting BFI globally.
+
 The mixed influences of Soil Type A may indicate complexities in how infiltration processes are represented in the model or variability in hydrologic conditions within similar soil classes.
@@
 This complexity emphasizes the importance of integrated climate and land-management strategies to sustain groundwater-dependent ecosystems.
+
+## Conclusions
+
+-   The long-term average BFI across Arizona is \~0.32, with strong spatial variability tied to hydrogeology and climate.
+-   The XGBoost model predicts BFI well (R² = 0.764), using elevation, land cover, and soil type as key predictors.
+-   High BFI regions include the Grand Canyon and Mogollon Rim; low BFI areas include the Defiance Plateau and southern Arizona.
+-   Decreasing BFI trends are more prevalent than increasing ones, especially in monsoon-dominated, warm-dry, and low-slope regions.
+-   Precipitation is the dominant driver of BFI and base-flow variability, while ETO and temperature trends add complexity.
+-   The streamgage network in Arizona has limitations for base-flow monitoring; future design should include low-flow-capable instrumentation.
+-   This study demonstrates how machine learning and observational data can be integrated to improve base-flow estimates in data-scarce environments.
+-   The modeling framework is applicable to other dryland regions globally and can inform strategies for climate adaptation and water resource management.

### 2025-03-24 elsevier_MANUSCRIPT.qmd
EDITS

@@
 ---
-title: "elsevier_MANUSCRIPT"
-format:
-  pdf: default
-  elsevier-pdf:
-    keep-tex: true
-    journal:
-      name: Journal of Hydrology
-      formatting: preprint
-      model: 1p
-      layout: onecolumn
-      cite-style: number
-      highlights:
+title: Regional Base-Flow Index in Arid Landscapes Using Machine Learning and Instrumented Records
+author:
+  - name: Caelum Mroczek
+    orcid: 0009-0005-6695-0134
+    corresponding: true
+    email: csm428@nau.edu
+    roles:
+      - Conceptualization
+      - Investigation
+      - Methodology
+      - Formal analysis
+      - Visualization
+      - Software
+      - Data curation
+      - Writing - Original Draft
+      - Writing – Review & Editing 
+    affiliations:
+      - 'School of Earth and Sustainability, Northern Arizona University, Flagstaff, AZ, USA'
+
+  - name: Abraham Springer
+    orcid: 0000-0003-4826-9124
+    corresponding: false
+    roles:
+      - Conceptualization
+      - Funding acquisition
+      - Project administration
+      - Resources
+      - Supervision
+      - Writing – Review & Editing 
+    affiliations:
+      - 'School of Earth and Sustainability, Northern Arizona University, Flagstaff, AZ, USA'
+
+  - name: Neha Gupta
+    orcid: 0000-0002-2247-3509
+    corresponding: false
+    roles: 
+      - Conceptualization
+      - Funding acquisition
+      - Project administration
+      - Resources
+      - Supervision
+      - Writing – Review & Editing 
+    affiliations:
+      - 'Department of Hydrology and Atmospheric Sciences, University of Arizona, Tucson, AZ, USA'
+
+  - name: Temuulen Sankey
+    orcid: 0000-0002-6919-8428
+    corresponding: false
+    roles: 
+      - Writing – Review & Editing 
+    affiliations:
+      - 'School of Informatics, Computing, and Cyber Systems, Northern Arizona University, Flagstaff, AZ, USA'
+
+  - name: Benjamin Lucas
+    orcid: 0000-0002-2021-3076
+    corresponding: false
+    roles: 
+      - Conceptualization
+      - Resources
+      - Supervision
+      - Software
+      - Writing – Review & Editing 
+    affiliations:
+      - 'Department of Mathematics and Statistics, Northern Arizona University, Flagstaff, AZ, USA'
+abstract: |
+  Base flow, sustained by groundwater discharge, is a vital component of river ecosystems, particularly in drylands, where water resources are limited.
+This study analyzes the instrumented streamflow record in Arizona to assess long-term base-flow index (BFI) trends across gauged catchments.
+Results indicate that approximately 32% of Arizona’s streamflow originates from groundwater discharge, with significant spatial variability driven by landscape and climatic factors.
+Base flow  relationships are analyzed with coincident trends in climate variables such as precipitation, evapotranspiration, and temperature.
+Spatial and climatic trends reveal  variability in base-flow contributions, providing insight into groundwater-surface water interactions in arid and semi-arid landscapes.
+Building on this analysis, we applied machine learning methods to predict BFI in ungauged basins, addressing the challenges of Arizona’s sparse streamgage network.
+Using the eXtreme Gradient Boosting (XGBoost) algorithm trained on hydroclimate and physiographic predictors, we estimated long-term BFI from 1991 to 2020.
+This combined approach integrates observational data with predictive modeling to enhance our understanding of base-flow processes and provide a framework for water resource management in data-limited regions.
+
+bibliography: references.bib
+---
+
+### Key Points
+
 -   Approximately 32% of Arizona's streamflow originates from groundwater discharge, with significant spatial and climatic variability.
@@
 -   Base-flow trends align most closely with precipitation changes, emphasizing the role of climate in shaping groundwater-surface water dynamics in dryland regions.
-bibliography: references.bib
----
 
@@
 This water scarcity is intensifying due to the compounding effects of climate variability and increased groundwater extraction [@taylor2013].
-In drylands, groundwater serves as a vital resource for sustaining ecosystems and meeting human needs [@scanlon2006; @yao2018].
+Groundwater serves as a vital resource in drylands for sustaining ecological functions and supporting human livelihoods [@scanlon2006; @yao2018].
 
@@
 @ahiablame2013 found that using a regression model to estimate annual base flow of ungauged catchments was reasonably easy and accurate.
-@beck2013 overcame the nonlinearity of basin characteristics and improved results of multivariate analyses by using artificial neural networks to estimate BFI globally.
+@beck2013 overcame the nonlinearity of basin characteristics and improved results of multivariate analyses by using artificial neural networks (ANN) to estimate BFI globally.
 @singh2018 implemented a random forest algorithm to predict long-term BFI for ungauged catchments across New Zealand.
@@
 
-This study develops a technique for estimating BFI in ungauged basins across Arizona and evaluates the state’s long-term BFI.
-Regional trends in base flow and BFI at instrumented sites are also analyzed, with these trends being linked to coincident trends in precipitation, reference evapotranspiration (ET~O~), and temperature over the same periods.
-Using a machine learning model trained on the hydrogeologic characteristics of surface water basins, we estimate the mean BFI for ungauged basins from 1991 to 2020.
-This approach helps address the spatial gaps in the streamgage network, which is relatively sparse across the state.
-The results offer novel insights into low-flow processes in both gauged and ungauged basins, enhancing our understanding of climate controls on consistent flows in Arizona.
-This study contributes to a more comprehensive view of hydrological dynamics in the context of arid and semi-arid landscapes.
+Previous studies have examined base-flow regionalization and synthesis across various spatial scales, from global to continental [@beck2013; @santhi2008; @ayers2022; @singh2018].
+Such large-scale analyses often utilize generalized datasets and methodologies, resulting in limited applicability to regions with unique hydrogeologic and climatic conditions.
+Additionally, global and continental-scale studies tend to rely on streamgage networks that disproportionately represent large perennial rivers and watersheds with dense human populations, while underrepresenting arid and semi-arid regions characterized by non-perennial flow regimes and smaller streams [@krabbenhoft-2022].
+Thus, their effectiveness in accurately capturing groundwater-surface water interactions, particularly in critically water-stressed dryland regions, remains constrained.
+
+Despite advancements, existing large-scale studies have limited applicability to the unique hydrogeologic conditions in arid and semi-arid regions like Arizona, leaving uncertainty regarding groundwater-surface water interactions in these critically water-stressed areas.
+This study addresses this crucial gap by developing a regionally tailored machine learning model specifically designed to estimate BFI for Arizona's ungauged basins.
+Utilizing hydrogeologic characteristics and hydroclimatic data, we estimate long-term mean BFI (1991–2020) in ungauged catchments, filling spatial gaps in the sparse streamgage network.
+Furthermore, regional trends in base flow and BFI at instrumented sites are analyzed alongside coincident climate trends in precipitation, reference evapotranspiration (ETO), and temperature.
+The outcomes offer novel insights into groundwater contributions to streamflow, supporting water resource management in an area increasingly vulnerable to drought and climate change impacts, and contributing broadly to the understanding of hydrologic processes in dryland environments.
 
@@
 
-While our data selection ensures a robust record for this analysis, it also highlights gaps in spatial coverage that the machine learning model can address.
-Arizona has 184 active USGS streamflow stations (as of 2024) across an area of 295,253 km², whereas Indiana (a humid state in the U.S.), for example, has 189 active stations within an area of 94,326 km².
-This translates to a streamgage density of approximately 2.004 gages per 1,000 km² in Indiana, over three times higher than Arizona's density of 0.623 gages per 1,000 km².
-This stark difference underscores the relative sparsity of Arizona's streamgage network, particularly in the context of its larger geographic area and the unique hydrological challenges posed by its arid and semi-arid landscapes.
-<!--# Add something here to discuss the application in broader semi-arid low streamgage regions globally -->
+Our data selection criteria ensure a robust analysis but also highlight notable spatial gaps that our machine learning model can address.
+Arizona has 184 active USGS streamflow stations (as of 2024) covering an area of 295,253 km².
+For comparison, Indiana (a humid state in the U.S.)maintains 189 active stations within a significantly smaller area of 94,326 km².
+This results in a streamgage density of approximately 2.004 gauges per 1,000 km² in Indiana, more than three times greater than Arizona's density of 0.623 gauges per 1,000 km².
+Such disparities in gauge coverage are typical for dryland regions globally [@krabbenhoft-2022], underscoring the necessity and relevance of this type of modeling approach in arid and semi-arid environments.
 
@@
 
+### Observed BFI in Gauged Catchments
+
+The long-term BFI for the 205 gauged reaches across Arizona is illustrated in @fig-instrumented-bfi .
+The long-term mean BFI is 0.32, indicating that \~32% of long-term streamflow in Arizona likely originates from groundwater discharge and other delayed sources.
+The highest BFI values (\>0.9) are found along the Grand Canyon in northwestern Arizona, where highly karstic geology facilitates rapid subsurface flow to surface water and spring outlets [@chambless_deep-karst_2023].
+High BFI values (\>0.8) are also found at the spring-fed headwaters of the Verde River and Fossil Creek.
+
+The stream reaches of the Little Colorado River Basin (northeastern Arizona) indicate consistently low BFI values (\< 0.2).
+This is likely due to low-yielding perched aquifers underlying the Defiance Plateau in northeastern Arizona, which are hydrologically connected to surface streams, while the high-yield, confined regional aquifer is much deeper [@blanchard02].
+
+![Long-term BFI for the period of record from instrumented stream flow data.](images/BFI_Instrumented_20241203.png){#fig-instrumented-bfi}
+
+#### Trends in Base Flow and BFI {#sec-bfi-trends}
+
+Trends in BFI over the period of record for each streamgage are illustrated in @fig-instrumented-trend and summarized in @tbl-trends.
+Base flow and BFI trends were analyzed across all instrumented sites over their respective periods of record using the Mann–Kendall test.
+Statistically significant trends were observed in both metrics, with a 72.2% coincidence rate between significant base flow and BFI trends, indicating a strong dependence of BFI on base-flow dynamics.
+
+@fig-instrumented-trend illustrates the spatial variation in BFI trends across the study area.
+Statistically significant decreasing trends are observed at 16.1% of sites, while increasing trends are found at 8.8% of sites.
+No consistent regional patterns are evident.
+
+![Trends in BFI over full period of record for instrumented sites used in this study.
+Red upward (blue downward) arrows indicate an increasing (decreasing) trend at a significance level of 5%.
+White circles represent sites with no statistically significant trends.](images/BFITrends_20241203.png){#fig-instrumented-trend}
+
+#### Classification Trends
+
+Classifications presented in @tbl-trends were determined based on precipitation regime, physiographic region, climate, and slope.
+The dominant precipitation regime (monsoon vs.
+snowmelt) was identified by analyzing streamflow hydrographs for each station, focusing on peak flow periods during the monsoon season (July–September) and the snowmelt season (March–June).
+Physiographic region was assigned based on which region the streamgage is located.
+Climate classifications were defined as warm (above the long-term median temperature of Arizona), cool (below the long-term median temperature), wet (above the long-term median precipitation), and dry (below the long-term median precipitation).
+Slope was categorized as high (above the median slope) and low (below the median slope).
+
+Statistically significant decreasing trends in BFI were more common than increasing trends across all site classifications (@tbl-trends).
+While decreasing trends dominate, both increasing and decreasing trends are observed within each classification.
+Monsoon-dominated regions exhibit a higher proportion of significant negative trends (24.1%) compared to snowmelt-dominated regions (10.2%), suggesting that monsoon-dominated systems are more consistently correlated with declining base flow.
+Among climate classifications, warm-dry climates have the highest proportion of negative trends (20.0%), followed by warm-wet climates (19.4%), indicating that regions with higher temperatures are more prone to base flow declines.
+Low-slope regions show a greater prevalence of negative trends (20.4%) compared to high-slope regions (11.8%).
+This suggests that flatter areas may be more susceptible to base-flow reductions, potentially due to differences in hydrologic connectivity and recharge dynamics.
+
+| **Classification Group** | **n** | **n_pos** | **n_neg** | **perc_pos** | **perc_neg** |
+|-------------|------------|------------|------------|------------|------------|
+| Precipitation - Monsoon Dominated | 87 | 8 | 21 | 0.092 | 0.241 |
+| Precipitation - Snowmelt Dominated | 118 | 9 | 12 | 0.076 | 0.102 |
+| Physiographic Region - Basin and Range | 156 | 14 | 26 | 0.090 | 0.167 |
+| Physiographic Region - Colorado Plateau | 49 | 3 | 7 | 0.061 | 0.143 |
+| Climate - Warm-Wet | 31 | 2 | 6 | 0.065 | 0.194 |
+| Climate - Warm-Dry | 55 | 6 | 11 | 0.109 | 0.200 |
+| Climate - Cool-Wet | 74 | 4 | 9 | 0.054 | 0.122 |
+| Climate - Cool-Dry | 45 | 5 | 7 | 0.111 | 0.156 |
+| Slope - High | 102 | 10 | 12 | 0.098 | 0.118 |
+| Slope - Low | 103 | 7 | 21 | 0.068 | 0.204 |
+
+: Comparison of trends for BFI for all sites split by various classifications.
+Only sites with a significant ($\rho \le 0.05$) trend are included here as established by a Mann-Kendall test for monotonic trends across the full period of record.
+n is the number of sites, n_pos (n_neg) is the number of sites with positive (negative) trends, perc_pos (perc_neg) is the percentage of n with a positive (negative) trend.
+{#tbl-trends}
+
+#### Coincident Climate Trends
+
+Coincident trends are defined as those in which base flow or BFI trends match in direction with climate variable trends over the same period.
+Results are summarized in @tbl-coincident-trends.
+This analysis includes both significant and non-significant trends, which is appropriate where the influence of complex, interconnected processes may not always manifest as statistically significant patterns over limited observational periods [@ficklin2016].
+Coincidence was highest with precipitation (64.88% for base flow; 53.17% for BFI), followed by ETO and then temperature.
+Temperature trends most often show an inverse relationship—i.e., warming associated with declining BFI and base flow.
+
+Given the variations in the period of record across the instrumented network (see Figure S2), we analyzed trends in base flow and BFI in relation to coincident trends in climate variables.
+Trends are classified as coincident when the direction of the climate variable trend (positive or negative) aligns with the trend observed in base flow or BFI (@tbl-coincident-trends).
+
+|           | **Climate Variable** | **Coincidence Percentage (%)** |
+|-----------|:--------------------:|:------------------------------:|
+| BFI       |        ET~O~         |             44.88              |
+|           |    Precipitation     |             53.17              |
+|           |     Temperature      |             47.32              |
+| Base Flow |        ET~O~         |             55.12              |
+|           |    Precipitation     |             64.88              |
+|           |     Temperature      |             39.02              |
+
+: Coincident trends of climate variables (ET~O~, precipitation, temperature) with base flow and BFI trends. {#tbl-coincident-trends}
+
 ### BFI of Ungauged Catchments
@@
 
-### Observed BFI in Gauged Catchments
-
-The long-term BFI for the 205 gauged reaches across Arizona is illustrated in @fig-instrumented-bfi .
-The long-term mean BFI is 0.32, indicating that \~32% of long-term streamflow in Arizona likely originates from groundwater discharge and other delayed sources.
-The highest BFI values (\>0.9) are found along the Grand Canyon in northwestern Arizona, where highly karstic geology facilitates rapid subsurface flow to surface water and spring outlets [@chambless_deep-karst_2023].
-High BFI values (\>0.8) are also found at the spring-fed headwaters of the Verde River and Fossil Creek.
-
-The stream reaches of the Little Colorado River Basin (northeastern Arizona) indicate consistently low BFI values (\< 0.2).
-This is likely due to low-yielding perched aquifers underlying the Defiance Plateau in northeastern Arizona, which are hydrologically connected to surface streams, while the high-yield, confined regional aquifer is much deeper [@blanchard02].
-
-![Long-term BFI for the period of record from instrumented stream flow data.](images/BFI_Instrumented_20241203.png){#fig-instrumented-bfi}
-
-#### Trends in Base Flow and BFI {#sec-bfi-trends}
-
-Trends in BFI over the period of record for each streamgage are illustrated in @fig-instrumented-trend and summarized in @tbl-trends.
-Base flow and BFI trends were analyzed across all instrumented sites over their respective periods of record using the Mann–Kendall test.
-Statistically significant trends were observed in both metrics, with a 72.2% coincidence rate between significant base flow and BFI trends, indicating a strong dependence of BFI on base-flow dynamics.
-
-@fig-instrumented-trend illustrates the spatial variation in BFI trends across the study area.
-Statistically significant decreasing trends are observed at 16.1% of sites, while increasing trends are found at 8.8% of sites.
-No consistent regional patterns are evident.
-
-![Trends in BFI over full period of record for instrumented sites used in this study.
-Red upward (blue downward) arrows indicate an increasing (decreasing) trend at a significance level of 5%.
-White circles represent sites with no statistically significant trends.](images/BFITrends_20241203.png){#fig-instrumented-trend}
-
-#### Classification Trends
-
-Classifications presented in @tbl-trends were determined based on precipitation regime, physiographic region, climate, and slope.
-The dominant precipitation regime (monsoon vs.
-snowmelt) was identified by analyzing streamflow hydrographs for each station, focusing on peak flow periods during the monsoon season (July–September) and the snowmelt season (March–June).
-Physiographic region was assigned based on which region the streamgage is located.
-Climate classifications were defined as warm (above the long-term median temperature of Arizona), cool (below the long-term median temperature), wet (above the long-term median precipitation), and dry (below the long-term median precipitation).
-Slope was categorized as high (above the median slope) and low (below the median slope).
-
-Statistically significant decreasing trends in BFI were more common than increasing trends across all site classifications (@tbl-trends).
-While decreasing trends dominate, both increasing and decreasing trends are observed within each classification.
-Monsoon-dominated regions exhibit a higher proportion of significant negative trends (24.1%) compared to snowmelt-dominated regions (10.2%), suggesting that monsoon-dominated systems are more consistently correlated with declining base flow.
-Among climate classifications, warm-dry climates have the highest proportion of negative trends (20.0%), followed by warm-wet climates (19.4%), indicating that regions with higher temperatures are more prone to base flow declines.
-Low-slope regions show a greater prevalence of negative trends (20.4%) compared to high-slope regions (11.8%).
-This suggests that flatter areas may be more susceptible to base-flow reductions, potentially due to differences in hydrologic connectivity and recharge dynamics.
-
-| **Classification Group** | **n** | **n_pos** | **n_neg** | **perc_pos** | **perc_neg** |
-|------------|------------|------------|------------|------------|------------|
-| Precipitation - Monsoon Dominated | 87 | 8 | 21 | 0.092 | 0.241 |
-| Precipitation - Snowmelt Dominated | 118 | 9 | 12 | 0.076 | 0.102 |
-| Physiographic Region - Basin and Range | 156 | 14 | 26 | 0.090 | 0.167 |
-| Physiographic Region - Colorado Plateau | 49 | 3 | 7 | 0.061 | 0.143 |
-| Climate - Warm-Wet | 31 | 2 | 6 | 0.065 | 0.194 |
-| Climate - Warm-Dry | 55 | 6 | 11 | 0.109 | 0.200 |
-| Climate - Cool-Wet | 74 | 4 | 9 | 0.054 | 0.122 |
-| Climate - Cool-Dry | 45 | 5 | 7 | 0.111 | 0.156 |
-| Slope - High | 102 | 10 | 12 | 0.098 | 0.118 |
-| Slope - Low | 103 | 7 | 21 | 0.068 | 0.204 |
-
-: Comparison of trends for BFI for all sites split by various classifications.
-Only sites with a significant ($\rho \le 0.05$) trend are included here as established by a Mann-Kendall test for monotonic trends across the full period of record.
-n is the number of sites, n_pos (n_neg) is the number of sites with positive (negative) trends, perc_pos (perc_neg) is the percentage of n with a positive (negative) trend.
-{#tbl-trends}
-
-#### Coincident Climate Trends
-
-Coincident trends are defined as those in which base flow or BFI trends match in direction with climate variable trends over the same period.
-Results are summarized in @tbl-coincident-trends.
-This analysis includes both significant and non-significant trends, which is appropriate where the influence of complex, interconnected processes may not always manifest as statistically significant patterns over limited observational periods [@ficklin2016].
-Coincidence was highest with precipitation (64.88% for base flow; 53.17% for BFI), followed by ETO and then temperature.
-Temperature trends most often show an inverse relationship—i.e., warming associated with declining BFI and base flow.
-
-Given the variations in the period of record across the instrumented network (see Figure S2), we analyzed trends in base flow and BFI in relation to coincident trends in climate variables.
-Trends are classified as coincident when the direction of the climate variable trend (positive or negative) aligns with the trend observed in base flow or BFI (@tbl-coincident-trends).
-
-|           | **Climate Variable** | **Coincidence Percentage (%)** |
-|-----------|:--------------------:|:------------------------------:|
-| BFI       |        ET~O~         |             44.88              |
-|           |    Precipitation     |             53.17              |
-|           |     Temperature      |             47.32              |
-| Base Flow |        ET~O~         |             55.12              |
-|           |    Precipitation     |             64.88              |
-|           |     Temperature      |             39.02              |
-
-: Coincident trends of climate variables (ET~O~, precipitation, temperature) with base flow and BFI trends. {#tbl-coincident-trends}
-
 ## Discussion
 
-### Model Performance and Implications
+Previous studies have examined base-flow regionalization and synthesis across various spatial scales, some overlapping with our study area.
+@beck2013 explored global patterns of BFI using streamflow observations extrapolated through an artificial neural network (ANN).
+The U.S.
+streamgages included in that analysis were drawn from the interim Model Parameter Estimation Experiment (MOPEX) dataset of 1862 USGS gauges rather than the finalized dataset of 431 gauges meeting precipitation gauge requirements [@beck2013; @mopex_2006].
+Only 10 of these interim gauges overlap geographically with our study area.
+Similarly, @ayers2022 assessed climatic influences on monthly baseflow across the continental United States (CONUS), utilizing a subset of GAGES-II streamflow gauges.
+Although exact gauge locations from that study are unavailable, published figures suggest minimal geographic overlap with our research domain.
+@santhi2008 employed interpolation methods for regional base-flow estimation across CONUS, but specific gauge locations or data are not publicly available, limiting direct regional comparisons.
 
-The regional model demonstrated strong predictive capability (overall R²=0.764) but showed systematic underprediction (negative pbias).
-The systematic bias suggests potential limitations in the model or underlying predictor data, particularly in monsoon-dominated and warm-dry climates.
-The performance differences across climatic and physiographic classifications highlight the need for targeted improvements, such as integrating additional predictors or refining the representation of regional hydrological processes.
-<!--# Discuss that we tested 3 models: regional (by phys region), full predictor set, and trimmed set.
-Trimmed set was the best performing, indicating that the regional separation isn't as beneficial as a more robust predictor dataset --> In the development of the final XGBoost model, we tested making two regional models (one for each physiographic region) to ensure best predictions.
-We found that the state-wide model outperformed the regional model, indicating that regional separation isn't as beneficial as a larger, state-wide dataset.
+The utility of existing streamgage networks for base-flow analyses in Arizona is limited by their design and monitoring objectives.
+While the number of streamgages has increased to meet regulatory imperatives, such as those driven by the Clean Water Act, many gauges prioritize peak flow monitoring rather than low-flow conditions critical for base-flow studies [@flood-control2020].
+Newer non-USGS gauges, such as those installed by flood control districts (e.g., the ALERT system), are generally tailored for flood detection rather than monitoring sustained base flow.
+Recent USGS installations focus on addressing in-stream flow rights and future monitoring improvements, though low-flow capabilities remain limited.
 
-### Interpretation of Predictor Variables
+This study addresses these monitoring gaps through a detailed analysis of observed base flow and BFI at instrumented streamgages, highlighting groundwater-surface water interactions within Arizona’s arid and semi-arid landscapes.
+Observed spatial patterns in BFI clearly align with known hydrogeological and climatic variations.
+High BFI values coincide with areas characterized by spring-fed streams and pronounced physiographic transitions, such as the Grand Canyon and Mogollon Rim.
+Conversely, lower observed BFI values correspond to regions with limited groundwater contributions, notably the Defiance Plateau and southern Arizona, emphasizing vulnerability in these arid environments.
 
-Elevation was found to be a critical predictor, reinforcing its effectiveness as a proxy for climate conditions influencing groundwater recharge.
-This finding agrees with previous studies across varied landscapes and scales.
-@singh2018 indicated that mean basin elevation was the most imporant predictor in their random forest model predicting BFI nationally in New Zealand.
-@santhi2008 found that relief (basin max elevation minus minimum elevation) was a key parameter in the best fit of their stepwise regression across hydrologic landscape regions in the conterminous US.
-@beck2013 indicated that basin elevation was the 3rd most important predictor of their artificial neural network in predicting BFI globally.
+Analysis of trends at instrumented sites revealed a predominance of declining BFI values, particularly within monsoon-dominated precipitation regimes, warm-dry climates, and low-slope basins.
+These negative trends likely reflect intensified climate stress, highlighting areas most vulnerable to groundwater depletion and surface-water scarcity.
+Precipitation was identified as the primary driver influencing base-flow variability, with temperature and evapotranspiration trends adding complexity.
+The weaker correlation between temperature and BFI trends suggests that temperature impacts are moderated by other environmental factors, underscoring the need for integrated climate and land-management strategies to protect groundwater-dependent ecosystems.
 
-The mixed influences of Soil Type A may indicate complexities in how infiltration processes are represented in the model or variability in hydrologic conditions within similar soil classes.
-Future research could benefit from explicit inclusion of temporal variables such as snowpack duration, soil moisture storage, and finer-scale land cover attributes.
+Given the limited coverage provided by existing gauges, a common issue in arid and semi-arid regions globally [@krabbenhoft-2022], we utilized a machine learning approach to estimate BFI in ungauged catchments.
+The regional model exhibited strong predictive capability (overall R²=0.764), though systematic underprediction (negaitve pbias) was noted, particularly in monsoon-dominated and warm-dry climates.
+These biases suggest areas for improvement in model performance, potentially through the incorporation of additional predictors or more refined regional hydrologic representations.
+Although creating separate regional models (by physiographic region) was explored to enhance predictive accuracy, the statewide model consistently outperformed these regionalized approaches, emphasizing the advantage of a comprehensive, statewide dataset.
 
-### Spatial Patterns of BFI
-
-The spatial distribution of predicted and observed BFI reflects known hydrogeological and climatic conditions.
-High BFI values correlate strongly with spring-fed streams and physiographic transitions, while low values align with arid regions.
-This spatial consistency validates the model’s utility for regional water resource assessments and highlights areas vulnerable to declining baseflows.
-
-### Trends and Climate Implications
-
-The prevalence of negative BFI trends, especially in warm-dry and low-slope regions, suggests that climate warming and reduced precipitation could disproportionately impact baseflow.
-The relatively weaker alignment of temperature trends with BFI suggests that temperature impacts are mediated through complex processes, potentially offset by concurrent precipitation or land-use changes.
-This complexity emphasizes the importance of integrated climate and land-management strategies to sustain groundwater-dependent ecosystems.
+Elevation emerged as the most significant predictor, reinforcing its established role as a proxy for climate conditions influencing groundwater recharge.
+This aligns with findings from previous studies conducted in diverse landscapes, including New Zealand [@singh2018] and across the conterminous U.S.
+[@santhi2008; @beck2013].
+The mixed influence observed for Soil Type A indicates complexity in infiltration dynamics, suggesting that future research could benefit from explicitly incorporating temporal variables such as snowpack duration, soil moisture, and detailed land-cover characteristics.
 
@@
 
--   The long-term average BFI across Arizona is \~0.32, with strong spatial variability tied to hydrogeology and climate.
--   The XGBoost model predicts BFI well (R² = 0.764), using elevation, land cover, and soil type as key predictors.
--   High BFI regions include the Grand Canyon and Mogollon Rim; low BFI areas include the Defiance Plateau and southern Arizona.
--   Decreasing BFI trends are more prevalent than increasing ones, especially in monsoon-dominated, warm-dry, and low-slope regions.
--   Precipitation is the dominant driver of BFI and base-flow variability, while ETO and temperature trends add complexity.
--   The streamgage network in Arizona has limitations for base-flow monitoring; future design should include low-flow-capable instrumentation.
--   This study demonstrates how machine learning and observational data can be integrated to improve base-flow estimates in data-scarce environments.
--   The modeling framework is applicable to other dryland regions globally and can inform strategies for climate adaptation and water resource management.
+The long-term average BFI across Arizona is approximately 0.32, highlighting that groundwater discharge significantly contributes to surface water flows but exhibits substantial spatial variability.
+Our machine learning model (XGBoost) effectively predicts BFI in ungauged catchments (R² = 0.764), primarily driven by basin elevation, land cover, and hydrologic soil characteristics.
+Regions such as the Grand Canyon and Mogollon Rim exhibit notably high BFI values, signifying robust groundwater-surface water interactions, whereas areas such as the Defiance Plateau and southern Arizona show consistently lower BFI values, reflecting limited groundwater contributions.
+
+Analysis of temporal trends indicates a prevailing pattern of declining base flow and BFI across the region, particularly in monsoon-dominated, warm-dry, and low-slope areas.
+Precipitation emerged as the strongest driver of base-flow variability, while temperature and evapotranspiration added additional layers of complexity.
+The observed trends underscore the sensitivity of Arizona's hydrological systems to climatic variability and suggest increasing vulnerability under projected climate change scenarios.
+
+This research highlights critical limitations in Arizona’s current streamgage network for monitoring base-flow processes, suggesting a need for targeted investments in instrumentation capable of capturing low-flow dynamics.
+Additionally, the demonstrated success of our modeling framework underscores its broader applicability to similarly data-scarce dryland regions globally, providing a valuable tool for water resource management and climate adaptation strategies.
+By improving our understanding of groundwater-surface water interactions under changing climatic conditions, this study contributes substantially to addressing hydrological uncertainties in arid and semi-arid environments.
+
+## References

### 2025-03-26 elsevier_MANUSCRIPT.qmd
updated

@@
 | **Classification Group** | **n** | **n_pos** | **n_neg** | **perc_pos** | **perc_neg** |
-|-------------|------------|------------|------------|------------|------------|
+|---------------------|----------|----------|----------|----------|----------|
 | Precipitation - Monsoon Dominated | 87 | 8 | 21 | 0.092 | 0.241 |
@@
 | **Classification Group** | **n** | **R2** | **MSE** | **RMSE** | **MAE** | **NSE** | **pbias** |
-|---------|---------|---------|---------|---------|---------|---------|---------|
+|-------------|---------|---------|---------|---------|---------|---------|---------|
 | Climate - Monsoon Dominated | 3039 | 0.633 | 0.016 | 0.126 | 0.074 | 0.619 | -13.7 |
@@
 The U.S.
-streamgages included in that analysis were drawn from the interim Model Parameter Estimation Experiment (MOPEX) dataset of 1862 USGS gauges rather than the finalized dataset of 431 gauges meeting precipitation gauge requirements [@beck2013; @mopex_2006].
+streamgages included in that analysis were drawn from the interim Model Parameter Estimation Experiment (MOPEX) dataset of 1862 USGS gauges [@beck2013; @mopex_2006].
 Only 10 of these interim gauges overlap geographically with our study area.
@@
 Although exact gauge locations from that study are unavailable, published figures suggest minimal geographic overlap with our research domain.
-@santhi2008 employed interpolation methods for regional base-flow estimation across CONUS, but specific gauge locations or data are not publicly available, limiting direct regional comparisons.
+@santhi2008 employed interpolation methods for regional base-flow estimation across CONUS, but specific gauge locations or data are not available, limiting direct regional comparisons.
 
-The utility of existing streamgage networks for base-flow analyses in Arizona is limited by their design and monitoring objectives.
+The existing streamgage networks Arizona are limited for base-flow analyses by their design and monitoring objectives.
 While the number of streamgages has increased to meet regulatory imperatives, such as those driven by the Clean Water Act, many gauges prioritize peak flow monitoring rather than low-flow conditions critical for base-flow studies [@flood-control2020].
@@
 
-This study addresses these monitoring gaps through a detailed analysis of observed base flow and BFI at instrumented streamgages, highlighting groundwater-surface water interactions within Arizona’s arid and semi-arid landscapes.
+This study addresses these monitoring gaps through an analysis of observed base flow and BFI at instrumented streamgages, highlighting groundwater-surface water interactions within Arizona’s arid and semi-arid landscapes.
 Observed spatial patterns in BFI clearly align with known hydrogeological and climatic variations.
@@
 These biases suggest areas for improvement in model performance, potentially through the incorporation of additional predictors or more refined regional hydrologic representations.
-Although creating separate regional models (by physiographic region) was explored to enhance predictive accuracy, the statewide model consistently outperformed these regionalized approaches, emphasizing the advantage of a comprehensive, statewide dataset.
+Although creating separate regional models (by physiographic region) was explored to enhance predictive accuracy, the state-wide model consistently outperformed these regionalized approaches, emphasizing the advantage of a comprehensive, state-wide dataset.
 
 Elevation emerged as the most significant predictor, reinforcing its established role as a proxy for climate conditions influencing groundwater recharge.
-This aligns with findings from previous studies conducted in diverse landscapes, including New Zealand [@singh2018] and across the conterminous U.S.
-[@santhi2008; @beck2013].
+This aligns with findings from previous studies conducted in diverse landscapes, including New Zealand [@singh2018], across CONUS [@santhi2008], and globally [@beck2013].
 The mixed influence observed for Soil Type A indicates complexity in infiltration dynamics, suggesting that future research could benefit from explicitly incorporating temporal variables such as snowpack duration, soil moisture, and detailed land-cover characteristics.
@@
 This research highlights critical limitations in Arizona’s current streamgage network for monitoring base-flow processes, suggesting a need for targeted investments in instrumentation capable of capturing low-flow dynamics.
-Additionally, the demonstrated success of our modeling framework underscores its broader applicability to similarly data-scarce dryland regions globally, providing a valuable tool for water resource management and climate adaptation strategies.
+Additionally, the demonstrated success of our modeling framework accentuates its broader applicability to similarly data-scarce dryland regions globally, providing a valuable tool for water resource management and climate adaptation strategies.
 By improving our understanding of groundwater-surface water interactions under changing climatic conditions, this study contributes substantially to addressing hydrological uncertainties in arid and semi-arid environments.

### 2025-03-31 elsevier_MANUSCRIPT.qmd
aes edits

@@
   Base flow, sustained by groundwater discharge, is a vital component of river ecosystems, particularly in drylands, where water resources are limited.
-This study analyzes the instrumented streamflow record in Arizona to assess long-term base-flow index (BFI) trends across gauged catchments.
+This novel study analyzes the instrumented streamflow record in Arizona to assess long-term base-flow index (BFI) trends across gauged catchments.
 Results indicate that approximately 32% of Arizona’s streamflow originates from groundwater discharge, with significant spatial variability driven by landscape and climatic factors.
@@
 -   Approximately 32% of Arizona's streamflow originates from groundwater discharge, with significant spatial and climatic variability.
--   A machine learning model (XGBoost) successfully predicted long-term BFI in ungauged basins, addressing gaps in a sparse streamgage network.
+-   A machine learning model (XGBoost) successfully predicted long-term BFI in ungauged basins, addressing gaps in a sparse low-flow streamgage network, typical of drylands.
 -   Base-flow trends align most closely with precipitation changes, emphasizing the role of climate in shaping groundwater-surface water dynamics in dryland regions.
@@
 Such large-scale analyses often utilize generalized datasets and methodologies, resulting in limited applicability to regions with unique hydrogeologic and climatic conditions.
-Additionally, global and continental-scale studies tend to rely on streamgage networks that disproportionately represent large perennial rivers and watersheds with dense human populations, while underrepresenting arid and semi-arid regions characterized by non-perennial flow regimes and smaller streams [@krabbenhoft-2022].
+Additionally, global and continental-scale studies tend to rely on streamgage networks that disproportionately represent large perennial rivers and regulated watersheds with dense human populations, while underrepresenting arid and semi-arid regions characterized by non-perennial flow regimes and smaller streams [@krabbenhoft-2022].
 Thus, their effectiveness in accurately capturing groundwater-surface water interactions, particularly in critically water-stressed dryland regions, remains constrained.
@@
 Arizona has 184 active USGS streamflow stations (as of 2024) covering an area of 295,253 km².
-For comparison, Indiana (a humid state in the U.S.)maintains 189 active stations within a significantly smaller area of 94,326 km².
+For comparison, Indiana (a humid state in the U.S.) maintains 189 active stations within a significantly smaller area of 94,326 km².
 This results in a streamgage density of approximately 2.004 gauges per 1,000 km² in Indiana, more than three times greater than Arizona's density of 0.623 gauges per 1,000 km².
@@
 Various geospatial variables, such as aspect, were disaggregated then averaged to assess the areal percentage of each sub-variable within individual HUC 8 basins.
-By calculating these percentages, we aimed to get a more comprehensive understanding of landscape composition across space.
+By calculating these percentages, we derived a more comprehensive understanding of landscape composition across space.
 Land cover was acquired from USGS-NLCD (National Land Cover Database), hydrologic soil group from SSURGO (Soil Survey Geographic Database), and underlying geology and karst from USGS were all similarly averaged across the basins.
@@
 The hyperparameters evaluated included the learning rate ($\eta$), minimum split loss ($\gamma$), maximum tree depth, minimum child weight, and the number of trees.
-While not an exhaustive list of all possible XGBoost hyperparameters, the range of values explored provided sufficient variation to ensure the selection of a high-performing model.
+While not an exhaustive list of all possible XGBoost hyperparameters, this range of values provided sufficient variation to ensure the selection of a high-performing model.
 The optimal hyperparameters were determined to be: 700 trees, a learning rate ($\eta$) of 0.05, a minimum split loss ($\gamma$) of 0.075, a maximum tree depth of 7, and a minimum child weight of 5.
@@
 
-To check for autocorrelation, we applied the Durbin-Watson test, which revealed significant autocorrelation at four streamgages on an annual basis.
+To check for autocorrelation, we applied the Durbin-Watson test [@durbin1950testing], which revealed significant autocorrelation at four streamgages on an annual basis.
 Of these, only one streamgage (09486500 - Santa Cruz River at Cortaro, AZ) showed a significant trend in BFI.
@@
 The highest BFI values (\>0.9) are found along the Grand Canyon in northwestern Arizona, where highly karstic geology facilitates rapid subsurface flow to surface water and spring outlets [@chambless_deep-karst_2023].
-High BFI values (\>0.8) are also found at the spring-fed headwaters of the Verde River and Fossil Creek.
+High BFI values (\>0.8) are also found at the spring-fed headwaters of the Verde River and Fossil Creek, which have similar highly-karstic, snowmelt-driven recharge areas.
 
@@
 | **Classification Group** | **n** | **n_pos** | **n_neg** | **perc_pos** | **perc_neg** |
-|---------------------|----------|----------|----------|----------|----------|
+|---------------|------------|------------|------------|------------|------------|
 | Precipitation - Monsoon Dominated | 87 | 8 | 21 | 0.092 | 0.241 |
@@
 | **Classification Group** | **n** | **R2** | **MSE** | **RMSE** | **MAE** | **NSE** | **pbias** |
-|-------------|---------|---------|---------|---------|---------|---------|---------|
+|---------|---------|---------|---------|---------|---------|---------|---------|
 | Climate - Monsoon Dominated | 3039 | 0.633 | 0.016 | 0.126 | 0.074 | 0.619 | -13.7 |
@@
 Only 10 of these interim gauges overlap geographically with our study area.
-Similarly, @ayers2022 assessed climatic influences on monthly baseflow across the continental United States (CONUS), utilizing a subset of GAGES-II streamflow gauges.
+Similarly, @ayers2022 assessed climatic influences on monthly base flow across the continental United States (CONUS), utilizing a subset of GAGES-II streamflow gauges.
 Although exact gauge locations from that study are unavailable, published figures suggest minimal geographic overlap with our research domain.
@@
 Precipitation emerged as the strongest driver of base-flow variability, while temperature and evapotranspiration added additional layers of complexity.
-The observed trends underscore the sensitivity of Arizona's hydrological systems to climatic variability and suggest increasing vulnerability under projected climate change scenarios.
+The observed trends underscore the sensitivity of Arizona's hydrological systems to climatic variability and suggest increasing vulnerability under projected climate change scenarios and other stresses to aquifers, such as increased groundwater pumping.
 

### 2025-04-02 elsevier_MANUSCRIPT.qmd
minor edits

@@
 
-  - name: Abraham Springer
+  - name: Abraham E Springer
     orcid: 0000-0003-4826-9124
@@
 Results indicate that approximately 32% of Arizona’s streamflow originates from groundwater discharge, with significant spatial variability driven by landscape and climatic factors.
-Base flow  relationships are analyzed with coincident trends in climate variables such as precipitation, evapotranspiration, and temperature.
+Base-flow  relationships are analyzed with coincident trends in climate variables such as precipitation, evapotranspiration, and temperature.
 Spatial and climatic trends reveal  variability in base-flow contributions, providing insight into groundwater-surface water interactions in arid and semi-arid landscapes.
@@
 ### Study Area {#sec-study-area}
+
+The study area represents a range of heterogeneous climate, elevation,
 
@@
 | **Classification Group** | **n** | **n_pos** | **n_neg** | **perc_pos** | **perc_neg** |
-|---------------|------------|------------|------------|------------|------------|
+|------------|------------|------------|------------|------------|------------|
 | Precipitation - Monsoon Dominated | 87 | 8 | 21 | 0.092 | 0.241 |
@@
 Conversely, lower observed BFI values correspond to regions with limited groundwater contributions, notably the Defiance Plateau and southern Arizona, emphasizing vulnerability in these arid environments.
+A trend of high BFI in upstream reaches to lower BFI in downstream reaches [@fig-instrumented-bfi] likely indicates a transition from zones of groundwater discharge to zones of groundwater recharge and may indicate a transition from gaining stream reaches to losing stream reaches [@winter2007].
 

### 2025-04-04 elsevier_MANUSCRIPT.qmd
updated figures

@@
 
-![Map of Arizona and US Geological Survey (USGS) streamgages used in this study. 8-digit HUC subbasin boundaries and physiographic regions shown.](images/StudyArea_20241203.png){#fig-study-area}
+![Map of Arizona and US Geological Survey (USGS) streamgages used in this study.
+8-digit HUC subbasin boundaries and physiographic regions shown.](images/updated/StudyArea_20250404.png){#fig-study-area}
 
@@
 
-![Long-term BFI for the period of record from instrumented stream flow data.](images/BFI_Instrumented_20241203.png){#fig-instrumented-bfi}
+![Long-term BFI for the period of record from instrumented stream flow data.](images/updated/BFI_Instrumented_20240404.png){#fig-instrumented-bfi}
 
@@
 Red upward (blue downward) arrows indicate an increasing (decreasing) trend at a significance level of 5%.
-White circles represent sites with no statistically significant trends.](images/BFITrends_20241203.png){#fig-instrumented-trend}
+White circles represent sites with no statistically significant trends.](images/updated/BFITrends_20250404.png){#fig-instrumented-trend}
 
@@
 
-![Predicted long-term BFI values for 8-digit HUC (1991-2020)](images/BFI_HUC8_20250129.png){#fig-bfi-huc}
+![Predicted long-term BFI values for 8-digit HUC (1991-2020)](images/updated/BFI_HUC8_20250404.png){#fig-bfi-huc}
 

### 2025-04-06 elsevier_MANUSCRIPT.qmd
prepping for submission

@@
 
--   Approximately 32% of Arizona's streamflow originates from groundwater discharge, with significant spatial and climatic variability.
--   A machine learning model (XGBoost) successfully predicted long-term BFI in ungauged basins, addressing gaps in a sparse low-flow streamgage network, typical of drylands.
--   Base-flow trends align most closely with precipitation changes, emphasizing the role of climate in shaping groundwater-surface water dynamics in dryland regions.
+-   Approximately 32% of Arizona's streamflow originates from groundwater discharge.
+-   XGBoost predicted long-term BFI in ungauged basins, filling dryland data gaps.
+-   Baseflow trends track precipitation, emphasizing climate’s role in dryland hydrology.
 
@@
 
-The study area represents a range of heterogeneous climate, elevation,
-
-The state of Arizona, located in the southwest United States, covers a total area of 295,253 km$^2$.
-Arizona is divided into two primary physiographic provinces: the Colorado Plateau in the northeast, and the Basin-and-Range region in the west and south.
-The Central Highlands is a transition zone consisting of scattered basins separated by the mountainous foothills of the Mogollon Rim.
-The Colorado Plateau is dominated by high-elevation desert with an average elevation of 1,936 masl (6,352 ft).
-Mean temperatures range from -6$^\circ$C (20$^\circ$F) to 26$^\circ$C (80$^\circ$F) and it averages 580 mm (23 in) of precipitation.
-The Basin-and-Range region has a semi-arid to arid climate with an average elevation of 490 masl (1600 ft).
-Average temperature ranges from 15$^\circ$C (60$^\circ$F) to 43$^\circ$C (110$^\circ$F) and the region averages 200 mm (8 in) of precipitation annually [@az_climate2024].
+Arizona, located in the southwestern United States, spans approximately 295,253 km² and encompasses a diverse range of landscapes, elevations, and climate regimes.
+The state includes portions of three major physiographic regions: the Colorado Plateau in the northeast, the Basin and Range province in the south and west, and the Central Highlands, a transitional zone between them.
+This heterogeneity results in substantial variation in environmental conditions across the state.
+The Colorado Plateau is characterized by high-elevation desert and mountain woodlands, averaging 1,936 masl (6,352 ft), with mean temperatures ranging from -6°C (20°F) to 26°C (80°F) and annual precipitation of about 580 mm (23 in).
+In contrast, the Basin and Range region is lower in elevation, averaging 490 masl (1,600 ft), and features a semi-arid to arid climate, with temperatures ranging from 15°C (60°F) to 43°C (110°F) and an average annual precipitation of 200 mm (8 in) [@az_climate2024].
+The Central Highlands feature a mix of mountainous terrain and interspersed basins, adding to the state's topographic and hydroclimatic complexity.
 
@@
 Arizona's hydrology varies seasonally and spatially between its physiographic regions.
-In the summer, localized and intense convective precipitation events are driven by the North American Monsoon, while in the winter, orographic precipitation comes from Pacific frontal systems [@eastoe2019mtnblock].
+In summer, localized and intense convective storms stem from the North American Monsoon; in winter, orographic precipitation is delivered by Pacific frontal systems [@eastoe2019mtnblock].
 While monsoonal precipitation can account for up to 50% of annual precipitation, evaporation and dry preceding soil properties leads to most precipitation becoming runoff [@sheppard2002].
@@
 Much of the hydrology of Arizona is snow-melt derived, driven by spring melt from the high-elevation Colorado Plateau winter snowpack.
-While winter precipitation provides only 30% of annual averages, it provides the majority of water for natural reservoirs [@sheppard2002].
+Though winter precipitation provides only 30% of annual averages, it provides the majority of water for natural reservoirs [@sheppard2002].
 
@@
 Instead of the water year, PRISM data uses a calendar-year format, which was adopted for consistency in the water balance.
-Although this may introduce challenges in the annual estimates due to inter-annual snow storage, the use of long-term annual averages is likely to reduce any potential errors [@reitz2017].
+Although this may introduce challenges in the annual estimates due to inter-annual snow storage, the use of long-term annual averages reduces potential errors [@reitz2017].
 
@@
 ![Trends in BFI over full period of record for instrumented sites used in this study.
-Red upward (blue downward) arrows indicate an increasing (decreasing) trend at a significance level of 5%.
+Red downward (blue upward) arrows indicate a decreasing (increasing) trend at a significance level of 5%.
 White circles represent sites with no statistically significant trends.](images/updated/BFITrends_20250404.png){#fig-instrumented-trend}
@@
 Previous studies have examined base-flow regionalization and synthesis across various spatial scales, some overlapping with our study area.
-@beck2013 explored global patterns of BFI using streamflow observations extrapolated through an artificial neural network (ANN).
+@beck2013 explored global patterns of BFI using streamflow observations extrapolated through an ANN.
 The U.S.

### 2025-04-25 elsevier_MANUSCRIPT.qmd
quick reviewer responses

@@
 Arizona, located in the southwestern United States, spans approximately 295,253 km² and encompasses a diverse range of landscapes, elevations, and climate regimes.
-The state includes portions of three major physiographic regions: the Colorado Plateau in the northeast, the Basin and Range province in the south and west, and the Central Highlands, a transitional zone between them.
+The state includes portions of two major physiographic regions: the Colorado Plateau in the northeast, the Basin and Range province in the south and west, and a transitional zone, the Central Highlands, between them.
 This heterogeneity results in substantial variation in environmental conditions across the state.
@@
 Much of the hydrology of Arizona is snow-melt derived, driven by spring melt from the high-elevation Colorado Plateau winter snowpack.
-Though winter precipitation provides only 30% of annual averages, it provides the majority of water for natural reservoirs [@sheppard2002].
+Though winter precipitation provides only 30% of annual average precipitation, it provides the majority of water for natural reservoirs [@sheppard2002].
 
@@
 Additionally, years with more than 30 missing days of streamflow data were excluded from the analysis.
-This study focuses on natural, streamflow-influencing dynamics, so streamgages affected by regulation or diversions were excluded.
-The regulated river streamgages were identified through annual reports on water data published by the USGS [@usgs2010].
-Furthermore, streamgages along the Colorado River were omitted because they represent managed flows governed by the Colorado River Compact.
+Streamgages directly downstream of major regulation (e.g., reservoirs, lakes, diversions) were excluded, based on USGS annual water data reports and site metadata \[@usgs2010\].
+While some flow alteration is widespread, the focus was on removing gauges with clear, immediate regulatory impacts.
+As such, streamgages along the Colorado River were omitted because they represent managed flows governed by the Colorado River Compact.
 After applying these selection criteria, 205 USGS streamgages with acceptable periods of record were included in the study (@fig-study-area).
@@
 
-A 30-meter resolution Digital Elevation Model (DEM) of Arizona was used to derive key basin characteristics: basin area, average slope, and the proportion of each basin oriented toward north or south aspects.
+Basin elevation was derived from a 30-meter resolution Digital Elevation Model (DEM) from the USGS National Elevation Dataset.
+The DEM of Arizona was used to derive key basin characteristics: basin area, average slope, and the proportion of each basin oriented toward north or south aspects.
 Various geospatial variables, such as aspect, were disaggregated then averaged to assess the areal percentage of each sub-variable within individual HUC 8 basins.
@@
 
-Our initial training dataset comprised 7,724 observations across 45 variables.
+Our initial training dataset comprised 7,724 observations (annual observations of BFI at streamgages) across 45 variables.
 To optimize the model's performance, we first conducted an exhaustive grid search combined with 5-fold cross-validation to identify the optimal hyperparameter values.
@@
 Feature effects provide an indication of the relationship between the value of a predicting feature and its impact on the prediction.
+
+![SHAP value plot of features used in final model.
+The number to the right of the feature name is the mean absolute SHAP value.
+Land cover features are indicated by the percentage of cover by each land cover type and soil types are defined by hydrologic soil group.](images/shap_summary.png){#fig-shap_values}
 
@@
 : Comparison of trends for BFI for all sites split by various classifications.
-Only sites with a significant ($\rho \le 0.05$) trend are included here as established by a Mann-Kendall test for monotonic trends across the full period of record.
+Only sites with a significant ($\alpha \le 0.05$) trend are included here as established by a Mann-Kendall test for monotonic trends across the full period of record.
 n is the number of sites, n_pos (n_neg) is the number of sites with positive (negative) trends, perc_pos (perc_neg) is the percentage of n with a positive (negative) trend.
@@
 
-![SHAP value plot of features used in final model.
-Land cover features are indicated by the percentage of cover by each land cover type and soil types are defined by hydrologic soil group.](images/shap_summary.png){#fig-shap_values}
-
 Land cover and land use predictors also play a significant role in BFI estimation.
@@
 Given the limited coverage provided by existing gauges, a common issue in arid and semi-arid regions globally [@krabbenhoft-2022], we utilized a machine learning approach to estimate BFI in ungauged catchments.
-The regional model exhibited strong predictive capability (overall R²=0.764), though systematic underprediction (negaitve pbias) was noted, particularly in monsoon-dominated and warm-dry climates.
+The regional model exhibited strong predictive capability (overall R²=0.764), though systematic underprediction (negative pbias) was noted, particularly in monsoon-dominated and warm-dry climates.
 These biases suggest areas for improvement in model performance, potentially through the incorporation of additional predictors or more refined regional hydrologic representations.
@@
 Our machine learning model (XGBoost) effectively predicts BFI in ungauged catchments (R² = 0.764), primarily driven by basin elevation, land cover, and hydrologic soil characteristics.
-Regions such as the Grand Canyon and Mogollon Rim exhibit notably high BFI values, signifying robust groundwater-surface water interactions, whereas areas such as the Defiance Plateau and southern Arizona show consistently lower BFI values, reflecting limited groundwater contributions.
+Regions such as the Grand Canyon and Mogollon Rim exhibit the highest BFI values, signifying robust groundwater-surface water interactions, whereas areas such as the Defiance Plateau and southern Arizona show consistently lower BFI values, reflecting limited groundwater contributions.
 

### 2025-05-01 elsevier_MANUSCRIPT.qmd
Update elsevier_MANUSCRIPT.qmd

@@
 Additionally, years with more than 30 missing days of streamflow data were excluded from the analysis.
-Streamgages directly downstream of major regulation (e.g., reservoirs, lakes, diversions) were excluded, based on USGS annual water data reports and site metadata \[@usgs2010\].
+Streamgages directly downstream of major regulation (e.g., reservoirs, lakes, diversions) were excluded, based on USGS annual water data reports and site metadata [@usgs2010].
 While some flow alteration is widespread, the focus was on removing gauges with clear, immediate regulatory impacts.
@@
 This streamgage was excluded from the trend analysis, as autocorrelation could inflate the variance of the Mann-Kendall statistic, potentially leading to biased trend estimates [@MK_Hamed1998].
-Trends with a $\rho \le 0.05$ are considered significant.
+Trends with a $\alpha \le 0.05$ are considered significant.
 

### 2025-05-26 elsevier_MANUSCRIPT.qmd
bulk edits from reviewers

--- 2025-04-25 elsevier_MANUSCRIPT.qmd
@@
 
-Dryland regions, encompassing arid, semi-arid, hyper-arid, and dry sub-humid systems, account for 40% of the Earth’s land surface.
+Dryland regions, encompassing arid, semi-arid, hyper-arid, and dry sub-humid systems, cover 40% of the Earth’s land surface.
 These regions are home to approximately 2 billion people globally and constitute the largest terrestrial biome [@iucn-drylands-2019].
@@
 Addressing this information gap is integral to approaching a comprehensive understanding of groundwater dynamics globally.
+This study addresses this gap by analyzing long-term BFI patterns across Arizona using instrumented streamflow records and applying a machine learning model to predict BFI in ungauged catchments.
 
 Advancements in machine learning provide tools to predict hydrologic indices in ungauged basins, addressing the limitations of sparse streamgage networks.
-To tackle the challenge of quantifying base flow in ungauged catchments, numerous studies have applied both regression and machine learning methods.
+To tackle the challenge of quantifying base-flow indices in ungauged catchments, numerous studies have applied both regression and machine learning methods.
 @ahiablame2013 found that using a regression model to estimate annual base flow of ungauged catchments was reasonably easy and accurate.
@@
 The outcomes offer novel insights into groundwater contributions to streamflow, supporting water resource management in an area increasingly vulnerable to drought and climate change impacts, and contributing broadly to the understanding of hydrologic processes in dryland environments.
+The objectives of this study are to (1) quantify long-term spatial and temporal trends in BFI across Arizona, and (2) develop a predictive model using eXtreme Gradient Boosting (XGBoost) to estimate BFI in ungauged basins based on hydroclimatic and geospatial variables.
 
@@
 Much of the hydrology of Arizona is snow-melt derived, driven by spring melt from the high-elevation Colorado Plateau winter snowpack.
-Though winter precipitation provides only 30% of annual average precipitation, it provides the majority of water for natural reservoirs [@sheppard2002].
+Though winter precipitation accounts for only 30% of annual precipitation totals, it provides the majority of water for natural reservoirs [@sheppard2002].
 
@@
 
-Directly measuring base flow and BFI presents unique challenges [@eckhardt2008].
-The technique chosen to separate base flow has been shown to affect results, and the choice of base-flow separation method is subjective since ‘true’ BFI values are not known [@beck2013].
-However, many methods have been developed to estimate these values.
-These methods include the use of tracers [@gonzales2009], graphical interpolation [@instituteofhydrology1980; @sloto1996], and digital filters [@arnold1995; @eckhardt2005; @lyne1979; @nathan1990].
-These techniques have varying levels of applicability depending on the spatial scale, time span, and the scope of the study.
-Comparisons of various base-flow separation techniques have been made in previous studies [e.g.
-@eckhardt2005; @eckhardt2008; @nathan1990]; this study does not explore the superiority of different methods.
+Directly measuring base flow and base-flow index (BFI) presents unique challenges [@eckhardt2008].
+Base flow is a physical hydrologic process representing groundwater contributions to streamflow, while BFI is a dimensionless, statistical metric that estimates the proportion of streamflow derived from base flow.
+Because BFI must be derived using a base-flow separation technique, its value is inherently method-dependent and sensitive to the separation approach used [@beck2013].
+A variety of techniques have been developed to estimate base flow, including tracer studies [@gonzales2009], graphical interpolation methods [@instituteofhydrology1980; @sloto1996], and digital filters [@arnold1995; @eckhardt2005; @lyne1979; @nathan1990].
+Each of these methods varies in its applicability depending on spatial scale, record length, and study objectives.
+Although the choice of separation method and parameterization introduces some subjectivity, these filters have been shown to yield reliable estimates when applied consistently within a study domain [@chapman_1999; @eckhardt2005; @instituteofhydrology1980; @ayers2022].
+Numerous studies have compared these separation techniques [e.g., @eckhardt2005; @eckhardt2008; @nathan1990]; however, this study does not evaluate the relative performance of different methods.
 
@@
 The filter was run three times (forward, backward, forward) to attenuate the base-flow signal.
+
+Two BFI metrics were used in this study: (1) annual BFI, calculated for each calendar year using daily streamflow records, and (2) long-term BFI, defined as the average of annual BFI values over the period of record at each site.
+Annual BFI reflects interannual variability in base-flow contributions, while long-term BFI represents a time-averaged metric of groundwater influence.
+The machine learning model was trained to predict annual BFI values, which were then aggregated to produce long-term BFI estimates.
 
@@
 
-Our initial training dataset comprised 7,724 observations (annual observations of BFI at streamgages) across 45 variables.
+Our training dataset consisted of 7,724 site-year observations of annual BFI, where each observation represents a single water year at a given streamgage.
+These observations were paired with 45 predictor variables derived from basin characteristics and climate data.
 To optimize the model's performance, we first conducted an exhaustive grid search combined with 5-fold cross-validation to identify the optimal hyperparameter values.
@@
 The long-term mean BFI is 0.32, indicating that \~32% of long-term streamflow in Arizona likely originates from groundwater discharge and other delayed sources.
-The highest BFI values (\>0.9) are found along the Grand Canyon in northwestern Arizona, where highly karstic geology facilitates rapid subsurface flow to surface water and spring outlets [@chambless_deep-karst_2023].
-High BFI values (\>0.8) are also found at the spring-fed headwaters of the Verde River and Fossil Creek, which have similar highly-karstic, snowmelt-driven recharge areas.
+The highest long-term BFI values (\>0.9) are found along the Grand Canyon in northwestern Arizona, where highly karstic geology facilitates rapid subsurface flow to surface water and spring outlets [@chambless_deep-karst_2023].
+High long-term BFI values (\>0.8) are also found at the spring-fed headwaters of the Verde River and Fossil Creek, which have similar highly-karstic, snowmelt-driven recharge areas.
 
-The stream reaches of the Little Colorado River Basin (northeastern Arizona) indicate consistently low BFI values (\< 0.2).
+The stream reaches of the Little Colorado River Basin (northeastern Arizona) indicate low long-term BFI values (\< 0.2).
 This is likely due to low-yielding perched aquifers underlying the Defiance Plateau in northeastern Arizona, which are hydrologically connected to surface streams, while the high-yield, confined regional aquifer is much deeper [@blanchard02].
@@
 
-#### Trends in Base Flow and BFI {#sec-bfi-trends}
+#### Trends in BFI {#sec-bfi-trends}
 
-Trends in BFI over the period of record for each streamgage are illustrated in @fig-instrumented-trend and summarized in @tbl-trends.
-Base flow and BFI trends were analyzed across all instrumented sites over their respective periods of record using the Mann–Kendall test.
-Statistically significant trends were observed in both metrics, with a 72.2% coincidence rate between significant base flow and BFI trends, indicating a strong dependence of BFI on base-flow dynamics.
+Trends in long-term BFI over the period of record for each streamgage are illustrated in @fig-instrumented-trend and summarized in @tbl-trends.
+Long-term BFI trends were assessed at all instrumented sites using the Mann–Kendall test.
+As BFI quantifies the proportion of streamflow attributed to base flow, its long-term trends inherently reflect changes in subsurface contributions relative to total flow.
+A 72.2% coincidence rate between significant base flow and BFI trends supports the interpretation that changes in BFI are largely driven by underlying base-flow dynamics.
+Statistically significant long-term BFI trends were observed across numerous sites, with spatial variation highlighting regional differences in hydrologic response.
 
-@fig-instrumented-trend illustrates the spatial variation in BFI trends across the study area.
+@fig-instrumented-trend illustrates the spatial variation in long-term BFI trends across the study area.
 Statistically significant decreasing trends are observed at 16.1% of sites, while increasing trends are found at 8.8% of sites.
@@
 
-Statistically significant decreasing trends in BFI were more common than increasing trends across all site classifications (@tbl-trends).
+Statistically significant decreasing trends in long-term BFI were more common than increasing trends across all site classifications (@tbl-trends).
 While decreasing trends dominate, both increasing and decreasing trends are observed within each classification.
-Monsoon-dominated regions exhibit a higher proportion of significant negative trends (24.1%) compared to snowmelt-dominated regions (10.2%), suggesting that monsoon-dominated systems are more consistently correlated with declining base flow.
-Among climate classifications, warm-dry climates have the highest proportion of negative trends (20.0%), followed by warm-wet climates (19.4%), indicating that regions with higher temperatures are more prone to base flow declines.
+Monsoon-dominated regions exhibit a higher proportion of significant negative trends (24.1%) compared to snowmelt-dominated regions (10.2%), suggesting that monsoon-dominated systems are more consistently correlated with declining long-term BFI.
+Among climate classifications, warm-dry climates have the highest proportion of negative trends (20.0%), followed by warm-wet climates (19.4%), indicating that regions with higher temperatures are more prone to BFI declines.
 Low-slope regions show a greater prevalence of negative trends (20.4%) compared to high-slope regions (11.8%).
@@
 
-Coincident trends are defined as those in which base flow or BFI trends match in direction with climate variable trends over the same period.
+Coincident trends are defined as those in which long-term BFI trends match in direction with climate variable trends over the same period.
 Results are summarized in @tbl-coincident-trends.
@@
 
-Given the variations in the period of record across the instrumented network (see Figure S2), we analyzed trends in base flow and BFI in relation to coincident trends in climate variables.
-Trends are classified as coincident when the direction of the climate variable trend (positive or negative) aligns with the trend observed in base flow or BFI (@tbl-coincident-trends).
+Given the variations in the period of record across the instrumented network (see Figure S2), we analyzed trends of long-term BFI in relation to coincident trends in climate variables.
+Trends are classified as coincident when the direction of the climate variable trend (positive or negative) aligns with the trend observed in BFI (@tbl-coincident-trends).
 
 |     | **Climate Variable** | **Coincidence Percentage (%)** |
-|-----------|:--------------------:|:------------------------------:|
+|-----|:--------------------:|:------------------------------:|
 | BFI |        ET~O~         |             44.88              |
@@
 |     |     Temperature      |             47.32              |
-| Base Flow |        ET~O~         |             55.12              |
-|           |    Precipitation     |             64.88              |
-|           |     Temperature      |             39.02              |
 
-: Coincident trends of climate variables (ET~O~, precipitation, temperature) with base flow and BFI trends. {#tbl-coincident-trends}
+: Coincident trends of climate variables (ET~O~, precipitation, temperature) with BFI trends. {#tbl-coincident-trends}
 
@@
 
-Predicted values of BFI are plotted against observed values for the entire period of record of the instrumented dataset in @fig-actual_predicted .
+Predicted values of annual BFI are plotted against observed values for the entire period of record of the instrumented dataset in @fig-actual_predicted .
 The agreement between "out-of-bag" predictions (blind cross-validation, treating each site as ungauged) and observed values indicate strong model performance across the full dataset (R^2^ = 0.764).
@@
 These metrics show consistent model performance across spatial and climatic classifications.
-However, the negative pbias values across all classifications indicate a systematic underprediction of BFI.
+However, the negative pbias values across all classifications indicate a systematic underprediction of annual BFI.
 Categories with relatively lower R^2^ and Nash-Sutcliffe Efficiency (NSE) values also exhibit higher biases, reflecting weaker model performance in those contexts.
@@
 
-: Performance of model predictions for BFI for all sites split by various classifications.
+: Performance of model predictions for annual BFI for all sites split by various classifications.
 n is number of observations, R^2^ is the coefficient of determination of a linear regression, MSE is mean-squared-error, RMSE is root-mean-squared-error, MAE is mean-absolute-error, NSE is Nash-Sutcliffe efficiency, and pbias is percent bias.
@@
 
-The predictors used to estimate BFI at ungauged sites were evaluated for their importance in the final XGBoost model, as illustrated in @fig-shap_values.
-The most influential feature for predicting long-term BFI is basin elevation.
+The predictors used to estimate annual BFI at ungauged sites were evaluated for their importance in the final XGBoost model, as illustrated in @fig-shap_values.
+The most influential feature for predicting annual BFI is basin elevation.
 While elevation itself does not directly affect base-flow characteristics, it has consistently been identified as a key predictor in previous BFI studies [@singh2018; @beck2013].
@@
 
-Land cover and land use predictors also play a significant role in BFI estimation.
+Land cover and land use predictors also play a significant role in annual BFI estimation.
 Analysis of SHAP values indicates that a higher percentage of evergreen forest positively influences BFI predictions, while higher proportions of shrubland and developed land exert a negative influence.
@@
 The regionalized (HUC-8) long-term BFI (1991–2020) is shown in @fig-bfi-huc.
-Basins with high BFIs, such as those along the Grand Canyon in the northwestern part of the study area, indicate greater surface water and groundwater interaction.
+Basins with high long-term BFIs, such as those along the Grand Canyon in the northwestern part of the study area, indicate greater surface water and groundwater interaction.
 Elevated BFI values are also observed along portions of the Mogollon Rim, a heavily forested region with high precipitation that marks the transition between physiographic regions.
-Additionally, headwater regions of perennial rivers tend to exhibit higher BFI values.
-In contrast, low BFI values are found in areas like the Defiance Plateau in northeastern Arizona and the arid southern regions of the state.
+Additionally, headwater regions of perennial rivers tend to exhibit higher long-term BFI values.
+In contrast, low long-term BFI values are found in areas like the Defiance Plateau in northeastern Arizona and the arid southern regions of the state.
 
-![Predicted long-term BFI values for 8-digit HUC (1991-2020)](images/updated/BFI_HUC8_20250404.png){#fig-bfi-huc}
+![Predicted long-term BFI values for 8-digit HUC (1991-2020). Bolded HUCs indicate those without a streamgage.](images/BFI_HUC8_20250525.png){#fig-bfi-huc}
 
@@
 
-This study addresses these monitoring gaps through an analysis of observed base flow and BFI at instrumented streamgages, highlighting groundwater-surface water interactions within Arizona’s arid and semi-arid landscapes.
+This study addresses these monitoring gaps through an analysis of observed BFI at instrumented streamgages, highlighting groundwater-surface water interactions within Arizona’s arid and semi-arid landscapes.
 Observed spatial patterns in BFI clearly align with known hydrogeological and climatic variations.
-High BFI values coincide with areas characterized by spring-fed streams and pronounced physiographic transitions, such as the Grand Canyon and Mogollon Rim.
+High long-term BFI values coincide with areas characterized by spring-fed streams and pronounced physiographic transitions, such as the Grand Canyon and Mogollon Rim.
 Conversely, lower observed BFI values correspond to regions with limited groundwater contributions, notably the Defiance Plateau and southern Arizona, emphasizing vulnerability in these arid environments.
@@
 
-Analysis of trends at instrumented sites revealed a predominance of declining BFI values, particularly within monsoon-dominated precipitation regimes, warm-dry climates, and low-slope basins.
+Analysis of trends at instrumented sites revealed a predominance of declining long-term BFI values, particularly within monsoon-dominated precipitation regimes, warm-dry climates, and low-slope basins.
 These negative trends likely reflect intensified climate stress, highlighting areas most vulnerable to groundwater depletion and surface-water scarcity.
@@
 
-Given the limited coverage provided by existing gauges, a common issue in arid and semi-arid regions globally [@krabbenhoft-2022], we utilized a machine learning approach to estimate BFI in ungauged catchments.
+Given the limited coverage provided by existing gauges, a common issue in arid and semi-arid regions globally [@krabbenhoft-2022], we utilized a machine learning approach to estimate annual BFI in ungauged catchments.
 The regional model exhibited strong predictive capability (overall R²=0.764), though systematic underprediction (negative pbias) was noted, particularly in monsoon-dominated and warm-dry climates.
@@
 The long-term average BFI across Arizona is approximately 0.32, highlighting that groundwater discharge significantly contributes to surface water flows but exhibits substantial spatial variability.
-Our machine learning model (XGBoost) effectively predicts BFI in ungauged catchments (R² = 0.764), primarily driven by basin elevation, land cover, and hydrologic soil characteristics.
-Regions such as the Grand Canyon and Mogollon Rim exhibit the highest BFI values, signifying robust groundwater-surface water interactions, whereas areas such as the Defiance Plateau and southern Arizona show consistently lower BFI values, reflecting limited groundwater contributions.
+Our machine learning model (XGBoost) effectively predicts annual BFI in ungauged catchments (R² = 0.764), primarily driven by basin elevation, land cover, and hydrologic soil characteristics.
+Regions such as the Grand Canyon and Mogollon Rim exhibit the highest long-term BFI values, signifying robust groundwater-surface water interactions, whereas areas such as the Defiance Plateau and southern Arizona show consistently lower BFI values, reflecting limited groundwater contributions.
 
-Analysis of temporal trends indicates a prevailing pattern of declining base flow and BFI across the region, particularly in monsoon-dominated, warm-dry, and low-slope areas.
+Analysis of temporal trends indicates a prevailing pattern of declining long-term BFI across the region, particularly in monsoon-dominated, warm-dry, and low-slope areas.
 Precipitation emerged as the strongest driver of base-flow variability, while temperature and evapotranspiration added additional layers of complexity.

### 2025-05-28 elsevier_MANUSCRIPT.qmd
updates to full text

@@
 Building on this analysis, we applied machine learning methods to predict BFI in ungauged basins, addressing the challenges of Arizona’s sparse streamgage network.
-Using the eXtreme Gradient Boosting (XGBoost) algorithm trained on hydroclimate and physiographic predictors, we estimated long-term BFI from 1991 to 2020.
+Using the eXtreme Gradient Boosting (XGBoost) algorithm trained and validated on observed hydroclimate and physiographic variables, we estimate long-term BFI from 1991 to 2020.
 This combined approach integrates observational data with predictive modeling to enhance our understanding of base-flow processes and provide a framework for water resource management in data-limited regions.
@@
 bibliography: references.bib
+format: docx
 ---
@@
 
-Despite advancements, existing large-scale studies have limited applicability to the unique hydrogeologic conditions in arid and semi-arid regions like Arizona, leaving uncertainty regarding groundwater-surface water interactions in these critically water-stressed areas.
-This study addresses this crucial gap by developing a regionally tailored machine learning model specifically designed to estimate BFI for Arizona's ungauged basins.
-Utilizing hydrogeologic characteristics and hydroclimatic data, we estimate long-term mean BFI (1991–2020) in ungauged catchments, filling spatial gaps in the sparse streamgage network.
-Furthermore, regional trends in base flow and BFI at instrumented sites are analyzed alongside coincident climate trends in precipitation, reference evapotranspiration (ETO), and temperature.
-The outcomes offer novel insights into groundwater contributions to streamflow, supporting water resource management in an area increasingly vulnerable to drought and climate change impacts, and contributing broadly to the understanding of hydrologic processes in dryland environments.
-The objectives of this study are to (1) quantify long-term spatial and temporal trends in BFI across Arizona, and (2) develop a predictive model using eXtreme Gradient Boosting (XGBoost) to estimate BFI in ungauged basins based on hydroclimatic and geospatial variables.
+To address these limitations, this study focuses on Arizona’s arid and semi-arid basins, where groundwater–surface water interactions remain poorly characterized due to sparse monitoring and regionally distinct hydrologic behavior.
+We develop a region-specific machine learning model using eXtreme Gradient Boosting (XGBoost) to estimate long-term BFI (1991–2020) across ungauged basins, incorporating basin hydrogeology and hydroclimate predictors.
+In parallel, we assess trends in BFI at instrumented sites alongside coincident changes in precipitation, temperature, and reference evapotranspiration.
+Together, these analyses provide new insight into the spatial patterns and temporal dynamics of baseflow in a water-stressed region increasingly affected by climate change.
+The study objectives are to (1) quantify long-term spatial and temporal trends in BFI across Arizona and (2) develop a predictive model to estimate BFI in ungauged basins based on geospatial and climate variables.
 
@@
 
-The implementation of machine learning models to predict hydrologic indices has been successful in past studies [@singh2018; @schmidt_ML_2020; @Rozos2021Machine].
-In this work, we used the eXtreme Gradient Boosting (XGBoost) algorithm [@chen2016xgboost] to predict BFI at ungauged locations using catchment characteristics as predictors [@tbl-predictors].
-The XGBoost algorithm is a decision tree-based ensemble algorithm, which can be adapted for either regression or classification problems.
-This algorithm iteratively builds an ensemble of decision trees, where each tree corrects errors from previous trees to improve predictions [@chen2016xgboost].
-Its efficiency, scalability, and robustness have made it increasingly popular in recent years, with successful applications in environmental modeling tasks such as streamflow forecasting [@szczepanek2022daily; @ni2020streamflow] and land use/land cover classification [@Georganos2018Very].
-
-XGBoost operates by leveraging gradient boosting on decision tree algorithms, combining multiple low-variance models to produce a robust overall prediction.
-Gradient boosting works iteratively: the initial tree is trained on the target values, while subsequent trees are trained on the residual errors of the preceding tree.
-Each tree is assigned a weight based on its contribution to reducing error, and these weights are used to determine the influence of each tree in the final model.
-The ultimate prediction is made by aggregating the outputs of all $n$ weighted trees in the ensemble.
-In this study, the trained XGBoost model is used to predict BFI in ungauged catchments based on geospatial and hydroclimate predictor variables (@tbl-predictors).
-Certain features were further subdivided according to their areal coverage within each basin (e.g.
-land cover was divided into 16 subdivisions).
-This approach allowed the model to capture finer-scale spatial variability and improve predictive accuracy.
+Machine learning (ML) approaches have demonstrated strong utility in hydrologic prediction tasks, including the estimation of streamflow and base-flow indices [@singh2018; @schmidt_ML_2020; @Rozos2021Machine].
+In this study, we used the eXtreme Gradient Boosting (XGBoost) algorithm [@chen2016xgboost] to estimate annual BFI at ungauged locations across Arizona using catchment characteristics as predictors [@tbl-predictors].
+XGBoost is a gradient boosting algorithm that constructs an ensemble of decision trees, where each tree iteratively corrects errors from previous trees to improve predictive accuracy [@chen2016xgboost].
+The model was trained on annual BFI observations in order to estimate annual BFI in ungauged basins.
+These annual BFI estimates were averaged over time to represent long-term BFI, consistent with the time-integrated nature of our response metric.
 
 |              | Variable                      | Source | Geoprocessing     |
-|----|----|----|----|
-| Hydroclimate | Precipitation\* | PRISM | Basin average |
-|  | Mean Temperature\* | PRISM | Basin average |
-|  | Reference Evapotranspiration\* | TerraClimate | Basin average |
-| Geospatial | Elevation\* | DEM | Basin average |
-|  | Area | DEM | Basin average |
-|  | Slope | DEM | Basin average |
-|  | Aspect | DEM | Percent areal coverage |
-|  | Land Cover\* | NLCD | Percent areal coverage |
-|  | Hydrologic Soil Group\* | SSURGO | Percent areal coverage |
-|  | Geology | USGS | Percent areal coverage |
-|  | Karst | USGS | Percent areal coverage |
+|--------------|-------------------------------|--------|-------------------|
+| Hydroclimate | Temperature                   | PRISM  | Basin Mean        |
+|              | Mean Precip                   | PRISM  | Basin Mean        |
+|              | Reference ET                  | MODIS  | Basin Mean        |
+| Geospatial   | Elevation                     | DEM    | Basin Mean        |
+|              | Soil Type A                   | SSURGO | \% Areal Coverage |
+|              | Soil Type C                   | SSURGO | \% Areal Coverage |
+|              | Land Cover - Open Water       | NLCD   | \% Areal Coverage |
+|              | Land Cover - Low Develoment   | NLCD   | \% Areal Coverage |
+|              | Land Cover - Evergreen Forest | NLCD   | \% Areal Coverage |
+|              | Land Cover - Herbaceous       | NLCD   | \% Areal Coverage |
 
-: Basin-characteristic variables used as initial features in XGBoost model. Starred features are maintained in the final, dimensionality-reduced model. {#tbl-predictors}
+: Basin-characteristic variables used in reduced XGBoost model. A list of initial predictor variables used in the full model is provided in the Supplemental Information. {#tbl-predictors}
 
-Our training dataset consisted of 7,724 site-year observations of annual BFI, where each observation represents a single water year at a given streamgage.
+The training dataset consisted of 7,724 site-year observations, where each observation represents annual BFI for a single water year at a given USGS streamgage.
 These observations were paired with 45 predictor variables derived from basin characteristics and climate data.
-To optimize the model's performance, we first conducted an exhaustive grid search combined with 5-fold cross-validation to identify the optimal hyperparameter values.
-The hyperparameters evaluated included the learning rate ($\eta$), minimum split loss ($\gamma$), maximum tree depth, minimum child weight, and the number of trees.
-While not an exhaustive list of all possible XGBoost hyperparameters, this range of values provided sufficient variation to ensure the selection of a high-performing model.
-The optimal hyperparameters were determined to be: 700 trees, a learning rate ($\eta$) of 0.05, a minimum split loss ($\gamma$) of 0.075, a maximum tree depth of 7, and a minimum child weight of 5.
-Using these values, the XGBoost model was trained on the dataset with 10-fold cross-validation.
+Predictor selection was based on prior hydrologic studies, data availability across the study area, and pairwise correlation screening to minimize multicollinearity.
+The predictors included basin-averaged climate metrics (e.g., precipitation, temperature), topographic features (e.g., elevation, slope), and land surface attributes (e.g., land cover classes, soil type, karst presence).
 
-$K$-fold cross-validation provides an unbiased estimate of a model's accuracy on unseen data, while also insuring against overfitting or underfitting.
-In this approach, the data are randomly divided into $k$ folds of equal size.
-The model is trained on $k-1$ folds and tested on the remaining fold, referred to as the validation set.
-This process is repeated $k$ times, with each fold serving as the validation set exactly once.
-Each iteration trains an independent model with the same hyperparameters but using a different subset of training data.
-By averaging the model performance across all $k$ folds, we achieve a robust and reliable estimate of predictive accuracy.
-Root mean squared error (RMSE) was used as the performance metric for both model optimization and evaluation.
-RMSE provided a consistent and interpretable measure of the model’s accuracy throughout the training and validation process.
+We tested multiple model configurations, including versions with the full predictor set across the state, separate models for physiographic regions with full predictor set, and reduced subsets across the state, to evaluate the impact of different predictor combinations on model performance and to ensure robustness of the final selected model.
+Dimensionality reduction was performed to improve interpretability and reduce computational burden.
+We applied SHAP (SHapley Additive exPlanations) values [@SHAP_values] to assess global feature importance and rank predictors.
+SHAP values quantify both the magnitude and direction of each feature’s contribution to predictions.
+The ten most important features—based on mean absolute SHAP value—were retained in a reduced model, which showed improved predictive accuracy and efficiency over the full model and regional models.
+Sensitivity analyses tested region-specific models and smaller predictor subsets; however, the state-wide model using SHAP-prioritized features consistently outperformed alternatives.
+
+![SHAP value plot of features used in final model.
+The number to the right of the feature name is the mean absolute SHAP value.
+Land cover features are indicated by the percentage of cover by each land cover type and soil types are defined by hydrologic soil group.](images/shap_summary.png){#fig-shap_values alt="SHAP value plot of features used in final model.
+The number to the right of the feature name is the mean absolute SHAP value.
+Land cover features are indicated by the percentage of cover by each land cover type and soil types are defined by hydrologic soil group."}
+
+The model dataset was evaluated using 10-fold cross-validation, ensuring robust performance estimates and guarding against overfitting [@fig-k-fold] .
+In each iteration (fold), the model was trained on 90% of the data and validated on the remaining 10%, with folds rotated until all data had been used for validation in separate models.
+Root mean squared error (RMSE) was used as the primary performance metric.
+Model performance on unseen data was strong, with an overall R² of 0.764.
+Additional model validation results are presented in section 3.2.1.
 
@@
 Each fold serves as the validation set once, while the remaining folds are combined to create a training set for model development.
-Performance metrics for the test set are calculated and recorded, and this process is repeated for all $k$ folds.](images/k-fold_fig.png){#fig-k-fold fig-align="center" width="100%"}
+Performance metrics for the test set are calculated and recorded, and this process is repeated for all $k$ folds.](images/k-fold_fig.png){#fig-k-fold alt="Schematic of $k$-fold cross validation.
+The dataset is randomly divided into $k$ stratified folds.
+Each fold serves as the validation set once, while the remaining folds are combined to create a training set for model development.
+Performance metrics for the test set are calculated and recorded, and this process is repeated for all $k$ folds." fig-align="center" width="100%"}
 
-#### Feature Selection {#sec-feat-selection}
-
-Machine learning models are prone to overfitting, especially when provided with a large set of predictive features [@Ying_2019].
-Overfitting can degrade model performance on unseen data and increase the demand for computational resources and memory storage [@feat_selec2017].
-Dimensionality reduction offers a robust solution to these challenges and generally falls into two broad categories: feature extraction and feature selection.
-
-Feature extraction involves transforming the original dataset into a lower-dimensional feature space.
-However, this process generates new features that often lack the physical interpretability of the original variables.
-In contrast, feature selection identifies a subset of the original features, preserving their physical meaning while improving model readability and interpretability [@feat_selec2017].
-In this study, supervised feature selection was employed to reduce the number of predictors, which enhanced learning performance, reduced computational costs, and mitigated overfitting.
-
-To begin, an initial model was trained using the full feature set of 45 predictors (@tbl-predictors).
-A feature selection method based on feature importance was then applied to identify and remove less relevant and noisy features.
-Feature importance scores quantify the contribution of individual features—either positively or negatively—to the model's predictions [@ML_interpretability2019].
-In this analysis, SHAP (SHapley Additive exPlanations) values were used to compute feature importance scores [@SHAP_values].
-
-SHAP values is a method to explain the prediction of an individual instance by calculating the contribution of each feature to that prediction.
-The method is based on coalition game theory and is discussed further in @SHAP_values.
-Here, SHAP values are used for global interpretation of feature importance and feature effects on the model.
-Global feature importance is produced by the absolute Shapley values of each feature across the dataset, providing a list of features in order of most to least important.
-Feature effects provide an indication of the relationship between the value of a predicting feature and its impact on the prediction.
-
-![SHAP value plot of features used in final model.
-The number to the right of the feature name is the mean absolute SHAP value.
-Land cover features are indicated by the percentage of cover by each land cover type and soil types are defined by hydrologic soil group.](images/shap_summary.png){#fig-shap_values}
-
-The ten most important features [@fig-shap_values] were selected based on their SHAP values and used to train a subsequent model [@tbl-predictors].
-This refined model demonstrated improved performance and reduced computational time compared to the initial model.
-The final model, trained on this optimized feature subset, was ultimately used for the analysis presented here.
+Hyperparameter tuning was performed through an internal grid search.
+The search evaluated combinations of key parameters including the number of trees, learning rate ($\eta$), minimum split loss ($\gamma$), tree depth, and minimum child weight.
+A full discussion of these hyperparametes can be found in @chen2016xgboost.
+The final model used 700 trees, $\eta$ of 0.05, $\gamma$ of 0.075, a maximum depth of 7, and a minimum child weight of 5.
+All parameters were selected for their consistent reduction in RMSE across validation folds.
 
@@
 
-Statistical analyses were conducted on annual BFI and base-flow values from instrumented streamgages to identify temporal trends using the Mann-Kendall nonparametric trend test [@kendall1970; @mann1945nonparametric].
-This test detects monotonic trends in datasets that are non-parametric and assumes the absence of autocorrelation among observations.
-This test is widely used in studies of this nature [@ficklin2016; @ayers2019; @woodhouse_udall22].
+Statistical analyses were conducted on annual BFI values from instrumented streamgages to identify site-specific temporal trends using the Mann–Kendall nonparametric trend test [@kendall1970; @mann1945nonparametric].
+This test detects monotonic trends in non-normally distributed data and assumes the absence of autocorrelation.
+This test is widely used in hydrologic trend studies [@ficklin2016; @ayers2019; @woodhouse_udall22].
 
-To check for autocorrelation, we applied the Durbin-Watson test [@durbin1950testing], which revealed significant autocorrelation at four streamgages on an annual basis.
-Of these, only one streamgage (09486500 - Santa Cruz River at Cortaro, AZ) showed a significant trend in BFI.
-This streamgage was excluded from the trend analysis, as autocorrelation could inflate the variance of the Mann-Kendall statistic, potentially leading to biased trend estimates [@MK_Hamed1998].
+To account for potential autocorrelation in annual BFI time series at each instrumented site, we applied the Durbin–Watson test [@durbin1950testing].
+Four streamgages exhibited significant autocorrelation; these sites were excluded from the trend analysis to avoid inflated variance in the Mann–Kendall statistic and potential bias in trend detection [@MK_Hamed1998].
+As all autocorrelated series were removed, pre-whitening using the Hamed–Rao method was unnecessary [@MK_Hamed1998].
 Trends with a $\alpha \le 0.05$ are considered significant.
@@
 
-Previous studies have examined base-flow regionalization and synthesis across various spatial scales, some overlapping with our study area.
-@beck2013 explored global patterns of BFI using streamflow observations extrapolated through an ANN.
-The U.S.
-streamgages included in that analysis were drawn from the interim Model Parameter Estimation Experiment (MOPEX) dataset of 1862 USGS gauges [@beck2013; @mopex_2006].
-Only 10 of these interim gauges overlap geographically with our study area.
-Similarly, @ayers2022 assessed climatic influences on monthly base flow across the continental United States (CONUS), utilizing a subset of GAGES-II streamflow gauges.
-Although exact gauge locations from that study are unavailable, published figures suggest minimal geographic overlap with our research domain.
-@santhi2008 employed interpolation methods for regional base-flow estimation across CONUS, but specific gauge locations or data are not available, limiting direct regional comparisons.
+Several previous studies have examined regional and continental patterns of BFI, offering useful context for our findings despite limited geographic overlap.
+@beck2013 developed a global map of BFI using an artificial neural network trained on 1,862 U.S.
+streamgages from the MOPEX dataset [@mopex_2006].
+Only ten of these gauges overlap with our Arizona study area, limiting direct comparison.
+Nonetheless, the spatial distribution of BFI in Arizona from @beck2013 shows general agreement with our regional patterns.
+The higher BFI values reported by @beck2013 may reflect the influence of their global-scale ANN model and associated generalizations.
+Similarly, @ayers2022 examined base-flow dynamics using GAGES-II sites across CONUS, but their study included few gauges within Arizona.
+While exact locations are not specified, their published figures suggests regional agreement with our estimated BFI distributions.
+@santhi2008 employed interpolation methods for regional base-flow estimation across CONUS, but specific gauge locations or data are not available limiting comparison.\`\`\`\`
 
-The existing streamgage networks Arizona are limited for base-flow analyses by their design and monitoring objectives.
+In terms of long-term trends, our results align directionally with @ayers2022 , who found consistent monthly declines in base flow from 1989–2019 across the U.S.
+Southwest.
+Although our analysis focused on long-term trends in annual BFI, we observed a similar imbalance: 16% of sites showed significant decreasing trends compared to 8% with increasing trends.
+While we do not directly replicate their monthly analysis, the directional agreement supports the broader conclusion that groundwater contributions to streamflow are declining in many parts of the region.
+
+The existing streamgage networks in Arizona are limited for base-flow analyses by their design and monitoring objectives.
 While the number of streamgages has increased to meet regulatory imperatives, such as those driven by the Clean Water Act, many gauges prioritize peak flow monitoring rather than low-flow conditions critical for base-flow studies [@flood-control2020].
@@
 
+The combined use of long-term baseflow trend analysis and XGBoost modeling offers a transferable framework for understanding groundwater–surface water interactions in data-scarce regions globally.
+This approach is particularly relevant in dryland and seasonally variable climates, where direct baseflow observations are limited and hydrologic responses are strongly shaped by physiography and land cover.
+Critically, arid regions around the world remain under-instrumented, with sparse monitoring networks and limited long-term streamflow records, despite facing acute water scarcity and growing pressure on groundwater systems [@taylor2013].
+The success of such models depends on the availability and quality of climate, topographic, and soil datasets, as well as the representativeness of streamflow records used in training.
+Regions with differing hydrogeologic settings, such as glacial terrains, humid catchments with deep weathered profiles, or regions with inter-annual snowpack storage, may require alternative predictor variables or model structures.
+For instance, in humid regions with dense vegetation and shallow groundwater tables, evapotranspiration dynamics may play a more dominant role in baseflow variability, requiring region-specific calibration.
+Despite these caveats, the generalizable structure of this modeling framework, along with the interpretability offered by SHAP-based feature selection, makes it well-suited for adaptation to other geographic contexts, especially where water resources are vulnerable to climatic and land-use change
+
 ## Conclusions

### 2025-05-28 elsevier_MANUSCRIPT.tex
updates to full text

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
  authoryear,
  preprint,
  1p,
  onecolumn]{elsarticle}

\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else  
    % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\setcounter{secnumdepth}{5}
% Make \paragraph and \subparagraph free-standing
\makeatletter
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}{
    \@ifstar
      \xxxParagraphStar
      \xxxParagraphNoStar
  }
  \newcommand{\xxxParagraphStar}[1]{\oldparagraph*{#1}\mbox{}}
  \newcommand{\xxxParagraphNoStar}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}{
    \@ifstar
      \xxxSubParagraphStar
      \xxxSubParagraphNoStar
  }
  \newcommand{\xxxSubParagraphStar}[1]{\oldsubparagraph*{#1}\mbox{}}
  \newcommand{\xxxSubParagraphNoStar}[1]{\oldsubparagraph{#1}\mbox{}}
\fi
\makeatother

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother

\makeatletter
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{plain}
\@ifundefined{c@chapter}{\newfloat{suppfig}{h}{losuppfig}}{\newfloat{suppfig}{h}{losuppfig}[chapter]}
\floatname{suppfig}{Figure S}
\newcommand*\quartosuppfigref[1]{Figure \hyperref[#1]{S\ref{#1}}}
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\DeclareCaptionLabelFormat{quartosuppfigreflabelformat}{#1#2}
\captionsetup[suppfig]{labelformat=quartosuppfigreflabelformat}
\newcommand*\listofsuppfigs{\listof{suppfig}{List of Supplementary Figures}}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\journal{Journal of Hydrology}

\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage[]{natbib}
\bibliographystyle{elsarticle-harv}
\usepackage{bookmark}

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Regional Base-Flow Index in Arid Landscapes Using Machine Learning and Instrumented Records},
  pdfauthor={Caelum Mroczek; Abraham E Springer; Neha Gupta; Temuulen Sankey; Benjamin Lucas},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\setlength{\parindent}{6pt}
\begin{document}

\begin{frontmatter}
\title{Regional Base-Flow Index in Arid Landscapes Using Machine
Learning and Instrumented Records}
\author[1]{Caelum Mroczek%
\corref{cor1}%
}
 \ead{csm428@nau.edu} 
\author[1]{Abraham E Springer%
%
}

\author[2]{Neha Gupta%
%
}

\author[3]{Temuulen Sankey%
%
}

\author[4]{Benjamin Lucas%
%
}

\affiliation[1]{organization={School of Earth and Sustainability,
Northern Arizona University, Flagstaff, AZ, USA},,postcodesep={}}
\affiliation[2]{organization={Department of Hydrology and Atmospheric
Sciences, University of Arizona, Tucson, AZ, USA},,postcodesep={}}
\affiliation[3]{organization={School of Informatics, Computing, and
Cyber Systems, Northern Arizona University, Flagstaff, AZ,
USA},,postcodesep={}}
\affiliation[4]{organization={Department of Mathematics and Statistics,
Northern Arizona University, Flagstaff, AZ, USA},,postcodesep={}}

\cortext[cor1]{Corresponding author}

\begin{abstract}
Base flow, sustained by groundwater discharge, is a vital component of
river ecosystems, particularly in drylands, where water resources are
limited. This novel study analyzes the instrumented streamflow record in
Arizona to assess long-term base-flow index (BFI) trends across gauged
catchments. Results indicate that approximately 32\% of Arizona's
streamflow originates from groundwater discharge, with significant
spatial variability driven by landscape and climatic factors. Base-flow
relationships are analyzed with coincident trends in climate variables
such as precipitation, evapotranspiration, and temperature. Spatial and
climatic trends reveal variability in base-flow contributions, providing
insight into groundwater-surface water interactions in arid and
semi-arid landscapes. Building on this analysis, we applied machine
learning methods to predict BFI in ungauged basins, addressing the
challenges of Arizona's sparse streamgage network. Using the eXtreme
Gradient Boosting (XGBoost) algorithm trained on hydroclimate and
physiographic predictors, we estimated long-term BFI from 1991 to 2020.
This combined approach integrates observational data with predictive
modeling to enhance our understanding of base-flow processes and provide
a framework for water resource management in data-limited regions.
\end{abstract}

\end{frontmatter}

\subsection{Key Points}\label{key-points}

\begin{itemize}
\tightlist
\item
  Approximately 32\% of Arizona's streamflow originates from groundwater
  discharge.
\item
  XGBoost predicted long-term BFI in ungauged basins, filling dryland
  data gaps.
\item
  Baseflow trends track precipitation, emphasizing climate's role in
  dryland hydrology.
\end{itemize}

\section{Introduction}\label{sec-intro}

Dryland regions, encompassing arid, semi-arid, hyper-arid, and dry
sub-humid systems, cover 40\% of the Earth's land surface. These regions
are home to approximately 2 billion people globally and constitute the
largest terrestrial biome \citep{iucn-drylands-2019}. Despite supporting
diverse ecosystems and human populations, dryland regions face mounting
hydrologic challenges exacerbated by increasing urbanization, expanding
agricultural activities, and climate-induced amplification of
precipitation patterns \citep{taylor2013}. This water scarcity is
intensifying due to the compounding effects of climate variability and
increased groundwater extraction \citep{taylor2013}. Groundwater serves
as a vital resource in drylands for sustaining ecological functions and
supporting human livelihoods \citep{scanlon2006, yao2018}.

Base flow is the sustained portion of streamflow in the absence of
runoff that is derived from groundwater discharge
\citep{usgs-glossary-2018}. Base flow is critical to maintaining
seasonal low-flow regimes, supporting aquatic ecosystems, and
facilitating the transport of nutrients and chemicals. Base-flow
contribution to streamflow can be highly variable spatially
\citep{singh2018, bosch2017, beck2013}, and temporally
\citep{ficklin2016, tan2020}. Increasing groundwater extraction, changes
in land cover/land use, and changes in precipitation patterns due to
climate change affect the timing and volumes of base flow
\citep{tan2020, taylor2013}. Effective management of water quantity and
quality depends on understanding seasonal and interannual base-flow
patterns and long-term changes in base-flow behavior.

The Base-Flow Index (BFI) is the ratio of the long-term mean base-flow
volume to the long-term total streamflow volume expressed as a
percentage. BFI serves as a normalized measure of groundwater
contribution interannually or between basins. BFI is determined by
hydrograph separation and is influenced by the climate and physiographic
characteristics of a catchment \citep{neff2005, beck2013, singh2018}.
Between catchments, base flow fluctuates according to changes in the
moisture content of the vadose zone, influenced by varying levels of
evapotranspiration and aquifer storage dynamics \citep{bosch2017}. Since
BFI calculations rely on instrumented stream records, it remains unknown
for ungauged catchments, which encompass most of the earth's land
surface \citep{fekete2007}. Addressing this information gap is integral
to approaching a comprehensive understanding of groundwater dynamics
globally. This study addresses this gap by analyzing long-term BFI
patterns across Arizona using instrumented streamflow records and
applying a machine learning model to predict BFI in ungauged catchments.

Advancements in machine learning provide tools to predict hydrologic
indices in ungauged basins, addressing the limitations of sparse
streamgage networks. To tackle the challenge of quantifying base-flow
indices in ungauged catchments, numerous studies have applied both
regression and machine learning methods. \citet{ahiablame2013} found
that using a regression model to estimate annual base flow of ungauged
catchments was reasonably easy and accurate. \citet{beck2013} overcame
the nonlinearity of basin characteristics and improved results of
multivariate analyses by using artificial neural networks (ANN) to
estimate BFI globally. \citet{singh2018} implemented a random forest
algorithm to predict long-term BFI for ungauged catchments across New
Zealand. These applications demonstrate the versatility and
effectiveness of machine learning in capturing complex ecohydrologic
dynamics and improving our understanding of groundwater contributions to
streamflow.

Previous studies have examined base-flow regionalization and synthesis
across various spatial scales, from global to continental
\citep{beck2013, santhi2008, ayers2022, singh2018}. Such large-scale
analyses often utilize generalized datasets and methodologies, resulting
in limited applicability to regions with unique hydrogeologic and
climatic conditions. Additionally, global and continental-scale studies
tend to rely on streamgage networks that disproportionately represent
large perennial rivers and regulated watersheds with dense human
populations, while underrepresenting arid and semi-arid regions
characterized by non-perennial flow regimes and smaller streams
\citep{krabbenhoft-2022}. Thus, their effectiveness in accurately
capturing groundwater-surface water interactions, particularly in
critically water-stressed dryland regions, remains constrained.

Despite advancements, existing large-scale studies have limited
applicability to the unique hydrogeologic conditions in arid and
semi-arid regions like Arizona, leaving uncertainty regarding
groundwater-surface water interactions in these critically
water-stressed areas. This study addresses this crucial gap by
developing a regionally tailored machine learning model specifically
designed to estimate BFI for Arizona's ungauged basins. Utilizing
hydrogeologic characteristics and hydroclimatic data, we estimate
long-term mean BFI (1991--2020) in ungauged catchments, filling spatial
gaps in the sparse streamgage network. Furthermore, regional trends in
base flow and BFI at instrumented sites are analyzed alongside
coincident climate trends in precipitation, reference evapotranspiration
(ETO), and temperature. The outcomes offer novel insights into
groundwater contributions to streamflow, supporting water resource
management in an area increasingly vulnerable to drought and climate
change impacts, and contributing broadly to the understanding of
hydrologic processes in dryland environments. The objectives of this
study are to (1) quantify long-term spatial and temporal trends in BFI
across Arizona, and (2) develop a predictive model using eXtreme
Gradient Boosting (XGBoost) to estimate BFI in ungauged basins based on
hydroclimatic and geospatial variables.

\section{Methods}\label{sec-methods}

\subsection{Study Area}\label{sec-study-area}

Arizona, located in the southwestern United States, spans approximately
295,253 km² and encompasses a diverse range of landscapes, elevations,
and climate regimes. The state includes portions of two major
physiographic regions: the Colorado Plateau in the northeast, the Basin
and Range province in the south and west, and a transitional zone, the
Central Highlands, between them. This heterogeneity results in
substantial variation in environmental conditions across the state. The
Colorado Plateau is characterized by high-elevation desert and mountain
woodlands, averaging 1,936 masl (6,352 ft), with mean temperatures
ranging from -6°C (20°F) to 26°C (80°F) and annual precipitation of
about 580 mm (23 in). In contrast, the Basin and Range region is lower
in elevation, averaging 490 masl (1,600 ft), and features a semi-arid to
arid climate, with temperatures ranging from 15°C (60°F) to 43°C (110°F)
and an average annual precipitation of 200 mm (8 in)
\citep{az_climate2024}. The Central Highlands feature a mix of
mountainous terrain and interspersed basins, adding to the state's
topographic and hydroclimatic complexity.

\begin{figure}

\centering{

\includegraphics{images/updated/StudyArea_20250404.png}

}

\caption{\label{fig-study-area}Map of Arizona and US Geological Survey
(USGS) streamgages used in this study. 8-digit HUC subbasin boundaries
and physiographic regions shown.}

\end{figure}%

Arizona's hydrology varies seasonally and spatially between its
physiographic regions. In summer, localized and intense convective
storms stem from the North American Monsoon; in winter, orographic
precipitation is delivered by Pacific frontal systems
\citep{eastoe2019mtnblock}. While monsoonal precipitation can account
for up to 50\% of annual precipitation, evaporation and dry preceding
soil properties leads to most precipitation becoming runoff
\citep{sheppard2002}. As such, 94\% of streams in Arizona are ephemeral
or intermittent \citep{epa_AZephemeral}. Much of the hydrology of
Arizona is snow-melt derived, driven by spring melt from the
high-elevation Colorado Plateau winter snowpack. Though winter
precipitation accounts for only 30\% of annual precipitation totals, it
provides the majority of water for natural reservoirs
\citep{sheppard2002}.

\subsection{Data}\label{sec-data}

Daily observed streamflow data obtained from the United States
Geological Survey (USGS) National Water Information System (NWIS) were
used in this study. Streamgages were selected depending on criteria to
ensure the applicability of each site. Following the findings of
\citet{odonnell2016}, which determined that 8--10 years of calibration
data are necessary to account for climate variability in paired
watershed studies in the region, a minimum record length of 10 years was
required. Additionally, years with more than 30 missing days of
streamflow data were excluded from the analysis. Streamgages directly
downstream of major regulation (e.g., reservoirs, lakes, diversions)
were excluded, based on USGS annual water data reports and site metadata
\citep{usgs2010}. While some flow alteration is widespread, the focus
was on removing gauges with clear, immediate regulatory impacts. As
such, streamgages along the Colorado River were omitted because they
represent managed flows governed by the Colorado River Compact. After
applying these selection criteria, 205 USGS streamgages with acceptable
periods of record were included in the study
(Figure~\ref{fig-study-area}). Periods of record ranged from 10 to 112
years, with a median of 28 years.

Our data selection criteria ensure a robust analysis but also highlight
notable spatial gaps that our machine learning model can address.
Arizona has 184 active USGS streamflow stations (as of 2024) covering an
area of 295,253 km². For comparison, Indiana (a humid state in the U.S.)
maintains 189 active stations within a significantly smaller area of
94,326 km². This results in a streamgage density of approximately 2.004
gauges per 1,000 km² in Indiana, more than three times greater than
Arizona's density of 0.623 gauges per 1,000 km². Such disparities in
gauge coverage are typical for dryland regions globally
\citep{krabbenhoft-2022}, underscoring the necessity and relevance of
this type of modeling approach in arid and semi-arid environments.

Watersheds across the United States are delineated by the USGS using a
hydrologically-defined network. This system delineates the country using
hierarchical hydrologic unit codes (HUCs), where each subsequent basin
includes the digits of the enclosing basin. Here, 8-digit HUCs (HUC 8s)
are used to divide Arizona into 84 sub-basins that are fully or
partially in the state (Figure~\ref{fig-study-area}). These HUC 8
sub-basins are analogous to medium-sized river basins and are defined by
surface water characteristics.

Annual precipitation and temperature data came from the PRISM climate
group at Oregon State University at a resolution of 4 km
(\url{https://prism.oregonstate.edu;} \citep{daly2008}). The PRISM
dataset provides valuable insights into regional climate in ungauged
regions and has been shown to perform well across the southwestern US
\citep{buban_PRISM}. Instead of the water year, PRISM data uses a
calendar-year format, which was adopted for consistency in the water
balance. Although this may introduce challenges in the annual estimates
due to inter-annual snow storage, the use of long-term annual averages
reduces potential errors \citep{reitz2017}.

Annual reference evapotranspiration (ET\textsubscript{O}) data came from
TerraClimate, a 4-km grid climatological data set
\citep{abatzoglou2018}. TerraClimate uses a Penman-Monteith approach to
generate a reference evapotranspiration. The ET\textsubscript{O} values
were calculated assuming a reference grass surface across the landscape
with unlimited water. In the drylands of the southwestern US,
ET\textsubscript{O} typically exceeds precipitation annually
\citep{zomer2022}.

Basin elevation was derived from a 30-meter resolution Digital Elevation
Model (DEM) from the USGS National Elevation Dataset. The DEM of Arizona
was used to derive key basin characteristics: basin area, average slope,
and the proportion of each basin oriented toward north or south aspects.
Various geospatial variables, such as aspect, were disaggregated then
averaged to assess the areal percentage of each sub-variable within
individual HUC 8 basins. By calculating these percentages, we derived a
more comprehensive understanding of landscape composition across space.
Land cover was acquired from USGS-NLCD (National Land Cover Database),
hydrologic soil group from SSURGO (Soil Survey Geographic Database), and
underlying geology and karst from USGS were all similarly averaged
across the basins. Aggregating variables to align with the HUC 8
boundaries allowed for more precise predictions of BFI by integrating
spatial variations within each basin.

\subsection{Base-flow separation}\label{sec-bf_sep}

Directly measuring base flow and base-flow index (BFI) presents unique
challenges \citep{eckhardt2008}. Base flow is a physical hydrologic
process representing groundwater contributions to streamflow, while BFI
is a dimensionless, statistical metric that estimates the proportion of
streamflow derived from base flow. Because BFI must be derived using a
base-flow separation technique, its value is inherently method-dependent
and sensitive to the separation approach used \citep{beck2013}. A
variety of techniques have been developed to estimate base flow,
including tracer studies \citep{gonzales2009}, graphical interpolation
methods \citep{instituteofhydrology1980, sloto1996}, and digital filters
\citep{arnold1995, eckhardt2005, lyne1979, nathan1990}. Each of these
methods varies in its applicability depending on spatial scale, record
length, and study objectives. Although the choice of separation method
and parameterization introduces some subjectivity, these filters have
been shown to yield reliable estimates when applied consistently within
a study domain
\citep{chapman_1999, eckhardt2005, instituteofhydrology1980, ayers2022}.
Numerous studies have compared these separation techniques
\citep[e.g.,][]{eckhardt2005, eckhardt2008, nathan1990}; however, this
study does not evaluate the relative performance of different methods.

Base flow was calculated using a single-parameter, recursive digital
filter technique from \citet{nathan1990}. This base-flow separation
technique is based on a recursive digital filter used in signal analysis
that separates high-frequency signals (quickflow) from low-frequency
signals (base flow) \citep{lyne1979}. \citet{eckhardt2023} noted that
recursive digital filters lack a physical basis, but as the method is
easy to automate, objective, and repeatable, it is appropriate for a
regional-scale study. The Lyne-Hollick filter has been used in multiple
studies \citep{ARNOLD200021, santhi2008, bloomfield2009, singh2018}, and
it takes the form of

\begin{equation}\phantomsection\label{eq-bf-separate}{
b = \alpha b_{k-1} + \frac{1-\alpha}{2}(Q_k + Q_{k-1})
}\end{equation}

where \(b\) is base flow, \(\alpha\) is the filter parameter, \(Q\) is
the total streamflow, and \(k\) is the time step. A filter parameter
\(\alpha\) of 0.925 was used as in \citet{nathan1990} and
\citet{fuka2014ecohydrology}. The filter was run three times (forward,
backward, forward) to attenuate the base-flow signal.

Two BFI metrics were used in this study: (1) annual BFI, calculated for
each calendar year using daily streamflow records, and (2) long-term
BFI, defined as the average of annual BFI values over the period of
record at each site. Annual BFI reflects interannual variability in
base-flow contributions, while long-term BFI represents a time-averaged
metric of groundwater influence. The machine learning model was trained
to predict annual BFI values, which were then aggregated to produce
long-term BFI estimates.

\subsection{Machine Learning}\label{machine-learning}

The implementation of machine learning models to predict hydrologic
indices has been successful in past studies
\citep{singh2018, schmidt_ML_2020, Rozos2021Machine}. In this work, we
used the eXtreme Gradient Boosting (XGBoost) algorithm
\citep{chen2016xgboost} to predict BFI at ungauged locations using
catchment characteristics as predictors Table~\ref{tbl-predictors}. The
XGBoost algorithm is a decision tree-based ensemble algorithm, which can
be adapted for either regression or classification problems. This
algorithm iteratively builds an ensemble of decision trees, where each
tree corrects errors from previous trees to improve predictions
\citep{chen2016xgboost}. Its efficiency, scalability, and robustness
have made it increasingly popular in recent years, with successful
applications in environmental modeling tasks such as streamflow
forecasting \citep{szczepanek2022daily, ni2020streamflow} and land
use/land cover classification \citep{Georganos2018Very}.

XGBoost operates by leveraging gradient boosting on decision tree
algorithms, combining multiple low-variance models to produce a robust
overall prediction. Gradient boosting works iteratively: the initial
tree is trained on the target values, while subsequent trees are trained
on the residual errors of the preceding tree. Each tree is assigned a
weight based on its contribution to reducing error, and these weights
are used to determine the influence of each tree in the final model. The
ultimate prediction is made by aggregating the outputs of all \(n\)
weighted trees in the ensemble. In this study, the trained XGBoost model
is used to predict BFI in ungauged catchments based on geospatial and
hydroclimate predictor variables (Table~\ref{tbl-predictors}). Certain
features were further subdivided according to their areal coverage
within each basin (e.g.~land cover was divided into 16 subdivisions).
This approach allowed the model to capture finer-scale spatial
variability and improve predictive accuracy.

\begin{longtable}[]{@{}llll@{}}
\caption{Basin-characteristic variables used as initial features in
XGBoost model. Starred features are maintained in the final,
dimensionality-reduced model.}\label{tbl-predictors}\tabularnewline
\toprule\noalign{}
& Variable & Source & Geoprocessing \\
\midrule\noalign{}
\endfirsthead
\toprule\noalign{}
& Variable & Source & Geoprocessing \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
Hydroclimate & Precipitation* & PRISM & Basin average \\
& Mean Temperature* & PRISM & Basin average \\
& Reference Evapotranspiration* & TerraClimate & Basin average \\
Geospatial & Elevation* & DEM & Basin average \\
& Area & DEM & Basin average \\
& Slope & DEM & Basin average \\
& Aspect & DEM & Percent areal coverage \\
& Land Cover* & NLCD & Percent areal coverage \\
& Hydrologic Soil Group* & SSURGO & Percent areal coverage \\
& Geology & USGS & Percent areal coverage \\
& Karst & USGS & Percent areal coverage \\
\end{longtable}

Our training dataset consisted of 7,724 site-year observations of annual
BFI, where each observation represents a single water year at a given
streamgage. These observations were paired with 45 predictor variables
derived from basin characteristics and climate data. To optimize the
model's performance, we first conducted an exhaustive grid search
combined with 5-fold cross-validation to identify the optimal
hyperparameter values. The hyperparameters evaluated included the
learning rate (\(\eta\)), minimum split loss (\(\gamma\)), maximum tree
depth, minimum child weight, and the number of trees. While not an
exhaustive list of all possible XGBoost hyperparameters, this range of
values provided sufficient variation to ensure the selection of a
high-performing model. The optimal hyperparameters were determined to
be: 700 trees, a learning rate (\(\eta\)) of 0.05, a minimum split loss
(\(\gamma\)) of 0.075, a maximum tree depth of 7, and a minimum child
weight of 5. Using these values, the XGBoost model was trained on the
dataset with 10-fold cross-validation.

\(K\)-fold cross-validation provides an unbiased estimate of a model's
accuracy on unseen data, while also insuring against overfitting or
underfitting. In this approach, the data are randomly divided into \(k\)
folds of equal size. The model is trained on \(k-1\) folds and tested on
the remaining fold, referred to as the validation set. This process is
repeated \(k\) times, with each fold serving as the validation set
exactly once. Each iteration trains an independent model with the same
hyperparameters but using a different subset of training data. By
averaging the model performance across all \(k\) folds, we achieve a
robust and reliable estimate of predictive accuracy. Root mean squared
error (RMSE) was used as the performance metric for both model
optimization and evaluation. RMSE provided a consistent and
interpretable measure of the model's accuracy throughout the training
and validation process.

\begin{figure}

\centering{

\includegraphics[width=1\textwidth,height=\textheight]{images/k-fold_fig.png}

}

\caption{\label{fig-k-fold}Schematic of \(k\)-fold cross validation. The
dataset is randomly divided into \(k\) stratified folds. Each fold
serves as the validation set once, while the remaining folds are
combined to create a training set for model development. Performance
metrics for the test set are calculated and recorded, and this process
is repeated for all \(k\) folds.}

\end{figure}%

\subsubsection{Feature Selection}\label{sec-feat-selection}

Machine learning models are prone to overfitting, especially when
provided with a large set of predictive features \citep{Ying_2019}.
Overfitting can degrade model performance on unseen data and increase
the demand for computational resources and memory storage
\citep{feat_selec2017}. Dimensionality reduction offers a robust
solution to these challenges and generally falls into two broad
categories: feature extraction and feature selection.

Feature extraction involves transforming the original dataset into a
lower-dimensional feature space. However, this process generates new
features that often lack the physical interpretability of the original
variables. In contrast, feature selection identifies a subset of the
original features, preserving their physical meaning while improving
model readability and interpretability \citep{feat_selec2017}. In this
study, supervised feature selection was employed to reduce the number of
predictors, which enhanced learning performance, reduced computational
costs, and mitigated overfitting.

To begin, an initial model was trained using the full feature set of 45
predictors (Table~\ref{tbl-predictors}). A feature selection method
based on feature importance was then applied to identify and remove less
relevant and noisy features. Feature importance scores quantify the
contribution of individual features---either positively or
negatively---to the model's predictions \citep{ML_interpretability2019}.
In this analysis, SHAP (SHapley Additive exPlanations) values were used
to compute feature importance scores \citep{SHAP_values}.

SHAP values is a method to explain the prediction of an individual
instance by calculating the contribution of each feature to that
prediction. The method is based on coalition game theory and is
discussed further in \citet{SHAP_values}. Here, SHAP values are used for
global interpretation of feature importance and feature effects on the
model. Global feature importance is produced by the absolute Shapley
values of each feature across the dataset, providing a list of features
in order of most to least important. Feature effects provide an
indication of the relationship between the value of a predicting feature
and its impact on the prediction.

\begin{figure}

\centering{

\includegraphics{images/shap_summary.png}

}

\caption{\label{fig-shap_values}SHAP value plot of features used in
final model. The number to the right of the feature name is the mean
absolute SHAP value. Land cover features are indicated by the percentage
of cover by each land cover type and soil types are defined by
hydrologic soil group.}

\end{figure}%

The ten most important features Figure~\ref{fig-shap_values} were
selected based on their SHAP values and used to train a subsequent model
Table~\ref{tbl-predictors}. This refined model demonstrated improved
performance and reduced computational time compared to the initial
model. The final model, trained on this optimized feature subset, was
ultimately used for the analysis presented here.

\subsection{Statistical Analyses}\label{statistical-analyses}

Statistical analyses were conducted on annual BFI values from
instrumented streamgages to identify site-specific temporal trends using
the Mann--Kendall nonparametric trend test
\citep{kendall1970, mann1945nonparametric}. This test detects monotonic
trends in non-normally distributed data and assumes the absence of
autocorrelation. This test is widely used in hydrologic trend studies
\citep{ficklin2016, ayers2019, woodhouse_udall22}.

To account for potential autocorrelation in annual BFI time series at
each instrumented site, we applied the Durbin--Watson test
\citep{durbin1950testing}. Four streamgages exhibited significant
autocorrelation; these sites were excluded from the trend analysis to
avoid inflated variance in the Mann--Kendall statistic and potential
bias in trend detection \citep{MK_Hamed1998}. As all autocorrelated
series were removed, pre-whitening using the Hamed--Rao method was
unnecessary \citep{MK_Hamed1998}. Trends with a \(\alpha \le 0.05\) are
considered significant.

\section{Results}\label{results}

\subsection{Observed BFI in Gauged
Catchments}\label{observed-bfi-in-gauged-catchments}

The long-term BFI for the 205 gauged reaches across Arizona is
illustrated in Figure~\ref{fig-instrumented-bfi} . The long-term mean
BFI is 0.32, indicating that \textasciitilde32\% of long-term streamflow
in Arizona likely originates from groundwater discharge and other
delayed sources. The highest long-term BFI values (\textgreater0.9) are
found along the Grand Canyon in northwestern Arizona, where highly
karstic geology facilitates rapid subsurface flow to surface water and
spring outlets \citep{chambless_deep-karst_2023}. High long-term BFI
values (\textgreater0.8) are also found at the spring-fed headwaters of
the Verde River and Fossil Creek, which have similar highly-karstic,
snowmelt-driven recharge areas.

The stream reaches of the Little Colorado River Basin (northeastern
Arizona) indicate low long-term BFI values (\textless{} 0.2). This is
likely due to low-yielding perched aquifers underlying the Defiance
Plateau in northeastern Arizona, which are hydrologically connected to
surface streams, while the high-yield, confined regional aquifer is much
deeper \citep{blanchard02}.

\begin{figure}

\centering{

\includegraphics{images/updated/BFI_Instrumented_20240404.png}

}

\caption{\label{fig-instrumented-bfi}Long-term BFI for the period of
record from instrumented stream flow data.}

\end{figure}%

\subsubsection{Trends in BFI}\label{sec-bfi-trends}

Trends in long-term BFI over the period of record for each streamgage
are illustrated in Figure~\ref{fig-instrumented-trend} and summarized in
Table~\ref{tbl-trends}. Long-term BFI trends were assessed at all
instrumented sites using the Mann--Kendall test. As BFI quantifies the
proportion of streamflow attributed to base flow, its long-term trends
inherently reflect changes in subsurface contributions relative to total
flow. A 72.2\% coincidence rate between significant base flow and BFI
trends supports the interpretation that changes in BFI are largely
driven by underlying base-flow dynamics. Statistically significant
long-term BFI trends were observed across numerous sites, with spatial
variation highlighting regional differences in hydrologic response.

Figure~\ref{fig-instrumented-trend} illustrates the spatial variation in
long-term BFI trends across the study area. Statistically significant
decreasing trends are observed at 16.1\% of sites, while increasing
trends are found at 8.8\% of sites. No consistent regional patterns are
evident.

\begin{figure}

\centering{

\includegraphics{images/updated/BFITrends_20250404.png}

}

\caption{\label{fig-instrumented-trend}Trends in BFI over full period of
record for instrumented sites used in this study. Red downward (blue
upward) arrows indicate a decreasing (increasing) trend at a
significance level of 5\%. White circles represent sites with no
statistically significant trends.}

\end{figure}%

\subsubsection{Classification Trends}\label{classification-trends}

Classifications presented in Table~\ref{tbl-trends} were determined
based on precipitation regime, physiographic region, climate, and slope.
The dominant precipitation regime (monsoon vs.~snowmelt) was identified
by analyzing streamflow hydrographs for each station, focusing on peak
flow periods during the monsoon season (July--September) and the
snowmelt season (March--June). Physiographic region was assigned based
on which region the streamgage is located. Climate classifications were
defined as warm (above the long-term median temperature of Arizona),
cool (below the long-term median temperature), wet (above the long-term
median precipitation), and dry (below the long-term median
precipitation). Slope was categorized as high (above the median slope)
and low (below the median slope).

Statistically significant decreasing trends in long-term BFI were more
common than increasing trends across all site classifications
(Table~\ref{tbl-trends}). While decreasing trends dominate, both
increasing and decreasing trends are observed within each
classification. Monsoon-dominated regions exhibit a higher proportion of
significant negative trends (24.1\%) compared to snowmelt-dominated
regions (10.2\%), suggesting that monsoon-dominated systems are more
consistently correlated with declining long-term BFI. Among climate
classifications, warm-dry climates have the highest proportion of
negative trends (20.0\%), followed by warm-wet climates (19.4\%),
indicating that regions with higher temperatures are more prone to BFI
declines. Low-slope regions show a greater prevalence of negative trends
(20.4\%) compared to high-slope regions (11.8\%). This suggests that
flatter areas may be more susceptible to base-flow reductions,
potentially due to differences in hydrologic connectivity and recharge
dynamics.

\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1667}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1667}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1667}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1667}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1667}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1667}}@{}}
\caption{Comparison of trends for BFI for all sites split by various
classifications. Only sites with a significant (\(\alpha \le 0.05\))
trend are included here as established by a Mann-Kendall test for
monotonic trends across the full period of record. n is the number of
sites, n\_pos (n\_neg) is the number of sites with positive (negative)
trends, perc\_pos (perc\_neg) is the percentage of n with a positive
(negative) trend.}\label{tbl-trends}\tabularnewline
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
\textbf{Classification Group}
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
\textbf{n}
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
\textbf{n\_pos}
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
\textbf{n\_neg}
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
\textbf{perc\_pos}
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
\textbf{perc\_neg}
\end{minipage} \\
\midrule\noalign{}
\endfirsthead
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
\textbf{Classification Group}
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
\textbf{n}
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
\textbf{n\_pos}
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
\textbf{n\_neg}
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
\textbf{perc\_pos}
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
\textbf{perc\_neg}
\end{minipage} \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
Precipitation - Monsoon Dominated & 87 & 8 & 21 & 0.092 & 0.241 \\
Precipitation - Snowmelt Dominated & 118 & 9 & 12 & 0.076 & 0.102 \\
Physiographic Region - Basin and Range & 156 & 14 & 26 & 0.090 &
0.167 \\
Physiographic Region - Colorado Plateau & 49 & 3 & 7 & 0.061 & 0.143 \\
Climate - Warm-Wet & 31 & 2 & 6 & 0.065 & 0.194 \\
Climate - Warm-Dry & 55 & 6 & 11 & 0.109 & 0.200 \\
Climate - Cool-Wet & 74 & 4 & 9 & 0.054 & 0.122 \\
Climate - Cool-Dry & 45 & 5 & 7 & 0.111 & 0.156 \\
Slope - High & 102 & 10 & 12 & 0.098 & 0.118 \\
Slope - Low & 103 & 7 & 21 & 0.068 & 0.204 \\
\end{longtable}

\subsubsection{Coincident Climate
Trends}\label{coincident-climate-trends}

Coincident trends are defined as those in which long-term BFI trends
match in direction with climate variable trends over the same period.
Results are summarized in Table~\ref{tbl-coincident-trends}. This
analysis includes both significant and non-significant trends, which is
appropriate where the influence of complex, interconnected processes may
not always manifest as statistically significant patterns over limited
observational periods \citep{ficklin2016}. Coincidence was highest with
precipitation (64.88\% for base flow; 53.17\% for BFI), followed by ETO
and then temperature. Temperature trends most often show an inverse
relationship---i.e., warming associated with declining BFI and base
flow.

Given the variations in the period of record across the instrumented
network (see Figure S2), we analyzed trends of long-term BFI in relation
to coincident trends in climate variables. Trends are classified as
coincident when the direction of the climate variable trend (positive or
negative) aligns with the trend observed in BFI
(Table~\ref{tbl-coincident-trends}).

\begin{longtable}[]{@{}lcc@{}}
\caption{Coincident trends of climate variables (ET\textsubscript{O},
precipitation, temperature) with BFI
trends.}\label{tbl-coincident-trends}\tabularnewline
\toprule\noalign{}
& \textbf{Climate Variable} & \textbf{Coincidence Percentage (\%)} \\
\midrule\noalign{}
\endfirsthead
\toprule\noalign{}
& \textbf{Climate Variable} & \textbf{Coincidence Percentage (\%)} \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
BFI & ET\textsubscript{O} & 44.88 \\
& Precipitation & 53.17 \\
& Temperature & 47.32 \\
\end{longtable}

\subsection{BFI of Ungauged
Catchments}\label{bfi-of-ungauged-catchments}

\subsubsection{Model Validation}\label{model-validation}

Predicted values of annual BFI are plotted against observed values for
the entire period of record of the instrumented dataset in
Figure~\ref{fig-actual_predicted} . The agreement between ``out-of-bag''
predictions (blind cross-validation, treating each site as ungauged) and
observed values indicate strong model performance across the full
dataset (R\textsuperscript{2} = 0.764). The overall RMSE is 0.129 and
the overall percent bias (pbias) is -5.6\%.

Model performance metrics across various classifications are summarized
in Table~\ref{tbl-performance}. These metrics show consistent model
performance across spatial and climatic classifications. However, the
negative pbias values across all classifications indicate a systematic
underprediction of annual BFI. Categories with relatively lower
R\textsuperscript{2} and Nash-Sutcliffe Efficiency (NSE) values also
exhibit higher biases, reflecting weaker model performance in those
contexts.

\begin{figure}

\centering{

\includegraphics{images/actual-predicted.png}

}

\caption{\label{fig-actual_predicted}Linear relationship between
observed BFI and predicted BFI. The solid line is the 1:1 line, the
dashed, green line is regressed to the data.}

\end{figure}%

\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.1250}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.1250}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.1250}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.1250}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.1250}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.1250}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.1250}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.1250}}@{}}
\caption{Performance of model predictions for annual BFI for all sites
split by various classifications. n is number of observations,
R\textsuperscript{2} is the coefficient of determination of a linear
regression, MSE is mean-squared-error, RMSE is root-mean-squared-error,
MAE is mean-absolute-error, NSE is Nash-Sutcliffe efficiency, and pbias
is percent bias.}\label{tbl-performance}\tabularnewline
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
\textbf{Classification Group}
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
\textbf{n}
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
\textbf{R2}
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
\textbf{MSE}
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
\textbf{RMSE}
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
\textbf{MAE}
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
\textbf{NSE}
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
\textbf{pbias}
\end{minipage} \\
\midrule\noalign{}
\endfirsthead
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
\textbf{Classification Group}
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
\textbf{n}
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
\textbf{R2}
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
\textbf{MSE}
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
\textbf{RMSE}
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
\textbf{MAE}
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
\textbf{NSE}
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
\textbf{pbias}
\end{minipage} \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
Climate - Monsoon Dominated & 3039 & 0.633 & 0.016 & 0.126 & 0.074 &
0.619 & -13.7 \\
Climate - Snowmelt Dominated & 4685 & 0.733 & 0.015 & 0.121 & 0.087 &
0.725 & -3.5 \\
PhysRegion - Basin\&Range & 6147 & 0.733 & 0.016 & 0.127 & 0.084 & 0.724
& -6.3 \\
PhysRegion - CO Plateau & 1577 & 0.846 & 0.011 & 0.104 & 0.073 & 0.843 &
-3.8 \\
Climate - Warm-Wet & 1506 & 0.693 & 0.014 & 0.117 & 0.077 & 0.685 &
-8.2 \\
Climate - Warm-Dry & 2351 & 0.693 & 0.022 & 0.147 & 0.092 & 0.675 &
-11.9 \\
Climate - Cool-Wet & 2350 & 0.738 & 0.011 & 0.106 & 0.078 & 0.736 &
-1.7 \\
Climate - Cool-Dry & 1517 & 0.831 & 0.012 & 0.111 & 0.078 & 0.827 &
-4.3 \\
Slope - High & 3795 & 0.776 & 0.012 & 0.111 & 0.079 & 0.771 & -3.3 \\
Slope - Low & 3929 & 0.724 & 0.018 & 0.133 & 0.085 & 0.713 & -9.1 \\
\end{longtable}

\subsubsection{Predictor Importance}\label{sec-predictor-importance}

The predictors used to estimate annual BFI at ungauged sites were
evaluated for their importance in the final XGBoost model, as
illustrated in Figure~\ref{fig-shap_values}. The most influential
feature for predicting annual BFI is basin elevation. While elevation
itself does not directly affect base-flow characteristics, it has
consistently been identified as a key predictor in previous BFI studies
\citep{singh2018, beck2013}. The importance of elevation aligns with
findings from \citet{beck2013}, highlighting its role as a proxy for
climate variables such as temperature, precipitation, and snowpack
duration. Seasonal snowpack duration, in particular, has been shown to
strongly correlate with springflow and groundwater recharge in this
region \citep{donovan_karst_2022}.

Land cover and land use predictors also play a significant role in
annual BFI estimation. Analysis of SHAP values indicates that a higher
percentage of evergreen forest positively influences BFI predictions,
while higher proportions of shrubland and developed land exert a
negative influence. Similarly, hydrologic soil types show distinct
trends in their impact on BFI. Soil Type C, characterized by moderately
high runoff potential (20-40\% clay), tends to negatively influence BFI.
In contrast, Soil Type A, which has low runoff potential and facilitates
rapid water infiltration, exhibits a mixed influence \citep{nrcs2009}.

\subsubsection{Predicted BFI in Ungauged
Basins}\label{predicted-bfi-in-ungauged-basins}

The regionalized (HUC-8) long-term BFI (1991--2020) is shown in
Figure~\ref{fig-bfi-huc}. Basins with high long-term BFIs, such as those
along the Grand Canyon in the northwestern part of the study area,
indicate greater surface water and groundwater interaction. Elevated BFI
values are also observed along portions of the Mogollon Rim, a heavily
forested region with high precipitation that marks the transition
between physiographic regions. Additionally, headwater regions of
perennial rivers tend to exhibit higher long-term BFI values. In
contrast, low long-term BFI values are found in areas like the Defiance
Plateau in northeastern Arizona and the arid southern regions of the
state.

\begin{figure}

\centering{

\includegraphics{images/BFI_HUC8_20250525.png}

}

\caption{\label{fig-bfi-huc}Predicted long-term BFI values for 8-digit
HUC (1991-2020). Bolded HUCs indicate those without a streamgage.}

\end{figure}%

\section{Discussion}\label{discussion}

Several previous studies have examined regional and continental patterns
of BFI, offering useful context for our findings despite limited
geographic overlap. \citet{beck2013} developed a global map of BFI using
an artificial neural network trained on 1,862 U.S. streamgages from the
MOPEX dataset \citep{mopex_2006}. Only ten of these gauges overlap with
our Arizona study area, limiting direct comparison. Nonetheless, the
spatial distribution of BFI in Arizona from \citet{beck2013} shows
general agreement with our regional patterns. The higher BFI values
reported by \citet{beck2013} may reflect the influence of their
global-scale ANN model and associated generalizations. Similarly,
\citet{ayers2022} examined base-flow dynamics using GAGES-II sites
across CONUS, but their study included few gauges within Arizona. While
exact locations are not specified, their published figures suggests
regional agreement with our estimated BFI distributions.
\citet{santhi2008} employed interpolation methods for regional base-flow
estimation across CONUS, but specific gauge locations or data are not
available limiting comparison.````

In terms of long-term trends, our results align directionally with
\citet{ayers2022} , who found consistent monthly declines in base flow
from 1989--2019 across the U.S. Southwest. Although our analysis focused
on long-term trends in annual BFI, we observed a similar imbalance: 16\%
of sites showed significant decreasing trends compared to 8\% with
increasing trends. While we do not directly replicate their monthly
analysis, the directional agreement supports the broader conclusion that
groundwater contributions to streamflow are declining in many parts of
the region.

The existing streamgage networks in Arizona are limited for base-flow
analyses by their design and monitoring objectives. While the number of
streamgages has increased to meet regulatory imperatives, such as those
driven by the Clean Water Act, many gauges prioritize peak flow
monitoring rather than low-flow conditions critical for base-flow
studies \citep{flood-control2020}. Newer non-USGS gauges, such as those
installed by flood control districts (e.g., the ALERT system), are
generally tailored for flood detection rather than monitoring sustained
base flow. Recent USGS installations focus on addressing in-stream flow
rights and future monitoring improvements, though low-flow capabilities
remain limited.

This study addresses these monitoring gaps through an analysis of
observed BFI at instrumented streamgages, highlighting
groundwater-surface water interactions within Arizona's arid and
semi-arid landscapes. Observed spatial patterns in BFI clearly align
with known hydrogeological and climatic variations. High long-term BFI
values coincide with areas characterized by spring-fed streams and
pronounced physiographic transitions, such as the Grand Canyon and
Mogollon Rim. Conversely, lower observed BFI values correspond to
regions with limited groundwater contributions, notably the Defiance
Plateau and southern Arizona, emphasizing vulnerability in these arid
environments. A trend of high BFI in upstream reaches to lower BFI in
downstream reaches Figure~\ref{fig-instrumented-bfi} likely indicates a
transition from zones of groundwater discharge to zones of groundwater
recharge and may indicate a transition from gaining stream reaches to
losing stream reaches \citep{winter2007}.

Analysis of trends at instrumented sites revealed a predominance of
declining long-term BFI values, particularly within monsoon-dominated
precipitation regimes, warm-dry climates, and low-slope basins. These
negative trends likely reflect intensified climate stress, highlighting
areas most vulnerable to groundwater depletion and surface-water
scarcity. Precipitation was identified as the primary driver influencing
base-flow variability, with temperature and evapotranspiration trends
adding complexity. The weaker correlation between temperature and BFI
trends suggests that temperature impacts are moderated by other
environmental factors, underscoring the need for integrated climate and
land-management strategies to protect groundwater-dependent ecosystems.

Given the limited coverage provided by existing gauges, a common issue
in arid and semi-arid regions globally \citep{krabbenhoft-2022}, we
utilized a machine learning approach to estimate annual BFI in ungauged
catchments. The regional model exhibited strong predictive capability
(overall R²=0.764), though systematic underprediction (negative pbias)
was noted, particularly in monsoon-dominated and warm-dry climates.
These biases suggest areas for improvement in model performance,
potentially through the incorporation of additional predictors or more
refined regional hydrologic representations. Although creating separate
regional models (by physiographic region) was explored to enhance
predictive accuracy, the state-wide model consistently outperformed
these regionalized approaches, emphasizing the advantage of a
comprehensive, state-wide dataset.

Elevation emerged as the most significant predictor, reinforcing its
established role as a proxy for climate conditions influencing
groundwater recharge. This aligns with findings from previous studies
conducted in diverse landscapes, including New Zealand
\citep{singh2018}, across CONUS \citep{santhi2008}, and globally
\citep{beck2013}. The mixed influence observed for Soil Type A indicates
complexity in infiltration dynamics, suggesting that future research
could benefit from explicitly incorporating temporal variables such as
snowpack duration, soil moisture, and detailed land-cover
characteristics.

\section{Conclusions}\label{conclusions}

The long-term average BFI across Arizona is approximately 0.32,
highlighting that groundwater discharge significantly contributes to
surface water flows but exhibits substantial spatial variability. Our
machine learning model (XGBoost) effectively predicts annual BFI in
ungauged catchments (R² = 0.764), primarily driven by basin elevation,
land cover, and hydrologic soil characteristics. Regions such as the
Grand Canyon and Mogollon Rim exhibit the highest long-term BFI values,
signifying robust groundwater-surface water interactions, whereas areas
such as the Defiance Plateau and southern Arizona show consistently
lower BFI values, reflecting limited groundwater contributions.

Analysis of temporal trends indicates a prevailing pattern of declining
long-term BFI across the region, particularly in monsoon-dominated,
warm-dry, and low-slope areas. Precipitation emerged as the strongest
driver of base-flow variability, while temperature and
evapotranspiration added additional layers of complexity. The observed
trends underscore the sensitivity of Arizona's hydrological systems to
climatic variability and suggest increasing vulnerability under
projected climate change scenarios and other stresses to aquifers, such
as increased groundwater pumping.

This research highlights critical limitations in Arizona's current
streamgage network for monitoring base-flow processes, suggesting a need
for targeted investments in instrumentation capable of capturing
low-flow dynamics. Additionally, the demonstrated success of our
modeling framework accentuates its broader applicability to similarly
data-scarce dryland regions globally, providing a valuable tool for
water resource management and climate adaptation strategies. By
improving our understanding of groundwater-surface water interactions
under changing climatic conditions, this study contributes substantially
to addressing hydrological uncertainties in arid and semi-arid
environments.

\renewcommand\refname{References}
  \bibliography{references.bib}

\end{document}

### 2025-05-31 elsevier_MANUSCRIPT.qmd
Update elsevier_MANUSCRIPT.qmd

@@
 Using the eXtreme Gradient Boosting (XGBoost) algorithm trained and validated on observed hydroclimate and physiographic variables, we estimate long-term BFI from 1991 to 2020.
+Results indicate spatial heterogeneity in BFI trends, with decreasing baseflow most pronounced in warm-dry and monsoon-dominated catchments.
 This combined approach integrates observational data with predictive modeling to enhance our understanding of base-flow processes and provide a framework for water resource management in data-limited regions.
@@
 
-Basin elevation was derived from a 30-meter resolution Digital Elevation Model (DEM) from the USGS National Elevation Dataset.
-The DEM of Arizona was used to derive key basin characteristics: basin area, average slope, and the proportion of each basin oriented toward north or south aspects.
+Basin elevation was derived from a 30-meter resolution Digital Elevation Model (DEM) from the USGS National Elevation Dataset (NED).
+The DEM of Arizona was used to derive key basin characteristics: basin area, basin slope, and the proportion of each basin oriented toward north or south aspects.
 Various geospatial variables, such as aspect, were disaggregated then averaged to assess the areal percentage of each sub-variable within individual HUC 8 basins.
@@
 In this study, we used the eXtreme Gradient Boosting (XGBoost) algorithm [@chen2016xgboost] to estimate annual BFI at ungauged locations across Arizona using catchment characteristics as predictors [@tbl-predictors].
-XGBoost is a gradient boosting algorithm that constructs an ensemble of decision trees, where each tree iteratively corrects errors from previous trees to improve predictive accuracy [@chen2016xgboost].
+XGBoost is a gradient boosting algorithm that builds an ensemble of decision trees, where each tree iteratively corrects errors from previous trees to improve predictive accuracy [@chen2016xgboost].
 The model was trained on annual BFI observations in order to estimate annual BFI in ungauged basins.
@@
 
-The training dataset consisted of 7,724 site-year observations, where each observation represents annual BFI for a single water year at a given USGS streamgage.
-These observations were paired with 45 predictor variables derived from basin characteristics and climate data.
-Predictor selection was based on prior hydrologic studies, data availability across the study area, and pairwise correlation screening to minimize multicollinearity.
-The predictors included basin-averaged climate metrics (e.g., precipitation, temperature), topographic features (e.g., elevation, slope), and land surface attributes (e.g., land cover classes, soil type, karst presence).
+The training dataset consisted of 7,724 site-year observations, where each observation represents annual BFI for a single year at a given USGS streamgage.
+These observations are paired with 45 predictor variables derived from basin characteristics and climate data.
+Predictor selection was based on prior hydrologic studies and data availability across the study area.
+The predictors included basin-averaged climate metrics (e.g., precipitation, temperature), topographic features (e.g., elevation, slope), and land surface attributes (e.g., land cover classes, soil type, presence of karst).
 
@@
 SHAP values quantify both the magnitude and direction of each feature’s contribution to predictions.
-The ten most important features—based on mean absolute SHAP value—were retained in a reduced model, which showed improved predictive accuracy and efficiency over the full model and regional models.
-Sensitivity analyses tested region-specific models and smaller predictor subsets; however, the state-wide model using SHAP-prioritized features consistently outperformed alternatives.
+The ten most important features, based on SHAP value, were retained in a reduced model.
+The reduced model showed improved predictive accuracy and efficiency over the full model and regional models.
+Sensitivity analyses tested region-specific models and smaller predictor subsets; however, the state-wide model using SHAP-prioritized features outperformed alternatives.
+The final model was applied to all HUC8 basins across Arizona using the same predictor layers to estimate long-term BFI values.
 
@@
 The model dataset was evaluated using 10-fold cross-validation, ensuring robust performance estimates and guarding against overfitting [@fig-k-fold] .
-In each iteration (fold), the model was trained on 90% of the data and validated on the remaining 10%, with folds rotated until all data had been used for validation in separate models.
+In each iteration (fold), the model was trained on 90% of the data and validated on the remaining 10%, with folds being rotated until all of the data had been used for validation in separate models.
 Root mean squared error (RMSE) was used as the primary performance metric.
@@
 Hyperparameter tuning was performed through an internal grid search.
-The search evaluated combinations of key parameters including the number of trees, learning rate ($\eta$), minimum split loss ($\gamma$), tree depth, and minimum child weight.
-A full discussion of these hyperparametes can be found in @chen2016xgboost.
+The search evaluated combinations of parameters including the number of trees, learning rate ($\eta$), minimum split loss ($\gamma$), tree depth, and minimum child weight.
+A description of these hyperparametes can be found in @chen2016xgboost.
 The final model used 700 trees, $\eta$ of 0.05, $\gamma$ of 0.075, a maximum depth of 7, and a minimum child weight of 5.
-All parameters were selected for their consistent reduction in RMSE across validation folds.
+All parameters were selected for their reduction in RMSE across validation folds.
 
@@
 
-#### Trends in BFI {#sec-bfi-trends}
+#### Trends in Annual BFI {#sec-bfi-trends}
 
-Trends in long-term BFI over the period of record for each streamgage are illustrated in @fig-instrumented-trend and summarized in @tbl-trends.
-Long-term BFI trends were assessed at all instrumented sites using the Mann–Kendall test.
+Trends in annual BFI over the period of record for each streamgage are illustrated in @fig-instrumented-trend and summarized in @tbl-trends.
+Annual BFI trends were assessed at all instrumented sites using the Mann–Kendall test.
 As BFI quantifies the proportion of streamflow attributed to base flow, its long-term trends inherently reflect changes in subsurface contributions relative to total flow.
-A 72.2% coincidence rate between significant base flow and BFI trends supports the interpretation that changes in BFI are largely driven by underlying base-flow dynamics.
+A 72.2% coincidence rate between significant base-flow volume and BFI trends supports the interpretation that changes in BFI are largely driven by underlying base-flow dynamics.
 Statistically significant long-term BFI trends were observed across numerous sites, with spatial variation highlighting regional differences in hydrologic response.
@@
 This suggests that flatter areas may be more susceptible to base-flow reductions, potentially due to differences in hydrologic connectivity and recharge dynamics.
+Negative trends in low-slope regions may also be a result of reductions in water table height, assumed to follow topography, leading to a change in flow direction.
 
@@
 This analysis includes both significant and non-significant trends, which is appropriate where the influence of complex, interconnected processes may not always manifest as statistically significant patterns over limited observational periods [@ficklin2016].
-Coincidence was highest with precipitation (64.88% for base flow; 53.17% for BFI), followed by ETO and then temperature.
+Long-term BFI coincided most with precipitation (53.17%), followed by ET~O~ and then temperature.
 Temperature trends most often show an inverse relationship—i.e., warming associated with declining BFI and base flow.
@@
 
-The regionalized (HUC-8) long-term BFI (1991–2020) is shown in @fig-bfi-huc.
-Basins with high long-term BFIs, such as those along the Grand Canyon in the northwestern part of the study area, indicate greater surface water and groundwater interaction.
-Elevated BFI values are also observed along portions of the Mogollon Rim, a heavily forested region with high precipitation that marks the transition between physiographic regions.
-Additionally, headwater regions of perennial rivers tend to exhibit higher long-term BFI values.
-In contrast, low long-term BFI values are found in areas like the Defiance Plateau in northeastern Arizona and the arid southern regions of the state.
+Predicted long-term BFI (1991–2020) across Arizona's HUC-8 basins is shown in @fig-bfi-huc.
+Higher values were observed along the Grand Canyon corridor and parts of the Mogollon Rim, while lower values were found across the Defiance Plateau and in southern Arizona.
+Spatial variation generally aligns with physiographic regions and precipitation gradients, with upstream reaches of perennial rivers typically exhibiting higher BFI.
 
@@
 ## Discussion
+
+The existing streamgage networks in Arizona are limited for base-flow analyses by their design and monitoring objectives.
+While the number of streamgages has increased to meet regulatory imperatives, such as those driven by the Clean Water Act, many gauges prioritize peak flow monitoring rather than low-flow conditions critical for base-flow studies [@flood-control2020].
+Newer non-USGS gauges, such as those installed by flood control districts (e.g., the ALERT system), are generally tailored for flood detection rather than monitoring sustained base flow.
+Recent USGS installations focus on addressing in-stream flow rights and future monitoring improvements, though low-flow capabilities remain limited.
+
+Given the limited coverage provided by existing gauges, a common issue in arid and semi-arid regions globally [@krabbenhoft-2022], we utilized a machine learning approach to estimate annual BFI in ungauged catchments.
+The regional model exhibited strong predictive capability (overall R²=0.764), though systematic underprediction (negative pbias) was noted, particularly in monsoon-dominated and warm-dry climates.
+These biases highlight opportunities to improve model performance through incorporating additional predictors or refining regional hydrologic representations.
+Although creating separate regional models (by physiographic region) was explored to enhance predictive accuracy, the state-wide model consistently outperformed these regionalized approaches, emphasizing the advantage of a comprehensive, state-wide dataset.
+
+The spatial patterns in BFI reflect underlying differences in hydrogeologic and climatic drivers.
+In high-BFI regions such as the Mogollon Rim and Grand Canyon, fractured bedrock, karst systems, and deep regional aquifers provide substantial groundwater discharge to surface waters.
+These areas benefit from high orographic precipitation and forested landscapes that promote infiltration and recharge.
+In contrast, low-BFI regions such as the Defiance Plateau and southern Arizona feature shallow or disconnected groundwater systems, low precipitation, and high evaporative demand, all of which limit baseflow contributions.
+Land cover and soil properties further modulate these processes; areas with high-infiltration soils and perennial vegetation are more conducive to sustained baseflow, while impervious surfaces and fine-grained soils suppress it.
+The downstream decline in BFI [@fig-instrumented-bfi] is consistent with the transition from headwater recharge zones to more arid, losing stream reaches, highlighting the importance of hydrogeologic setting in shaping baseflow dynamics [@winter2007].
+
+Analysis of trends at instrumented sites revealed a predominance of declining long-term BFI values, particularly within monsoon-dominated precipitation regimes, warm-dry climates, and low-slope basins.
+These negative trends likely reflect intensified climate stress, highlighting areas most vulnerable to groundwater depletion and surface-water scarcity.
+Precipitation was identified as the primary driver influencing base-flow variability, with temperature and evapotranspiration trends adding complexity.
+The weaker correlation between temperature and BFI trends suggests that temperature impacts are moderated by other environmental factors, underscoring the need for integrated climate and land-management strategies to protect groundwater-dependent ecosystems.
+
+Elevation emerged as the most significant predictor, reinforcing its established role as a proxy for climate conditions influencing groundwater recharge.
+This aligns with findings from previous studies conducted in diverse landscapes, including New Zealand [@singh2018], across CONUS [@santhi2008], and globally [@beck2013].
+The mixed influence observed for Soil Type A indicates complexity in infiltration dynamics, suggesting that future research could benefit from explicitly incorporating temporal variables such as snowpack duration, soil moisture, and detailed land-cover characteristics.
 
@@
 While exact locations are not specified, their published figures suggests regional agreement with our estimated BFI distributions.
-@santhi2008 employed interpolation methods for regional base-flow estimation across CONUS, but specific gauge locations or data are not available limiting comparison.\`\`\`\`
+@santhi2008 employed interpolation methods for regional base-flow estimation across CONUS; however, the lack of gauge-level detail limits direct comparison.
 
@@
 While we do not directly replicate their monthly analysis, the directional agreement supports the broader conclusion that groundwater contributions to streamflow are declining in many parts of the region.
-
-The existing streamgage networks in Arizona are limited for base-flow analyses by their design and monitoring objectives.
-While the number of streamgages has increased to meet regulatory imperatives, such as those driven by the Clean Water Act, many gauges prioritize peak flow monitoring rather than low-flow conditions critical for base-flow studies [@flood-control2020].
-Newer non-USGS gauges, such as those installed by flood control districts (e.g., the ALERT system), are generally tailored for flood detection rather than monitoring sustained base flow.
-Recent USGS installations focus on addressing in-stream flow rights and future monitoring improvements, though low-flow capabilities remain limited.
-
-This study addresses these monitoring gaps through an analysis of observed BFI at instrumented streamgages, highlighting groundwater-surface water interactions within Arizona’s arid and semi-arid landscapes.
-Observed spatial patterns in BFI clearly align with known hydrogeological and climatic variations.
-High long-term BFI values coincide with areas characterized by spring-fed streams and pronounced physiographic transitions, such as the Grand Canyon and Mogollon Rim.
-Conversely, lower observed BFI values correspond to regions with limited groundwater contributions, notably the Defiance Plateau and southern Arizona, emphasizing vulnerability in these arid environments.
-A trend of high BFI in upstream reaches to lower BFI in downstream reaches [@fig-instrumented-bfi] likely indicates a transition from zones of groundwater discharge to zones of groundwater recharge and may indicate a transition from gaining stream reaches to losing stream reaches [@winter2007].
-
-Analysis of trends at instrumented sites revealed a predominance of declining long-term BFI values, particularly within monsoon-dominated precipitation regimes, warm-dry climates, and low-slope basins.
-These negative trends likely reflect intensified climate stress, highlighting areas most vulnerable to groundwater depletion and surface-water scarcity.
-Precipitation was identified as the primary driver influencing base-flow variability, with temperature and evapotranspiration trends adding complexity.
-The weaker correlation between temperature and BFI trends suggests that temperature impacts are moderated by other environmental factors, underscoring the need for integrated climate and land-management strategies to protect groundwater-dependent ecosystems.
-
-Given the limited coverage provided by existing gauges, a common issue in arid and semi-arid regions globally [@krabbenhoft-2022], we utilized a machine learning approach to estimate annual BFI in ungauged catchments.
-The regional model exhibited strong predictive capability (overall R²=0.764), though systematic underprediction (negative pbias) was noted, particularly in monsoon-dominated and warm-dry climates.
-These biases suggest areas for improvement in model performance, potentially through the incorporation of additional predictors or more refined regional hydrologic representations.
-Although creating separate regional models (by physiographic region) was explored to enhance predictive accuracy, the state-wide model consistently outperformed these regionalized approaches, emphasizing the advantage of a comprehensive, state-wide dataset.
-
-Elevation emerged as the most significant predictor, reinforcing its established role as a proxy for climate conditions influencing groundwater recharge.
-This aligns with findings from previous studies conducted in diverse landscapes, including New Zealand [@singh2018], across CONUS [@santhi2008], and globally [@beck2013].
-The mixed influence observed for Soil Type A indicates complexity in infiltration dynamics, suggesting that future research could benefit from explicitly incorporating temporal variables such as snowpack duration, soil moisture, and detailed land-cover characteristics.
 

### 2025-05-31 elsevier_MANUSCRIPT.qmd
new draft

@@
 We applied SHAP (SHapley Additive exPlanations) values [@SHAP_values] to assess global feature importance and rank predictors.
-SHAP values quantify both the magnitude and direction of each feature’s contribution to predictions.
+SHAP was chosen for its interpretability and ability to account for feature interactions [@SHAP_values].
+The values quantify both the magnitude and direction of each feature’s contribution to predictions.
 The ten most important features, based on SHAP value, were retained in a reduced model.
@@
 The number to the right of the feature name is the mean absolute SHAP value.
-Land cover features are indicated by the percentage of cover by each land cover type and soil types are defined by hydrologic soil group.](images/shap_summary.png){#fig-shap_values alt="SHAP value plot of features used in final model.
-The number to the right of the feature name is the mean absolute SHAP value.
-Land cover features are indicated by the percentage of cover by each land cover type and soil types are defined by hydrologic soil group."}
+Land cover features are indicated by the percentage of cover by each land cover type and soil types are defined by hydrologic soil group.](images/shap_summary.png){#fig-shap_values}
 
@@
 Model performance on unseen data was strong, with an overall R² of 0.764.
-Additional model validation results are presented in section 3.2.1.
 
-![Schematic of $k$-fold cross validation.
-The dataset is randomly divided into $k$ stratified folds.
+To contextualize model accuracy, we compared XGBoost to two baseline methods: a simple linear regression and an inverse distance weighting (IDW) interpolation.
+XGBoost reduced RMSE by 43% relative to the linear model and 30% relative to IDW, while improving NSE from 0.25 (linear) and 0.45 (IDW) to 0.75, indicating substantially stronger predictive accuracy and spatial generalization.
+
+![Schematic of \$k\$-fold cross validation.
+The dataset is randomly divided into \$k\$ stratified folds.
 Each fold serves as the validation set once, while the remaining folds are combined to create a training set for model development.
-Performance metrics for the test set are calculated and recorded, and this process is repeated for all $k$ folds.](images/k-fold_fig.png){#fig-k-fold alt="Schematic of $k$-fold cross validation.
-The dataset is randomly divided into $k$ stratified folds.
-Each fold serves as the validation set once, while the remaining folds are combined to create a training set for model development.
-Performance metrics for the test set are calculated and recorded, and this process is repeated for all $k$ folds." fig-align="center" width="100%"}
+Performance metrics for the test set are calculated and recorded, and this process is repeated for all \$k\$ folds.](images/k-fold_fig.png){#fig-k-fold fig-align="center" width="100%"}
 
@@
 | **Classification Group** | **n** | **R2** | **MSE** | **RMSE** | **MAE** | **NSE** | **pbias** |
-|---------|---------|---------|---------|---------|---------|---------|---------|
+|-------------------------|-------|-------|-------|-------|-------|-------|-------|
+| Full Dataset | 7724 | 0.764 | 0.015 | 0.122 | 0.081 | 0.752 | -5.6 |
+| linear model | 7724 | 0.248 | 0.046 | 0.214 | 0.173 | 0.247 | \~ 0 |
+| Inverse Distance Weighted (LOO CV) | 7724 | 0.449 | 0.029 | 0.170 | 0.131 | 0.448 | 1.86 |
 | Climate - Monsoon Dominated | 3039 | 0.633 | 0.016 | 0.126 | 0.074 | 0.619 | -13.7 |
@@
 
+The reduced predictor set was selected to maximize interpretability and minimize overfitting.
+However, incorporating additional variables could improve model performance, especially in capturing regionally complex baseflow dynamics.
+Subsurface hydrologic characteristics, such as aquifer properties, groundwater depth, and spring density, are known to influence baseflow but were excluded due to limited availability across the study domain.
+Temporal predictors, including snowpack duration, antecedent soil moisture, and wetness indices, could improve annual BFI predictions by capturing lagged hydrologic responses to climate variability.
+Additionally, incorporating human influences (for example, groundwater withdrawals, land-use change, or reservoir operations) would be important in regions with modified hydrologic regimes.
+Future efforts may benefit from integrating these datasets where available, although such improvements would require region-specific data processing and validation strategies.
+
 Several previous studies have examined regional and continental patterns of BFI, offering useful context for our findings despite limited geographic overlap.

### 2025-05-31 submission_docs/2nd GO/supp_info.qmd
prepping new versions

---
title: "supp_info"
format: docx
---

| Variable | Source | Sub Variables | Unit | Geoprocessing |
|----------------|------------|----------------------|------------|------------|
| Precipitation | PRISM |  | mm | Basin average |
| Temperature | PRISM |  | C | Basin average |
| Reference Evapotranspiration | MODIS |  | mm | Basin average |
| Elevation | DEM |  | m | Basin average |
| Area | DEM |  | km\^2 | Basin average |
| Slope | DEM |  | degrees | Basin average |
| Aspect | DEM | Flat | \% | Percent areal coverage |
|  |  | North | \% | Percent areal coverage |
|  |  | South | \% | Percent areal coverage |
| Land Cover | NLCD | Open Water | \% | Percent areal coverage |
|  |  | Developed - Open Space | \% | Percent areal coverage |
|  |  | Developed - High Intensity | \% | Percent areal coverage |
|  |  | Developed - Medium Intensity | \% | Percent areal coverage |
|  |  | Developed - Low Intensity | \% | Percent areal coverage |
|  |  | Barren Land | \% | Percent areal coverage |
|  |  | Deciduous Forest | \% | Percent areal coverage |
|  |  | Evergreen Forest | \% | Percent areal coverage |
|  |  | Mixed Forest | \% | Percent areal coverage |
|  |  | Shrub | \% | Percent areal coverage |
|  |  | Herbaceous | \% | Percent areal coverage |
|  |  | Hay Pasture | \% | Percent areal coverage |
|  |  | Cultivated Crops | \% | Percent areal coverage |
|  |  | Woody Wetlands | \% | Percent areal coverage |
|  |  | Emergent Herbaceous Wetlands | \% | Percent areal coverage |
| Hydrologic Soil Group | SSURGO | Group A | \% | Percent areal coverage |
|  |  | Group B | \% | Percent areal coverage |
|  |  | Group C | \% | Percent areal coverage |
|  |  | Group D | \% | Percent areal coverage |
|  |  | Group A/D | \% | Percent areal coverage |
|  |  | Group B/D | \% | Percent areal coverage |
|  |  | Group C/D | \% | Percent areal coverage |
| Underlying Geology | USGS | Igneous Intrusive | \% | Percent areal coverage |
|  |  | Sedimentary Clastic | \% | Percent areal coverage |
|  |  | Metamorphic & Sedimentary Undifferentiated | \% | Percent areal coverage |
|  |  | Igneous & Sedimentary Undifferentiated | \% | Percent areal coverage |
|  |  | Metamorphic Schist | \% | Percent areal coverage |
|  |  | Sedimentary Undifferentiated | \% | Percent areal coverage |
|  |  | Igneous Volcanic | \% | Percent areal coverage |
|  |  | Metamorphic Undifferentiated | \% | Percent areal coverage |
|  |  | Unconsolidated Undifferentiated | \% | Percent areal coverage |
|  |  | Igneous Undifferentiated | \% | Percent areal coverage |
|  |  | Metamorphic Gneiss | \% | Percent areal coverage |
|  |  | Metamorphic Volcanic | \% | Percent areal coverage |
|  |  | Metamorphic Sedimentary Clastic | \% | Percent areal coverage |
| Karst | USGS |  | \% | Percent areal coverage |

### 2025-06-24 elsevier_MANUSCRIPT.qmd

@@
 | **Classification Group** | **n** | **R2** | **MSE** | **RMSE** | **MAE** | **NSE** | **pbias** |
-|-------------------------|-------|-------|-------|-------|-------|-------|-------|
+|----|----|----|----|----|----|----|----|
 | Full Dataset | 7724 | 0.764 | 0.015 | 0.122 | 0.081 | 0.752 | -5.6 |

### 2025-06-26 elsevier_MANUSCRIPT.qmd
new submission docs

--- 2025-05-31 elsevier_MANUSCRIPT.qmd
@@
 We develop a region-specific machine learning model using eXtreme Gradient Boosting (XGBoost) to estimate long-term BFI (1991–2020) across ungauged basins, incorporating basin hydrogeology and hydroclimate predictors.
+To our knowledge, this is the first map of spatially distributed BFI in Arizona.
 In parallel, we assess trends in BFI at instrumented sites alongside coincident changes in precipitation, temperature, and reference evapotranspiration.
@@
 XGBoost is a gradient boosting algorithm that builds an ensemble of decision trees, where each tree iteratively corrects errors from previous trees to improve predictive accuracy [@chen2016xgboost].
+To our knowledge this is the first time that this advanced, decision-tree algorithm has been used in hydrologic prediction.
 The model was trained on annual BFI observations in order to estimate annual BFI in ungauged basins.
@@
 The final model was applied to all HUC8 basins across Arizona using the same predictor layers to estimate long-term BFI values.
-
-![SHAP value plot of features used in final model.
-The number to the right of the feature name is the mean absolute SHAP value.
-Land cover features are indicated by the percentage of cover by each land cover type and soil types are defined by hydrologic soil group.](images/shap_summary.png){#fig-shap_values}
 
@@
 | **Classification Group** | **n** | **R2** | **MSE** | **RMSE** | **MAE** | **NSE** | **pbias** |
-|-------------------------|-------|-------|-------|-------|-------|-------|-------|
+|---------|---------|---------|---------|---------|---------|---------|---------|
 | Full Dataset | 7724 | 0.764 | 0.015 | 0.122 | 0.081 | 0.752 | -5.6 |
@@
 Seasonal snowpack duration, in particular, has been shown to strongly correlate with springflow and groundwater recharge in this region [@donovan_karst_2022].
+
+![SHAP value plot of features used in final model.
+The number to the right of the feature name is the mean absolute SHAP value.
+Land cover features are indicated by the percentage of cover by each land cover type and soil types are defined by hydrologic soil group.](images/shap_summary.png){#fig-shap_values}
 

### 2025-09-07 submission_docs/2nd GO/supp_info.qmd
edits to edits

@@
 title: "supp_info"
-format: docx
+format: 
+  docx: default
 ---
@@
 | Variable | Source | Sub Variables | Unit | Geoprocessing |
-|----------------|------------|----------------------|------------|------------|
+|--------------|--------------|-----------------|--------------|--------------|
 | Precipitation | PRISM |  | mm | Basin average |
